---
title: "Practice Problems · Session 12"
subtitle: "Logistic regression: odds, odds ratios, and predicted probabilities"
---

::: {.callout-note appearance="simple"}
**Ungraded self-checks.** Work each problem in your own R session *before* clicking the solution. Homework 12 is your own final-project model, so this bank is where you drill the logistic workflow on neutral data first — if you can do these cold, Checkpoint 4 (and the presentation questions afterward) will feel routine.
:::

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(brms)
set.seed(705)
crashes <- read_csv("https://smourtgos.github.io/crju705/data/crash-ak.csv",
                    show_col_types = FALSE) |>
  mutate(injury = as.integer(injuries > 0))
```

Problems that use data need only this setup — the 20,000 Alaska crashes from Labs 5, 7, 8, and 12. In lab you modeled *whether the driver was intoxicated*; here the question flips: **does an intoxicated driver make a crash more likely to injure someone?** The outcome is a flag we build ourselves:

```{r}
#| label: setup-shown
#| eval: false
library(tidyverse)
library(brms)   # only if you re-fit Problem 6's model yourself

crashes <- read_csv("https://smourtgos.github.io/crju705/data/crash-ak.csv") |>
  mutate(injury = as.integer(injuries > 0))   # 1 = someone was injured
```

## Problem 1 · Odds ↔ probability, three conversions

No data, just the ladder. (a) About **12%** of released defendants in a jurisdiction fail to appear: convert that probability to **odds**, and phrase it as "1 to ___." (b) A program director says the odds a participant completes supervision are **4**: what *probability* is that? (c) A logistic model reports a baseline **log-odds of −1.99**: use `plogis()` to recover the probability — and notice where you've landed.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p1
0.12 / (1 - 0.12)   # (a) probability -> odds

4 / (4 + 1)         # (b) odds -> probability

plogis(-1.99)       # (c) log-odds -> probability

qlogis(0.12)        # and back up the ladder: probability -> log-odds
```

(a) Odds ≈ **0.136** — about 1 to 7.3: for every failure to appear, roughly seven appearances. (b) Odds of 4 ("four to one") is a probability of **0.80**. (c) `plogis(-1.99)` ≈ **0.12** — you've climbed back down to (a), because −1.99 is exactly `qlogis(0.12)`. Probability, odds, and log-odds are three readouts of one quantity; `plogis()` and `qlogis()` move you up and down the ladder, and the model does its arithmetic on the log-odds rung.
:::

## Problem 2 · Fit it: `glm(family = binomial)`

First check the base rate of `injury` (you built the flag in Setup — always know the baseline you're modeling). Then fit `injury ~ intox_any + num_vehicles`. What do the two coefficients say *before* any transformation — and what scale are they on?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p2
mean(crashes$injury)

injury_model <- glm(injury ~ intox_any + num_vehicles,
                    data = crashes, family = binomial)
summary(injury_model)
```

Baseline: about **48%** of crashes injure someone — this outcome is nowhere near rare, which will matter in Problem 5. Both coefficients are positive with tiny p-values: intoxication (+0.65) and each additional vehicle (+0.24) raise the **log-odds** of an injury crash. Sign and significance are readable right off this table; the *sizes* are not, because nobody thinks in log-odds — which is exactly why the next move is always `exp()`.
:::

## Problem 3 · Exponentiate: log-odds → odds ratios

Turn the model into odds ratios with confidence intervals (`exp(cbind(OR = coef(...), confint(...)))`). Interpret each predictor in one sentence with the holding-constant clause — and state what the "no effect" value is for an odds ratio.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p3
exp(cbind(OR = coef(injury_model), confint(injury_model)))
```

- **Intoxication: OR ≈ 1.91** [1.72, 2.13] — a crash involving an intoxicated driver has about **91% higher odds** of injuring someone than a sober-driver crash, holding the number of vehicles constant.
- **Vehicles: OR ≈ 1.27** [1.18, 1.36] — each additional vehicle multiplies the odds of injury by about 1.27. (In these data `num_vehicles` is only ever 1 or 2, so this is really the two-vehicle vs. single-vehicle contrast — check your predictor's actual values before narrating a "per unit" story.)
- For odds ratios the no-effect value is **1**, not 0 — both intervals sit clearly above it.

`exp()` converts every log-odds coefficient into a multiplier on the *odds*, and 1 is the number that means "nothing happening."
:::

## Problem 4 · From effects to cases: predicted probabilities

Odds ratios describe effects; a briefing needs *cases*. Use `predict(type = "response")` for two crash profiles — **(i)** sober driver, single vehicle; **(ii)** intoxicated driver, two vehicles — then reproduce profile (i) by hand with `plogis()` and the coefficients. How do the two probabilities compare with the 48% baseline?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p4
profiles <- tibble(intox_any = c(0, 1), num_vehicles = c(1, 2))

profiles |>
  mutate(p_injury = predict(injury_model, newdata = profiles, type = "response"))

b <- coef(injury_model)
unname(plogis(b[1] + b[2] * 0 + b[3] * 1))   # profile (i), by hand
```

A sober single-vehicle crash carries about a **42%** predicted probability of injury; an intoxicated two-vehicle crash about **64%** — one below the 48% baseline, one well above it. And the by-hand version demystifies the machinery: plug the profile into the log-odds equation, then `plogis()` climbs down the ladder to a probability. Report odds ratios for *effects* and predicted probabilities for *cases* — decision-makers think in probabilities, and `predict()` is how you hand them one.
:::

## Problem 5 · Audit the memo (the classic mistake)

A draft memo about your model reads:

> "The odds ratio for intoxication is 1.91, so a crash with an intoxicated driver is **91% more likely to injure someone**. Since about 48% of crashes cause injury, that works out to roughly **0.48 × 1.91 ≈ 92% of intoxicated-driver crashes causing injury**."

Both bolded claims are wrong — and the second one should already smell wrong (what would an OR of 2.5 imply?). Name each error, then compute what the model *actually* says happens to a two-vehicle crash's injury probability when the driver is intoxicated, working through the odds.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p5
p_sober <- unname(plogis(b[1] + b[3] * 2))          # sober, two vehicles
p_intox <- unname(plogis(b[1] + b[2] + b[3] * 2))   # intoxicated, two vehicles
c(p_sober = p_sober, p_intox = p_intox)

odds_sober <- p_sober / (1 - p_sober)
odds_intox <- p_intox / (1 - p_intox)
c(odds_sober = odds_sober, odds_intox = odds_intox, ratio = odds_intox / odds_sober)
```

**Error 1:** "91% more likely" reads the odds ratio as a *probability* ratio. An OR of 1.91 multiplies the **odds** by 1.91 — nothing else. **Error 2:** multiplying the baseline probability by the OR (0.48 × 1.91) is arithmetic on the wrong scale; with an OR of 2.5 it would "predict" 120% of crashes causing injury, which is the tell that probabilities don't multiply this way.

What actually happens: the sober two-vehicle probability of .48 converts to odds of ≈ 0.91; multiply *those* by 1.91 to get odds ≈ 1.75; convert back and the probability is ≈ **.64**. The real change is 48% → 64% — about 16 percentage points, a ~33% relative rise, nowhere near 91%. When the baseline probability is small, odds and probability move almost together; at a baseline near 50%, they part company badly — so multiply odds, never probabilities, and use `predict()` when you want the probability itself.
:::

## Problem 6 · Reading a `brms` bernoulli output

A classmate fits the Bayesian twin of your Problem 2 model and sends you the summary below. **Without fitting anything**, answer: (a) what do `Family: bernoulli` and `Links: mu = logit` tell you about the scale of the `Estimate` column? (b) Exponentiate the `intox_any` estimate and both CI bounds — what odds ratio and 95% credible interval do you get, and does it agree with `glm()`'s? (c) Are there signs of sampling trouble? (d) Write the one Bayesian sentence about intoxication you're now licensed to say.

```
 Family: bernoulli
  Links: mu = logit
Formula: injury ~ intox_any + num_vehicles
   Data: crashes (Number of observations: 20000)
  Draws: 2 chains, each with iter = 2000; warmup = 1000; thin = 1;
         total post-warmup draws = 2000

Regression Coefficients:
             Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept       -0.56      0.07    -0.70    -0.44 1.00     2056     1378
intox_any        0.65      0.06     0.54     0.76 1.00     1795     1378
num_vehicles     0.24      0.04     0.17     0.31 1.00     2081     1415

Draws were sampled using sampling(NUTS). For each parameter, Bulk_ESS
and Tail_ESS are effective sample size measures, and Rhat is the potential
scale reduction factor on split chains (at convergence, Rhat = 1).
```

*(Want to reproduce it? The code is below — but budget for the "Compiling Stan program..." pause plus a couple of minutes of sampling on 20,000 rows. The reading practice is the point; the refit is optional.)*

```{r}
#| label: p6-refit
#| eval: false
injury_bayes <- brm(injury ~ intox_any + num_vehicles, data = crashes,
                    family = bernoulli(),
                    chains = 2, iter = 2000, seed = 705, refresh = 0)
summary(injury_bayes)
```

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p6
exp(c(estimate = 0.65, ci_low = 0.54, ci_high = 0.76))
```

(a) `bernoulli` is `brms`'s name for a 0/1 outcome and `logit` is the link — so the `Estimate` column is in **log-odds**, exactly like `glm()`'s coefficient table, and needs exponentiating before anyone speaks.

(b) exp(0.65) ≈ **1.92**, with a 95% credible interval of roughly **[1.72, 2.14]** — landing almost exactly on `glm()`'s OR of 1.91 [1.72, 2.13], as expected with default priors and 20,000 crashes.

(c) No trouble: **Rhat = 1.00** on every row and effective sample sizes in the 1,400–2,100 range mean the two chains agreed and mixed well.

(d) "Given the data and priors, there is a 95% probability that intoxication multiplies a crash's odds of injuring someone by between roughly 1.7 and 2.1, holding the number of vehicles constant." Same machinery as every `brm()` you've run — bernoulli is just `family = binomial` in Bayesian clothing, and the coefficients only become English after `exp()`.
:::
