---
title: "Practice Problems · Session 6"
subtitle: "Hypothesis testing, both ways: p-values, Bayes factors, and posteriors"
---

::: {.callout-note appearance="simple"}
**Ungraded self-checks.** Work each problem in your own R session *before* clicking the solution. These use different questions and numbers than the homework on purpose — if you can run *and read* both traditions cold here, Homework 6 (and the midterm's inference questions) will feel routine.
:::

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(BayesFactor)
set.seed(705)
crimes <- read_csv("https://smourtgos.github.io/crju705/data/chicago-crimes-2025.csv",
                   show_col_types = FALSE)
```

All problems that use data need only:

```{r}
#| label: setup-shown
#| eval: false
library(tidyverse)
library(BayesFactor)   # install.packages("BayesFactor") once, if needed
crimes <- read_csv("https://smourtgos.github.io/crju705/data/chicago-crimes-2025.csv")
```

## Problem 1 · What a p-value promises

A study compares case-processing times in two courts and reports **p = 0.03**. Three readings:

> (a) "There is a 3% chance the courts truly process cases at the same speed."
>
> (b) "If the courts truly processed cases at the same speed, a gap this large or larger would show up in about 3% of studies like this one."
>
> (c) "There is a 97% chance the courts differ."

Pick the correct one. Then verify the machinery's actual promise by simulation: run 5,000 t-tests comparing two groups drawn from the *same* normal distribution (n = 40 each) and compute how often p < .05.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p1
set.seed(705)
null_world_p <- replicate(5000, t.test(rnorm(40), rnorm(40))$p.value)
mean(null_world_p < 0.05)
```

**(b)** — and only (b). A **p-value** is $P(\text{data this extreme} \mid H_0)$, a statement about *data under an assumed null*; (a) and (c) both flip it into $P(H \mid \text{data})$, which frequentism never computes (that flip is Bayes' job, and it costs a prior). The simulation shows the promise being kept: in a world where the null is *true*, about 5.5% of our 5,000 tests came up "significant" — right around the advertised 5% false-alarm rate, and no better.
:::

## Problem 2 · Translate the evidence

Four studies report Bayes factors: $BF_{10}$ = **0.24**, **2.6**, **18**, and **240**. For each, compute $BF_{01}$ and classify the evidence on the extended scale (1–3 barely worth mentioning · 3–10 moderate · 10–30 strong · 30–100 very strong · 100+ extreme). Which study's data actually *favor the null* — and how strongly?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p2
bf10 <- c(0.24, 2.6, 18, 240)
round(1 / bf10, 3)   # BF01: the same evidence, read for the null
```

A **Bayes factor** is a ratio of predictions, so below 1 means the *null* predicted the data better: the first study's $BF_{10}$ = 0.24 is $BF_{01}$ ≈ 4.2 — **moderate evidence for no effect** (something a p-value can never report). The rest, on the scale: 2.6 is barely worth mentioning, 18 is strong, and 240 is extreme evidence for a difference. Same scale in both directions — just flip the fraction.
:::

## Problem 3 · Both tests, one new question

Do crimes on the **street** happen at different hours than crimes inside a **residence**? Filter `crimes` to `location_description %in% c("STREET", "RESIDENCE")`, compare mean `hour` by group, then run **both** tests: `t.test()` and `ttestBF()`. Watch each tool's direction, and say whether the two traditions agree here — in class, on violent-vs-property, they didn't.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p3
sr <- crimes |> filter(location_description %in% c("STREET", "RESIDENCE"))

sr |> summarize(mean_hour = mean(hour), sd_hour = sd(hour), n = n(),
                .by = location_description)

t.test(hour ~ location_description, data = sr)

street_hours    <- sr$hour[sr$location_description == "STREET"]
residence_hours <- sr$hour[sr$location_description == "RESIDENCE"]

bf <- ttestBF(x = street_hours, y = residence_hours)
extractBF(bf)$bf
```

Street incidents average hour 12.6 vs 11.9 in residences — about 47 minutes later. Directions differ by tool: `t.test()` sorts alphabetically (RESIDENCE − STREET, hence the negative t ≈ −5.5 and CI of −1.06 to −0.50 hours), while we handed `ttestBF()` x = street, y = residence, so its difference runs positive. This time the two traditions **agree**: p ≈ 5 × 10⁻⁸ and $BF_{10}$ ≈ 38,000 — extreme evidence for a real difference. Unlike class's 15-minute violent-vs-property gap, this one is large enough (relative to its precision) that "some difference" out-predicts "no difference" by a mile.
:::

## Problem 4 · The number the star hides

Both frameworks just endorsed the street-vs-residence difference. Before anyone re-draws a patrol schedule: compute **Cohen's d** (pooled-SD version, as in class) and the gap in **minutes**. What's the full, honest verdict?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p4
sr <- crimes |> filter(location_description %in% c("STREET", "RESIDENCE"))
street_hours    <- sr$hour[sr$location_description == "STREET"]
residence_hours <- sr$hour[sr$location_description == "RESIDENCE"]

sd_pooled <- sqrt((var(street_hours) + var(residence_hours)) / 2)
(mean(street_hours) - mean(residence_hours)) / sd_pooled   # Cohen's d
(mean(street_hours) - mean(residence_hours)) * 60          # gap in minutes
```

d ≈ 0.11 — below even the 0.2 "small" benchmark: a 47-minute shift against SDs of about 7 *hours*. The full verdict has three parts and needs all of them: the difference is almost certainly real (both tests), and it is modest in size (**effect size**). "Significant" and "decisive" describe our *certainty*, never the *magnitude* — report the units a human cares about, every time.
:::

## Problem 5 · The sentence you're licensed to say

Session 5 forbade "there's a 95% probability the truth is in this interval." Buy the license: draw 10,000 posterior samples from Problem 3's Bayes-factor model with `posterior()`, get the 95% **credible interval** for the difference, and compute the posterior probability the difference is positive. Then write the sentence — in minutes.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p5
sr <- crimes |> filter(location_description %in% c("STREET", "RESIDENCE"))
street_hours    <- sr$hour[sr$location_description == "STREET"]
residence_hours <- sr$hour[sr$location_description == "RESIDENCE"]
bf <- ttestBF(x = street_hours, y = residence_hours)

set.seed(705)
post <- suppressMessages(posterior(bf, iterations = 10000))
diff_post <- as.numeric(post[, "beta (x - y)"])

quantile(diff_post, c(0.025, 0.975))
mean(diff_post > 0)
```

The licensed sentence: *"Given the model and prior, there is a 95% probability that street crimes occur, on average, between about 30 and 64 minutes later in the day than in-residence crimes."* That is exactly what a **credible interval** means — probability attached to the parameter itself, the reading everyone *wants* to give a confidence interval. And every one of the 10,000 posterior draws came out positive, so the direction is about as settled as this model can make it.
:::

## Problem 6 · The diversion-program trial

A county pilots a **pretrial diversion program** to reduce failures to appear (FTA). Historically, about **25%** of defendants fail to appear — encode that as a **Beta(3, 9)** prior (mean 0.25, worth about 12 observations). The pilot randomizes 120 defendants: of **60 diversion clients, 12** fail to appear; of **60 controls, 19** do.

1. Run the frequentist test: `prop.test(x = c(12, 19), n = c(60, 60))`.
2. Update the prior for each group with `rbeta()` (as in the class recidivism example) and compute **P(diversion FTA rate < control FTA rate)**.
3. Write one sentence for the county board.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p6
prop.test(x = c(12, 19), n = c(60, 60))

set.seed(705)
theta_div     <- rbeta(10000, 3 + 12, 9 + 48)   # posterior, diversion group
theta_control <- rbeta(10000, 3 + 19, 9 + 41)   # posterior, control group

mean(theta_div < theta_control)                 # P(program reduces FTA)
c(mean(theta_div), mean(theta_control))         # posterior means
```

The frequentist verdict: p ≈ 0.21 — an 11.7-point gap like this could plausibly arise if the program did nothing, so "not significant," and the grant report probably ends there. The **posterior probability** tells the board what it actually asked: given prior experience and the pilot data, there is about a **91% probability the diversion program lowers failure-to-appear** (posterior means ≈ 21% vs 31%). One board-ready sentence: *"The pilot is too small for a definitive verdict, but the evidence currently runs about 9-to-1 that diversion reduces failures to appear."* Both statements are true of the same data — they answer different questions, and only one of them helps a decision-maker.
:::
