---
title: "Practice Problems · Session 10"
subtitle: "Correlation: covariance to r, what r can't see, and inference in both traditions"
---

::: {.callout-note appearance="simple"}
**Ungraded self-checks.** Work each problem in your own R session *before* clicking the solution. These use different data and variables than the homework on purpose — if you can do these cold, Homework 10 (and Checkpoint 2 on your own project) will feel routine.
:::

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(correlation)
set.seed(705)
cities <- read_csv("https://smourtgos.github.io/crju705/data/cities-wide.csv",
                   show_col_types = FALSE)
```

Problems that use data need only this setup — `cities` is the 100 simulated-but-realistic cities you pivoted in Lab 4; today we stay in one year and go cross-sectional. One row = one **city**, not one person. Keep that in mind: it decides what your sentences are allowed to claim (Problem 6 will test you on it).

```{r}
#| label: setup-shown
#| eval: false
library(tidyverse)
library(correlation)

cities <- read_csv("https://smourtgos.github.io/crju705/data/cities-wide.csv")
```

## Problem 1 · Covariance's units problem

Compute the covariance of `vcr_2025` (violent crimes per 100,000 residents) with **`population_2020`**, and then with **`poverty_rate`**. One of these covariances is roughly 3,800 times the size of the other. Is that association 3,800 times stronger? Settle it: divide each covariance by the two standard deviations involved, then confirm with `cor()`.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p1
cov(cities$population_2020, cities$vcr_2025)
cov(cities$poverty_rate,    cities$vcr_2025)

# Divide out the units: cov / (sd_x * sd_y)
cov(cities$population_2020, cities$vcr_2025) /
  (sd(cities$population_2020) * sd(cities$vcr_2025))

cov(cities$poverty_rate, cities$vcr_2025) /
  (sd(cities$poverty_rate) * sd(cities$vcr_2025))

cor(cities$population_2020, cities$vcr_2025)
cor(cities$poverty_rate,    cities$vcr_2025)
```

The ~3.4-million covariance turns into **r = 0.04** — essentially nothing — while the "little" 914 becomes **r = 0.61**. The first number was huge only because population is measured in *hundreds of thousands of people*; covariance is glued to its units, so its size is uninterpretable across pairs. Dividing by both SDs cancels the units and puts every pair on the same −1 to +1 ruler.
:::

## Problem 2 · What r can't see

A training study measures 80 recruits' pre-qualification **anxiety** (0–10 scale) and their **marksmanship score** (100-point course). Simulate it, compute r, and then plot it:

```{r}
#| label: p2-data
set.seed(705)   # inline seed: keeps this particular draw stable
recruits <- tibble(
  anxiety      = seq(0, 10, length.out = 80),
  marksmanship = 88 - 2.1 * (anxiety - 5)^2 + rnorm(80, 0, 6)
)
```

A colleague computes `cor()`, gets roughly −0.04, and writes: *"anxiety and marksmanship are unrelated."* What did they miss, and what habit would have caught it?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p2
#| fig-height: 3.5
cor(recruits$anxiety, recruits$marksmanship)

ggplot(recruits, aes(anxiety, marksmanship)) +
  geom_point(alpha = 0.6, color = "grey30") +
  geom_smooth(method = "loess", se = FALSE, color = "firebrick") +
  labs(x = "Pre-qualification anxiety (0-10)", y = "Marksmanship score")
```

r ≈ −0.04, and the plot shows anything but "unrelated": scores peak around mid-scale anxiety and collapse at both extremes — a strong inverted-U (the classic arousal-performance curve). Because the upslope and downslope cancel, the *linear* summary lands near zero. **r measures straight-line association only; r ≈ 0 means no linear relationship, not no relationship** — which is why you plot before you correlate, every time.
:::

## Problem 3 · One unit can manufacture a correlation

Twenty patrol beats, each with an average 911 **response time** (minutes) and a resident **satisfaction** score (0–100). Then the analyst appends beat 21 — the beat covering the airport, a giant low-density parcel: response time 28.4 minutes, satisfaction 31.

```{r}
#| label: p3-data
set.seed(705)   # inline seed: keeps this particular draw stable
beats <- tibble(
  response_min = rnorm(20, 10, 1.5),
  satisfaction = rnorm(20, 70, 6)
)

beats_all <- beats |>
  add_row(response_min = 28.4, satisfaction = 31)
```

Compute **Pearson's r and Spearman's rho** for the 20 beats, and again with the airport included. What happens to each — and which coefficient is telling the more honest story about the *typical* beat?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p3
c(pearson_20  = cor(beats$response_min, beats$satisfaction),
  spearman_20 = cor(beats$response_min, beats$satisfaction, method = "spearman"))

c(pearson_21  = cor(beats_all$response_min, beats_all$satisfaction),
  spearman_21 = cor(beats_all$response_min, beats_all$satisfaction, method = "spearman"))
```

Twenty beats show essentially nothing (r ≈ 0.04). Add **one** unrepresentative unit and Pearson leaps to ≈ **−0.81** — a "very strong" negative correlation manufactured entirely by a single point sitting far out in both variables. Spearman moves only from ≈ −0.02 to ≈ −0.15, because to a rank-based measure the airport is just "rank 21," not "a dozen standard deviations of leverage." Neither number substitutes for the scatterplot — but when one wild unit is in the data, Spearman describes the typical beats far more honestly, and a Pearson-Spearman gap this large is itself a red flag to go look.
:::

## Problem 4 · Inference: the matrix, then the star pair

(a) Run `correlation()` on `vcr_2025`, `poverty_rate`, `unemployment_rate`, and `police_per_10k`. Which pairs survive the Holm adjustment — and why do these p-values need adjusting at all?

(b) The star pair is poverty and violence. Run `cor.test()` on it and write the one-sentence lecture-style report: r, df, p, CI, direction, strength.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p4
cities |>
  select(vcr_2025, poverty_rate, unemployment_rate, police_per_10k) |>
  correlation()

cor.test(cities$poverty_rate, cities$vcr_2025)
```

(a) Exactly **one** pair survives: poverty with violent crime (r = .61, adjusted p < .001). Every other CI crosses 0, with adjusted p-values from .59 to essentially 1. The adjustment exists because this table runs **six tests at once** — six chances at a 5% false alarm is Session 8's pile-of-t-tests problem wearing correlation clothes, and Holm restores the family-wise honesty.

(b) "Across these 100 cities, poverty was moderately associated with the violent-crime rate, r = .61, t(98) = 7.59, p < .001, 95% CI [.47, .72] — cities with higher poverty rates recorded more violent crime per 100,000 residents." (By the lecture's bands, .61 is a *moderate* correlation — and in messy social-science data, a moderate r is a major finding.) A correlation matrix is the map; `cor.test()` is the full readout you pull for the one relationship your analysis is actually about.
:::

## Problem 5 · Bayesian correlation: a strong pair and an honest null

Run the Bayesian version (`bayesian = TRUE`) twice: once on `poverty_rate` + `vcr_2025`, once on `unemployment_rate` + `vcr_2025`. For each, read out **rho, the 95% credible interval, pd, % in ROPE, and the Bayes factor**. For the second pair, also compute 1/BF. What can you say about unemployment that a p-value never could?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p5
set.seed(705)   # inline seed: the Bayesian version samples from the posterior
cities |>
  select(vcr_2025, poverty_rate) |>
  correlation(bayesian = TRUE)

set.seed(705)
bayes_unemp <- cities |>
  select(vcr_2025, unemployment_rate) |>
  correlation(bayesian = TRUE)
bayes_unemp

1 / bayes_unemp$BF
```

**Poverty:** rho ≈ .58 (a touch below the frequentist .61 — the mildly skeptical prior's pull), 95% CI ≈ [.46, .70], pd = 100%, 0% in ROPE, BF ≈ 4 × 10⁸ — evidence far past "extreme" on Session 6's scale.

**Unemployment:** rho ≈ .08, CI ≈ [−.11, .25] straddling 0, pd ≈ 80% (a coin flip barely leaning positive), and **over half the posterior sits inside the ROPE** — the "practically zero" band. The BF ≈ 0.32 is *below 1*: taking 1/BF, the data are about **3 times more consistent with no correlation than with one**. That's the sentence a p-value can never earn — the frequentist p ≈ .40 only says "failed to reject"; the Bayes factor gives you *positive evidence for the null*, stated as such.
:::

## Problem 6 · The sentence audit

Your Problems 4–5 results end up in a briefing. Five draft sentences below — **which are defensible, and what exactly is wrong with the rest?**

> (a) "Cities with higher poverty rates recorded higher violent-crime rates in 2025 (r = .61)."
>
> (b) "Poor residents commit most of the violent crime in these cities."
>
> (c) "Poverty causes violent crime."
>
> (d) "City-level poverty statistically accounts for about 37% of the city-to-city variance in violent-crime rates."
>
> (e) "We found no reliable city-level association between unemployment and violent crime; if anything, the data modestly favor there being none."

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p6
cor(cities$poverty_rate, cities$vcr_2025)^2   # the 37% in sentence (d)
```

**Defensible: (a), (d), and (e).** (a) stays at the level of the rows — cities — and claims only association. (d) is r² = .61² ≈ .37, correctly framed as *variance accounted for*, still at the city level. (e) matches Problem 5's output, including the BF ≈ 0.32 nuance.

**(b) fails** — the **ecological fallacy**. These rows are cities; every resident of a city shares one poverty rate and one crime rate, so the data contain nothing about *which residents* are involved in violence. **(c) fails** — a causal claim from a cross-sectional correlation; disinvestment, housing, policing patterns, and more lurk behind both variables. The audit habit: check every sentence for the *level* the data measure (areas vs. people) and the *strength* the design supports (association vs. cause).
:::
