---
title: "Practice Problems · Session 4"
subtitle: "The law of large numbers, sampling distributions, and the CLT"
---

::: {.callout-note appearance="simple"}
**Ungraded self-checks.** Work each problem in your own R session *before* clicking the solution. These use different numbers and variables than the homework on purpose — if you can do these cold, Homework 4 (and the midterm's sampling questions) will feel routine. Your random numbers will differ from the solutions' unless you run `set.seed(705)` immediately before each simulation — close is what you're checking for, not digit-for-digit matches.
:::

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(705)
nbhd <- read_csv("https://smourtgos.github.io/crju705/data/neighborhoods.csv",
                 show_col_types = FALSE)
```

All problems that use data need only:

```{r}
#| label: setup-shown
#| eval: false
library(tidyverse)
nbhd <- read_csv("https://smourtgos.github.io/crju705/data/neighborhoods.csv")
```

## Problem 1 · The long run arrives slowly

Suppose **12%** of intake drug screens at a jail come back positive. Simulate one week (**25 screens**), one quarter (**250**), and one decade (**25,000**) with `sample()`, and compute the positive proportion for each. How close does each get to 0.12?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p1
set.seed(705)
mean(sample(c(TRUE, FALSE), 25,    replace = TRUE, prob = c(0.12, 0.88)))
mean(sample(c(TRUE, FALSE), 250,   replace = TRUE, prob = c(0.12, 0.88)))
mean(sample(c(TRUE, FALSE), 25000, replace = TRUE, prob = c(0.12, 0.88)))
```

This run: ≈ 0.20, ≈ 0.16, ≈ 0.122. A week of screens can sit far from the truth; a decade practically *is* the truth — the **law of large numbers**. The practical warning runs the other way: any rate computed from 25 cases deserves deep suspicion.
:::

## Problem 2 · Build a sampling distribution

Treat the 500 neighborhoods in `nbhd` as a **population**, and imagine an analyst who can only audit **40 neighborhoods at a time**. Using the `tibble()` + `map_dbl()` pattern from the Session 4 slides, draw **1,000 samples of 40** and keep each sample's mean of `assaults`. Histogram the 1,000 means and mark the true population mean with a `geom_vline()`. Where is the distribution centered?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p2
#| fig-height: 3.5
set.seed(705)
sample_means <- tibble(draw = 1:1000) |>
  mutate(mean_assaults = map_dbl(draw, \(d) mean(sample(nbhd$assaults, size = 40))))

mean(nbhd$assaults)                 # the truth
mean(sample_means$mean_assaults)    # center of the 1,000 sample means

ggplot(sample_means, aes(mean_assaults)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = mean(nbhd$assaults), color = "red", linewidth = 1) +
  labs(x = "Sample mean assaults (n = 40)", y = "Number of samples")
```

The 1,000 means pile up in a bell centered almost exactly on the true mean (≈ 1.06 vs 1.07 this run) — a **sampling distribution** in action: individual samples wander, but the *collection* of sample means is centered on the truth.
:::

## Problem 3 · The standard error, both ways

Session 4's rule: the formula must survive a simulation check. Compute the SD of your 1,000 sample means from Problem 2, then the formula version `sd(nbhd$assaults) / sqrt(40)`. How close?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p3
set.seed(705)
sample_means <- tibble(draw = 1:1000) |>
  mutate(mean_assaults = map_dbl(draw, \(d) mean(sample(nbhd$assaults, size = 40))))

sd(sample_means$mean_assaults)     # SE from the simulation
sd(nbhd$assaults) / sqrt(40)       # SE from the formula
```

≈ 0.20 simulated vs ≈ 0.21 from the formula — the **standard error** is just the SD of the sampling distribution, and `SD / sqrt(n)` predicts it without ever running the simulation. That's why one sample plus the formula is enough for the inference we start next session.
:::

## Problem 4 · Spot the error

A report on the Problem 2 audit states:

> "The standard error of about 0.21 tells us how much individual neighborhoods vary around the citywide average."

Fix the sentence. Then: if the analyst's budget grew to **n = 160** neighborhoods per audit, what would happen to the SE? Verify with the formula.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p4
c(se_40  = sd(nbhd$assaults) / sqrt(40),
  se_160 = sd(nbhd$assaults) / sqrt(160))
```

Wrong quantity: describing how individual neighborhoods vary is the **standard deviation's** job (≈ 1.3 here). The **standard error** describes how much the *sample mean* would move across repeated audits of 40 — a fix: "If we re-drew the 40-neighborhood audit, the estimated mean would typically shift by about 0.21 assaults." And quadrupling n to 160 *halves* the SE (≈ 0.21 → ≈ 0.10) — precision grows with √n, so it gets expensive fast.
:::

## Problem 5 · The CLT on a yes/no variable

The variable `hoa` is about as non-normal as data gets — every neighborhood is simply `"Yes"` or `"No"`. Compute the true population proportion of HOA neighborhoods, then draw **1,000 samples of 60** and keep each sample's proportion (`mean(sample(nbhd$hoa, size = 60) == "Yes")`). Histogram the 1,000 proportions. What shape appears, and which theorem predicted it?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p5
#| fig-height: 3.5
mean(nbhd$hoa == "Yes")            # true population proportion

set.seed(705)
hoa_props <- tibble(draw = 1:1000) |>
  mutate(prop_hoa = map_dbl(draw, \(d) mean(sample(nbhd$hoa, size = 60) == "Yes")))

ggplot(hoa_props, aes(prop_hoa)) +
  geom_histogram(binwidth = 0.02, fill = "steelblue", color = "white") +
  geom_vline(xintercept = mean(nbhd$hoa == "Yes"), color = "red", linewidth = 1) +
  labs(x = "Sample proportion with an HOA (n = 60)", y = "Number of samples")
```

A two-category population, yet the sample proportions form an approximately normal bell centered on the truth (0.354) — the **central limit theorem**. It's the same magic as the homework's arrest proportions, and it's why we can attach a standard error to *any* rate without the raw data ever looking normal.
:::

## Problem 6 · Why n − 1, re-proven on counts

Session 4 paid the n − 1 IOU with test scores; make it pay again with skewed count data. Draw **2,000 tiny samples of n = 4** from `nbhd$assaults` and compute each sample's variance both ways — dividing by n, and with `var()` (which divides by n − 1). Compare the two averages to the true population variance, `mean((x - mean(x))^2)`. *(The slides' list-column pattern — one `map()`, then two `map_dbl()`s — is the move.)*

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p6
set.seed(705)
var_sims <- tibble(draw = 1:2000) |>
  mutate(
    s           = map(draw, \(d) sample(nbhd$assaults, size = 4)),
    var_div_n   = map_dbl(s, \(x) sum((x - mean(x))^2) / length(x)),
    var_div_nm1 = map_dbl(s, \(x) var(x))
  )

var_sims |>
  summarize(
    true_variance           = mean((nbhd$assaults - mean(nbhd$assaults))^2),
    avg_divide_by_n         = mean(var_div_n),
    avg_divide_by_n_minus_1 = mean(var_div_nm1)
  )
```

Dividing by n lands around 1.28 — about 25% below the true variance (≈ 1.69) — while `var()` averages ≈ 1.7, right on target. Nothing about this trick required bell-shaped data: a sample clusters around *its own* mean, understating spread, and n − 1 is exactly the correction that makes the estimator **unbiased**.
:::
