---
title: "Practice Problems · Session 8"
subtitle: "Correlation: covariance to r, what r can't see, and testing r"
---

::: {.callout-note appearance="simple"}
**Ungraded self-checks.** Work each problem in your own R session *before* clicking the solution. These use different data and variables than the homework on purpose: if you can do these cold, Homework 8, Part 6 of the midterm, and Checkpoint 2 on your own project will feel routine.
:::

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(705)
cities <- read_csv("https://smourtgos.github.io/crju705/data/cities-wide.csv",
                   show_col_types = FALSE)
```

Problems that use data need only this setup. `cities` is 100 simulated-but-realistic cities, and we use one year of it. One row = one **city**, not one person. Keep that in mind: it decides what your sentences are allowed to claim (Problem 6 will test you on it).

```{r}
#| label: setup-shown
#| eval: false
library(tidyverse)

cities <- read_csv("https://smourtgos.github.io/crju705/data/cities-wide.csv")
```

## Problem 1 · Covariance's units problem

Compute the covariance of `vcr_2025` (violent crimes per 100,000 residents) with **`population_2020`**, and then with **`poverty_rate`**. One of these covariances is roughly 3,800 times the size of the other. Is that association 3,800 times stronger? Settle it: divide each covariance by the two standard deviations involved, then confirm with `cor()`.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p1
cov(cities$population_2020, cities$vcr_2025)
cov(cities$poverty_rate,    cities$vcr_2025)

# Divide out the units: cov / (sd_x * sd_y)
cov(cities$population_2020, cities$vcr_2025) /
  (sd(cities$population_2020) * sd(cities$vcr_2025))

cov(cities$poverty_rate, cities$vcr_2025) /
  (sd(cities$poverty_rate) * sd(cities$vcr_2025))

cor(cities$population_2020, cities$vcr_2025)
cor(cities$poverty_rate,    cities$vcr_2025)
```

The covariance in the millions turns into **r = `r sprintf("%.2f", cor(cities$population_2020, cities$vcr_2025))`**, essentially nothing, while the "little" one becomes **r = `r sprintf("%.2f", cor(cities$poverty_rate, cities$vcr_2025))`**. The first number was huge only because population is measured in *hundreds of thousands of people*; covariance is glued to its units, so its size can't be compared across pairs. Dividing by both SDs cancels the units and puts every pair on the same -1 to +1 ruler.
:::

## Problem 2 · What r can't see

A training study measures 80 recruits' pre-qualification **anxiety** (0 to 10 scale) and their **marksmanship score** (100-point course). Simulate it, compute r, and then plot it:

```{r}
#| label: p2-data
set.seed(705)   # inline seed: keeps this particular draw stable
recruits <- tibble(
  anxiety      = seq(0, 10, length.out = 80),
  marksmanship = 88 - 2.1 * (anxiety - 5)^2 + rnorm(80, 0, 6)
)
```

A colleague computes `cor()`, gets a number near zero, and writes: *"anxiety and marksmanship are unrelated."* What did they miss, and what habit would have caught it?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p2
#| fig-height: 3.5
cor(recruits$anxiety, recruits$marksmanship)

ggplot(recruits, aes(anxiety, marksmanship)) +
  geom_point(alpha = 0.6, color = "grey30") +
  geom_smooth(method = "loess", se = FALSE, color = "firebrick") +
  labs(x = "Pre-qualification anxiety (0-10)", y = "Marksmanship score")
```

r = `r sprintf("%.2f", cor(recruits$anxiety, recruits$marksmanship))`, and the plot shows anything but "unrelated": scores peak around mid-scale anxiety and collapse at both extremes, a strong upside-down U (the classic arousal and performance curve). Because the upslope and downslope cancel, the *straight-line* summary lands near zero. **r measures straight-line association only; r near 0 means no straight-line relationship, not no relationship**, which is why you plot before you correlate, every time.
:::

## Problem 3 · One unit can manufacture a correlation

Twenty patrol beats, each with an average 911 **response time** (minutes) and a resident **satisfaction** score (0 to 100). Then the analyst adds beat 21, the beat covering the airport, a giant low-density parcel: response time 28.4 minutes, satisfaction 31.

```{r}
#| label: p3-data
set.seed(705)   # inline seed: keeps this particular draw stable
beats <- tibble(
  response_min = rnorm(20, 10, 1.5),
  satisfaction = rnorm(20, 70, 6)
)

beats_all <- beats |>
  add_row(response_min = 28.4, satisfaction = 31)
```

Compute **Pearson's r and Spearman's rho** for the 20 beats, and again with the airport included. What happens to each, and which coefficient tells the more honest story about the *typical* beat?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p3
c(pearson_20  = cor(beats$response_min, beats$satisfaction),
  spearman_20 = cor(beats$response_min, beats$satisfaction, method = "spearman"))

c(pearson_21  = cor(beats_all$response_min, beats_all$satisfaction),
  spearman_21 = cor(beats_all$response_min, beats_all$satisfaction, method = "spearman"))
```

Twenty beats show essentially nothing (r = `r sprintf("%.2f", cor(beats$response_min, beats$satisfaction))`). Add **one** unusual unit and Pearson leaps to **`r sprintf("%.2f", cor(beats_all$response_min, beats_all$satisfaction))`**, a "very strong" negative correlation manufactured entirely by a single point sitting far out on both variables. Spearman moves only from `r sprintf("%.2f", cor(beats$response_min, beats$satisfaction, method = "spearman"))` to `r sprintf("%.2f", cor(beats_all$response_min, beats_all$satisfaction, method = "spearman"))`, because to a rank-based measure the airport is just "rank 21." Neither number substitutes for the scatterplot, but when one wild unit is in the data, Spearman describes the typical beats far more honestly, and a Pearson and Spearman gap this large is itself a red flag to go look.
:::

## Problem 4 · The matrix, then the star pair

(a) Run `cor()` on `vcr_2025`, `poverty_rate`, `unemployment_rate`, and `police_per_10k` (use `select()`, then `cor()`, then `round(2)`). Which variable has the strongest association with `vcr_2025`?

(b) The star pair is poverty and violence. Run `cor.test()` on it and write the one-sentence lecture-style report: r, the 95% confidence interval, the p-value, direction, and strength.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p4
cities |>
  select(vcr_2025, poverty_rate, unemployment_rate, police_per_10k) |>
  cor() |>
  round(2)

pov <- cor.test(cities$poverty_rate, cities$vcr_2025)
pov
```

(a) Poverty, at r = `r sprintf("%.2f", pov$estimate)`. Every other variable's correlation with `vcr_2025` is small.

(b) "Across these 100 cities, poverty was moderately associated with the violent-crime rate, r = `r sprintf("%.2f", pov$estimate)`, 95% CI `r sprintf("%.2f", pov$conf.int[1])` to `r sprintf("%.2f", pov$conf.int[2])`, p < 0.001: cities with higher poverty rates recorded more violent crime per 100,000 residents." By the lecture's bands this is a *moderate* correlation, and in messy social-science data a moderate r is a major finding. The matrix is the map; `cor.test()` is the full readout for the one relationship your analysis is actually about.
:::

## Problem 5 · An honest null

Run `cor.test()` on `unemployment_rate` and `vcr_2025`. Report r, the confidence interval, and the p-value. Where is zero? Then write the honest one-sentence conclusion, and say why "unemployment has nothing to do with violent crime" goes further than these data allow.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p5
unemp <- cor.test(cities$unemployment_rate, cities$vcr_2025)
unemp
```

r = `r sprintf("%.2f", unemp$estimate)`, 95% CI `r sprintf("%.2f", unemp$conf.int[1])` to `r sprintf("%.2f", unemp$conf.int[2])`, p = `r sprintf("%.2f", unemp$p.value)`. **Zero is inside the interval**, so we fail to reject. The honest sentence: "These 100 cities show no evidence of a straight-line association between unemployment and violent crime." Saying "nothing to do with" goes further in two ways: failing to reject is not proof of zero (the interval still allows a modest positive or a small negative correlation), and r only sees straight lines, so check the scatterplot for a curve.
:::

## Problem 6 · The sentence audit

Your Problem 4 and 5 results end up in a briefing. Five draft sentences below: **which are defensible, and what exactly is wrong with the rest?**

> (a) "Cities with higher poverty rates recorded higher violent-crime rates in 2025 (r = `r sprintf("%.2f", pov$estimate)`)."
>
> (b) "Poor residents commit most of the violent crime in these cities."
>
> (c) "Poverty causes violent crime."
>
> (d) "We found no evidence of a city-level association between unemployment and violent crime."
>
> (e) "We proved that unemployment and violent crime are unrelated."

::: {.callout-tip collapse="true"}
**Solution**

**Defensible: (a) and (d).** (a) stays at the level of the rows (cities) and claims only association. (d) matches Problem 5: no evidence, stated at the city level.

**(b) fails**: the **ecological fallacy**. These rows are cities; every resident of a city shares one poverty rate and one crime rate, so the data contain nothing about *which residents* are involved in violence. **(c) fails**: a claim about cause from a correlation measured at one point in time; disinvestment, housing, policing patterns, and more lurk behind both variables. **(e) fails**: failing to reject never proves the null. The audit habit: check every sentence for the *level* the data measure (places vs. people) and the *strength* the design supports (association vs. cause, no evidence vs. proof of none).
:::
