---
title: "Practice Problems · Session 3"
subtitle: "Probability, conditional thinking, and Bayes' theorem"
---

::: {.callout-note appearance="simple"}
**Ungraded self-checks.** Work each problem in your own R session *before* clicking the solution. These use different numbers and variables than the homework on purpose — if you can do these cold, the homework (and the midterm's probability section) will feel routine.
:::

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(705)
crimes <- read_csv("https://smourtgos.github.io/crju705/data/chicago-crimes-2025.csv",
                   show_col_types = FALSE)
```

All problems that use data need only:

```{r}
#| label: setup-shown
#| eval: false
library(tidyverse)
crimes <- read_csv("https://smourtgos.github.io/crju705/data/chicago-crimes-2025.csv")
```

## Problem 1 · Contraband in the long run

Suppose 8% of traffic stops turn up contraband. Simulate 10,000 nights of **10 stops each** with `rbinom()`. In what proportion of nights does the squad find contraband **2 or more times**?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p1
set.seed(705)
nights <- rbinom(n = 10000, size = 10, prob = 0.08)
mean(nights >= 2)
```

About 19% of nights. Note the move: simulate the process, then take a `mean()` of a logical condition — the same shape as every probability we computed in class.
:::

## Problem 2 · Build the table, read the margins

Using `crimes`, make a contingency table of `primary_type` × `arrest` for **just THEFT and BATTERY** incidents (`filter(primary_type %in% c("THEFT", "BATTERY"))`). Then compute: (a) the **joint** probability that a randomly chosen incident *from this subset* is a BATTERY **and** ended in arrest; (b) the **marginal** probability of BATTERY.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p2
tb <- crimes |> filter(primary_type %in% c("THEFT", "BATTERY"))

tb |>
  count(primary_type, arrest) |>
  pivot_wider(names_from = arrest, values_from = n)

# (a) joint — divide the cell by the subset's grand total:
mean(tb$primary_type == "BATTERY" & tb$arrest)

# (b) marginal:
mean(tb$primary_type == "BATTERY")
```

The joint divides by *everything in the subset* (both crime types, both arrest outcomes). If you divided the BATTERY-and-arrest count by the number of BATTERY rows, you computed a conditional — that's Problem 3.
:::

## Problem 3 · Both directions of the conditional

Still in the THEFT/BATTERY subset: compute **P(arrest | BATTERY)** and **P(BATTERY | arrest)**. Then, in one sentence, explain why the two numbers differ so much when they come from the same cells.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p3
tb <- crimes |> filter(primary_type %in% c("THEFT", "BATTERY"))

# P(arrest | BATTERY): shrink to BATTERY rows, take the mean
tb |> filter(primary_type == "BATTERY") |> summarize(p = mean(arrest))

# P(BATTERY | arrest): shrink to arrest rows, ask how many are BATTERY
tb |> filter(arrest) |> summarize(p = mean(primary_type == "BATTERY"))
```

Same cells, different **denominators**: the first re-proportions within battery incidents, the second within arrests. Direction matters — P(A | B) ≠ P(B | A) except by coincidence.
:::

## Problem 4 · Is time of day independent of arrest?

Create a logical variable `night = hour >= 20 | hour < 4` on the full `crimes` data. Compare P(arrest | night), P(arrest | not night), and the overall P(arrest). Closer to independent or dependent?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p4
crimes |>
  mutate(night = hour >= 20 | hour < 4) |>
  summarize(p_arrest = mean(arrest), n = n(), .by = night)

mean(crimes$arrest)
```

The conditional probabilities sit close to the marginal — knowing an incident happened at night moves the arrest probability only slightly. In these data, `night` is *nearly* independent of arrest (remember the definition: independence means P(A | B) = P(A)). Hedge accordingly: "essentially independent *in this sample*."
:::

## Problem 5 · The burglar alarm (Bayes' theorem)

A home-security company advertises that its alarm detects **99%** of break-ins. The alarm also goes off on **2%** of days with no break-in (cats, wind, deliveries). Suppose the daily probability of a break-in attempt on a given house is **0.5%** (5 in 1,000).

The alarm is ringing. Build the counting table for 10,000 days and compute **P(break-in | alarm)**.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p5
break_ins    <- 10000 * 0.005          # 50 break-in days
caught       <- break_ins * 0.99       # 49.5 -> alarm rings
quiet_days   <- 10000 - break_ins      # 9,950 days without a break-in
false_alarms <- quiet_days * 0.02      # 199 false rings

caught / (caught + false_alarms)       # P(break-in | alarm)
```

**About 20%.** Four out of five rings are cats and wind, because quiet days outnumber break-in days 199:1. The "99% accurate" figure is P(alarm | break-in) — the *transposed* conditional. Ignoring the 0.5% base rate is exactly the mistake the Ianfluenza example warned about.
:::

## Problem 6 · Verify it by simulation

Session 3's rule: don't take a probability claim on faith. Simulate 100,000 days of Problem 5's alarm and check your answer.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p6
set.seed(705)
days <- tibble(
  break_in = sample(c(TRUE, FALSE), 100000, replace = TRUE, prob = c(0.005, 0.995))
) |>
  mutate(alarm = if_else(
    break_in,
    sample(c(TRUE, FALSE), 100000, replace = TRUE, prob = c(0.99, 0.01)),
    sample(c(TRUE, FALSE), 100000, replace = TRUE, prob = c(0.02, 0.98))
  ))

days |> filter(alarm) |> summarize(p_break_in = mean(break_in), rings = n())
```

The long run lands where the counting table said it would — right around 20%.
:::
