---
title: "Practice Problems · Session 8"
subtitle: "ANOVA: two variances, the F-ratio, Tukey, and posterior differences"
---

::: {.callout-note appearance="simple"}
**Ungraded self-checks.** Work each problem in your own R session *before* clicking the solution. These use different data and numbers than the homework on purpose — if you can do these cold, Homework 8 (and the midterm's ANOVA questions) will feel routine.
:::

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(brms)
library(emmeans)
set.seed(705)
```

Every problem here runs on typed-in or built-in data, so the only setup you need is:

```{r}
#| label: setup-shown
#| eval: false
library(tidyverse)
library(brms)      # Problem 5 only — expect the usual compile pause
library(emmeans)
```

## Problem 1 · The pile of t-tests, quantified

Your department compares mean response times across **five patrol sectors**. A colleague proposes "keeping it simple": a separate t-test on every pair of sectors, each at the usual 5% false-alarm risk.

(a) How many pairwise tests is that? (`choose()` counts them.)
(b) If all five sectors truly have the **same** mean, what is the probability that at least one test comes up "significant" anyway?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p1
choose(5, 2)          # pairwise tests among 5 groups

1 - 0.95^choose(5, 2) # P(at least one false alarm across all of them)
```

Ten tests, and roughly a **40% chance** of at least one bogus "significant" difference when nothing differs at all. That's why ANOVA asks *one* omnibus question — are all five means plausibly equal? — at a single 5% risk, and saves the pairwise digging for an adjusted follow-up (Problem 4).
:::

## Problem 2 · Two variances, by hand

Dispatch pulls five response times (minutes) from each of three sectors — small on purpose, so you can see every number:

```{r}
#| label: p2-data
sectors <- tibble(
  minutes = c(7, 9, 8, 10, 8,       # Adams
              12, 14, 13, 15, 13,   # Baker
              8, 10, 9, 7, 9),      # Charlie
  sector  = factor(rep(c("Adams", "Baker", "Charlie"), each = 5))
)
```

Compute each sector's mean and variance, then the **within-group variance** (average of the three variances), the **between-group variance** (variance of the three means), and their ratio. What is the ratio telling you, in signal-and-noise terms?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p2
sector_stats <- sectors |>
  summarize(sector_mean = mean(minutes), sector_var = var(minutes), .by = sector)
sector_stats

within_var  <- mean(sector_stats$sector_var)   # noise: spread inside sectors
between_var <- var(sector_stats$sector_mean)   # signal: spread OF the means

within_var
between_var
between_var / within_var
```

The sector means (8.4, 13.4, 8.6) are about **6 times** more spread out than the chance bouncing inside sectors would produce — between dwarfs within, which is the signature of a real group difference. Between-group variance is the signal, within-group variance is the noise, and the ratio is your first, unscaled look at F.
:::

## Problem 3 · `aov()` and reading the F table

Run `aov(minutes ~ sector, ...)` on the same data and look at `summary()`. Then connect it to Problem 2: which table entry **is** your within-group variance? Why is `Mean Sq` for sector *not* 8.01? And what do F and p say?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p3
sector_model <- aov(minutes ~ sector, data = sectors)
summary(sector_model)
```

- **Mean Sq Residuals = 1.30** — the within-group variance from Problem 2, exactly.
- **Mean Sq sector = 40.07 = 5 × 8.01** — the between-group variance scaled by the **group size**. A 5-minute gap between means built on 5 calls each is stronger evidence than the same gap built on 2, so `aov()` bakes the sample size into the signal.
- **F(2, 12) = 30.8** — which is 5 × our hand ratio of 6.16 — with **p ≈ 0.00002**: an F this large essentially never happens when the three sectors truly share a mean.

The F table is Problem 2's two variances dressed up: same logic, with the group size folded into the signal.
:::

## Problem 4 · Real(ish) data: `warpbreaks` and Tukey

R ships `warpbreaks`: 54 industrial loom runs, recording how many times the yarn **broke** during each run under low, medium, or high `tension` (L/M/H, 18 runs each). Quality-control question: does mean breakage differ by tension setting?

Compute the group means, run the ANOVA, then `TukeyHSD()`. Which pairs differ? **One pair shows no reliable difference** — name it and report the interval that told you so.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p4
data("warpbreaks")

warpbreaks |>
  summarize(mean_breaks = mean(breaks), sd_breaks = sd(breaks), n = n(),
            .by = tension)

warp_model <- aov(breaks ~ tension, data = warpbreaks)
summary(warp_model)

TukeyHSD(warp_model)
```

**F(2, 51) = 7.2, p ≈ .002** — the omnibus F says *something* differs. Tukey says *what*: M − L (−10.0, interval [−19.6, −0.4], p ≈ .04) and H − L (−14.7, interval [−24.3, −5.2], p ≈ .001) both exclude 0 — low tension breaks more yarn than either alternative. But **H − M spans 0** ([−14.3, 4.8], p ≈ .46): no evidence medium and high differ. A significant omnibus F never means *every* pair differs — Tukey names the pairs while holding the family-wise false-alarm rate at 5%.
:::

## Problem 5 · The Bayesian version: `brm()` + `emmeans()` + `pairs()`

Fit the Bayesian twin of Problem 4's model (expect the *Compiling Stan program...* pause — that's normal, not a crash), then get the posterior pairwise differences. Which 95% HPD intervals **cross zero**, and does the Bayesian verdict match Tukey's?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p5-fit
#| output: false
warp_bayes <- brm(breaks ~ tension, data = warpbreaks,
                  chains = 2, iter = 2000, seed = 705, refresh = 0)
```

```{r}
#| label: p5-pairs
warp_post <- emmeans(warp_bayes, ~ tension)
warp_post

pairs(warp_post)
```

```{r}
#| label: p5-plot
#| fig-height: 2.5
plot(pairs(warp_post)) +
  geom_vline(xintercept = 0, linetype = 2, color = "firebrick")
```

L − M (≈ +10, HPD roughly [2, 18]) and L − H (≈ +14.7, roughly [7, 23]) **exclude 0** — credible differences. M − H (≈ +4.7, roughly [−3, 13]) **crosses 0** — zero remains a credible value, so no credible difference. Same verdict as Tukey, pair for pair. One nuance worth noticing: these Bayesian intervals are *per-comparison*, while Tukey deliberately widens its intervals to hold the *family-wise* rate — with only 3 contrasts the two land in the same place, but on borderline pairs they can disagree — worth watching for whenever you run both. When an HPD interval crosses zero, "no difference" is still on the table; when it doesn't, the difference is credible.
:::

## Problem 6 · Audit the memo

A crime-analyst memo about the `warpbreaks`-style analysis contains these four sentences. **Two are wrong.** Find them, say why, and fix them.

> (a) "The omnibus test was significant, F(2, 51) = 7.2, p = .002, so all three tension settings differ from one another."
>
> (b) "To be thorough, we also ran all three pairwise t-tests without adjustment and report those p-values instead of Tukey's."
>
> (c) "Tukey's HSD keeps the family-wise false-alarm rate at 5% across all three comparisons."
>
> (d) "Low tension averaged about 15 more breaks per run than high tension, 95% CI [5.2, 24.3]."

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p6
# Unadjusted pairwise p-values vs. Tukey's family-wise-adjusted ones:
pairwise.t.test(warpbreaks$breaks, warpbreaks$tension, p.adjust.method = "none")

TukeyHSD(aov(breaks ~ tension, data = warpbreaks))$tension[, "p adj"]
```

**(a) is wrong.** A significant omnibus F licenses only "at least one mean differs" — and Tukey showed H vs. M does *not* differ. Fix: "...so not all three tension means are equal; pairwise comparisons follow."

**(b) is wrong.** Reporting *unadjusted* pairwise p-values quietly triples the false-alarm exposure — Problem 1's pile-of-t-tests problem in miniature. Compare the two outputs above: every unadjusted p is smaller than its Tukey counterpart (e.g., M vs. L: .015 unadjusted vs. .038 adjusted) precisely because the unadjusted version ignores the other tests being run. Fix: report the Tukey-adjusted results.

**(c) and (d) are fine** — (d) is just Tukey's H − L row with the sign flipped, magnitude and interval intact. The concept to keep: per-comparison p-values and family-wise p-values answer different questions, and after a significant F, the family-wise one is the honest one.
:::
