---
title: "Practice Problems · Session 2"
subtitle: "Central tendency, dispersion, and the shapes of distributions"
---

::: {.callout-note appearance="simple"}
**Ungraded self-checks.** Work each problem in your own R session *before* clicking the solution. These use different numbers and variables than the homework on purpose — if you can do these cold, Homework 2 (and the midterm's descriptives section) will feel routine.
:::

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(modeest)
set.seed(705)
officers <- read_csv("https://smourtgos.github.io/crju705/data/officers.csv",
                     show_col_types = FALSE)
nbhd     <- read_csv("https://smourtgos.github.io/crju705/data/neighborhoods.csv",
                     show_col_types = FALSE)
```

All problems that use data need only:

```{r}
#| label: setup-shown
#| eval: false
library(tidyverse)
library(modeest)   # for mfv(); install.packages("modeest") once if needed
officers <- read_csv("https://smourtgos.github.io/crju705/data/officers.csv")
nbhd     <- read_csv("https://smourtgos.github.io/crju705/data/neighborhoods.csv")
```

## Problem 1 · Three kinds of typical

Response times (in minutes) for 10 priority calls one Friday night — nine routine, plus one call that sat in the queue through a shift change:

```r
response <- c(4, 5, 5, 6, 6, 6, 7, 8, 9, 42)
```

Compute the **mean**, **median**, and **mode**. Which of the three would you *not* want on the department's public dashboard, and why?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p1
response <- c(4, 5, 5, 6, 6, 6, 7, 8, 9, 42)

mean(response)
median(response)
mfv(response)
```

The one 42-minute call drags the mean to 9.8 minutes even though *no* routine call took longer than 9 — classic central-tendency behavior: the mean chases outliers, while the median (6) and mode (6) stay with the bulk of the data.
:::

## Problem 2 · Same mean, different story

Two detective units both average 30 open cases per detective — but look at the individual caseloads:

```r
unit_a <- c(18, 30, 42)
unit_b <- c(29, 30, 31)
```

Compute for each unit: the **range** (with `range()` and `diff()`), the **deviations from the mean**, the **sum of squared deviations**, and the **population variance and SD** by hand. Then run `var()` and `sd()` — why don't the built-ins match your hand calculations?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p2
unit_a <- c(18, 30, 42)
unit_b <- c(29, 30, 31)

diff(range(unit_a)); diff(range(unit_b))

unit_a - mean(unit_a)
unit_b - mean(unit_b)

# Sum of squares
sum((unit_a - mean(unit_a))^2)
sum((unit_b - mean(unit_b))^2)

# Population variance and SD (divide by n)
sum((unit_a - mean(unit_a))^2) / 3; sqrt(sum((unit_a - mean(unit_a))^2) / 3)
sum((unit_b - mean(unit_b))^2) / 3; sqrt(sum((unit_b - mean(unit_b))^2) / 3)

# R's built-ins
var(unit_a); sd(unit_a)
var(unit_b); sd(unit_b)
```

Identical centers, radically different **dispersion**: unit A's SD (~9.8 population / 12 sample) says workloads genuinely vary; unit B's (~1) says everyone carries the same load. The built-ins disagree with your hand math because R divides by n − 1, not n — Session 4 pays off that IOU.
:::

## Problem 3 · Audit the briefing note

A draft briefing memo about Problem 1's data contains two sentences:

> (a) "The typical priority call waited about 10 minutes for an officer."
>
> (b) "Half of priority calls had an officer on scene within 6 minutes."

One is defensible, one is misleading. Which is which — and rewrite the misleading one in a single honest sentence.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p3
response <- c(4, 5, 5, 6, 6, 6, 7, 8, 9, 42)
c(mean = mean(response), median = median(response))
```

Sentence (b) is defensible — it is exactly what the median (6) means. Sentence (a) launders the mean (9.8) into the word "typical," but *nine of ten* calls beat 10 minutes; only the queued outlier pushes the average there. An honest rewrite: "Half of priority calls had an officer within 6 minutes; one call held during a shift change took 42." With skewed data, "typical" belongs to the median — and outliers deserve their own sentence, not a hiding place inside the mean.
:::

## Problem 4 · Real data: how much do officers sleep?

Using `officers`, summarize `sleep_hours` — mean, median, SD, min, max — and make a histogram (`binwidth = 0.5`). Is the distribution roughly symmetric or skewed? How do you know from the *numbers alone*?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p4
#| fig-height: 3.5
officers |>
  summarize(mean_sleep = mean(sleep_hours), median_sleep = median(sleep_hours),
            sd_sleep = sd(sleep_hours), min = min(sleep_hours), max = max(sleep_hours))

ggplot(officers, aes(sleep_hours)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "white") +
  labs(x = "Sleep per night (hours)", y = "Officers")
```

Mean (6.40) and median (6.4) are essentially identical — the numeric signature of a roughly **symmetric distribution**, which the histogram confirms. (Compare Problem 1, where mean and median disagreed badly: that gap *is* the skew detector.)
:::

## Problem 5 · The empirical rule, checked against reality

If `sleep_hours` were perfectly normal, about **68%** of officers would fall within ±1 SD of the mean and about **95%** within ±2 SD. Compute the *actual* shares. (Hint: `mean(abs(x - m) <= s)` — the same mean-of-a-logical move as always.)

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p5
m <- mean(officers$sleep_hours)
s <- sd(officers$sleep_hours)

mean(abs(officers$sleep_hours - m) <= s)       # within +/- 1 SD
mean(abs(officers$sleep_hours - m) <= 2 * s)   # within +/- 2 SD
```

About 69.5% and 95.5% — within a couple points of the **empirical rule's** 68/95 benchmarks. That is what "the SD tells you how typical or extreme a value is" means in practice: for bell-shaped data, mean ± 2 SD brackets nearly everyone.
:::

## Problem 6 · A Poisson in the wild

Using `nbhd` (500 neighborhoods), make a bar chart of `assaults` (last year's count per neighborhood), then compute its **mean and variance**. Now simulate 500 draws from a Poisson with the same λ — `rpois(500, lambda = mean(nbhd$assaults))` — and compare the two count tables. Does the textbook distribution fit the real one?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p6
#| fig-height: 3.5
ggplot(nbhd, aes(assaults)) +
  geom_bar(fill = "grey40") +
  labs(x = "Assaults last year", y = "Neighborhoods")

nbhd |> summarize(mean_assaults = mean(assaults), var_assaults = var(assaults))

set.seed(705)
table(nbhd$assaults)
table(rpois(500, lambda = mean(nbhd$assaults)))
```

The broad shape matches — right-skewed counts of a rare event, exactly the **Poisson's** territory. But look closer: the real data have *more zeros* (219 vs ~148) *and* a longer tail (counts up to 7 vs 5), and the variance (1.69) exceeds the mean (1.07), where a true Poisson would have them equal. Real CJ counts are usually messier than the textbook distribution — a mismatch called overdispersion that you'll meet again later in the course.
:::
