---
title: "Practice Problems · Session 11"
subtitle: "Multiple regression: slopes with units, R², lurking variables, and diagnostics"
---

::: {.callout-note appearance="simple"}
**Ungraded self-checks.** Work each problem in your own R session *before* clicking the solution. These use different data and variables than the homework on purpose — if you can do these cold, Homework 11 (and Checkpoint 3's first regression on your own data) will feel routine.
:::

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(brms)
library(car)
set.seed(705)
cities <- read_csv("https://smourtgos.github.io/crju705/data/cities-wide.csv",
                   show_col_types = FALSE)
```

Problems that use data need only this setup — the same 100 cities as the Session 10 practice bank, where poverty and the violent-crime rate ran r = .61. Correlation said the two travel together; regression asks **how much**, and what a second predictor changes. One row = one **city** — every sentence you write stays at that level.

```{r}
#| label: setup-shown
#| eval: false
library(tidyverse)
library(brms)   # Problem 5 only — expect the usual compile pause
library(car)    # for vif()

cities <- read_csv("https://smourtgos.github.io/crju705/data/cities-wide.csv")
```

## Problem 1 · One predictor: make the slope speak units

Regress `vcr_2025` (violent crimes per 100,000 residents) on `poverty_rate` (percentage points). Interpret the slope **with units**, say what the intercept means here (and whether any real city lives there), and connect R² to Session 10's r = .61.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p1
city_m1 <- lm(vcr_2025 ~ poverty_rate, data = cities)
summary(city_m1)

cor(cities$poverty_rate, cities$vcr_2025)^2
```

- **Slope ≈ 14.5:** each additional percentage point of city poverty is associated with about **14.5 more violent crimes per 100,000 residents**, in this sample. Units on both ends or the number means nothing.
- **Intercept ≈ 110:** the predicted rate for a city with 0% poverty — no city in the data comes close (the minimum is 4.4%), so it anchors the line rather than describing anywhere real.
- **R² = .37**, which is exactly .61² — with one predictor, R² is last session's correlation squared: poverty accounts for about 37% of the city-to-city variance in violent-crime rates, leaving 63% to everything else.

A slope is a *rate* with units (outcome units per predictor unit); R² is the share of outcome variance the model captures — two different questions about the same line.
:::

## Problem 2 · A second predictor: does it earn its keep?

Cities differ in police staffing too. Add `police_per_10k` (sworn officers per 10,000 residents) to the model. Compare the poverty slope across the two models, interpret the staffing coefficient with its 95% CI, and check what happens to **adjusted** R². Then write the honest one-sentence conclusion about staffing.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p2
city_m2 <- lm(vcr_2025 ~ poverty_rate + police_per_10k, data = cities)
summary(city_m2)
confint(city_m2)

c(adj_m1 = summary(city_m1)$adj.r.squared,
  adj_m2 = summary(city_m2)$adj.r.squared)
```

The poverty slope barely moves (14.46 → 14.59) — poverty and staffing are nearly uncorrelated here (r ≈ .05), so there was almost no shared variance to reapportion. The staffing coefficient is **−3.2 violent crimes per 100k per additional officer per 10k residents, 95% CI [−7.5, 1.1], p = .14**: the interval spans zero, and adjusted R² ticks up only trivially (.364 → .371). The honest sentence: "Holding poverty constant, police staffing showed no reliable association with the violent-crime rate in these cities." A new predictor has to earn its coefficient — a CI straddling zero and a flat adjusted R² is the data declining the offer, and *saying so* is a legitimate finding.
:::

## Problem 3 · The lurking third variable, caught in the act

Across 90 towns, an analyst finds that towns with more **licensed pawn shops** report far more **burglaries**, and drafts a memo proposing pawn-shop license caps. Simulate the analyst's data — note what actually generates both variables:

```{r}
#| label: p3-data
set.seed(705)   # inline seed: keeps this particular draw stable
towns <- tibble(
  population = runif(90, 10, 120),                        # thousands of residents
  pawn_shops = 1 + 0.09 * population + rnorm(90, 0, 1.4),
  burglaries = 30 + 5.5 * population + rnorm(90, 0, 55)   # per year
)
```

(a) Fit `burglaries ~ pawn_shops`. How does the coefficient look — and does anything in the output warn you it's misleading?
(b) Now add `population`. What happens to the pawn-shop coefficient, and what does that tell the memo's author?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p3
lm(burglaries ~ pawn_shops, data = towns) |> summary() |> coef() |> round(3)

lm(burglaries ~ pawn_shops + population, data = towns) |> summary() |> coef() |> round(3)
```

(a) About **48 more burglaries per additional pawn shop**, p < .001 by a mile — and *nothing in the output warns you*. The model can only see the variables you gave it.

(b) Holding population constant, the pawn-shop coefficient collapses to ≈ 1.4 with **p ≈ .68** — nothing — while population comes in at ≈ 5.3 burglaries per additional thousand residents. Bigger towns simply have more of *both*. That collapse is "holding constant" doing real work: comparing towns of the same size, pawn-shop counts tell you nothing about burglary. And the sobering corollary: regression can only hold constant the lurkers you **measured and included** — the ones you didn't are still at large.
:::

## Problem 4 · Diagnostics: the residual plot and VIF

Back to the real model, `city_m2` from Problem 2. Make the residuals-vs-fitted plot and run `vif()`. For each: what is it checking, and what does it show *here*?

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p4
#| fig-height: 3.5
tibble(fitted = fitted(city_m2), residual = resid(city_m2)) |>
  ggplot(aes(fitted, residual)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point(alpha = 0.7, color = "grey30") +
  labs(x = "Fitted violent-crime rate", y = "Residual")

vif(city_m2)
```

**The residual plot** checks whether the model's misses are patternless: no curve (the straight-line shape isn't visibly wrong) and roughly even spread across the fitted range. This one is what healthy looks like — a shapeless band centered on zero, no bend, no widening, with the biggest misses (reaching roughly ±400) scattered across the range rather than piling up at one end. Compare Lab 11's Chicago plot, where the spread visibly widened at high fitted values; that's the kind of pattern this plot exists to catch.

**VIF** checks a different disease — predictor redundancy. Both values sit at ≈ 1.00 (poverty and staffing share almost nothing, r ≈ .05), so each coefficient is estimated on its own information. The residual plot audits the *misses*; VIF audits the *predictors* — one model, two separate health checks.
:::

## Problem 5 · Both traditions, one sentence each

Fit the Bayesian twin of `city_m2` with `brm()` (expect the *Compiling Stan program...* pause — normal, not frozen), look at the posterior summaries, and then write the pair of reporting sentences: one frequentist (estimate, CI, p) and one Bayesian (the credible-interval sentence you're allowed to say) about the **poverty** coefficient.

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p5-fit
#| output: false
city_bayes <- brm(vcr_2025 ~ poverty_rate + police_per_10k, data = cities,
                  chains = 2, iter = 2000, seed = 705, refresh = 0)
```

```{r}
#| label: p5-fixef
fixef(city_bayes) |> round(2)
```

The posterior lands almost exactly on `lm()`'s estimates — expected with `brms` default priors (say so when you report: "brms default priors, 2 chains × 2,000 iterations, seed 705") and 100 cities of data.

> **Frequentist:** "Holding police staffing constant, each additional percentage point of poverty was associated with about 14.6 more violent crimes per 100,000 residents (b = 14.59, 95% CI [10.8, 18.4], p < .001)."

> **Bayesian:** "Given the data and priors, there is a 95% probability the poverty coefficient lies between roughly 10.9 and 18.3 — zero is not a credible value."

Same formula, nearly identical numbers — what changes is the sentence each tradition licenses, and the Bayesian one makes the direct probability statement the confidence interval never could.
:::

## Problem 6 · The sentence audit

Four sentences drafted from the Problem 2 model. **Two are defensible; two are not.** Sort them, and fix the broken ones.

> (a) "Holding police staffing constant, each additional percentage point of poverty was associated with about 14.6 more violent crimes per 100,000 residents (95% CI [10.8, 18.4])."
>
> (b) "Poverty drives up violent crime in American cities."
>
> (c) "R² = .38, so the model correctly predicts 38% of cities."
>
> (d) "Police staffing showed no reliable association with the violent-crime rate in these cities (b = −3.2, 95% CI [−7.5, 1.1])."

::: {.callout-tip collapse="true"}
**Solution**

```{r}
#| label: p6
confint(city_m2) |> round(1)
summary(city_m2)$r.squared
```

**Defensible: (a) and (d).** Both state an *adjusted association*, with magnitude, units, uncertainty, and the holding-constant clause — and (d) shows that "no reliable association" is a reportable result, not a failure.

**(b) fails twice:** "drives up" is a causal verb no cross-sectional model earns (lurkers we never measured are still at large — Problem 3's lesson), and "American cities" generalizes a 100-city sample to a population the data can't speak for. Fix: "In these 100 cities, higher poverty was associated with higher violent-crime rates, holding staffing constant."

**(c) misreads R²:** .38 is the **share of variance in the outcome** the model accounts for — not an accuracy rate, and not a count of cities predicted "correctly." Fix: "The two predictors together account for about 38% of the city-to-city variance in violent-crime rates." The audit habit: check every regression sentence for causal verbs, level of analysis, units, and what R² actually is.
:::
