---
title: "The Logic of Inference: Confidence Intervals"
subtitle: "CRJU 705 · Session 5 · Fall 2026"
author: "Scott M. Mourtgos"
format:
  revealjs:
    slide-number: true
    preview-links: auto
    chalkboard: true
    footer: "CRJU 705 · Session 5"
    theme: [default, custom-slides.scss]
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(ggrepel)
library(HistData)
source(here::here("R/setup.R"))
theme_set(theme_minimal(base_size = 18))
set.seed(705)
crimes <- read_csv(here::here("data/chicago-crimes-2025.csv"), show_col_types = FALSE)
```

## This Week

**Topic:** The logic of inference, using confidence intervals

- What statistical *inference* actually is (and isn't)
- Point estimates and why they're not enough
- Confidence intervals: building them, reading them, and — hardest of all — *saying what they mean out loud without lying*

**Learning goals:** explain in your own words what a 95% confidence interval does and does **not** mean; compute CIs for a proportion and a difference in means from real data.

**R4DS thread:** you read Ch. 7 (Data import) & Ch. 8 (Getting help) — today's lab puts both to work.

## Where We've Been

::: incremental
- **Session 3:** probability — the rules randomness follows
- **Session 4:** sampling distributions — take many samples, and the sample statistics themselves form a distribution with predictable behavior
- **Today:** we cash that in. If we know how sample statistics *behave*, we can attach an honest statement of **uncertainty** to a single sample's estimate.
:::

# The Logic of Inference {background-color="#73000A"}

## What Does "Inference" Mean?

**Inference:** a reasoning process that begins with some information and leads to some conclusion.

**Inductive inference:** reasons from *specific cases* to the *more general* — a cat jumps from a tree and lands on its feet; you infer something about cats generally.

**Statistical inference:** dealing systematically with situations where we have *uncertain or incomplete* information. Conclusions drawn inductively from samples are **never fixed or firm**.

. . .

::: {.callout-important appearance="simple"}
You **cannot PROVE** anything from samples or by using statistical inference.
:::

## If We Can't Prove Anything, What Are We Doing?

We collect data and analyze it with inferential statistics **to build up a weight of evidence that influences our certainty about one conclusion or another**.

. . .

**Let's look at the classic example:**

- 1854 — cholera breaks out in Soho, London
- The debate: bad air ("miasma") or contaminated water?
- John Snow couldn't test *every household in London* — that's the whole population
- So he mapped the **sample of evidence** he could get: deaths, and where people got their water

## Sampled Evidence: Deaths and Pumps

```{r}
#| label: snow-setup
#| output-location: slide
#| code-line-numbers: "1-2|4-8|10-16"
#| fig-height: 6
deaths <- as_tibble(Snow.deaths) |> mutate(did = row_number())
pumps  <- as_tibble(Snow.pumps)  |> mutate(pid = row_number())

# Every death-pump distance, then each death's NEAREST pump
pairs <- crossing(did = deaths$did, pid = pumps$pid) |>
  left_join(deaths |> select(did, dx = x, dy = y), by = "did") |>
  left_join(pumps  |> select(pid, px = x, py = y), by = "pid") |>
  mutate(dist = sqrt((dx - px)^2 + (dy - py)^2))

nearest <- pairs |>
  slice_min(dist, n = 1, by = did)

by_pump <- nearest |>
  count(pid, name = "deaths_nearest") |>
  arrange(desc(deaths_nearest))
by_pump
```

## Sampled Evidence: The Map

```{r}
#| label: snow-map
#| echo: false
#| fig-height: 6.5
top_pid <- by_pump$pid[1]
broad   <- pumps |> filter(pid == top_pid)
radius  <- 1.0

circle <- tibble(
  x = broad$x + radius * cos(seq(0, 2 * pi, length.out = 360)),
  y = broad$y + radius * sin(seq(0, 2 * pi, length.out = 360))
)
n_within <- deaths |>
  mutate(d = sqrt((x - broad$x)^2 + (y - broad$y)^2)) |>
  summarize(n = sum(d <= radius)) |>
  pull(n)
pumps_lab <- pumps |>
  mutate(label = if_else(pid == top_pid, "Broad St. (most deaths)", paste0("Pump ", pid)))

ggplot() +
  geom_point(data = deaths, aes(x, y), color = crju_colors["garnet"], alpha = 0.6, size = 1.8) +
  geom_point(data = pumps, aes(x, y), color = crju_colors["navy"], size = 3, shape = 17) +
  geom_path(data = circle, aes(x, y), linetype = 2) +
  geom_label_repel(data = pumps_lab, aes(x, y, label = label), size = 3) +
  annotate("label", x = broad$x, y = broad$y - 0.8,
           label = paste0(n_within, " deaths within r = ", radius), fill = "white") +
  coord_equal() +
  labs(title = "Cholera deaths cluster around a single pump",
       subtitle = "Dotted circle: chosen radius around the pump with the most nearby deaths",
       x = NULL, y = NULL)
```

## Sampled Evidence: Broad Street Accumulates Deaths Fastest

```{r}
#| label: snow-cumulative
#| echo: false
#| fig-height: 5.5
r_seq <- seq(0, 2.5, by = 0.1)

cum_for_pump <- function(px, py) {
  d <- sqrt((deaths$x - px)^2 + (deaths$y - py)^2)
  tibble(radius = r_seq, cum_deaths = map_int(r_seq, \(r) sum(d <= r)))
}

cum_broad <- cum_for_pump(broad$x, broad$y) |>
  rename(`Broad St. (culprit)` = cum_deaths)

cum_others <- pumps |>
  filter(pid != top_pid) |>
  mutate(cum = map2(x, y, cum_for_pump)) |>
  select(pid, cum) |>
  unnest(cum) |>
  summarize(`Median other pump` = median(cum_deaths), .by = radius)

cum_broad |>
  left_join(cum_others, by = "radius") |>
  pivot_longer(-radius, names_to = "series", values_to = "cum_deaths") |>
  ggplot(aes(radius, cum_deaths, color = series)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = unname(crju_colors[c("garnet", "gray")])) +
  labs(title = "Cumulative cholera deaths as we widen the radius around each pump",
       x = "Radius around pump", y = "Cumulative deaths", color = NULL)
```

## Why This Is Statistical Inference

- Why wasn't Snow just *describing* the map?
- What **bigger claim** was he making about the whole city — and about *cholera itself*?

**That jump — from the data you have to the world you don't — is what we mean by inference.**

. . .

**Review from previous weeks — why do we sample?**

We almost never have access to the totality of the population data. At its *best*, a sample is **representative** of a population.

# Point Estimates and Their Discontents {background-color="#1B2A4A"}

## A Rare Luxury: We Know the Truth

Our Chicago file holds ~30,000 incidents from 2025. Today, *pretend the file is the whole population* — then we can check our inferences against the truth.

```{r}
#| label: pop-truth
crimes |> summarize(pop_arrest_rate = mean(arrest), n = n())
```

The "true" citywide arrest proportion is **16.1%**. Lock that in — in real research you never get to see this number.

## Now Act Like a Researcher: Draw One Sample

A researcher can rarely afford all 30,000 case files. Suppose we pull a random sample of **250** incidents:

```{r}
#| label: one-sample
sample_250 <- crimes |> slice_sample(n = 250)

sample_250 |> summarize(sample_arrest_rate = mean(arrest))
```

. . .

This is a **point estimate** — the sample's best single guess at the population value.

**14.8% vs the true 16.1%.** Close-ish. But how good is this estimate? How *certain* are we?

## Caution! Samples Fluctuate

Draw a fresh sample of 250 and the estimate moves. Do it 1,000 times:

```{r}
#| label: many-samples
#| output-location: slide
#| fig-height: 5
many_estimates <- map_dbl(1:1000, \(i) {
  crimes |> slice_sample(n = 250) |> summarize(p = mean(arrest)) |> pull(p)
})

ggplot(tibble(p = many_estimates), aes(p)) +
  geom_histogram(bins = 30, fill = crju_colors["blue"], color = "white") +
  geom_vline(xintercept = mean(crimes$arrest),
             color = crju_colors["garnet"], linewidth = 1.2) +
  labs(title = "1,000 sample estimates of the arrest rate (n = 250 each)",
       subtitle = "Garnet line = the true population value",
       x = "Sample arrest proportion", y = "Number of samples")
```

## What the Fluctuation Tells Us

```{r}
#| label: quantile-bridge
quantile(many_estimates, c(0.025, 0.975))   # what is this telling us?
```

::: incremental
- 95% of sample estimates landed between about **12% and 21%** — that's the *sampling distribution* from Session 4, live
- Any *single* sample's estimate could easily be a couple points off the truth
- So reporting "14.8%" alone is **overconfident** — we owe our reader a range that reflects the fluctuation
- Problem: this required 1,000 samples and a known population. A real researcher has **one sample**. Now what?
:::

# The Confidence Interval {background-color="#73000A"}

## Our First Inferential Tool

The **confidence interval** turns *one* sample into an honest range, using what Session 4 taught us about how sample statistics behave:

$$\text{estimate} \; \pm \; \underbrace{1.96 \times SE}_{\text{margin of error}}$$

- **SE** (standard error) = the standard deviation of the sampling distribution, *estimated from the one sample you have*
- **1.96** comes from the normal distribution: ±1.96 SDs covers the middle 95%

Bigger samples → smaller SE → narrower interval.

## R Does This For Us: `prop.test()`

```{r}
#| label: prop-test
prop.test(sum(sample_250$arrest), n = 250)
```

. . .

Our one sample of 250 says: point estimate **14.8%**, 95% CI **10.8% to 20.0%**. The true value (16.1%) is inside — this time. But *"95% confident"* of what, exactly?

## The Classic Picture: 100 Samples, 100 Intervals

```{r}
#| label: ci-sim
#| output-location: slide
#| code-line-numbers: "1-7|9-10|12-18"
#| fig-height: 6
ci_100 <- map(1:100, \(i) {
  s  <- crimes |> slice_sample(n = 250)
  ci <- prop.test(sum(s$arrest), n = 250)$conf.int
  tibble(sample = i, estimate = mean(s$arrest), lo = ci[1], hi = ci[2])
}) |>
  list_rbind() |>
  mutate(captured = lo <= mean(crimes$arrest) & mean(crimes$arrest) <= hi)

sum(ci_100$captured)   # how many of the 100 intervals caught the truth?

ggplot(ci_100, aes(sample, estimate, color = captured)) +
  geom_hline(yintercept = mean(crimes$arrest), color = "grey30", linewidth = 1) +
  geom_pointrange(aes(ymin = lo, ymax = hi), linewidth = 0.7, size = 0.25) +
  scale_color_manual(values = c("TRUE" = "grey65", "FALSE" = "#73000A")) +
  labs(title = paste0("100 samples, 100 confidence intervals — ",
                      sum(ci_100$captured), " capture the truth"),
       subtitle = "Horizontal line = true population arrest rate; garnet = intervals that missed",
       x = "Sample number", y = "Arrest proportion", color = "Captured?")
```

## What Just Happened

::: incremental
- Each of the 100 samples produced a *different* interval — the interval is a **random quantity**, the population value is **fixed**
- **`r sum(ci_100$captured)` of our 100** intervals captured the true arrest rate; `r 100 - sum(ci_100$captured)` missed (the garnet ones)
- In the long run, the 95% procedure captures the truth in **~95% of replications** — that long-run capture rate is *the entire meaning* of "95% confident"
- The researchers who drew the garnet samples did nothing wrong — and their data give **no warning** that theirs are the unlucky intervals
:::

## In Practice

If we reran our whole study many times — drawing a new sample each time and constructing a new confidence interval each time — **in 95% of those replications the interval would contain the true population value.**

::: {.callout-important appearance="simple"}
**IMPORTANT:**

1. Our 95% CI does **not** necessarily contain the true population value
2. Our 95% CI is **not** a statement about the probability that *this particular* interval is correct
3. It is a **long-run prediction** based on numerous replications
:::

. . .

**WTF?!?** — yes, everyone reacts this way. Point at the 100-intervals plot until it clicks.

## Saying It Out Loud

✅ **Defensible:**

> "We estimate the citywide arrest rate at 14.8% (95% CI: 10.8% to 20.0%). Values in this range are consistent with our data."

> "The method we used captures the true value 95% of the time."

❌ **Not defensible:**

> "There is a 95% probability the true arrest rate is between 10.8% and 20.0%." *(The truth is fixed — it's either in there or it isn't.)*

> "95% of arrest rates fall in this interval." *(The CI is about the population parameter, not individual observations.)*

## Sample Size Buys Precision

Same procedure, three sample sizes, one draw each:

```{r}
#| label: width-by-n
#| output-location: fragment
map(c(100, 250, 1000, 4000), \(n) {
  s  <- crimes |> slice_sample(n = n)
  ci <- prop.test(sum(s$arrest), n = n)$conf.int
  tibble(n = n, estimate = mean(s$arrest),
         lo = ci[1], hi = ci[2], width = ci[2] - ci[1])
}) |>
  list_rbind()
```

. . .

Quadrupling *n* roughly **halves** the width — precision grows with $\sqrt{n}$, so it gets expensive fast.

# Comparing Two Groups {background-color="#1B2A4A"}

## A Real Question: Do Violent Crimes Happen Later in the Day?

Street lore says violence is a nighttime phenomenon. Property crime, the daytime hustle. Our data:

```{r}
#| label: two-group-est
vp <- crimes |> filter(crime_category %in% c("Violent", "Property"))

vp |> summarize(mean_hour = mean(hour), sd_hour = sd(hour),
                n = n(), .by = crime_category)
```

. . .

Two **point estimates**: violent crimes average hour 12.8, property 12.5. Is that ~15-minute gap real, or sample fluctuation?

## Let's Look at It Visually

```{r}
#| label: two-group-box
#| fig-height: 5
ggplot(vp, aes(crime_category, hour, fill = crime_category)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = unname(crju_colors[c("blue", "garnet")])) +
  labs(x = NULL, y = "Hour of day (0–23)")
```

The boxes nearly coincide — the eye can't settle this one. We need inference.

## Resample the Difference

Same trick as before — resample each group, compute the difference in means, repeat:

```{r}
#| label: mean-diffs
#| output-location: slide
#| fig-height: 5
violent_hours  <- vp$hour[vp$crime_category == "Violent"]
property_hours <- vp$hour[vp$crime_category == "Property"]

mean_diffs <- replicate(1000, {
  mean(sample(violent_hours,  length(violent_hours),  replace = TRUE)) -
  mean(sample(property_hours, length(property_hours), replace = TRUE))
})

quantile(mean_diffs, c(0.025, 0.975))   # what is this telling us?

ggplot(tibble(mean_diffs), aes(mean_diffs)) +
  geom_histogram(bins = 30, fill = crju_colors["navy"], color = "white") +
  geom_vline(xintercept = 0, color = crju_colors["garnet"], linewidth = 1.2) +
  labs(title = "1,000 resampled differences in mean hour (Violent − Property)",
       x = "Difference in mean hour", y = "Count")
```

## The Two-Sample t-Test Gives the CI Directly

```{r}
#| label: t-test
t.test(hour ~ crime_category, data = vp)
```

Watch the direction: R orders groups alphabetically, so this tests **Property − Violent** — the mirror image of our resampled Violent − Property differences. Same finding, flipped sign.

## Reading That Output Honestly

```{r}
#| label: t-test-ci
t.test(hour ~ crime_category, data = vp)$conf.int
```

::: incremental
- The 95% CI for the difference (Property − Violent) runs from **−0.44 to −0.06 hours** — it excludes zero, so the data are consistent with violent crimes genuinely averaging later
- But look at the *magnitude*: the difference is somewhere between about **4 and 27 minutes**. Detectable ≠ dramatic.
- With ~19,500 incidents, we can detect tiny differences. The CI hands you both facts at once: *direction* (probably real) and *size* (small). **Never report the first without the second.**
- This is your vaccination against next week's topic — a "significant" result can be substantively trivial
:::

## Your Turn (5 minutes, with a neighbor)

Using `crimes`, compute a 95% CI for the proportion of incidents flagged as **domestic**:

```r
crimes |> summarize(mean(domestic))          # point estimate
prop.test(sum(crimes$domestic), nrow(crimes))
```

1. Write one **defensible** sentence reporting your interval.
2. Trade sentences with your neighbor and audit each other: does the sentence claim a probability about *this* interval? If so, fix it.

## Today's Lab & Homework

- **Lab 5:** importing real data — `read_csv()`, `read_excel()`, parsing dates, and what to do when R yells at you ([lab page](../labs/lab-05-import.qmd))
- **Homework 5:** confidence intervals on Alaska crash data (posted on the [Session 5 page](../weeks/week-05.qmd))
- **Read before Session 6:** Stanton Ch. 5 · R4DS Ch. 19 (Joins) & skim Ch. 18 (Missing values)
