---
title: "Correlation, Then Midterm Review"
subtitle: "CRJU 705 · Session 8 · Fall 2026"
author: "Scott M. Mourtgos"
format:
  revealjs:
    slide-number: true
    preview-links: auto
    chalkboard: true
    footer: "CRJU 705 · Session 8"
    theme: [default, custom-slides.scss]
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(patchwork)
library(tidyverse)
source(here::here("R/setup.R"))
source(here::here("R/bayes.R"))
options(scipen = 0)   # print p-values the way students' consoles do (2.2e-16)
theme_set(theme_minimal(base_size = 18))
set.seed(705)
areas   <- read_csv(here::here("data/chicago-areas.csv"), show_col_types = FALSE)
crashes <- read_csv(here::here("data/crash-ak.csv"), show_col_types = FALSE) |>
  mutate(intox = factor(if_else(intox_any == 1, "Intoxicated", "Sober"),
                        levels = c("Intoxicated", "Sober")))
stops   <- read_csv(here::here("data/mpls-stops.csv"), show_col_types = FALSE)
```

## This Week

**Topic:** one new tool, then we get ready for next week's midterm.

- **First half:** correlation. Both variables are numbers, no groups. Do two measures **move together**?
- **Lab 8:** you run the correlation recipe yourself
- **Second half:** midterm review, part by part

**Learning goals:** read r's sign and strength; test it with `cor.test()`; explain why a correlation between **areas** says nothing about **people**.

## How Today Works

:::: columns
::: {.column width="33%"}
**Lecture**

Correlation: what r is, what it cannot see, and how to test it
:::

::: {.column width="33%"}
**Lab 8**

The recipe on two datasets, and you write the code
:::

::: {.column width="33%"}
**Midterm review**

What's on the exam, and one practice problem per part
:::
::::

Everything I run tonight is written up in the [demo walkthrough](../demos/demo-10-correlation.qmd).

## Which Test? One New Row

| The outcome is... | Compared across... | Test |
|---|---|---|
| **yes/no** | two groups | `prop.test()` and `bayes_prop_test()` |
| **a number** | two groups | `t.test(outcome ~ group, data = )` |
| **a number** | **another number** | `cor.test(x, y)` |

. . .

Sessions 5 to 7 were all about groups. Tonight every row has two numbers: do they rise and fall together? This is the doorway to regression (Sessions 10 and 11).

# A New Kind of Question {background-color="#73000A"}

## What Does Correlation Tell Us?

When two variables move together, how can we describe that relationship?

- One goes up when the other goes up: **positive correlation**
- One goes up when the other goes down: **negative correlation**
- Changes in one don't track changes in the other: **no correlation**

. . .

Direction is only half the story. The **strength** of the relationship tells how tightly the changes are linked. That's the number we build today.

## Which Is Which?

Positive, negative, or nothing? Make your call for each panel:

```{r}
#| label: guess-panels
#| echo: false
#| fig-height: 4.5
set.seed(705)   # inline seed: keeps these particular clouds (and their r values) stable
guess <- bind_rows(
  tibble(panel = "A", x = 1:30, y = 30:1 + rnorm(30, 0, 3)),
  tibble(panel = "B", x = 1:30, y = 1:30 + rnorm(30, 0, 3)),
  tibble(panel = "C", x = 1:30, y = rnorm(30, 15, 8))
)

ggplot(guess, aes(x, y)) +
  geom_point(size = 2.5, color = crju_colors["navy"]) +
  facet_wrap(~panel) +
  labs(x = "Variable X", y = "Variable Y")
```

## Which Is Which? Revealed

```{r}
#| label: guess-reveal
#| echo: false
#| fig-height: 4.5
guess_r <- guess |>
  summarize(r = cor(x, y), .by = panel) |>
  mutate(label = paste0(panel, ":  r = ", sprintf("%.2f", r)))

guess |>
  left_join(guess_r, by = "panel") |>
  ggplot(aes(x, y)) +
  geom_point(size = 2.5, color = crju_colors["navy"]) +
  geom_smooth(method = "lm", se = FALSE, color = crju_colors["garnet"]) +
  facet_wrap(~label) +
  labs(x = "Variable X", y = "Variable Y")
```

. . .

The lines make it obvious, and the r values put a number on what your eye already knew.

## Real-World Examples

- **Positive:** hours studied ↑, exam scores ↑
- **Negative:** officers on shift ↑, response times ↓
- **Near zero:** shoe size and GPA

. . .

**Your turn:** give me a CJ example of each, *and* one pair you suspect people **believe** is correlated but isn't.

## Correlation Does NOT Mean Causation

::: incremental
- Correlation shows **association**, not cause and effect
- Ice cream sales and violent crime rise together every year. Does ice cream cause violence?
- Two variables can move together because of a **third variable** (summer heat drives both), or by pure coincidence
- When you see a correlation, ask: does X cause Y? Y cause X? Does something else drive both? Or is it luck?
- Correlation is a **pattern**, not proof of cause. **Beware the lurking third variable!**
:::

# From Covariance to r {background-color="#1B2A4A"}

## Covariance: Moving Together, in Raw Units

Row by row: do x and y sit on the *same side* of their averages?

$$\text{cov}(x, y) \;=\; \frac{\sum_i (x_i - \bar{x})(y_i - \bar{y})}{n - 1}$$

::: incremental
- Both above average (or both below): positive products, **positive covariance**. Opposite sides: **negative covariance**
- The **sign** is trustworthy: covariance tells you *whether* two variables move together
- The **size** is not. Watch what happens on real data
:::

## The Units Problem

Two covariances with violent crime, from Chicago's 77 community areas:

```{r}
#| label: cov-units
cov(areas$unemployment_rate, areas$violent_rate)
cov(areas$median_income,     areas$violent_rate)
```

. . .

Is -482,660 a "stronger" relationship than 108? No idea. The first number is in *dollars times crimes per 1,000*; the second in *percentage points times crimes per 1,000*. **Covariance is glued to its units**, so you can't compare across pairs.

## Correlation Rescales It

Divide the covariance by both standard deviations and the units cancel:

$$r \;=\; \frac{\text{cov}(x, y)}{s_x \, s_y}$$

```{r}
#| label: r-standardize
cov(areas$unemployment_rate, areas$violent_rate) /
  (sd(areas$unemployment_rate) * sd(areas$violent_rate))

cor(areas$unemployment_rate, areas$violent_rate)   # same thing, one function
```

. . .

Always between **-1 and +1**, no units, comparable. Income runs r = -0.71 with violence, so unemployment (0.84) tracks violence more tightly than income does, direction aside.

## Reading r {.smaller}

| r (either sign) | Strength | What the scatterplot looks like |
|---|---|---|
| 0.8 to 1.0 | Very strong | Dots hug a line |
| 0.4 to 0.7 | Moderate | A clear trend, with plenty of scatter |
| 0.1 to 0.3 | Weak | A slight tilt you might not see without the line |
| about 0 | None (no *straight-line* pattern) | A shapeless cloud |

: {tbl-colwidths="[22,28,50]"}

. . .

The **sign** gives the direction: + means both rise together; - means one rises as the other falls. These bands are **conventions, not laws**: in messy social-science data, a "moderate" r can be a major finding.

## Calibrate Your Eye

```{r}
#| label: r-grid
#| echo: false
#| fig-height: 5.5
set.seed(705)   # inline seed: stable panels
r_targets <- c(0.9, 0.5, 0.2, 0, -0.5, -0.9)

r_grid <- map(r_targets, \(r) {
  x <- rnorm(150)
  y <- r * x + sqrt(1 - r^2) * rnorm(150)
  tibble(label = paste0("r = ", sprintf("%.2f", cor(x, y))), x = x, y = y)
}) |>
  list_rbind() |>
  mutate(label = fct_inorder(label))

ggplot(r_grid, aes(x, y)) +
  geom_point(alpha = 0.5, size = 1.6, color = crju_colors["navy"]) +
  facet_wrap(~label, nrow = 2) +
  labs(title = "Six simulated datasets, 150 points each",
       x = NULL, y = NULL) +
  theme(axis.text = element_blank())
```

By 0.5 (either sign) the trend is visible; below 0.2 your eye mostly sees noise, and it's right.

# What r Cannot See {background-color="#73000A"}

## Anscombe's Quartet: Four Identical Summaries

```{r}
#| label: anscombe-stats
anscombe_long <- anscombe |>
  pivot_longer(everything(), names_to = c(".value", "set"), names_pattern = "(.)(.)")

anscombe_long |>
  summarize(mean_x = mean(x), mean_y = mean(y), r = cor(x, y), .by = set)
```

. . .

Four little datasets that ship with R (Anscombe, 1973). Same averages, same r = 0.82. Statistically interchangeable. Right?

## Anscombe's Quartet, Seen

```{r}
#| label: anscombe-plot
#| echo: false
#| fig-height: 5.5
ggplot(anscombe_long, aes(x, y)) +
  geom_point(size = 2.5, color = crju_colors["navy"]) +
  geom_smooth(method = "lm", se = FALSE, color = crju_colors["garnet"], linewidth = 0.8) +
  facet_wrap(~set, labeller = labeller(set = \(s) paste0("Set ", s, ": r = 0.82"))) +
  labs(title = "Same r, four completely different stories",
       x = NULL, y = NULL)
```

Only Set 1 is what you picture when someone says "r = 0.82." **Always plot before you correlate.**

## Limitation 1: r Only Sees Straight Lines

```{r}
#| label: u-shape
#| echo: false
#| fig-height: 5
set.seed(705)   # inline seed: stable curve
u_data <- tibble(
  x = seq(-4, 4, length.out = 50),
  y = x^2 + rnorm(50, 0, 1)
)

ggplot(u_data, aes(x, y)) +
  geom_point(size = 2.5, color = crju_colors["rust"]) +
  geom_smooth(method = "loess", se = FALSE, color = crju_colors["rust"]) +
  labs(title = "A clear relationship that r cannot see",
       subtitle = paste0("Pearson's r = ", sprintf("%.2f", cor(u_data$x, u_data$y)),
                         ": the data follow a U-shaped curve"),
       x = "Variable X", y = "Variable Y")
```

Think stress and performance: an upside-down U. r near 0 means no **straight-line** relationship, not no relationship.

## Limitation 2: One Point Can Manufacture a Correlation

```{r}
#| label: outlier-demo
#| echo: false
#| fig-height: 5
set.seed(705)   # inline seed: stable demo (numbers cited on the next slide)
base_data    <- tibble(x = 1:20, y = rnorm(20, 10, 4))   # pure noise
outlier_data <- bind_rows(base_data, tibble(x = 25, y = 100))

p_base <- ggplot(base_data, aes(x, y)) +
  geom_point(size = 2.5, color = "grey35") +
  geom_smooth(method = "lm", se = FALSE, color = crju_colors["navy"]) +
  labs(title = "20 points of pure noise",
       subtitle = paste0("r = ", sprintf("%.2f", cor(base_data$x, base_data$y))),
       x = "Variable X", y = "Variable Y")

p_out <- ggplot(outlier_data, aes(x, y)) +
  geom_point(size = 2.5, color = "grey35") +
  geom_point(data = filter(outlier_data, y > 50), color = crju_colors["garnet"], size = 4.5) +
  geom_smooth(method = "lm", se = FALSE, color = crju_colors["navy"]) +
  labs(title = "Same noise + ONE outlier",
       subtitle = paste0("r = ", sprintf("%.2f", cor(outlier_data$x, outlier_data$y))),
       x = "Variable X", y = "Variable Y")

p_base + p_out
```

## What Just Happened

::: incremental
- Twenty points of pure noise happened to give r = `r sprintf("%.2f", cor(base_data$x, base_data$y))`. **Small samples produce correlations out of nothing**; that's why we test, not just compute
- Adding **one** extreme point swung r to **`r sprintf("%+.2f", cor(outlier_data$x, outlier_data$y))`**: it flipped the sign *and* manufactured a "moderate positive" correlation
- Neither number describes a real relationship
- Outliers can also work the other way, *hiding* a real correlation
- Defenses: **plot first** (you'd catch this instantly), and a correlation that doesn't care about wild values...
:::

## A Rank-Based Backup: Spearman's rho

Replace each value with its **rank** (1st, 2nd, 3rd...), then correlate the ranks. Spearman asks only: *is the relationship consistently up, or consistently down?*

```{r}
#| label: spearman-outlier
c(pearson  = cor(outlier_data$x, outlier_data$y),
  spearman = cor(outlier_data$x, outlier_data$y, method = "spearman"))
```

. . .

To Spearman, the wild point is just "rank 21": rho drifts from `r sprintf("%.2f", cor(base_data$x, base_data$y, method = "spearman"))` to `r sprintf("%.2f", cor(outlier_data$x, outlier_data$y, method = "spearman"))`, still weak, while Pearson flipped sign. When data are skewed or have wild values, check Spearman alongside Pearson.

# Real Data: Structure and Violence Across Chicago {background-color="#1B2A4A"}

## From Incidents to Areas

The anchor data's **second level** ([codebook](../chicago-data.qmd)): all 237,472 of 2025's incidents, added up to the **77 community areas** and joined with Census measures:

```{r}
#| label: areas-peek
areas |>
  select(community, violent_rate, unemployment_rate, median_income, pct_vacant) |>
  head(3)
```

. . .

A classic question about places: **which features of a neighborhood track its violence?** Correlation is the right first tool.

## Look First: Unemployment and Violence

```{r}
#| label: unemp-scatter
#| echo: false
#| fig-height: 5.5
ggplot(areas, aes(unemployment_rate, violent_rate)) +
  geom_point(size = 3, alpha = 0.7, color = crju_colors["garnet"]) +
  geom_smooth(method = "lm", se = FALSE, color = crju_colors["navy"]) +
  labs(title = "Violent-crime rate and unemployment rate, 77 community areas",
       subtitle = paste0("r = ", sprintf("%.2f", cor(areas$unemployment_rate, areas$violent_rate))),
       x = "Unemployment rate (%)", y = "Violent crimes per 1,000 residents")
```

One dot = one community area. This is about as strong as area-level associations get in criminology.

## The Full Menu, in One Dataset

```{r}
#| label: four-covariates
#| echo: false
#| fig-height: 5.5
areas |>
  select(violent_rate, unemployment_rate, pct_renter, median_income, median_age) |>
  pivot_longer(-violent_rate, names_to = "covariate", values_to = "value") |>
  mutate(covariate = case_match(covariate,
    "unemployment_rate" ~ paste0("Unemployment rate: r = ",
      sprintf("%.2f", cor(areas$unemployment_rate, areas$violent_rate))),
    "pct_renter" ~ paste0("% renter-occupied: r = ",
      sprintf("%.2f", cor(areas$pct_renter, areas$violent_rate))),
    "median_income" ~ paste0("Median income: r = ",
      sprintf("%.2f", cor(areas$median_income, areas$violent_rate))),
    "median_age" ~ paste0("Median age: r = ",
      sprintf("%.2f", cor(areas$median_age, areas$violent_rate))))) |>
  ggplot(aes(value, violent_rate)) +
  geom_point(alpha = 0.6, size = 2, color = crju_colors["navy"]) +
  geom_smooth(method = "lm", se = FALSE, color = crju_colors["garnet"], linewidth = 0.8) +
  facet_wrap(~covariate, scales = "free_x") +
  labs(title = "Violent-crime rate and four neighborhood measures",
       x = NULL, y = "Violent crimes per 1,000")
```

Strong positive, moderate positive, strong negative, and one genuine *nothing* (median age). Real data span the whole menu.

## These Are Areas, Not People

Our r = 0.84 says: **community areas** with higher unemployment rates record more violent crime per 1,000 residents.

::: {.callout-important appearance="simple"}
**The ecological fallacy:** using an association between *groups* (areas, states, cities) to make claims about *individuals*. Our data contain **no individuals**. Nothing here says *unemployed people* commit violence.
:::

. . .

The classic example (Robinson, 1950): across U.S. **states**, more immigrants went with *higher* literacy. Across **individuals**, immigrants had *lower* literacy. Immigrants had settled in states with high literacy. Adding people up into areas can weaken, exaggerate, or even **flip** a relationship.

## Saying It Out Loud (Area Data)

✅ **Defensible:**

> "Community areas with higher unemployment rates recorded higher violent-crime rates (r = 0.84)."

❌ **Not defensible:**

> "Unemployed residents commit more violent crime." *(A claim about individuals. These data can't see individuals.)*

> "Unemployment drives up violence." *(A claim about cause. Concentrated poverty, disinvestment, and policing patterns all lurk behind both variables.)*

# Putting Uncertainty on r {background-color="#73000A"}

## From Description to a Test

We can *see* r = 0.84. The same question as every test since Session 6:

::: incremental
- **Null hypothesis:** the true correlation is 0. No straight-line association in the process that produced these data
- *If* it were 0, how surprising would an r like ours be, with 77 areas?
- One wrinkle: these 77 areas are **all of Chicago**, not a sample. Read the null as a claim about the *process* that produces neighborhood outcomes, not about a bigger population we sampled from
:::

## `cor.test()`

```{r}
#| label: cor-test
cor.test(areas$unemployment_rate, areas$violent_rate)
```

## Reading It

::: incremental
- **cor = 0.84** at the bottom: that's r
- **95% confidence interval: 0.76 to 0.90.** An interval for the *correlation*, so it lives between -1 and +1. **Where is zero?** Far outside it
- **p-value < 2.2e-16:** reject the null. A world with no correlation essentially never lines up 77 areas this tightly
- With 77 areas, any r above about **0.22** (either sign) clears p < 0.05. Passing the test is a low bar; **the strength is the story**
- Nothing in this output turns association into cause, or areas into individuals
:::

## When the Answer Is "Nothing Here"

Median age and violence, from the menu slide:

```{r}
#| label: cor-test-null
cor.test(areas$median_age, areas$violent_rate)
```

```{r}
#| label: cor-test-null-save
#| include: false
age_test <- cor.test(areas$median_age, areas$violent_rate)
```

. . .

Interval `r sprintf("%.2f", age_test$conf.int[1])` to `r sprintf("%+.2f", age_test$conf.int[2])`: **zero is inside**. Fail to reject. Say *no evidence of a straight-line association*, which is **not** proof there is none.

## Several Pairs at Once

Real analyses rarely stop at one pair. `cor()` takes every numeric column you give it:

```{r}
#| label: cor-matrix
areas |>
  select(violent_rate, unemployment_rate, median_income, pct_vacant) |>
  cor() |>
  round(2)
```

. . .

Read it like a mileage chart: each pair appears twice, and the 1s down the diagonal are each variable with itself. Handy for a first look at your project variables. Then plot and `cor.test()` the pairs that matter.

## How to Report It

> "Across Chicago's 77 community areas, unemployment was strongly associated with the violent-crime rate, r = 0.84, 95% CI 0.76 to 0.90, p < 0.001. This is an area-level association; it does not speak to individuals."

Direction, strength, uncertainty, and the level of analysis. Every time.

## Your Turn (5 Minutes)

`pct_no_hs_diploma` is the share of an area's adults without a high-school diploma. **Before you compute anything:** write down your guess for the sign and rough size of its correlation with `violent_rate`.

```r
areas <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv")

ggplot(areas, aes(pct_no_hs_diploma, violent_rate)) + geom_point()
cor.test(areas$pct_no_hs_diploma, areas$violent_rate)
```

. . .

Most people guess a strong positive. The data: **r = `r sprintf("%.2f", cor(areas$pct_no_hs_diploma, areas$violent_rate))`, p = `r sprintf("%.2f", cor.test(areas$pct_no_hs_diploma, areas$violent_rate)$p.value)`**, no evidence of an association. Yet `pct_bachelors` runs **`r sprintf("%.2f", cor(areas$pct_bachelors, areas$violent_rate))`**. Sensible-sounding measures can fail, and related measures tangle together. Untangling them is *exactly* what regression does in Session 10.

# Now You: Lab 8 {background-color="#1B2A4A"}

## Lab 8

- The same recipe, twice: **question, look, number and test, say it**
- Neighborhoods: poverty and assaults
- Correctional officers: time on the job and job satisfaction
- Submit your Your Turn script to Canvas before you leave

[Lab 8 page](../labs/lab-10-correlation.qmd)

# Midterm Review {background-color="#73000A"}

## The Midterm: Next Week, Session 9

- **In class, monitored.** You get an `.R` script and the data files; you answer in the script and submit it
- **Allowed:** your *printed* notes and homework. **Not allowed:** internet, AI tools, messaging
- Code answers run in R; interpretation answers go in `#` comments
- Everything on it is something we've already done. **Nothing new**
- Full list of skills: the [Midterm Study Guide](../midterm.qmd)

## Six Parts {.smaller}

| Part | Topic | Points | Where we did it |
|---|---|---|---|
| 1 | Data and levels of measurement | 10 | Sessions 1 and 2 |
| 2 | Descriptive statistics | 20 | Session 2 |
| 3 | Probability, sampling, and Bayes counting tables | 20 | Sessions 3 and 4 |
| 4 | Confidence intervals and `t.test()` | 20 | Sessions 5 to 7 |
| 5 | Frequentist and Bayesian: p-values, `bayes_prop_test()` | 15 | Sessions 6 and 7 |
| 6 | Correlation | 15 | Tonight |

: {tbl-colwidths="[8,52,12,28]"}

. . .

The next slides are one practice problem per part. **Different data from the exam, same skills.**

## Part 1: Levels of Measurement

Crash data: nominal, ordinal, interval, or ratio?

- `road_condition` (normal, wet, icy)
- `severity` (coded 1, 2, 3, from least to most severe)
- `driver_age`
- `vehicle1_year`

. . .

**Nominal** (no order) · **ordinal** (ordered, uneven spacing, even when stored as 1, 2, 3) · **ratio** (true zero) · **interval** (equal steps, no true zero: year 0 is not "no year")

On the exam: explain each level, with your own CJ example. **Order, spacing, zero.**

## Part 2: Describe a Small Vector

```{r}
#| label: rev-descriptives
days_held <- c(1, 1, 2, 2, 2, 3, 4, 20)
mean(days_held)
median(days_held)
sd(days_held)
```

. . .

- One long stay (20 days) pulls the **mean** up to `r round(mean(days_held), 1)`, while the **median** stays at `r median(days_held)`. Which one describes a *typical* stay? The median. The tail names the skew: this is **right-skewed**
- The SD, in words: *how far a typical value sits from the mean*. Here, about `r round(sd(days_held))` days, and most of that comes from one person

## Part 3: The Counting Table

A risk tool flags people as "likely to reoffend." **10%** of people actually reoffend. The tool flags **80%** of those who do, and **20%** of those who don't. Out of 1,000 people: of everyone flagged, what share actually reoffends?

. . .

| | Flagged | Not flagged | Total |
|---|---|---|---|
| Reoffends | 80 | 20 | 100 |
| Does not | 180 | 720 | 900 |
| **Total** | **260** | 740 | 1,000 |

P(reoffends | flagged) = 80 / **260** = **0.31**. The "80%" is P(flagged | reoffends), the *other* conditional. **Read the bar out loud: whatever follows it is the denominator.**

## Part 4: A Two-Group `t.test()`

Were intoxicated drivers younger than sober drivers, on average? (Intoxicated first, set by a `factor()` line.)

```{r}
#| label: rev-ttest
#| output-location: slide
t.test(driver_age ~ intox, data = crashes)
```

## Part 4: Reading It

```{r}
#| label: rev-ttest-save
#| include: false
age_t <- t.test(driver_age ~ intox, data = crashes)
```

- Averages: intoxicated drivers `r sprintf("%.1f", age_t$estimate[1])`, sober drivers `r sprintf("%.1f", age_t$estimate[2])`. The interval is **Intoxicated minus Sober**, in **years**, because age is in years: `r sprintf("%.1f", age_t$conf.int[1])` to `r sprintf("%.1f", age_t$conf.int[2])`
- **Where is zero?** `r if (age_t$conf.int[1] > 0 | age_t$conf.int[2] < 0) "Outside the interval, so the difference is detectable" else "Inside the interval, so no detectable difference"` (p = `r format.pval(age_t$p.value, digits = 2)`)
- **Detectable is not the same as large.** The exam asks both: is there a difference, and is it big in real units?
- A boxplot first (`ggplot(crashes, aes(intox, driver_age)) + geom_boxplot()`) shows you the answer before the test does

## Part 5: Which Sentence Belongs to Which Output?

Lab 7: suspicious-person stops vs traffic stops, searched or not.

```{r}
#| label: rev-bayes
#| output-location: slide
bayes_prop_test(x = c(3214, 2023), n = c(3214 + 17271, 2023 + 21191))
```

. . .

- *"If there were truly no difference, a gap this large would almost never happen."* That's the **p-value** (`prop.test()`)
- *"Given these data, there is more than a 99% probability that suspicious-person stops have the higher search rate."* That's **`bayes_prop_test()`**
- A p-value is **never** the probability that the null is true
- On the exam: `source("bayes.R")` (the file is in the exam folder)

## Part 6: Correlation

You did it tonight. On the exam, the same four steps:

1. **Look:** a scatterplot
2. **Number and test:** `cor.test(x, y)`. Report r, the confidence interval, and the p-value
3. **Where is zero?** Inside the interval or outside?
4. **Say it:** direction and strength in words, and what the correlation **cannot** tell you (cause; and if the rows are places, anything about individuals)

## How to Study This Week

- Work through the [Midterm Study Guide](../midterm.qmd): every skill on the exam is listed there
- Re-run your own labs and homework from Sessions 2 to 8 until the code feels routine
- Do the practice pages: Sessions [2](../practice/practice-02.qmd), [3](../practice/practice-03.qmd), [4](../practice/practice-04.qmd), [5](../practice/practice-05.qmd), [6](../practice/practice-06.qmd), and [8](../practice/practice-10.qmd)
- **Print** the notes you want to have. No internet during the exam
- Charge your laptop, and make sure `library(tidyverse)` and `library(modeest)` load

## Homework and Next Week

- **Homework 8:** correlation on the Chicago areas, **plus Project Checkpoint 1**: describe your project's outcome and two or three predictors ([Session 8 page](../weeks/week-10.qmd)). Due Sunday
- **Session 9:** the midterm
- **Checkpoint 2** is due the Sunday after the midterm: correlations in *your* data, plus installing `brms` for Session 10
- **Read before Session 10:** Stanton Ch. 8 (multiple regression)
