---
title: "Logistic Regression: Yes/No Outcomes"
subtitle: "CRJU 705 · Session 12 · Fall 2026"
author: "Scott M. Mourtgos"
format:
  revealjs:
    slide-number: true
    preview-links: auto
    chalkboard: true
    footer: "CRJU 705 · Session 12"
    theme: [default, custom-slides.scss]
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(brms)
library(plotly)
source(here::here("R/setup.R"))
theme_set(theme_minimal(base_size = 18))
set.seed(705)
areas <- read_csv(here::here("data/chicago-areas.csv"), show_col_types = FALSE)
```

## This Week

**Topic:** logistic regression — modeling outcomes that are *yes or no*

- Why `lm()` breaks when the outcome is 0/1
- Odds, log-odds, and the **odds ratio** — the currency of logistic regression
- `glm(family = binomial)`, predicted probabilities, and the Bayesian version with `brms`

**Learning goals:** fit a logistic regression in both traditions; translate coefficients into odds ratios *and* predicted probabilities; say the results in plain language without confusing odds with probability.

**Also today:** the course pays off a promise it made in Session 1.

## Where We've Been

::: incremental
- **Session 10:** correlation — how two continuous variables move together
- **Session 11:** multiple regression — a *plane* through a 3D cloud of points, minimizing squared error, each coefficient's effect holding the others constant (and lurking variables like the ice-cream/drowning lurker handled by putting them *in* the model)
- Every model so far predicted a **number**: a crime rate, an incident hour, a chick's weight
- **Today:** what happens when the outcome only has two values — arrest or no arrest, reoffend or desist, hit again or not?
:::

# The Problem: Yes/No Outcomes {background-color="#73000A"}

## Criminal Justice Runs on Binary Outcomes

- Was an **arrest** made?
- Does the defendant **fail to appear**?
- Is the parolee **re-arrested** within three years?
- Did the crash involve an **intoxicated driver**?
- Is this neighborhood a **high-violence** area?

. . .

Linear regression predicts *continuous* outcomes (crime rates, response times). **Logistic regression predicts binary outcomes** — its currency is **probability (0 to 1)**, expressed along the way as odds and log-odds.

## Session 1 Made You a Promise

Remember the opening case study? **40 bank robberies, five zones**, one question: *where will they hit next?*

::: incremental
- We built `future_hit` = 1 if the same zone was hit again within 10 days, 0 otherwise
- We fit a model, showed you a coefficient table nobody could read yet, and sent the patrol cars to the **West Zone** — where the robber was caught
- That slide said: *"Don't worry about the machinery today — logistic regression is Session 12."*
:::

. . .

**Today is Session 12.** Let's reopen the machinery.

```{r}
#| label: robbery-sim
#| include: false
# Same simulation as the Session 1 deck (seed 58 preserved so the numbers match)
set.seed(58)
n <- 40
zones <- c("North", "South", "East", "West", "Downtown")

robberies <- tibble(
  robbery_id = 1:n,
  date = as.Date("2024-01-01") + cumsum(rpois(n, 5)),
  zone = sample(zones, size = n, replace = TRUE,
                prob = c(0.1, 0.1, 0.1, 0.5, 0.2))
) |>
  arrange(date) |>
  mutate(
    day_of_year = yday(date),
    zone = factor(zone, levels = zones),
    prior_robberies_in_zone = row_number() - 1,
    .by = zone
  ) |>
  arrange(date)

robberies <- robberies |>
  arrange(date) |>
  mutate(
    future_hit = map_int(row_number(), \(i) {
      if (i == nrow(robberies)) return(NA_integer_)
      later <- robberies[(i + 1):nrow(robberies), ]
      as.integer(any(later$zone == zone[i] & later$date <= date[i] + 10))
    })
  ) |>
  filter(!is.na(future_hit))
```

## The Machinery, Reopened

The exact model from Session 1 — one function, one `family` argument:

```{r}
#| label: robbery-fit
robbery_model <- glm(
  future_hit ~ 0 + scale(day_of_year) + prior_robberies_in_zone + zone,
  data = robberies, family = binomial
)
```

- `glm()` = *generalized* linear model — regression's family business
- `family = binomial` is the switch that makes it **logistic**: predict a 0/1 outcome
- Everything else looks exactly like the `lm()` formulas from Session 11

::: aside
Two formula details you can safely skim: `0 +` drops the global intercept so each zone gets its own baseline, and `scale()` puts day-of-year on a standard scale. Read the *table*, not the formula — today is about the `family` switch.
:::

## The Table You Couldn't Read in August

```{r}
#| label: robbery-table
#| echo: false
broom::tidy(robbery_model) |>
  mutate(odds_ratio = exp(estimate)) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable()
```

## Reading It With Today's Eyes

::: incremental
- The `estimate` column is in **log-odds** — that's why it looked like gibberish in August; by the end of today it won't
- `exp(estimate)` turns each one into an **odds ratio** — West's is by far the largest, which is why West had the highest predicted probability of a repeat hit. *That's* why the cars went west.
- Those −17.5s with standard errors in the thousands? North and South never saw a repeat hit within 10 days — with **zero events**, the model can't estimate their log-odds. 39 robberies is small data, and the output says so honestly.
:::

## Logistic vs. Linear: The Road and the Hill

Linear regression is a **straight road** — it keeps going forever in both directions. Logistic regression is a **curved hill** that flattens at 0 and 1, because probabilities can't go below 0 or above 1.

```{r}
#| label: road-hill
#| echo: false
#| fig-height: 4.2
analogy <- tibble(x = seq(-6, 6, length.out = 200)) |>
  mutate(
    Linear   = 0.1 * x + 0.5,
    Logistic = 1 / (1 + exp(-x))
  ) |>
  pivot_longer(c(Linear, Logistic), names_to = "model", values_to = "predicted")

ggplot(analogy, aes(x, predicted, color = model)) +
  geom_hline(yintercept = c(0, 1), linetype = 2, color = "grey60") +
  geom_line(linewidth = 1.3) +
  scale_color_manual(values = unname(crju_colors[c("garnet", "blue")])) +
  scale_y_continuous(limits = c(-0.2, 1.2), breaks = seq(0, 1, 0.2)) +
  labs(x = "Predictor (X)", y = "Predicted value / probability", color = NULL)
```

## Watch Linear Regression Fail

Predicting *graduated the academy (yes/no)* from study hours:

```{r}
#| label: linear-fails
#| echo: false
#| fig-height: 5
set.seed(42)   # same simulated recruits as the 2025 lecture
recruits <- tibble(study_hours = runif(100, 0, 10)) |>
  mutate(
    prob_grad = 1 / (1 + exp(-(-4 + 0.8 * study_hours))),
    graduated = rbinom(100, 1, prob_grad)
  )

model_linear <- lm(graduated ~ study_hours, data = recruits)
model_logit  <- glm(graduated ~ study_hours, data = recruits, family = binomial)

grid_hours <- tibble(study_hours = seq(-5, 15, length.out = 400)) |>
  mutate(
    Linear   = predict(model_linear, newdata = pick(study_hours)),
    Logistic = predict(model_logit, newdata = pick(study_hours), type = "response")
  ) |>
  pivot_longer(c(Linear, Logistic), names_to = "model", values_to = "pred")

ggplot(recruits, aes(study_hours, graduated)) +
  geom_jitter(height = 0.05, width = 0, alpha = 0.5, color = "grey40") +
  geom_hline(yintercept = c(0, 1), linetype = 2, color = "grey60") +
  geom_line(data = grid_hours, aes(y = pred, color = model), linewidth = 1.2) +
  scale_color_manual(values = unname(crju_colors[c("garnet", "blue")])) +
  scale_y_continuous(limits = c(-1.5, 2.5), breaks = seq(0, 1, 0.5)) +
  annotate("text", x = -3, y = 2.1, label = "Impossible (>1)",
           color = crju_colors["garnet"], fontface = "bold", size = 5) +
  annotate("text", x = 13, y = -1.1, label = "Impossible (<0)",
           color = crju_colors["garnet"], fontface = "bold", size = 5) +
  labs(x = "Study hours", y = "Predicted probability of graduating", color = NULL)
```

## What Went Wrong

::: incremental
- The linear model happily predicts values like **1.2** and **−0.3** — probabilities that cannot exist
- Inside the middle of the data the two lines nearly agree; the damage is at the edges — exactly where predictions get used ("what about a recruit who studies *a lot*?")
- The fix is the **logit transformation**: instead of modeling the probability directly, model something built from it that a straight line *can* live on
- (The full derivation is Ph.D.-stats material — what you need is the ladder on the next slides)
:::

# Odds and Log-Odds {background-color="#1B2A4A"}

## The Model You Already Know

$$Y = b_0 + b_1 X_1 + b_2 X_2 + \varepsilon$$

- $Y$: outcome (dependent variable)
- $b_0$: intercept (value of $Y$ when all X's are zero)
- $b_1, b_2$: slope coefficients (effect of each X, **holding the others constant**)
- $\varepsilon$: error term (what's left unexplained)

. . .

The right-hand side is an unbounded straight line. Probability is trapped in [0, 1]. Something has to give.

## Step 1 — Probability to Odds

Suppose 3 out of 4 parolees complete supervision: $p = 0.75$.

$$\text{odds} = \frac{p}{1-p} = \frac{0.75}{0.25} = 3 \quad \text{("three to one")}$$

| Probability | Odds | Log-odds |
|:---:|:---:|:---:|
| 0.25 | 0.33 | −1.10 |
| 0.50 | 1.00 | 0.00 |
| 0.75 | 3.00 | +1.10 |

**Step 2 — odds to log-odds:** odds run 0 to ∞; *log*-odds run −∞ to +∞. That's a scale a straight line can live on.

## The Logistic Model

Just treat logit($p$) — the log-odds — as your Y. Then...

$$\text{logit}(p) = \ln\!\left(\frac{p}{1-p}\right) = b_0 + b_1 X_1 + b_2 X_2$$

- Same intercept, same slopes, same "holding the others constant"
- The line now predicts **log-odds**, which convert back to a probability that politely stays between 0 and 1 — the curved hill

. . .

::: {.callout-important appearance="simple"}
**But the coefficients are log-odds!!!** Nobody thinks in log-odds. So what do we do with them?
:::

## What Do We Do With Log-Odds?

<br>

::: {.r-fit-text}
**EXPONENTIATE!**
:::

<br>

`exp(coefficient)` = the **odds ratio**: *multiply the odds of the outcome by this much for each one-unit increase in X, holding the other predictors constant.*

## The Odds-Ratio Rosetta Stone: Odds Shrink {.smaller}

| Log-odds coefficient (β) | Odds ratio (e^β^) | Effect on odds | Intuitive interpretation |
|:---:|:---:|:---:|:---|
| −2.0 | 0.14 | ↓ 86% | Odds are 86% lower than baseline |
| −1.0 | 0.37 | ↓ 63% | Odds are 63% lower than baseline |
| −0.5 | 0.61 | ↓ 39% | Odds are 39% lower than baseline |
| 0.0 | 1.00 | No change | Odds unchanged |

Negative log-odds → ratio **below 1** → odds shrink. Zero → ratio of exactly 1 → nothing happens.

## The Odds-Ratio Rosetta Stone: Odds Grow {.smaller}

| Log-odds coefficient (β) | Odds ratio (e^β^) | Effect on odds | Intuitive interpretation |
|:---:|:---:|:---:|:---|
| 0.0 | 1.00 | No change | Odds unchanged |
| 0.3 | 1.35 | ↑ 35% | Odds are 35% higher than baseline |
| 0.7 | 2.01 | ↑ 101% | Odds are about twice as high |
| 1.0 | 2.72 | ↑ 172% | Odds are 2.7× higher |
| 2.0 | 7.39 | ↑ 639% | Odds are over seven times higher |

Positive log-odds → ratio **above 1** → odds grow.

# Fitting It: glm() {background-color="#73000A"}

## Two Traditions, Same Curve

| Approach | What it estimates | Interpretation |
|:---|:---|:---|
| **Frequentist** | Point estimates and confidence intervals | "If we repeated this 100 times, 95% of the intervals would contain the true β." |
| **Bayesian** | Posterior distributions for each β | "Given our data and priors, there's a 95% probability β lies in this range." |

Exactly the split you know from Sessions 6 and 8 — nothing new, just a new model underneath it.

## The Frequentist Version

Toy data first — 32 cars from a 1974 *Motor Trend* issue, predicting **manual transmission (1) vs. automatic (0)** from horsepower and weight. Small enough to check every number; real CJ data in a few minutes.

```{r}
#| label: mtcars-glm
#| output-location: slide
data(mtcars)
cars_df <- mtcars |>
  rownames_to_column("car") |>
  mutate(am = as.factor(am))   # 0 = automatic, 1 = manual

model_freq <- glm(am ~ hp + wt, data = cars_df, family = binomial)
summary(model_freq)
```

## Reading the Output: Exponentiate

```{r}
#| label: mtcars-or
exp(coef(model_freq))
```

::: incremental
- `hp`: log-odds 0.036 → **odds ratio 1.037** — each additional horsepower multiplies the odds of a manual transmission by about 1.04 (+3.7%), *holding weight constant*
- `wt`: log-odds −8.08 → **odds ratio 0.0003** — each additional 1,000 lbs cuts the odds by ~99.97%, *holding horsepower constant*
- Wildly different sizes partly because the **units** differ: one horsepower is a nudge; a thousand pounds is a different car
:::

## From Effects to Cases: predict()

Odds ratios describe *effects*. For an actual *case*, ask for the probability directly:

```{r}
#| label: mtcars-pred
two_cars <- cars_df |>
  filter(car %in% c("Honda Civic", "Cadillac Fleetwood")) |>
  select(car, hp, wt)

two_cars |>
  mutate(p_manual = predict(model_freq, newdata = two_cars, type = "response"))
```

. . .

`type = "response"` returns probabilities: the light Civic is a near-certain manual (99.95%); the 5,250-lb Fleetwood, near-certain automatic. **Report both:** odds ratios for effects, predicted probabilities for cases.

# Seeing It: The Curved Surface {background-color="#1B2A4A"}

## Session 11's Plane, Bent

::: incremental
- Last session: two predictors + linear regression = a **flat plane** floating through the 3D cloud
- Same two predictors + logistic regression = a **curved sheet**: a plateau near probability 1, a floor near 0, and an S-shaped cliff between them
- The surface *is* the model: its height at any (hp, wt) combination is the predicted probability of a manual transmission there
- Let's build it from nothing but `predict()`
:::

## Building the Surface

Ask the model for a prediction at every combination of horsepower and weight — a 40 × 40 grid, 1,600 predictions:

```{r}
#| label: surface-grid
hp_seq <- seq(min(cars_df$hp), max(cars_df$hp), length.out = 40)
wt_seq <- seq(min(cars_df$wt), max(cars_df$wt), length.out = 40)
grid   <- expand.grid(hp = hp_seq, wt = wt_seq)

grid$prob <- predict(model_freq, newdata = grid, type = "response")

# Reshape the 1,600 predictions into a grid; t() puts rows on the
# wt (y) axis and columns on the hp (x) axis, which is what plotly expects
z_matrix <- t(matrix(grid$prob, nrow = length(hp_seq), ncol = length(wt_seq)))
```

The observed cars sit at height 0 (automatic) or 1 (manual); the surface passes between them.

## The 3D Logistic Surface

```{r}
#| label: surface-plot
#| output-location: slide
data_points <- cars_df |>
  mutate(prob = as.numeric(as.character(am)))   # 0 = auto, 1 = manual

plot_ly(width = 950, height = 520) |>
  add_markers(data = data_points,
              x = ~hp, y = ~wt, z = ~prob,
              color = ~am, colors = c("steelblue", "firebrick"),
              symbol = I("circle"), size = I(50),
              name = "Observed Data") |>
  add_surface(x = hp_seq, y = wt_seq, z = z_matrix,
              colorscale = list(c(0, 'lightblue'), c(1, 'firebrick')),
              opacity = 0.6,
              name = "Predicted Probability Surface") |>
  layout(
    title = "Logistic Regression: Curved Probability Surface",
    scene = list(
      xaxis = list(title = "Horsepower (hp)"),
      yaxis = list(title = "Weight (1000 lbs)"),
      zaxis = list(title = "Probability (Manual Transmission)")
    )
  )
```

## The Guided Tour (Recorded)

Rotate the live surface yourself on the previous slide — or rewatch the recorded walkthrough anytime:

<video src="../media/week-12/3d-logistic.mp4" controls preload="metadata"></video>

# The Bayesian Version {background-color="#73000A"}

## Same Formula, Bayesian Machinery

```{r}
#| label: mtcars-brm
#| output: false
model_bayes <- brm(am ~ hp + wt, data = cars_df,
                   family = bernoulli(),
                   chains = 2, iter = 2000, seed = 705, refresh = 0)
```

::: incremental
- `family = bernoulli()` is brms's name for a 0/1 outcome — the switch that `family = binomial` flips in `glm()`
- Everything else is the Session 8 pattern: `chains = 2, iter = 2000`, the compile pause, `refresh = 0` to silence the log
- **The prior:** none stated → `brms` defaults again. Your project's model will use these too — **say so in your methods**: "brms default priors, 2 chains × 2,000 iterations, seed 705"
- Instead of one estimate per coefficient, we get a full **posterior distribution** for each log-odds coefficient
:::

## Posterior Odds Ratios

Exponentiate the posterior summaries — dropping the intercept row, whose posterior is astronomically wide with only 32 cars:

```{r}
#| label: mtcars-brm-or
exp(fixef(model_bayes)[-1, c("Estimate", "Q2.5", "Q97.5")])
```

::: incremental
- Same story as `glm()`: horsepower nudges the odds of a manual up; weight crushes them
- The payoff is the sentence you're *allowed to say*: "Given the data and priors, there's a 95% probability the true odds ratio lies in this interval" — and with 32 cars, those honest intervals are huge. Watch them tighten on real data next.
:::

# Real Data: Chicago Neighborhoods {background-color="#1B2A4A"}

## The Question

What separates Chicago's **high-violence** community areas from the rest?

```{r}
#| label: chicago-outcome
areas |> count(high_violence)
```

. . .

`high_violence` ships in the file (it's in the [codebook](../chicago-data.qmd)): is the area's violent-crime rate above the **citywide median**? 38 above, 39 at or below.

## Making a Binary Outcome

The line that built it:

```r
areas |> mutate(high_violence = violent_rate > median(violent_rate))
```

::: incremental
- R treats `TRUE` as 1 and `FALSE` as 0, so a logical column drops straight into `glm()`
- If your project outcome isn't binary yet, it may be one `mutate()` away too
- One caution: cutting a continuous variable in half **throws away information** — defensible when the yes/no version is the question you actually care about
:::

## Look First (Always)

One predictor at a time before the model — unemployment against the outcome, with the fitted logistic curve:

```{r}
#| label: chicago-curve
#| echo: false
#| fig-height: 5
ggplot(areas, aes(unemployment_rate, as.numeric(high_violence))) +
  geom_jitter(height = 0.04, width = 0, size = 2.5, alpha = 0.6,
              color = crju_colors["garnet"]) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              se = FALSE, color = crju_colors["navy"], linewidth = 1.2) +
  labs(title = "The S-curve, on real data",
       subtitle = "77 community areas; curve = single-predictor logistic fit",
       x = "Unemployment rate (%)", y = "High-violence area (0/1)")
```

## Fit It

Two predictors: unemployment and vacant housing — both plausible markers of concentrated disadvantage:

```{r}
#| label: chicago-glm
#| output-location: slide
v_model <- glm(high_violence ~ unemployment_rate + pct_vacant,
               data = areas, family = binomial)
summary(v_model)
```

## The Coefficients, Translated

```{r}
#| label: chicago-or
exp(cbind(OR = coef(v_model), confint(v_model)))
```

::: incremental
- **Unemployment: OR 1.45** [1.19, 1.85] — each additional point of unemployment multiplies an area's odds of high-violence status by ~1.45 (+45%), holding vacancy constant
- **Vacancy: OR 1.33** [1.04, 1.84] — +33% odds per point, holding unemployment constant
- For a *ratio*, no effect = **1**, not 0 — both intervals exclude it. And these are associations, **not causes**: nobody randomly assigned unemployment to neighborhoods
:::

## Coefficients Into Cases

```{r}
#| label: chicago-pred
two_areas <- areas |>
  filter(community %in% c("Forest Glen", "West Garfield Park")) |>
  select(community, unemployment_rate, pct_vacant)

two_areas |>
  mutate(p_high_violence = predict(v_model, newdata = two_areas,
                                   type = "response"))
```

. . .

In plain language: for an area like Forest Glen (4.2% unemployment, 2.6% vacant), the model puts the probability of high-violence status at about **3%**. For one like West Garfield Park (18.9%, 20.0%), it's over **99%**. That's a sentence a chief — or a city council — can actually use.

## The Bayesian Version, Real Data

```{r}
#| label: chicago-brm
#| output: false
v_bayes <- brm(high_violence ~ unemployment_rate + pct_vacant,
               data = areas, family = bernoulli(),
               chains = 2, iter = 2000, seed = 705, refresh = 0)
```

```{r}
#| label: chicago-brm-or
exp(fixef(v_bayes)[, c("Estimate", "Q2.5", "Q97.5")])
```

. . .

Both traditions land in the same place: posterior odds ratios near the `glm()` estimates, credible intervals excluding 1. With 77 areas instead of 32 cars, the intervals are usably tight.

## How to Report It: Frequentist

> "Each one-point increase in the unemployment rate was associated with 45% higher odds of being a high-violence area (OR = 1.45, 95% CI [1.19, 1.85]), holding vacancy constant. Predicted probabilities ranged from about 3% for the least-disadvantaged areas to over 99% for the most."

. . .

The effect (as an odds ratio), the uncertainty, and a **probability a reader can feel** — all three, every time.

## How to Report It: Bayesian

> "Given the data and priors, each one-point increase in unemployment multiplied the odds of high-violence status by about 1.5 (posterior mean OR 1.49, 95% credible interval [1.22, 1.92])."

. . .

Same three ingredients — and this one gets to make a direct probability statement about the odds ratio itself.

## Your Turn (3 minutes, with a neighbor)

::: {.your-turn}
A reentry study models **re-arrest within 12 months** and reports two log-odds coefficients: **risk score (per point): 0.7** · **completed programming: −0.5**

1. Convert both to odds ratios (Rosetta Stone, or `exp()` in R)
2. Say each in a sentence a police chief would understand
3. Audit your neighbor: did they say ***odds*** — or slip into *probability*?
:::

. . .

`exp(0.7)` ≈ 2.0 — each risk point roughly **doubles the odds**. `exp(−0.5)` ≈ 0.61 — about **39% lower odds**. Neither is a statement about *probability*.

## The Toolbox Is Complete

::: incremental
- In Session 1 we predicted a robber's next move with a table you couldn't read. **You've now earned the tool the course opened with** — that table is just log-odds, and log-odds are just `exp()` away from sense
- Descriptives → probability → inference → group comparison → correlation → regression → **logistic regression**
- Nothing in your final project requires a method you haven't now practiced in both traditions
- **Checkpoint 4 — the final one — is in this week's homework:** fit and interpret *your* final model
:::

## Today's Lab & Homework

- **Lab 12:** the whole logistic workflow — Chicago neighborhoods, then Alaska crashes at night ([lab page](../labs/lab-12-logistic.qmd))
- **Homework 12:** your final-project model + **Project Checkpoint 4 (final)** — cleaned variables, model equation, fitted model, and the three sentences you'd say about it ([assignment](../homework/hw-12.qmd))
- **Session 13 is the project help session** — no new reading; bring your project code, your data, and your two most stuck-on problems
- Full project due via Canvas **before midnight Sunday** ahead of Session 14 ([details](../final-project.qmd))
