---
title: "Statistics is Awesome!"
subtitle: "CRJU 705 · Session 1 · Fall 2026"
author: "Scott M. Mourtgos"
format:
  revealjs:
    slide-number: true
    preview-links: auto
    chalkboard: true
    footer: "CRJU 705 · Session 1"
    theme: [default, custom-slides.scss]
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
source(here::here("R/setup.R"))
theme_set(theme_minimal(base_size = 18))
set.seed(705)
zone_pal <- unname(palette.colors(6, palette = "Okabe-Ito"))[2:6]
```

# Welcome {background-color="#73000A"}

## Class Introductions

- Name
- Place in program
- Area of interest
- Experience with statistics (*anxiety* about statistics?)
- Why are you here?
- Future plans / goals

# A City Under Siege {background-color="#1B2A4A"}

## The Situation

Over three months, **40 bank robberies** strike across a city's five zones: North, South, East, West, and Downtown.

::: incremental
- No clear pattern. No suspects. No leads.
- The robber hits banks, disappears, and resurfaces days later in a different part of the city.
- Detectives are stumped. The media is escalating. Public concern is growing.
- Then someone makes a call to the PD's new crime analyst team:
:::

. . .

> "Can you help us predict where they'll hit next?"

## What Does the Data Look Like?

40 bank robberies across 5 city zones over several months:

| Variable | Meaning |
|----------|---------|
| `robbery_id` | Robbery sequence |
| `date` | When the robbery occurred |
| `zone` | Which part of the city was hit |
| `day_of_year` | Numeric version of the date (for modeling time) |
| `lat` & `lon` | Spatial location (for mapping) |
| `prior_robberies_in_zone` | How many robberies already happened in that zone? |

## Before We Model, We Look

Very little information is available: no suspect descriptions, no MO patterns — just **when** and **where**.

Can we use this simple dataset to build a predictive model with statistics?

. . .

But before we even build a model, we can use basic plots to spot patterns.

**Good data analysts always visualize before they model.**

```{r}
#| label: simulate-robberies
#| include: false
# Same simulation as Fall 2025 lecture (seed preserved)
set.seed(58)
n <- 40
zones <- c("North", "South", "East", "West", "Downtown")

robberies <- tibble(
  robbery_id = 1:n,
  date = as.Date("2024-01-01") + cumsum(rpois(n, 5)),
  zone = sample(zones, size = n, replace = TRUE,
                prob = c(0.1, 0.1, 0.1, 0.5, 0.2))
) |>
  arrange(date) |>
  mutate(
    day_of_year = yday(date),
    zone = factor(zone, levels = zones),
    prior_robberies_in_zone = row_number() - 1,
    .by = zone
  ) |>
  arrange(date)
```

## Timeline of Robberies by Zone — any patterns?

```{r}
#| label: timeline-plot
#| echo: false
#| fig-width: 10
#| fig-height: 5
ggplot(robberies, aes(x = date, y = zone, color = zone)) +
  geom_point(size = 4, alpha = 0.85) +
  scale_color_manual(values = zone_pal) +
  labs(title = "Timeline of Bank Robberies by Zone", x = "Date", y = NULL) +
  theme(legend.position = "none")
```

## Watching It Unfold Over Time

<video src="../media/week-01/robbery_timeline.mp4" controls preload="metadata"></video>

## Mapped Locations — any patterns?

<video src="../media/week-01/animated_robberies_map.mp4" controls preload="metadata"></video>

## Building a Predictive Model

We want to predict **where the next robbery will occur** within a short time window, based on patterns in *time*, *location*, and *prior activity*.

**Key predictors:**

- Day of year (captures seasonal or other unobserved time patterns)
- Prior robberies in the same zone
- Zone

## What Are We Trying to Predict?

We use **logistic regression** to predict whether the same zone will be hit again within 10 days:

- **Outcome:** `future_hit` = 1 if another robbery occurs in the *same zone* within the next 10 days; otherwise 0
- This creates a **binary classification** problem

*(Don't worry about the machinery today — logistic regression is Session 12. This is a preview of where we're headed.)*

```{r}
#| label: fit-model
#| include: false
robberies <- robberies |>
  arrange(date) |>
  mutate(
    future_hit = map_int(row_number(), \(i) {
      if (i == nrow(robberies)) return(NA_integer_)
      later <- robberies[(i + 1):nrow(robberies), ]
      as.integer(any(later$zone == zone[i] & later$date <= date[i] + 10))
    })
  ) |>
  filter(!is.na(future_hit))

model <- glm(future_hit ~ 0 + scale(day_of_year) + prior_robberies_in_zone + zone,
             data = robberies, family = binomial)
```

## Logistic Regression Results

```{r}
#| label: model-summary
#| echo: false
broom::tidy(model) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable()
```

## Predicted Risk by Zone

<video src="../media/week-01/predicted_prob_growth.mp4" controls preload="metadata"></video>

## Predicted Risk by Zone

Based on the model, the **West Zone** has the highest likelihood of being targeted next.

. . .

**So what should we do with the information derived from the model?**

## Strategic Deployment

:::: {.columns}
::: {.column width="62%"}
<video src="../media/week-01/deploy_cars_east_in.mp4" controls preload="metadata" style="max-height:430px;"></video>
:::
::: {.column width="38%"}
<br>

- 📍 **Where:** West Zone had the highest predicted probability
- 🚔 **Action:** Deploy units there for surveillance and hopeful capture
- 🎯 **Goal:** Increase arrest probability
:::
::::

## Case Study Wrap-Up & Discussion

**Outcome:** Within days, officers apprehended the suspect in the predicted area as he was approaching a bank.

Statistics turned scattered crime data into a clear lead — saving time, resources, and preventing further victimization.

::: incremental
- What data did we *not* use? How could we update this model as new robberies occur in real time?
- What are the risks of over-relying on a model? How do we communicate uncertainty to decision-makers?
- What other areas that interest you could statistics be applied to?
:::

# Course Structure {background-color="#73000A"}

## How This Course Works

Everything lives on the course website — bookmark it:

**<https://smourtgos.github.io/crju705/>**

- **[Schedule](https://smourtgos.github.io/crju705/schedule.html)** — topics, readings, slides, labs, homework for every session
- **[Syllabus](https://smourtgos.github.io/crju705/syllabus.html)** — walk-through now (grading, policies, AI policy, the mid-term warning)
- **Two textbooks:** Stanton (required) + *R for Data Science* (free online)
- **Both traditions:** you will learn frequentist **and** Bayesian inference — few courses teach both
- Submissions happen via Canvas; materials live here

# Types of Data & Levels of Measurement {background-color="#1B2A4A"}

## Types of Data

**Qualitative (categorical)**

- Describe groups or labels (zone, sex, race)
- Text / interviews

**Quantitative (numerical)**

- Measurable amounts (age, prior robberies)
- **Discrete** = whole counts (number of robberies)
- **Continuous** = can take any value (income, age)

## Levels of Measurement

| Level | Definition | Examples |
|-------|-----------|----------|
| **Nominal** | Categories without order | Race, sex, zone |
| **Ordinal** | Ordered categories, unequal gaps | Likert scales |
| **Interval** | Meaningful differences, no true zero | Temperature (°C) |
| **Ratio** | Meaningful differences **and** true zero | Robbery counts, age |

## Why Does the Level Matter?

::: incremental
- Your variable's scale determines **the math you're allowed to do**
- It also determines **the plot you should make**
- Using the wrong scale leads to **wrong conclusions**
:::

```{r}
#| label: rob-sim
#| include: false
set.seed(58)
rob <- tibble(
  zone = factor(sample(c("North", "South", "East", "West", "Downtown"), 200, TRUE)),
  prior_robberies_in_zone = rpois(200, 3),    # discrete (ratio)
  risk_likert = sample(1:7, 200, TRUE),       # ordinal
  temp_c = rnorm(200, 20, 5),                 # interval
  age = pmax(18, round(rnorm(200, 40, 12))),  # ratio
  caught = factor(rbinom(200, 1, 0.35), labels = c("No", "Yes"))
)
```

## Scale Determines the Math: Nominal

```{r}
#| label: nominal-wrong
# WRONG: the "mean zone" is meaningless
rob$zone |> as.numeric() |> mean()
```

. . .

```{r}
#| label: nominal-right
# RIGHT: counts and percentages
rob |>
  count(zone) |>
  mutate(pct = 100 * n / sum(n))
```

## Scale Determines the Math: Ordinal

```{r}
#| label: ordinal-summaries
# Summarize a Likert item (ordinal): median & IQR
rob |>
  summarize(med = median(risk_likert), iqr = IQR(risk_likert))

# Association with a numeric variable: Spearman (rank-based)
cor(rob$risk_likert, rob$prior_robberies_in_zone, method = "spearman")
```

## Interval vs. Ratio

```{r}
#| label: interval-demo
mean(rob$temp_c)             # ~20 °C
mean(rob$temp_c * 1.8 + 32)  # same temperatures in °F
```

Temperature has **no true zero** — 0 °C is just where water freezes.

- "30 °C is 10 °C warmer than 20 °C" ✔ (differences are fine)
- "20 °C is *twice as hot* as 10 °C" ✘ (ratios are meaningless)

. . .

```{r}
#| label: ratio-demo
mean(rob$age)
```

Age **has** a true zero — a 40-year-old *is* twice as old as a 20-year-old. Differences **and** ratios both work.

## Plots That Match the Scale

```{r}
#| label: scale-plots
#| echo: false
#| fig-width: 12
#| fig-height: 6
library(patchwork)
p1 <- ggplot(rob, aes(zone)) + geom_bar(fill = crju_colors["garnet"]) +
  labs(title = "Nominal: bar chart", x = NULL)
p2 <- ggplot(rob, aes(factor(risk_likert, levels = 1:7))) +
  geom_bar(fill = crju_colors["navy"]) +
  labs(title = "Ordinal: ordered bars", x = "Risk (1–7)")
p3 <- ggplot(rob, aes(age)) + geom_histogram(binwidth = 5, fill = crju_colors["blue"]) +
  labs(title = "Continuous (ratio): histogram")
p4 <- ggplot(rob, aes(prior_robberies_in_zone)) + geom_bar(fill = crju_colors["gold"]) +
  labs(title = "Discrete counts: bars", x = "Prior robberies")
(p1 + p2) / (p3 + p4)
```

## Choosing the Right Test (Preview of the Semester)

```{r}
#| label: test-preview
#| output-location: fragment
# Categorical × categorical → chi-squared (compare proportions)
chisq.test(table(rob$zone, rob$caught))
```

```{r}
#| label: test-preview-2
#| output-location: fragment
# Numeric by category → ANOVA (compare group means)  [Session 8]
aov(age ~ zone, data = rob) |> summary()
```

```{r}
#| label: test-preview-3
#| output-location: fragment
# Numeric × numeric → Pearson correlation             [Session 10]
cor(rob$age, rob$prior_robberies_in_zone)
```

## Treating Ordinal as Numeric Can Mislead

```{r}
#| label: ordinal-trap
#| fig-height: 4
educ <- rep(1:5, each = 50)                 # ordinal codes 1..5
income <- rep(c(1, 2, 3, 4, 1000), each = 50)  # HUGE jump at the top
cor(educ, income)                           # Pearson looks "strong"
```

```{r}
#| label: ordinal-trap-plot
#| echo: false
#| fig-height: 4
ggplot(tibble(educ, income), aes(educ, income)) +
  geom_point(size = 3, alpha = 0.6, color = crju_colors["garnet"]) +
  geom_smooth(method = lm, se = FALSE, color = crju_colors["navy"]) +
  scale_y_log10() +
  labs(x = "Education (1–5 coded as numbers)", y = "Income (log scale)",
       title = "Ordinal codes hide wildly unequal spacing")
```

The correlation is "large," but the categories are **not equally spaced** — code numbers ≠ measurements.

# Getting You Running in R {background-color="#73000A"}

## Install R + RStudio (Before Session 2!)

Follow the step-by-step guide on the course website:

**[smourtgos.github.io/crju705/resources.html](https://smourtgos.github.io/crju705/resources.html)**

1. Install **R** (from CRAN)
2. Install **RStudio** (from Posit)
3. Test: type `2 + 2` in the Console → you should see `[1] 4`
4. Install the course packages:

```{r}
#| label: install-pkgs
#| eval: false
install.packages(c("tidyverse", "swirl", "BayesFactor", "here"))
```

**Check:** open RStudio and verify you see four panes (Source, Console, Environment/History, Files/Plots/Help).

## First Steps: Arithmetic & Assignment

*(Stanton Appendix A)*

```{r}
#| label: appendix-a1
3 + 5
x <- 1:10
sum(x); mean(x); sd(x)
```

```{r}
#| label: appendix-a-help
#| eval: false
?mean   # every function has a help page — use them constantly
```

## First Steps: Logical Indexing

```{r}
#| label: appendix-a2
x > 5
x[x > 5]
```

## First Steps: Characters & Factors

```{r}
#| label: appendix-a3
student_names <- c("Alice", "Bob", "Carol")
is.character(student_names)

f <- factor(c("red", "blue", "red"))
f
```

::: {.your-turn}
**Your turn:** build a vector of seven ages; compute the mean & SD; subset the values > 30; then use `?seq` to generate `seq(0, 1, by = 0.2)`.
:::

## Data Frames & dplyr

*(Stanton Appendix B — modern form)*

```{r}
#| label: appendix-b1
library(tidyverse)
cars_data <- as_tibble(mtcars, rownames = "model")
head(cars_data, 4)
```

::: aside
The `|>` symbol is the **pipe**: `x |> f()` means `f(x)` — "take x, *then* do f." Older code (including the Stanton book) writes it `%>%`; they work the same way for everything we do.
:::

## select — pick columns

```{r}
#| label: appendix-b2
cars_data |> select(model, mpg, cyl, hp) |> head(4)
```

## filter — pick rows

```{r}
#| label: appendix-b3
cars_data |> filter(cyl == 6) |> head(4)
```

## mutate & summarize

```{r}
#| label: appendix-b4
cars_data |>
  mutate(mpg_per_hp = mpg / hp) |>
  select(model, mpg, hp, mpg_per_hp) |>
  head(4)

cars_data |>
  summarize(avg_mpg = mean(mpg), n = n(), .by = cyl)
```

## Your Turn

::: {.your-turn}
Using the built-in `mpg` data:

1. Find the top-3 models by `hwy`.
2. Compute the average `cty` by `class`.
:::

*(Hints: `arrange(desc(...))`, `slice_head()`, and `summarize(..., .by = ...)` — all from your R4DS Ch. 3 reading next week.)*

## Coming Attractions: Real Data, All Semester

Starting next week, our running example is **real**: every crime reported in Chicago in 2025.

```{r}
#| label: chicago-teaser
#| echo: false
#| fig-height: 4.5
crimes <- read_csv(here::here("data/chicago-crimes-2025.csv"), show_col_types = FALSE)
crimes |>
  count(primary_type, sort = TRUE) |>
  slice_head(n = 10) |>
  ggplot(aes(n, fct_reorder(primary_type, n))) +
  geom_col(fill = crju_colors["garnet"]) +
  labs(title = "Ten most common reported crimes, Chicago 2025",
       subtitle = "30,000-incident sample · Chicago Police Department data portal",
       x = "Incidents", y = NULL)
```

## Homework (Due Before Noon Tuesday)

**swirl tutorials** — interactive R lessons *inside* R:

```{r}
#| label: swirl-hw
#| eval: false
install.packages("swirl")
library(swirl)
swirl()
```

Complete these lessons from *R Programming*: **1)** Basic Building Blocks · **2)** Workspace and Files · **3)** Sequences of Numbers · **4)** Vectors · **5)** Missing Values · **6)** Subsetting Vectors · **7)** Matrices and Data Frames · **8)** Logic · **12)** Looking at Data · **15)** Base Graphics

Submit your R script files via **Canvas**. Then read **Stanton Ch. 1** before Session 2.
