---
title: "Comparing Groups and Analyzing Experiments"
subtitle: "CRJU 705 · Session 8 · Fall 2026"
author: "Scott M. Mourtgos"
format:
  revealjs:
    slide-number: true
    preview-links: auto
    chalkboard: true
    footer: "CRJU 705 · Session 8"
    theme: [default, custom-slides.scss]
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(brms)
library(emmeans)
source(here::here("R/setup.R"))
theme_set(theme_minimal(base_size = 18))
set.seed(705)
crimes <- read_csv(here::here("data/chicago-crimes-2025.csv"), show_col_types = FALSE)
```

## This Week

**Topic:** ANOVA — comparing three or more group means at once

- Why "just run a bunch of t-tests" backfires
- The core idea: two kinds of variance and their ratio, F
- `aov()` + Tukey's HSD, then Bayesian `brms` + `emmeans`
- Real data: does Chicago's clock differ by crime category?

**Learning goals:** explain what F compares; run and read a one-way ANOVA in both traditions; follow a significant F with an honest pairwise answer.

**Also this week:** Homework 8 includes **Project Checkpoint 1** — your first graded step on your final-project dataset.

## Where We've Been

::: incremental
- **Session 6:** two groups — is *this* mean different from *that* one? t-tests, in both the frequentist and Bayesian traditions
- **Session 7:** the applied workshop — you took a messy file through the whole pipeline and **picked your final-project dataset**
- **Today:** the question that comes up constantly in real research — what if there are *three* groups? Or six?
:::

# Experiments and Group Comparisons {background-color="#73000A"}

## Experiments: IVs and DVs

**Independent variable (IV):** what we change or compare — the *groups*.

**Dependent variable (DV):** what we measure — the *outcome* that (maybe) depends on the IV.

| Independent variable | Dependent variable |
|---|---|
| Type of training program | Test score |
| Type of fertilizer | Plant height |
| Region of the country | Annual precipitation |
| Patrol strategy assigned to a beat | Response time (minutes) |

: {tbl-colwidths="[50,50]"}

## Why ANOVA?

::: incremental
- A **t-test** compares **2** means
- **ANOVA** compares **3+** means
- Use it when the **IV is categorical** (the groups) and the **DV is numeric** (the outcome)
- The name is strange: **AN**alysis **O**f **VA**riance — we're asking about *means*, but the test works by comparing *variances*. That's the whole trick, and it's coming up.
:::

## Why Not Just Run a Pile of t-Tests?

Three groups means 3 pairwise tests. Six groups means 15. Each test carries its own 5% false-alarm risk, and the risks pile up:

```{r}
#| label: alpha-inflation
tibble(groups = c(2, 3, 4, 6, 10)) |>
  mutate(
    pairwise_tests   = choose(groups, 2),
    p_false_alarm    = round(1 - 0.95^pairwise_tests, 2)
  )
```

## ANOVA's Answer: One Omnibus Question

By six groups you have a coin-flip's chance of at least one bogus "significant" difference.

. . .

ANOVA asks **one omnibus question** — *are all the group means plausibly equal?* — at a single 5% risk. (When we go pairwise afterward, we'll adjust for this. Hold that thought.)

# The Core Idea: Two Kinds of Variance {background-color="#1B2A4A"}

## Imagine Three Classrooms

```{r}
#| label: classrooms
classrooms <- tibble(
  score     = c(75, 80, 78, 74, 77,    # Class 1
                88, 90, 92, 85, 87,    # Class 2
                70, 65, 68, 72, 66),   # Class 3
  classroom = factor(rep(c("Class 1", "Class 2", "Class 3"), each = 5))
)

class_stats <- classrooms |>
  summarize(class_mean = mean(score), class_var = var(score), .by = classroom)
class_stats
```

Two different kinds of "spread" are hiding in this table.

## Two Kinds of Variance

**Within-group variance:** how much scores bounce around *inside* each classroom.

**Between-group variance:** how much the classroom *averages* differ from each other.

```{r}
#| label: classrooms-two-vars
mean(class_stats$class_var)   # WITHIN: average spread inside a class
var(class_stats$class_mean)   # BETWEEN: spread of the class means
```

. . .

Between (**102.8**) dwarfs within (**7.1**) — the class averages are far more spread out than chance bouncing inside classes would suggest. Signal-to-noise ratio ≈ 14.5. These classes look genuinely different.

## The Picture

```{r}
#| label: classrooms-plot
#| echo: false
#| fig-height: 5.5
ggplot(classrooms, aes(classroom, score)) +
  geom_boxplot(alpha = 0.2, width = 0.3) +
  geom_jitter(width = 0.08, size = 3, alpha = 0.6, color = crju_colors["navy"]) +
  geom_point(data = class_stats, aes(y = class_mean),
             shape = 23, size = 5, fill = crju_colors["garnet"], color = "white") +
  labs(title = "Within vs. between group variance",
       subtitle = "Within = spread inside each box · Between = spread OF the garnet diamonds (class means)",
       x = NULL, y = "Test score")
```

## Stanton's Rain Example (From the Book)

Now flip the situation: three groups where **nothing differs** — by construction. `precip` holds annual precipitation for 70 U.S. cities. Draw three random groups of 20 from the *same* 70 cities:

```{r}
#| label: rain-groups
data("precip")
set.seed(705)   # so we all draw the same "random" groups
group1 <- sample(precip, 20)
group2 <- sample(precip, 20)
group3 <- sample(precip, 20)

c(mean(group1), mean(group2), mean(group3))
```

. . .

The means differ a little — but any differences here are **pure sampling luck**. The null hypothesis is *true by construction*.

## Rain: Within vs. Between

```{r}
#| label: rain-two-vars
mean(c(var(group1), var(group2), var(group3)))    # WITHIN
var(c(mean(group1), mean(group2), mean(group3)))  # BETWEEN
```

. . .

Within ≈ **133**, between ≈ **2.3**. The classrooms gave us 7.1 vs. 102.8 — completely flipped. Why?

## Why So Lopsided?

::: incremental
- City-to-city rainfall varies a *lot* (Phoenix vs. Mobile) → big **within**-group variance
- But each group's **mean** averages 20 cities — and Session 4 taught us means fluctuate far less than raw values ($SE = \sigma/\sqrt{n}$)
- With no real group differences, between-group variance is nothing but the tiny sampling noise of three means
- So the diagnostic: **between ≫ within → something real is separating the groups. Between ≪ within → nothing but noise.**
- All we need is a ruler for "how much bigger is big enough" —
:::

## The F-Statistic

$$F \;=\; \frac{\text{between-group variance}}{\text{within-group variance}} \;=\; \frac{\text{signal}}{\text{noise}}$$

::: incremental
- When the null is true, F bounces around low values by luck of the draw
- If our F lands **far in the tail** of the F distribution → our data are *unlikely under the null* — same logic as every test we've run
- The **F distribution** describes the F values a null world produces. Its shape depends on two df: **groups − 1** and **N − groups**
:::

## What F Looks Like When the Null Is True

```{r}
#| label: f-dist
#| output-location: slide
#| fig-height: 6
set.seed(705)
f_null <- tibble(
  design = rep(c("F(2, 57) — our rain design",
                 "F(8, 101) — more groups, bigger N"), each = 10000) |> fct_inorder(),
  f      = c(rf(10000, df1 = 2, df2 = 57), rf(10000, df1 = 8, df2 = 101))
)
cutoffs <- f_null |> summarize(cut = quantile(f, 0.95), .by = design)

ggplot(f_null, aes(f)) +
  geom_histogram(bins = 60, fill = crju_colors["blue"], color = "white") +
  geom_vline(data = cutoffs, aes(xintercept = cut),
             color = crju_colors["garnet"], linewidth = 1.1) +
  facet_wrap(~design, ncol = 1) +
  labs(title = "10,000 F values from a null world, for two designs",
       subtitle = "Garnet line = 95th percentile — F values beyond it are rare when nothing differs",
       x = "F value", y = "Count")
```

# Running the ANOVA {background-color="#73000A"}

## First, We Need a Data Frame

`aov()` wants the data in the long format we've used all semester — one row per observation, one column saying which group it belongs to:

```{r}
#| label: rain-df
rain <- tibble(
  rainfall = c(group1, group2, group3),
  group    = factor(rep(c("Group 1", "Group 2", "Group 3"), each = 20))
)
rain
```

## `aov()` Does the Whole Thing

```{r}
#| label: rain-aov
rain_model <- aov(rainfall ~ group, data = rain)
summary(rain_model)
```

. . .

**F = 0.34, p = 0.71.** That F sits squarely in the fat part of F(2, 57) — nowhere near the 95% cutoff (≈ 3.2). No evidence the groups differ. *Correct answer* — they're the same population by construction.

## The Table Is Our Two Variances, Dressed Up

::: incremental
- **Mean Sq Residuals = 132.7** — exactly the within-group variance we computed by hand ✓
- **Mean Sq group = 45.2 = 20 × 2.26** — our between-group variance, *scaled by the group size*
- Why × 20? A 2-inch gap between means built on **20 cities each** is far stronger evidence than the same gap built on 2. `aov()` bakes the sample size into the signal.
- **F = 45.2 / 132.7 = 0.34.** So when your hand-computed ratio and `aov()`'s F disagree, this scaling is why — same logic, different units.
:::

## Let's Make a Difference Artificially

Shift Group 3 down 15 inches — now there *is* a real difference, and we know exactly where it lives:

```{r}
#| label: rain-shift
rain_shifted <- rain |>
  mutate(rainfall = if_else(group == "Group 3", rainfall - 15, rainfall))

rain_shifted |> summarize(mean_rain = mean(rainfall), .by = group)
```

## Original vs. Shifted, Seen

```{r}
#| label: rain-shift-plot
#| echo: false
#| fig-height: 5.5
bind_rows(
  rain         |> mutate(version = "Original: all three from the same population"),
  rain_shifted |> mutate(version = "Group 3 shifted down 15 inches")
) |>
  mutate(version = fct_inorder(version)) |>
  ggplot(aes(group, rainfall, fill = group)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = unname(crju_colors[c("blue", "gold", "garnet")])) +
  facet_wrap(~version) +
  labs(x = NULL, y = "Annual precipitation (inches)")
```

## Redo the ANOVA

```{r}
#| label: rain-aov-shifted
shifted_model <- aov(rainfall ~ group, data = rain_shifted)
summary(shifted_model)
```

. . .

The within-group noise didn't move (132.7 — shifting a whole group doesn't change the spread *inside* it). The between-group signal exploded. **F = 14.2, p ≈ 0.00001** — deep in the tail.

## A Significant F Only Says "*Something* Differs"

Which pairs? **Tukey's HSD**: all pairwise comparisons, with the family-wise 5% false-alarm rate restored — the fix for the t-test pile-up problem from the top of class:

```{r}
#| label: rain-tukey
TukeyHSD(shifted_model)
```

## Reading the Tukey Output

::: incremental
- **Group 3 differs from both others** — intervals exclude 0, adjusted p-values tiny. Exactly where we put the difference.
- **Group 1 vs. Group 2:** interval spans 0, p = 0.81 — no evidence, *as designed*
- The omnibus F said "something's there"; Tukey pointed at *what*, without inflating the false-alarm rate
:::

# The Bayesian Version {background-color="#1B2A4A"}

## Two New Packages

- **brms** — **B**ayesian **R**egression **M**odels using **S**tan: writes the model, compiles it, samples the posterior
- **emmeans** — **e**stimated **m**arginal **means**: takes any fitted model and returns group means and group *differences*

```r
install.packages(c("emmeans", "brms"))
```

::: {.callout-warning appearance="simple"}
`brms` compiles models with your computer's C++ toolchain. Install **before** the lab, and expect the *first* model of each session to pause ~a minute at "Compiling Stan program..." — that's normal, not a crash.
:::

## Same Formula, Bayesian Machinery

```{r}
#| label: brm-rain
#| output: false
bayes_rain <- brm(rainfall ~ group, data = rain,
                  chains = 2, iter = 2000, seed = 705, refresh = 0)
```

::: incremental
- Identical formula syntax to `aov()` — this pattern carries through the rest of the course
- Priors: we didn't state any, so `brms` used its **defaults** — more on what that means on the next slide
- `chains = 2, iter = 2000`: two independent MCMC runs, 2,000 draws each → a full **posterior distribution** for every group mean
- `refresh = 0` just silences the sampling progress log
:::

## The Prior We Just Used

::: incremental
- `brms`'s defaults: essentially **flat on the group coefficients**, weakly informative on the intercept and spread
- Flat priors = "the data do the talking." Fine at this *n* — but *say so*: **"we used brms default priors"** belongs in any writeup that reports a posterior
- One logic, several badges: `ttestBF()` (S6), `brm()` (today), `correlation` (S10) — **different interfaces, same move**: prior + likelihood → posterior
- Session 10 shows what happens when you deliberately *change* the prior — and when it matters
:::

## Posterior Group Means

```{r}
#| label: emmeans-rain
rain_post <- emmeans(bayes_rain, ~ group)
rain_post
```

. . .

One row per group: the posterior median and a 95% **HPD interval** — the range holding the most credible values for that group's mean, given the data.

## Posterior Means, Seen

```{r}
#| label: plot-emmeans-rain
#| fig-height: 4.5
plot(rain_post) +
  geom_vline(xintercept = mean(rain$rainfall),
             linetype = 2, color = crju_colors["garnet"]) +
  labs(x = "Posterior mean rainfall (inches)", y = NULL)
```

All three intervals swallow the grand mean (dashed line). No group separates from the pack.

## What We Really Care About Are Differences

```{r}
#| label: pairs-rain
pairs(rain_post)
```

::: incremental
- Every 95% HPD interval for a pairwise difference **crosses 0**
- Bayesian reading: *zero — "no difference" — remains a credible value for every contrast*. No credible difference between any pair.
- Same verdict as F = 0.34, p = 0.71. Two traditions, one conclusion.
:::

## Differences, Seen

```{r}
#| label: plot-pairs-rain
#| fig-height: 4.5
plot(pairs(rain_post)) +
  geom_vline(xintercept = 0, linetype = 2, color = crju_colors["garnet"])
```

Zero sits comfortably inside every interval. When a difference is *real*, its interval abandons 0 — watch what real data do next.

# Real Data: Does the Clock Differ by Crime Type? {background-color="#73000A"}

## The Question

In Session 5 we found violent crimes average slightly *later* in the day than property crimes — two groups. But our Chicago data sort every incident into **three** categories. ANOVA takes all three in one shot:

```{r}
#| label: chicago-means
crimes |> summarize(mean_hour = mean(hour), sd_hour = sd(hour),
                    n = n(), .by = crime_category)
```

. . .

Violent runs latest (12.8), Other earliest (12.1). Gaps of fractions of an hour, with SDs near 7. Real structure, or noise? And remember: n = 30,000 — big samples detect *tiny* things.

## Look First

```{r}
#| label: chicago-box
#| echo: false
#| fig-height: 5.5
chicago_cat_means <- crimes |>
  summarize(mean_hour = mean(hour), .by = crime_category)

ggplot(crimes, aes(crime_category, hour)) +
  geom_boxplot(fill = "grey92", width = 0.45) +
  geom_point(data = chicago_cat_means, aes(y = mean_hour),
             shape = 23, size = 5, fill = crju_colors["garnet"], color = "white") +
  labs(title = "Incident hour by crime category — 30,000 incidents",
       subtitle = "Garnet diamonds = group means. The boxes nearly coincide; the eye can't settle this one.",
       x = NULL, y = "Hour of day (0–23)")
```

## Frequentist: `aov()`

```{r}
#| label: chicago-aov
chicago_model <- aov(hour ~ crime_category, data = crimes)
summary(chicago_model)
```

. . .

**F(2, 29997) = 25.3, p ≈ 10⁻¹¹.** *Something* differs, emphatically. But that's all the omnibus F will ever tell us — which pairs, and by how much?

## Which Pairs? Tukey

```{r}
#| label: chicago-tukey
TukeyHSD(chicago_model)
```

. . .

All three intervals exclude 0 — every category differs from every other. Now look at the **sizes**: 0.25 to 0.68 hours. That's **15 to 41 minutes**.

## Bayesian: Same Workflow, Real Data

```{r}
#| label: brm-chicago
#| output: false
bayes_chicago <- brm(hour ~ crime_category, data = crimes,
                     chains = 2, iter = 2000, seed = 705, refresh = 0)
```

The exact pattern from the rain example — `brm()`, then `emmeans()`, then `pairs()`. Nothing new to learn; only the data changed.

## Bayesian Differences

```{r}
#| label: pairs-chicago
chicago_post <- emmeans(bayes_chicago, ~ crime_category)
pairs(chicago_post)
```

. . .

Every HPD interval **excludes 0** — all three differences are credible. (Watch the signs: `emmeans` orders groups alphabetically, so these are Other − Property, Other − Violent, Property − Violent — the mirror image of Tukey's table. Same gaps.)

## Detectable ≠ Dramatic (Reprise)

::: incremental
- Both traditions agree: mean incident hour genuinely differs across all three categories
- But the biggest gap is **41 minutes** and the smallest is **15 minutes** — on a 24-hour clock, with SDs near 7 hours
- 30,000 incidents give us the power to detect differences this small. Detecting them is not the same as their *mattering*.
- Your job as the analyst: report **direction** *and* **size**, every time. "Significant" without magnitude is malpractice — Session 6's lesson, now with three groups.
:::

## How to Report It

> **Frequentist:** "Mean incident hour differed across crime categories, F(2, 29997) = 25.3, p < .001. Tukey's HSD: violent crimes averaged 0.25 hours (≈ 15 minutes) later than property crimes, 95% CI [0.02, 0.48], with similarly small gaps for the other pairs."

> **Bayesian:** "The posterior difference in mean hour (Violent − Property) was 0.26 hours, 95% HPD [0.06, 0.45], excluding zero — a credible but small difference."

Both sentences state the difference, the uncertainty, **and the magnitude**.

## Today's Lab & Homework

- **Lab 8:** the full ANOVA workflow, both traditions — three classrooms, then Alaska crash data ([lab page](../labs/lab-08-anova.qmd))
- **Homework 8:** ANOVA on Stanton's `chickwts` data — *plus* **Project Checkpoint 1**: descriptive statistics and one plot on **your** final-project dataset ([Session 8 page](../weeks/week-08.qmd))
- **Session 9 is the MIDTERM** — covers Sessions 1–8. Study guide on the site; no new reading. Review Stanton Ch. 1–6, the session pages, and your labs and homework.
