---
title: "Associations Between Variables: Correlation"
subtitle: "CRJU 705 · Session 10 · Fall 2026"
author: "Scott M. Mourtgos"
format:
  revealjs:
    slide-number: true
    preview-links: auto
    chalkboard: true
    footer: "CRJU 705 · Session 10"
    theme: [default, custom-slides.scss]
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(correlation)
library(patchwork)
source(here::here("R/setup.R"))
theme_set(theme_minimal(base_size = 18))
set.seed(705)
areas <- read_csv(here::here("data/chicago-areas.csv"), show_col_types = FALSE)
```

## This Week

**Topic:** measuring how two *numeric* variables move together

- Association vs. causation (and the lurking third variable)
- Covariance → **Pearson's r**: one number for direction and strength
- What r *cannot* see — and why you always plot first
- Inference on r: frequentist `cor.test()` **and** Bayesian, via the `correlation` package
- Real data: which structural features of Chicago's 77 areas track violence — and what that does **not** say about individuals

**Learning goals:** interpret r's sign and size; test it in both traditions; explain the ecological fallacy in one sentence.

## Where We've Been

::: incremental
- **Session 6:** two group means — is *this* mean different from *that* one?
- **Session 8:** three or more group means — ANOVA
- Both had the same shape: a **categorical** IV (the groups) and a **numeric** DV
- **Session 9:** the midterm. You survived. Welcome to the second half.
- **Today:** *both* variables are numeric — no groups at all. The question changes from "do groups differ?" to "**do two measures move together?**"
- This is the doorway to regression (Sessions 11–12), where your final projects live
:::

# A New Kind of Question {background-color="#73000A"}

## What Does Correlation Tell Us?

When two variables move together, how can we describe that relationship?

- One goes up when the other goes up → **positive correlation**
- One goes up when the other goes down → **negative correlation**
- Changes in one don't track changes in the other → **no correlation**

. . .

Direction is only half the story. The **strength** of the relationship tells how tightly the changes are linked — that's the number we build today.

## Which Is Which?

Positive, negative, or nothing? Make your call for each panel:

```{r}
#| label: guess-panels
#| echo: false
#| fig-height: 4.5
set.seed(705)   # inline seed: keeps these particular clouds (and their r values) stable
guess <- bind_rows(
  tibble(panel = "A", x = 1:30, y = 30:1 + rnorm(30, 0, 3)),
  tibble(panel = "B", x = 1:30, y = 1:30 + rnorm(30, 0, 3)),
  tibble(panel = "C", x = 1:30, y = rnorm(30, 15, 8))
)

ggplot(guess, aes(x, y)) +
  geom_point(size = 2.5, color = crju_colors["navy"]) +
  facet_wrap(~panel) +
  labs(x = "Variable X", y = "Variable Y")
```

## Which Is Which? — Revealed

```{r}
#| label: guess-reveal
#| echo: false
#| fig-height: 4.5
guess_r <- guess |>
  summarize(r = cor(x, y), .by = panel) |>
  mutate(label = paste0(panel, ":  r = ", sprintf("%.2f", r)))

guess |>
  left_join(guess_r, by = "panel") |>
  ggplot(aes(x, y)) +
  geom_point(size = 2.5, color = crju_colors["navy"]) +
  geom_smooth(method = "lm", se = FALSE, color = crju_colors["garnet"]) +
  facet_wrap(~label) +
  labs(x = "Variable X", y = "Variable Y")
```

. . .

The lines make it obvious — and the r values put a number on what your eye already knew.

## Real-World Examples

- **Positive:** hours studied ↑, exam scores ↑
- **Negative:** officers on shift ↑, response times ↓
- **Near zero:** shoe size vs. GPA

. . .

**Your turn:** give me a CJ example of each — *and* one pair you suspect people **believe** is correlated but isn't.

## Correlation Does NOT Mean Causation

::: incremental
- Correlation shows **association**, not cause and effect
- Ice cream sales and violent crime rise together every year. Does ice cream cause violence???
- Two variables can move together because of a **third variable** — summer heat drives both — or by pure coincidence
- When you see a correlation, ask: does X cause Y? Y cause X? Does Z drive both? Or is it luck?
- Correlation is a **pattern**, not proof of cause. **Beware the evil lurking third variable!**
:::

# From Covariance to r {background-color="#1B2A4A"}

## Covariance: Moving Together, in Raw Units

Observation by observation: do x and y sit on the *same side* of their means?

$$\text{cov}(x, y) \;=\; \frac{\sum_i (x_i - \bar{x})(y_i - \bar{y})}{n - 1}$$

::: incremental
- Both above average (or both below) → positive products → **positive covariance**; opposite sides → **negative covariance**
- The **sign** is trustworthy: covariance tells you *whether* two variables move together
- The **size** is not — watch what happens on real data
:::

## The Units Problem

Two covariances with violent crime, from our 77 Chicago community areas:

```{r}
#| label: cov-units
cov(areas$unemployment_rate, areas$violent_rate)
cov(areas$median_income,     areas$violent_rate)
```

. . .

Is −482,660 a "stronger" relationship than 108? No idea — the first number lives in *dollars × crimes per 1,000*, the second in *percentage points × crimes per 1,000*. **Covariance is glued to its units.** You can't compare across pairs.

## Correlation Rescales It

Divide the covariance by both standard deviations and the units cancel:

$$r \;=\; \frac{\text{cov}(x, y)}{s_x \, s_y}$$

```{r}
#| label: r-standardize
cov(areas$unemployment_rate, areas$violent_rate) /
  (sd(areas$unemployment_rate) * sd(areas$violent_rate))

cor(areas$unemployment_rate, areas$violent_rate)   # same thing, one function
```

. . .

Always between **−1 and +1**, unit-free, comparable: income runs r = −0.71 with violence — so unemployment (0.84) tracks violence more tightly than income does, direction aside.

## Reading r: The Conventional Bands {.smaller}

| Range of r | Strength | Direction | Example interpretation |
|---|---|---|---|
| +0.8 to +1.0 | Very strong | Positive | More prior arrests → much higher recidivism score |
| +0.4 to +0.7 | Moderate | Positive | More training hours → higher confidence |
| 0 to +0.3 | Weak | Positive | Slight upward trend |
| −0.3 to 0 | Weak | Negative | Slight inverse trend |
| −0.7 to −0.4 | Moderate | Negative | More absences → lower grades |
| −1.0 to −0.8 | Very strong | Negative | More speed → less accuracy |

: {tbl-colwidths="[22,18,18,42]"}

. . .

These bands are **conventions, not laws** — in messy social-science data, a "moderate" r can be a major finding.

## Calibrate Your Eye

```{r}
#| label: r-grid
#| echo: false
#| fig-height: 5.5
set.seed(705)   # inline seed: stable panels
r_targets <- c(0.9, 0.5, 0.2, 0, -0.5, -0.9)

r_grid <- map(r_targets, \(r) {
  x <- rnorm(150)
  y <- r * x + sqrt(1 - r^2) * rnorm(150)
  tibble(label = paste0("r = ", sprintf("%.2f", cor(x, y))), x = x, y = y)
}) |>
  list_rbind() |>
  mutate(label = fct_inorder(label))

ggplot(r_grid, aes(x, y)) +
  geom_point(alpha = 0.5, size = 1.6, color = crju_colors["navy"]) +
  facet_wrap(~label, nrow = 2) +
  labs(title = "Six simulated datasets, 150 points each",
       x = NULL, y = NULL) +
  theme(axis.text = element_blank())
```

By ±0.5 the trend is visible; below ±0.2 your eye mostly sees noise — and it's right.

# What r Cannot See {background-color="#73000A"}

## Anscombe's Quartet: Four Identical Summaries

```{r}
#| label: anscombe-stats
anscombe_long <- anscombe |>
  pivot_longer(everything(), names_to = c(".value", "set"), names_pattern = "(.)(.)")

anscombe_long |>
  summarize(mean_x = mean(x), mean_y = mean(y), r = cor(x, y), .by = set)
```

. . .

Four little datasets that ship with R (Anscombe, 1973). Same means, same r = 0.82 to two decimals — statistically interchangeable. Right?

## Anscombe's Quartet, Seen

```{r}
#| label: anscombe-plot
#| echo: false
#| fig-height: 5.5
ggplot(anscombe_long, aes(x, y)) +
  geom_point(size = 2.5, color = crju_colors["navy"]) +
  geom_smooth(method = "lm", se = FALSE, color = crju_colors["garnet"], linewidth = 0.8) +
  facet_wrap(~set, labeller = labeller(set = \(s) paste0("Set ", s, " — r = 0.82"))) +
  labs(title = "Same r, four completely different stories",
       x = NULL, y = NULL)
```

Only Set 1 is what you picture when someone says "r = 0.82." **Always plot before you correlate.**

## Limitation 1: r Only Sees Straight Lines

```{r}
#| label: u-shape
#| echo: false
#| fig-height: 5
set.seed(705)   # inline seed: stable curve
u_data <- tibble(
  x = seq(-4, 4, length.out = 50),
  y = x^2 + rnorm(50, 0, 1)
)

ggplot(u_data, aes(x, y)) +
  geom_point(size = 2.5, color = crju_colors["rust"]) +
  geom_smooth(method = "loess", se = FALSE, color = crju_colors["rust"]) +
  labs(title = "A clear relationship that r cannot see",
       subtitle = paste0("Pearson's r = ", sprintf("%.2f", cor(u_data$x, u_data$y)),
                         " — the data follow a curved (U-shaped) pattern"),
       x = "Variable X", y = "Variable Y")
```

Think stress vs. performance: an inverted U. r ≈ 0 means no **linear** relationship — not no relationship.

## Limitation 2: One Point Can Manufacture a Correlation

```{r}
#| label: outlier-demo
#| echo: false
#| fig-height: 5
set.seed(705)   # inline seed: stable demo (numbers cited on the next slide)
base_data    <- tibble(x = 1:20, y = rnorm(20, 10, 4))   # pure noise
outlier_data <- bind_rows(base_data, tibble(x = 25, y = 100))

p_base <- ggplot(base_data, aes(x, y)) +
  geom_point(size = 2.5, color = "grey35") +
  geom_smooth(method = "lm", se = FALSE, color = crju_colors["navy"]) +
  labs(title = "20 points of pure noise",
       subtitle = paste0("r = ", sprintf("%.2f", cor(base_data$x, base_data$y))),
       x = "Variable X", y = "Variable Y")

p_out <- ggplot(outlier_data, aes(x, y)) +
  geom_point(size = 2.5, color = "grey35") +
  geom_point(data = filter(outlier_data, y > 50), color = crju_colors["garnet"], size = 4.5) +
  geom_smooth(method = "lm", se = FALSE, color = crju_colors["navy"]) +
  labs(title = "Same noise + ONE outlier",
       subtitle = paste0("r = ", sprintf("%.2f", cor(outlier_data$x, outlier_data$y))),
       x = "Variable X", y = "Variable Y")

p_base + p_out
```

## What Just Happened

::: incremental
- Twenty points of pure noise happened to give r = −0.28 — **small samples produce correlations out of nothing**; that's why we'll test, not just compute
- Adding **one** extreme pair swung r from −0.28 to **+0.42** — a single point flipped the sign *and* manufactured a "moderate positive" correlation
- Neither number describes a real relationship
- Outliers can also work the other way, *hiding* a real correlation
- Defenses: **plot first** (you'd catch this instantly) — and a correlation that doesn't care about wild values —
:::

## A Rank-Based Backup: Spearman's rho

Replace each value with its **rank** (1st, 2nd, 3rd…), then correlate the ranks. Spearman asks only: *is the relationship consistently up or consistently down?*

```{r}
#| label: spearman-outlier
c(pearson  = cor(outlier_data$x, outlier_data$y),
  spearman = cor(outlier_data$x, outlier_data$y, method = "spearman"))
```

. . .

To Spearman, the wild point is just "rank 21": it drifts from `r sprintf("%.2f", cor(base_data$x, base_data$y, method = "spearman"))` to `r sprintf("%.2f", cor(outlier_data$x, outlier_data$y, method = "spearman"))` — sign intact, still weak — while Pearson swung 0.70 and flipped sign. When data are skewed or outlier-prone, check Spearman alongside Pearson.

# Real Data: Structure and Violence Across Chicago {background-color="#1B2A4A"}

## From Incidents to Areas

The anchor data's **second level** ([codebook](../chicago-data.qmd)): all 237,472 of 2025's incidents, aggregated to the **77 community areas** and joined with Census measures:

```{r}
#| label: areas-peek
areas |>
  select(community, violent_rate, unemployment_rate, median_income, pct_vacant) |>
  head(3)
```

. . .

A classic criminology-of-place question: **which structural features of a neighborhood track its violence?** Correlation is exactly the right first tool.

## Look First: Unemployment and Violence

```{r}
#| label: unemp-scatter
#| echo: false
#| fig-height: 5.5
ggplot(areas, aes(unemployment_rate, violent_rate)) +
  geom_point(size = 3, alpha = 0.7, color = crju_colors["garnet"]) +
  geom_smooth(method = "lm", se = FALSE, color = crju_colors["navy"]) +
  labs(title = "Violent-crime rate vs. unemployment rate — 77 community areas",
       subtitle = paste0("r = ", sprintf("%.2f", cor(areas$unemployment_rate, areas$violent_rate))),
       x = "Unemployment rate (%)", y = "Violent crimes per 1,000 residents")
```

One dot = one community area. This is about as strong as area-level associations get in criminology.

## The Full Menu, in One Dataset

```{r}
#| label: four-covariates
#| echo: false
#| fig-height: 5.5
areas |>
  select(violent_rate, unemployment_rate, pct_renter, median_income, median_age) |>
  pivot_longer(-violent_rate, names_to = "covariate", values_to = "value") |>
  mutate(covariate = case_match(covariate,
    "unemployment_rate" ~ paste0("Unemployment rate — r = ",
      sprintf("%.2f", cor(areas$unemployment_rate, areas$violent_rate))),
    "pct_renter" ~ paste0("% renter-occupied — r = ",
      sprintf("%.2f", cor(areas$pct_renter, areas$violent_rate))),
    "median_income" ~ paste0("Median income — r = ",
      sprintf("%.2f", cor(areas$median_income, areas$violent_rate))),
    "median_age" ~ paste0("Median age — r = ",
      sprintf("%.2f", cor(areas$median_age, areas$violent_rate))))) |>
  ggplot(aes(value, violent_rate)) +
  geom_point(alpha = 0.6, size = 2, color = crju_colors["navy"]) +
  geom_smooth(method = "lm", se = FALSE, color = crju_colors["garnet"], linewidth = 0.8) +
  facet_wrap(~covariate, scales = "free_x") +
  labs(title = "Violent-crime rate vs. four structural covariates",
       x = NULL, y = "Violent crimes per 1,000")
```

Strong positive, moderate positive, strong negative — and one genuine *nothing* (median age). Real data span the whole menu.

## These Are Areas, Not People

Our r = 0.84 says: **community areas** with higher unemployment rates record more violent crime per 1,000 residents.

::: {.callout-important appearance="simple"}
**The ecological fallacy:** using an aggregate-level association to make claims about individuals. Our data contain **no individuals** — every resident of an area gets the same 77 values. Nothing here says *unemployed people* commit violence.
:::

. . .

The classic demonstration (Robinson, 1950): across U.S. **states**, more foreign-born residents went with *higher* literacy; across **individuals**, foreign-born residents had *lower* literacy. Same data, opposite signs — immigrants had settled in high-literacy states. Aggregation can weaken, exaggerate, or even **flip** an individual-level relationship.

## Saying It Out Loud (Area-Data Edition)

✅ **Defensible:**

> "Community areas with higher unemployment rates recorded higher violent-crime rates (r = 0.84)."

❌ **Not defensible:**

> "Unemployed residents commit more violent crime." *(An individual-level claim — these data can't see individuals.)*

> "Unemployment drives up violence." *(A causal claim — concentrated disadvantage, disinvestment, and policing patterns all lurk behind both variables.)*

## Income: When Pearson and Spearman Disagree

```{r}
#| label: income-scatter
#| echo: false
#| fig-height: 5
ggplot(areas, aes(median_income, violent_rate)) +
  geom_point(size = 3, alpha = 0.7, color = crju_colors["navy"]) +
  geom_smooth(method = "loess", se = FALSE, color = crju_colors["garnet"]) +
  scale_x_continuous(labels = scales::dollar_format(scale = 1/1000, suffix = "k")) +
  labs(title = "Violent-crime rate vs. median household income",
       subtitle = paste0("Pearson r = ", sprintf("%.2f", cor(areas$median_income, areas$violent_rate)),
                         "  ·  Spearman rho = ",
                         sprintf("%.2f", cor(areas$median_income, areas$violent_rate, method = "spearman"))),
       x = "Median household income", y = "Violent crimes per 1,000")
```

The decline is real but **bent** — steep across the lower-income areas, nearly flat past ~$90k — and income is right-skewed. The rank-based Spearman (−0.82) runs stronger than Pearson (−0.71) because it only asks "downhill?", not "straight line?".

# Putting Uncertainty on r {background-color="#73000A"}

## From Description to Inference

We can *see* r = 0.84. Two honest questions remain — one per tradition:

::: incremental
- **Frequentist:** if nothing truly connected these variables, how often would 77 paired observations line up this tightly by luck?
- **Bayesian:** given the data, what *range* of correlation values is credible?
- One wrinkle: these 77 areas are **all of Chicago**, not a sample from a bag of Chicagos. Read the null as a claim about the *process generating neighborhood outcomes*, not about a larger population of areas we sampled from.
:::

## The Frequentist Test on r

- **Null hypothesis: ρ = 0** — no linear association in the process that generated the data
- Test statistic: a **t** with **n − 2** degrees of freedom — the same machinery as Sessions 6–8
- Same logic as every test we've run: *if ρ were zero, how surprising is an r like ours?*

. . .

In R, one function does the whole thing — and hands you a confidence interval too.

## `cor.test()`

```{r}
#| label: cor-test
cor.test(areas$unemployment_rate, areas$violent_rate)
```

## Reading It Honestly

::: incremental
- **r = 0.84, t(75) = 13.4, p < .001** — a null world essentially never makes 77 paired observations line up this tightly
- **95% CI [0.76, 0.90]** — Session 5's reading still applies: the *procedure* captures the true value in 95% of replications; this particular interval either does or doesn't
- Perspective: with n = 77, any |r| above about **0.22** clears p < .05. Significance is a low bar — **the strength is the story**
- And nothing in this output upgrades association to causation, or areas to individuals
:::

## The Whole Matrix at Once

Real analyses rarely stop at one pair. `cor()` takes every numeric column you feed it:

```{r}
#| label: cor-matrix
areas |>
  select(violent_rate, unemployment_rate, median_income, pct_vacant) |>
  cor() |>
  round(2)
```

. . .

Symmetric, 1s on the diagonal — a compact map of the association structure. But: no tests, no intervals.

## Same Matrix, With Uncertainty: `correlation()`

The **correlation** package (easystats) tests every pair and attaches CIs — and it **Holm-adjusts the p-values** for the number of pairs. Six tests at once is exactly Session 8's pile-of-tests problem, and the package handles it by default:

```{r}
#| label: corr-pkg
#| output-location: slide
areas |>
  select(violent_rate, unemployment_rate, median_income, pct_vacant) |>
  correlation()
```

# The Bayesian Version {background-color="#1B2A4A"}

## Bayesian Correlation: The Idea

::: incremental
- Instead of a single r and a p-value, you get a **posterior distribution** of plausible correlation values (ρ)
- You can ask direct questions: *given my data, what's the probability ρ > 0?*
- The prior encodes mild skepticism — real-world correlations are rarely extreme — gently pulling estimates toward 0
- Same package, one new argument: `bayesian = TRUE`
- And unlike `brms` in Session 8, **no compilation wait** — this samples in a blink
:::

## `correlation(..., bayesian = TRUE)`

```{r}
#| label: bayes-pair
set.seed(705)   # inline seed: posterior sampling, keeps printed values stable
bayes_pair <- areas |>
  select(violent_rate, unemployment_rate) |>
  correlation(bayesian = TRUE)
bayes_pair
```

## Reading the Bayesian Output

::: incremental
- **rho = `r sprintf("%.2f", bayes_pair$rho)`** — the posterior *median*, a touch below the frequentist 0.84: that's the prior's gentle pull toward zero
- **95% CI [`r sprintf("%.2f", bayes_pair$CI_low)`, `r sprintf("%.2f", bayes_pair$CI_high)`]** — a **credible interval**: given data and prior, 95% probability ρ lies in this range. *The direct statement Session 5 taught you the confidence interval can't make.*
- **pd = 100%** — the posterior probability the correlation is positive — and **0% in ROPE**: none of the posterior sits in a "practically zero" band
- **BF ≈ 10¹⁷** — the data are astronomically more likely under "correlated" than "not." On Session 6's scale, anything past 100 is extreme evidence; this is 10¹⁷.
:::

## The Posterior, Seen

```{r}
#| label: bayes-posterior
#| echo: false
#| fig-height: 5
set.seed(705)
post_sketch <- tibble(
  rho = rnorm(5000,
              mean = bayes_pair$rho,
              sd   = (bayes_pair$CI_high - bayes_pair$CI_low) / (2 * 1.96))
)

ggplot(post_sketch, aes(rho)) +
  geom_density(fill = crju_colors["blue"], alpha = 0.6, color = crju_colors["navy"]) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(-0.1, 1)) +
  labs(title = "Approximate posterior distribution of ρ",
       subtitle = paste0("Normal sketch from the posterior summary: median = ",
                         sprintf("%.2f", bayes_pair$rho),
                         ", 95% CI [", sprintf("%.2f", bayes_pair$CI_low),
                         ", ", sprintf("%.2f", bayes_pair$CI_high), "]"),
       x = "Correlation (ρ)", y = "Density")
```

The entire heap of credible values sits far from 0 (dashed line). *That picture* is what pd = 100% and the huge BF are summarizing.

## The Bayesian Matrix

```{r}
#| label: bayes-matrix
set.seed(705)   # inline seed: posterior sampling
areas |>
  select(violent_rate, unemployment_rate, median_income, pct_vacant) |>
  correlation(bayesian = TRUE) |>
  summary()
```

. . .

Same map as the frequentist matrix, but every entry is now a posterior median with full uncertainty machinery behind it.

## But the Priors…

The default prior is `"medium"` skepticism. Loosen it and see what moves:

```{r}
#| label: bayes-wide
set.seed(705)   # inline seed: posterior sampling
areas |>
  select(violent_rate, unemployment_rate) |>
  correlation(bayesian = TRUE, bayesian_prior = "wide")
```

## What Changed? Almost Nothing — Here

::: incremental
- The wide prior is *less* skeptical of big correlations, so it shrinks rho toward 0 slightly less — the estimate creeps closer to the raw r = 0.84
- With **77 areas of strongly correlated data, the data dominate**: the prior choice moves the second decimal
- With 15 areas and a modest correlation, the same choice could visibly move the estimate — priors matter most when data are thin
- Practical rule for this course: **defaults are fine — but say which prior you used.** If your conclusion survives medium *and* wide, say that too; it's a free robustness check.
:::

## Frequentist vs. Bayesian Correlation {.smaller}

| Feature | Frequentist (Pearson's r) | Bayesian (ρ) |
|---|---|---|
| Output | Single value, r | Posterior distribution of ρ |
| Uncertainty | 95% confidence interval | 95% credible interval |
| Evidence measure | p-value | Posterior probability, Bayes factor |
| Prior information | None | Included (by prior width) |
| Interpretation | "If no true correlation…" | "Given data, how probable is each value of ρ?" |

: {tbl-colwidths="[24,38,38]"}

## How to Report It

> **Frequentist:** "Across Chicago's 77 community areas, unemployment was strongly associated with the violent-crime rate, r = .84, t(75) = 13.4, p < .001, 95% CI [.76, .90]. This is an area-level association; it does not speak to individuals."

> **Bayesian:** "The posterior median correlation was ρ ≈ .82, 95% credible interval [.73, .87], BF₁₀ ≈ 10¹⁷ — overwhelming evidence for a strong positive area-level association."

Direction, strength, uncertainty, and the level of analysis — every time.

## Your Turn (5 Minutes)

`pct_no_hs_diploma` — the share of an area's adults without a high-school diploma. **Before you compute anything:** write down your predicted sign and rough size of its correlation with `violent_rate`.

```r
areas <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv")

ggplot(areas, aes(pct_no_hs_diploma, violent_rate)) + geom_point()
cor.test(areas$pct_no_hs_diploma, areas$violent_rate)
```

. . .

Most people predict a strong positive. The data: **r = 0.17, p = .14** — indistinguishable from nothing. Yet `pct_bachelors` runs a solid **−0.45**. Plausible-sounding covariates can fail, and correlated predictors tangle — untangling them is *exactly* Session 11's job.

## Today's Lab & Homework

- **Lab 10:** the full correlation workflow — Chicago areas, then neighborhoods and officers, including one dataset where the honest answer is *"there's nothing here"* ([lab page](../labs/lab-10-correlation.qmd))
- **Homework 10:** correlation in both traditions on the Chicago areas — *plus* **Project Checkpoint 2**: the first associations in **your** project dataset, feeding straight into Checkpoint 3's regression ([Session 10 page](../weeks/week-10.qmd))
- **Read before Session 11:** Stanton Ch. 8 — multiple regression, where today's scatterplots grow a second predictor
