---
title: "Descriptive Statistics"
subtitle: "CRJU 705 · Session 2 · Fall 2026"
author: "Scott M. Mourtgos"
format:
  revealjs:
    slide-number: true
    preview-links: auto
    chalkboard: true
    footer: "CRJU 705 · Session 2"
    theme: [default, custom-slides.scss]
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(modeest)
source(here::here("R/setup.R"))
theme_set(theme_minimal(base_size = 18))
set.seed(705)
crimes <- read_csv(here::here("data/chicago-crimes-2025.csv"), show_col_types = FALSE)
areas  <- read_csv(here::here("data/chicago-areas.csv"), show_col_types = FALSE)
```

## Week 1 Follow-Up

**swirl assignments** — experiences, issues, good stuff with R???

1. Basic Building Blocks · 2. Workspace and Files · 3. Sequences of Numbers · 4. Vectors · 5. Missing Values · 6. Subsetting Vectors · 7. Matrices and Data Frames · 8. Logic · 12. Looking at Data · 15. Base Graphics

## This Week

**Topic:** Descriptive statistics and statistical vocabulary

- Measures of central tendency
- Measures of dispersion
- Distributions and their shapes

**Learning goals:** define in your own words — *mean, median, mode, variance, standard deviation, histogram, normal distribution, Poisson distribution* — and use R to describe a small dataset and plot a distribution.

**R4DS thread:** you read Ch. 1 (Data visualization) & Ch. 2 (Workflow) — today's lab puts both to work.

# Measures of Central Tendency {background-color="#73000A"}

## Three Ways to Say "Typical"

| Measure | Meaning |
|---------|---------|
| **Mean** | the average |
| **Median** | the halfway point |
| **Mode** | the most frequently occurring value |

Measures of central tendency summarize a vector of data by locating the *typical*, the *middle*, or the *most common* point.

## Practice — The Mean

```{r}
#| label: mean-hard
rainfall <- c(0, 0, 0, 2, 1, 1, 4)   # rain each day this week

# The hard way:
sum(rainfall) / length(rainfall)
```

. . .

```{r}
#| label: mean-easy
# The easy way:
mean(rainfall)
```

## Practice — The Median

```{r}
#| label: median-hard
rainfall
sort(rainfall)   # middle value of the sorted data?
```

. . .

```{r}
#| label: median-easy
median(rainfall)
```

## Practice — The Mode

```{r}
#| label: mode
rainfall          # which value appears most often?

library(modeest)  # install.packages("modeest") once, library() every session
mfv(rainfall)     # mfv = "most frequent value"
```

## What Do These Look Like?

```{r}
#| label: rainfall-plot
#| code-line-numbers: "3-6|7-9"
#| output-location: slide
#| fig-height: 5
stats <- tibble(
  measure = c("Mean", "Median", "Mode"),
  value = c(mean(rainfall), median(rainfall), mfv(rainfall))
)
ggplot(tibble(rainfall), aes(rainfall)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  geom_vline(data = stats, aes(xintercept = value, color = measure),
             linewidth = 1.2, linetype = "dashed") +
  labs(title = "Rainfall with Mean, Median, and Mode",
       x = "Rainfall (inches)", y = "Frequency", color = NULL)
```

## Why We Care in CJ: Sentencing

```{r}
#| label: sentences
# Sentence lengths for 10 people convicted of robbery (years)
sentences <- c(2, 2, 3, 3, 3, 4, 4, 5, 5, 30)   # note the outlier!

mean(sentences)
median(sentences)
mfv(sentences)
```

One 30-year sentence drags the mean to **6.1** — nobody in the data actually got a "typical" 6-year sentence.

## Why We Care in CJ: Sentencing

```{r}
#| label: sentences-plot
#| echo: false
#| fig-height: 5
sent_stats <- tibble(
  measure = c("Mean", "Median", "Mode"),
  value = c(mean(sentences), median(sentences), mfv(sentences))
)
ggplot(tibble(sentences), aes(sentences)) +
  geom_histogram(binwidth = 1, fill = "grey85", color = "black") +
  geom_vline(data = sent_stats, aes(xintercept = value, color = measure),
             linewidth = 1.2, linetype = "dashed") +
  labs(title = "Sentence Lengths (in Years)",
       x = "Years in Prison", y = "Number of People", color = NULL)
```

## Thoughts?

- Why do these three values (mean = 6.1, median = 3.5, mode = 3) give such different impressions of "typical" sentence length?
- Which one might be most fair to use when describing a sentencing policy to the public?
- When might the mean be misleading?

## The Same Thing in Real Data

Community-level crime rates in Chicago, 2025:

```{r}
#| label: chicago-skew
areas |>
  summarize(
    mean_rate   = mean(crime_rate),
    median_rate = median(crime_rate),
    max_rate    = max(crime_rate)
  )
```

## Who's Dragging the Mean?

```{r}
#| label: chicago-worst
areas |> slice_max(crime_rate, n = 3) |> select(community, crime_rate)
```

The Loop's rate (driven by a small residential population and huge daytime crowds) pulls the mean above the median — same lesson as the 30-year sentence, in live data.

## The Same Thing in Real Data

```{r}
#| label: chicago-skew-plot
#| echo: false
#| fig-height: 5
ggplot(areas, aes(crime_rate)) +
  geom_histogram(bins = 25, fill = crju_colors["garnet"], color = "white") +
  geom_vline(xintercept = mean(areas$crime_rate), color = crju_colors["navy"],
             linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = median(areas$crime_rate), color = crju_colors["gold"],
             linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = mean(areas$crime_rate) + 12, y = 10.5,
           label = "mean", color = crju_colors["navy"], size = 6) +
  annotate("text", x = median(areas$crime_rate) - 12, y = 10.5,
           label = "median", color = "#8a6d1a", size = 6) +
  labs(title = "Crimes per 1,000 residents across 77 Chicago community areas (2025)",
       x = "Crime rate per 1,000 residents", y = "Community areas")
```

# Measures of Dispersion {background-color="#1B2A4A"}

## Central Tendency Isn't the Whole Story

Three people run for city council; **only the top two vote-getters win**:

:::: {.columns}
::: {.column width="50%"}
**Scenario 1**

- A = 200 votes
- B = 300 votes
- C = 400 votes
:::
::: {.column width="50%"}
**Scenario 2**

- X = 299 votes
- Y = 300 votes
- Z = 301 votes
:::
::::

In **both** cases the average is 300. But scenario 1 has *much greater spread* in support; scenario 2 is *tightly clustered* — it's hard to say who's truly "preferred."

→ We need **measures of dispersion**: range, deviations, variance, standard deviation.

## Practice — Range

```{r}
#| label: votes-setup
votes1 <- c(200, 300, 400)   # wide spread
votes2 <- c(299, 300, 301)   # tight cluster

mean(votes1); mean(votes2)   # identical means
```

. . .

```{r}
#| label: votes-range
range(votes1); diff(range(votes1))
range(votes2); diff(range(votes2))
```

## Practice — Deviations from the Mean

```{r}
#| label: votes-dev
votes1 - mean(votes1)
votes2 - mean(votes2)
```

. . .

**Why can't we just sum the deviations?**

```{r}
#| label: votes-sum-dev
sum(votes1 - mean(votes1))   # they always cancel to zero!
```

## Practice — Sum of Squares

Squaring makes every deviation positive (and emphasizes big misses):

```{r}
#| label: votes-ss
sum((votes1 - mean(votes1))^2)
sum((votes2 - mean(votes2))^2)
```

## Practice — Variance and SD

```{r}
#| label: votes-var
# Variance = mean squared deviation (population formula)
sum((votes1 - mean(votes1))^2) / length(votes1)

# Standard deviation = square root of variance (back to vote units!)
sqrt(sum((votes1 - mean(votes1))^2) / length(votes1))
```

. . .

```{r}
#| label: votes-var-fn
var(votes1); sd(votes1)   # R's built-ins use the SAMPLE formula (divide by n-1)
var(votes2); sd(votes2)   # more on WHY in Session 4
```

## Seeing the Machinery — Step 1: Raw Votes

```{r}
#| label: dispersion-step1
#| echo: false
#| fig-height: 5
votes_df <- tibble(
  scenario = rep(c("Wide Spread", "Tight Cluster"), each = 3),
  candidate = factor(rep(1:3, times = 2)),
  votes = c(votes1, votes2)
) |>
  mutate(dev = votes - mean(votes), sq_dev = dev^2, .by = scenario)

ggplot(votes_df, aes(candidate, votes, fill = scenario)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 300, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~scenario) +
  scale_fill_manual(values = unname(crju_colors[c("blue", "garnet")])) +
  labs(title = "Same mean (dashed line) — very different stories",
       x = "Candidate", y = "Votes")
```

## Seeing the Machinery — Step 2: Deviations

```{r}
#| label: dispersion-step2
#| echo: false
#| fig-height: 5
ggplot(votes_df, aes(candidate, dev, fill = scenario)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 0, color = "grey40") +
  facet_wrap(~scenario) +
  scale_fill_manual(values = unname(crju_colors[c("blue", "garnet")])) +
  labs(title = "How far is each value from the mean?",
       x = "Candidate", y = "Deviation from mean")
```

*(Notice the Tight Cluster panel: deviations of −1, 0, +1 are barely visible on the same scale.)*

## Seeing the Machinery — Step 3: Squared Deviations

```{r}
#| label: dispersion-step3
#| echo: false
#| fig-height: 5
ggplot(votes_df, aes(candidate, sq_dev, fill = scenario)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~scenario) +
  scale_fill_manual(values = unname(crju_colors[c("blue", "garnet")])) +
  labs(title = "Squaring: all positive, and big misses dominate",
       x = "Candidate", y = "Squared deviation")
```

*Average these and you have the variance; take the square root and you're back in vote units — the SD.*

## Standard Deviation, in Words

The **standard deviation** measures *how spread out the data are around the mean*.

> "On average, how far is each value from the mean?"

We square the deviations (to make them positive and emphasize larger differences), average them (that's the **variance**), then take the square root (back to the original units).

## Why the Standard Deviation?

::: incremental
- **Same units as the data** — variance is in squared units; SD is interpretable
- **Works with the normal distribution** — SD tells us how typical or extreme a value is
- **The basis for many tools** — confidence intervals, hypothesis tests, z-scores, regression all rely on it
- **Lets you compare variability** across datasets with similar scales
:::

## In CJ Scenarios

**Response times** for patrol officers:

- Low SD → officers respond in consistently similar times
- High SD → sometimes fast, sometimes slow — inconsistency that matters for policy, supervision, training

**Sentence lengths** for the same crime:

- High SD might suggest inconsistency — or bias — in sentencing

## Real Data: Consistency Across Communities

```{r}
#| label: chicago-sd
# Violent-crime rates: how much do Chicago communities differ?
areas |>
  summarize(
    mean_violent = mean(violent_rate),
    sd_violent   = sd(violent_rate),
    min_violent  = min(violent_rate),
    max_violent  = max(violent_rate)
  )
```

The **SD is nearly as large as the mean** — Chicago communities differ enormously in violence exposure. That inequality *is* a finding, and it took exactly one line of dispersion statistics to see it.

# Distributions {background-color="#73000A"}

## What *Is* a Distribution?

A **distribution** shows *how many observations* fall at different values — the **shape** of the data.

It answers:

- "Where do most of the values fall?"
- "How spread out are they?"
- "Are some values rare or common?"

We visualize distributions with a **histogram**.

## The Normal Distribution

```{r}
#| label: normal-hist
#| fig-height: 4
sims <- tibble(x = rnorm(n = 1000, mean = 100, sd = 10))
ggplot(sims, aes(x)) + geom_histogram(bins = 30, fill = "grey80", color = "white")
```

Symmetrical, bell-shaped; defined by **two numbers** — the mean (center) and SD (spread). **Try different values!**

## SD and the Normal: the Empirical Rule

```{r}
#| label: empirical-rule
#| echo: false
#| fig-height: 5.5
m <- mean(sims$x); s <- sd(sims$x)
bands <- tibble(
  k = 1:3,
  label = c("±1 SD ≈ 68%", "±2 SD ≈ 95%", "±3 SD ≈ 99.7%"),
  color = unname(crju_colors[c("blue", "gold", "garnet")])
)
ggplot(sims, aes(x)) +
  geom_histogram(bins = 30, fill = "grey85", color = "white") +
  geom_vline(xintercept = m, color = "red", linewidth = 1.2) +
  geom_vline(xintercept = m + c(-1, 1) * s, color = crju_colors["blue"], linewidth = 1, linetype = 2) +
  geom_vline(xintercept = m + c(-2, 2) * s, color = crju_colors["gold"], linewidth = 1, linetype = 3) +
  geom_vline(xintercept = m + c(-3, 3) * s, color = crju_colors["garnet"], linewidth = 1, linetype = 4) +
  annotate("text", x = m, y = Inf, label = "mean", vjust = 1.5, color = "red", size = 5) +
  labs(title = "±1 SD ≈ 68% · ±2 SD ≈ 95% · ±3 SD ≈ 99.7%", x = "Value", y = "Count")
```

```{r}
#| label: pct-within
pct_within <- function(k) 100 * (pnorm(k) - pnorm(-k))
tibble(k = 1:3, pct_within = round(pct_within(1:3), 2))
```

## The Poisson Distribution

```{r}
#| label: poisson-hist
#| fig-height: 4
pois_sims <- tibble(x = rpois(n = 1000, lambda = 1))
ggplot(pois_sims, aes(x)) + geom_bar(fill = crju_colors["navy"])
```

Often **skewed**, especially when λ is low. Models **counts of rare events**: 911 calls per hour, crimes per day, cars per minute. One parameter: **λ** = expected number of events. **Try different values!**

## The Poisson Is Not Hypothetical

Daily robbery counts in our Chicago sample, all of 2025:

```{r}
#| label: chicago-poisson
#| fig-height: 4.5
daily_robberies <- crimes |>
  filter(primary_type == "ROBBERY") |>
  count(date)

ggplot(daily_robberies, aes(n)) +
  geom_bar(fill = crju_colors["garnet"]) +
  labs(x = "Robberies per day", y = "Number of days",
       title = "Real crime counts look Poisson")
```

## Why We Care About Distributions in CJ

- The **normal distribution** helps us understand common variation (response times, test scores)
- The **Poisson distribution** models **event counts**, especially rare ones — crimes, incidents — and underlies many forecasting tools
- Recognizing the *shape* of your data is the first step in choosing the right analysis — a theme for the whole semester

## Today's Lab & Homework

- **Lab 2:** your first ggplots — histograms, boxplots, and bar charts with real corrections data ([lab page](../labs/lab-02-visualize.qmd))
- **Homework 2:** descriptive statistics in R (posted on the [Session 2 page](../weeks/week-02.qmd))
- **Read before Session 3:** Stanton Ch. 2 · R4DS Ch. 3 (Data transformation) & Ch. 4 (Code style)
