---
title: "Applied Workshop: Real Questions, Real Tests"
subtitle: "CRJU 705 · Session 7 · Fall 2026"
author: "Scott M. Mourtgos"
format:
  revealjs:
    slide-number: true
    preview-links: auto
    chalkboard: true
    footer: "CRJU 705 · Session 7"
    theme: [default, custom-slides.scss]
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
source(here::here("R/setup.R"))
source(here::here("R/bayes.R"))
options(scipen = 0)   # print p-values the way students' consoles do (2.2e-16)
theme_set(theme_minimal(base_size = 18))
set.seed(705)

rossi   <- read_csv(here::here("data/rossi.csv"), show_col_types = FALSE)
arrests <- read_csv(here::here("data/toronto-arrests.csv"), show_col_types = FALSE)
areas   <- read_csv(here::here("data/chicago-areas.csv"), show_col_types = FALSE) |>
  mutate(violence = factor(if_else(high_violence, "High violence", "Other areas"),
                           levels = c("High violence", "Other areas")))
shootings <- read_csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/v2/fatal-police-shootings-data.csv",
                      show_col_types = FALSE) |>
  mutate(period = if_else(year(date) >= 2020, "2020 to 2024", "2015 to 2019"))
```

## This Week

**Topic:** no new statistics. Tonight we *use* what you already have.

- Four real questions from four corners of criminal justice: reentry, pretrial release, neighborhoods, police use of force
- Every one answered with tools from Sessions 5 and 6: `count()`, `prop.test()`, `bayes_prop_test()`, `t.test()`
- Then: **what data is out there**, so you can choose your final-project dataset this week

**Learning goal:** given a two-group question on a dataset you have never seen, pick the test, run it, and say what it means (and what it does not).

## How Today Works

:::: columns
::: {.column width="33%"}
**Lecture**

Four worked examples, one recipe
:::

::: {.column width="33%"}
**Lab 7**

Two more, and you write the code
:::

::: {.column width="33%"}
**What's out there**

A menu of real datasets to explore this week
:::
::::

Everything I run tonight is written up in the [demo walkthrough](../demos/demo-07-workshop.qmd).

## The Recipe

Every question tonight, the same four steps:

1. **Question.** Name the two groups and the outcome
2. **Count or describe.** `count()` for yes/no; group averages for a number
3. **Test.** Frequentist always; Bayesian too when the outcome is yes/no
4. **Say it.** The honest sentence · how big, in real units · one thing this test **cannot** tell you

## Which Test?

| The outcome is... | Example | Test |
|---|---|---|
| **yes/no** | arrested again? searched? released? | `prop.test()` and `bayes_prop_test()` |
| **a number** | age, income, number of prior arrests | `t.test(outcome ~ group, data = )` |

. . .

That table covers every question in tonight's lecture and lab, and a surprising share of what gets published in criminal justice journals.

# Example 1 · Reentry: Does Money at Release Keep People Out of Jail? {background-color="#73000A"}

## The Study

- **432 men** released from Maryland state prisons in the 1970s, followed for one year
- Half were **randomly assigned** to receive financial aid after release; half were not
- Outcome: **arrested again** within the year (1 = yes)
- One row = one released man

. . .

**Question:** did the aid group get re-arrested less often? Two groups, yes/no outcome: `prop.test()` and `bayes_prop_test()`.

## Count

```{r}
#| label: rossi-count
rossi |> count(fin, arrest)
```

- **Aid** (`fin = yes`): 48 re-arrested out of 48 + 168 = **216** (22.2%)
- **No aid:** 66 out of 66 + 150 = **216** (30.6%)

Aid first all the way through, so every difference reads **Aid minus No aid**.

## Test: `prop.test()`

```{r}
#| label: rossi-prop
prop.test(x = c(48, 66), n = c(216, 216))
```

## Reading It

- Interval: about **17 points lower to 0.4 points higher**. Zero is (just) inside
- **p = 0.063**. Above 0.05: fail to reject the null

. . .

A funder reading only the p-value writes "no significant effect" and cuts the program. Sound familiar? It is Session 6's recidivism trial, except this one is real.

## Test: `bayes_prop_test()`

```{r}
#| label: rossi-bayes
bayes_prop_test(x = c(48, 66), n = c(216, 216))
```

## Say It

- **The sentence:** given these data, there is about a **97% probability** that men who got aid were re-arrested at a lower rate
- **How big:** about **22 of every 100** men with aid were re-arrested, against about **31 of every 100** without: roughly 8 fewer re-arrests per 100 people released
- **What this test cannot tell you:** whether the effect lasts past one year, or whether aid would work the same way today

. . .

**"Not significant" is not "no effect."** Here the best estimate is an 8-point drop, and the data say it is very probably real.

## Why We Can Say "Caused"

The aid was handed out **at random**, so the two groups should look alike before release. We can check, with a number: prior convictions.

```{r}
#| label: rossi-check
rossi <- rossi |> mutate(fin = factor(fin, levels = c("yes", "no")))
t.test(prio ~ fin, data = rossi)
```

## Reading the Check

- Aid: **2.98** prior convictions on average. No aid: **2.99**. p = 0.99
- The groups started out the same, so the gap afterward is about the aid

. . .

This is the rare test where you **hope** for a big p-value. Random assignment is what lets a difference mean "caused." None of tonight's other three examples has it.

# Example 2 · Pretrial Release: Does a Job Get You Released? {background-color="#1B2A4A"}

## The Data

- **5,226 arrests** by Toronto police for possessing small amounts of marijuana, 1997 to 2002
- Outcome: **released** with a summons (instead of held for a bail hearing)
- Group: was the person **employed**?
- One row = one arrest

## Count

```{r}
#| label: tor-count
arrests |> count(employed, released)
```

- **Employed:** 3,568 released out of 3,568 + 543 = **4,111** (86.8%)
- **Not employed:** 766 out of 766 + 349 = **1,115** (68.7%)

## Test

```{r}
#| label: tor-prop
#| output-location: slide
prop.test(x = c(3568, 766), n = c(4111, 1115))
```

## Reading It

- Employed minus not employed: about **15 to 21 points**, p below 2.2e-16
- `bayes_prop_test()` gives the same range (15.2 to 21.0 points) and more than a 99% probability the employed rate is higher. With 5,000 cases, the two traditions agree

. . .

**So a job gets you released?** Not so fast. What *else* is different about these two groups?

## The Groups Differ in Other Ways

`checks` counts how many police databases (prior arrests, convictions, parole, and so on) had the person's name in them.

```{r}
#| label: tor-checks
arrests |> summarize(avg_checks = mean(checks), people = n(), .by = employed)
```

People without a job had **more of a record**, and a record is a reason to hold someone. The 18-point gap could be partly about records, not jobs.

## Compare Like With Like

Keep only people with **no record at all** (`checks == 0`), and ask again:

```{r}
#| label: tor-clean
arrests |> filter(checks == 0) |> count(employed, released)
```

- **Employed:** 1,526 of 1,650 released (92.5%)
- **Not employed:** 166 of 201 (82.6%)

## Test, Again

```{r}
#| label: tor-clean-prop
#| output-location: slide
prop.test(x = c(1526, 166), n = c(1650, 201))
```

## Say It

- **Among people with clean records**, the employed were released about **10 points** more often (interval about 4 to 16 points, p = 0.000004)
- The gap **shrank** from 18 points to 10 when we compared like with like, but it did **not disappear**
- **What this cannot tell you:** whether some *other* difference (age, citizenship, where the arrest happened) explains the rest. Only random assignment rules everything out, and nobody randomizes jobs

. . .

`filter()` is the simplest honest move you have: **make the groups more alike, then ask again.**

# Example 3 · Neighborhoods: Violence and Income {background-color="#73000A"}

## The Data

- Chicago's **77 community areas** (our anchor data, 2025 crimes plus census numbers)
- Group: the 38 areas with the highest violent-crime rates vs the other 39
- Outcome: **median household income**, a number in dollars
- One row = one **area**, not one person

. . .

The outcome is a number, so: `t.test()`.

## Describe

```{r}
#| label: chi-describe
areas |>
  summarize(avg_income = mean(median_income), areas = n(), .by = violence) |>
  arrange(violence)
```

## Picture It

```{r}
#| label: chi-plot
#| echo: false
#| fig-height: 4.6
ggplot(areas, aes(x = violence, y = median_income)) +
  geom_boxplot(fill = crju_colors[["gold"]], width = 0.5) +
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(x = NULL, y = "Median household income",
       title = "Chicago community areas, 2025")
```

## Test

```{r}
#| label: chi-ttest
#| output-location: slide
t.test(median_income ~ violence, data = areas)
```

## Say It {.smaller}

- High-violence areas minus other areas: about **$25,000 to $47,000 lower** median income (averages $54,355 vs $90,088), p = 0.00000001
- **How big:** the typical household in a high-violence area earns roughly **$36,000 a year less**. The unit is dollars because the column is dollars
- **What this cannot tell you:**
  - which way it runs (does poverty breed violence, or does violence drive out income?) or what else explains both
  - anything about individual *people*. Each row is an area. "People with less income commit more violence" is a claim about people, and these data are not about people

# Example 4 · Use of Force: Body Cameras Over Time {background-color="#1B2A4A"}

## The Data

- The *Washington Post*'s database of **every person shot and killed by on-duty police** in the U.S., 2015 to 2024: **10,430 people**
- `body_camera` is TRUE when news or police reports said a body camera recorded the incident
- One row = one person killed

. . .

**Question:** after 2020's wave of police reform, did body camera footage exist more often? Group: 2015 to 2019 vs 2020 to 2024. Yes/no outcome.

## Count

```{r}
#| label: wapo-count
shootings |> count(period, body_camera)
```

- **2020 to 2024:** 1,205 of 1,205 + 4,302 = **5,507** (21.9%)
- **2015 to 2019:** 594 of 594 + 4,329 = **4,923** (12.1%)

## Test

```{r}
#| label: wapo-bayes
#| output-location: slide
bayes_prop_test(x = c(1205, 594), n = c(5507, 4923))
```

## Say It

- `prop.test()`: about **8 to 11 points** higher, p below 2.2e-16. Bayesian: 8.4 to 11.2 points, more than 0.99
- **How big:** about **12 in 100** fatal shootings had known camera footage before 2020, and about **22 in 100** after. Close to doubling, yet still fewer than 1 in 4
- **What this cannot tell you:** whether cameras changed anything about the shootings, or how many cameras *existed* but were never reported

. . .

With 10,430 cases, the p-value was never in doubt. **The size is the whole story.**

## Four Questions, One Recipe {.smaller}

| | Question | Test | Answer | The lesson |
|---|---|---|---|---|
| **Reentry** | Does aid lower re-arrest? | `prop.test()` + Bayes | 22% vs 31%; p = 0.063; 97% probably lower | "Not significant" is not "no effect." Randomization lets us say "caused" |
| **Pretrial** | Does a job get you released? | `prop.test()`, then again after `filter()` | 18 points; 10 among clean records | Groups differ in other ways. Compare like with like |
| **Neighborhoods** | Lower income where violence is high? | `t.test()` | $25k to $47k lower | Size in real units. Areas are not people |
| **Use of force** | More body camera footage after 2020? | `prop.test()` + Bayes | 12% to 22% | Huge n: the p-value adds nothing; the size is the story |

# Now You: Lab 7 {background-color="#73000A"}

## Lab 7

Two more real questions, same recipe, and this time you write the code ([lab page](../labs/lab-07-workshop.qmd)):

- **With me:** a pretrial risk-assessment tool. Do people it labels High risk get re-arrested more often than people it labels Low risk?
- **Your Turn:** Minneapolis police stops. Are people searched more often in suspicious-person stops than in traffic stops? Then a `t.test()` on the *Post* data

Submit your Your Turn script to Canvas before you leave.

# What's Out There {background-color="#1B2A4A"}

## Your Final Project Needs a Dataset

The **[Final-Project Data Options](../project-data.qmd)** page lists real criminal justice datasets that load with **one line of R**:

- Reentry, pretrial release, police stops, police shootings, risk assessment, public opinion about crime and police, state crime rates
- Each one: what a row is, how many rows, the key variables, a starter question, and whether it has the numeric variables regression will need (Sessions 10 and 11)
- You may use any of them, **including tonight's**, as long as your question goes beyond what we did in class

## Two Ways to Load a Dataset

Our course copy, the way you have loaded everything all semester:

```r
arrests <- read_csv("https://smourtgos.github.io/crju705/data/toronto-arrests.csv")
```

The original, from an R package (install once, then load it each time):

```r
install.packages("carData")   # once, in the console
library(carData)
data(Arrests)
```

Same data either way. Use whichever you like.

## How to Size Up a Dataset in Five Minutes

1. **What is one row?** A person? An arrest? A neighborhood? A year?
2. **How many rows?** Hundreds is plenty; millions is a headache
3. **What are the variables?** `glimpse()` tells you names and types
4. **Anything strange?** 999s, negative ages, lots of `NA`: `summary()` finds them
5. **Could it answer a question you care about?** Name the groups, or the variables

## Homework 7 and Next Week

- **Homework 7:** explore **two** datasets (from the menu, or one you found and cleared with me), then **choose one** for your final project. Due **Sunday by 11:59 pm** (posted on the [Session 7 page](../weeks/week-07.qmd))
- No R4DS reading this week. The wrangling thread is done
- **Before Session 8:** read Stanton Ch. 6 (ANOVA), and bring your chosen dataset. Project checkpoints begin
