---
title: "Applied Data Workshop"
subtitle: "CRJU 705 · Session 7 · Fall 2026"
author: "Scott M. Mourtgos"
format:
  revealjs:
    slide-number: true
    preview-links: auto
    chalkboard: true
    footer: "CRJU 705 · Session 7"
    theme: [default, custom-slides.scss]
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(BayesFactor)
source(here::here("R/setup.R"))
theme_set(theme_minimal(base_size = 18))
set.seed(705)
```

## This Week

**Topic:** the full pipeline, on data nobody cleaned for you

- No new statistics today — everything you need, you already have
- One worked pass through the pipeline together (fast)
- Then most of the session is **Lab 7, on your own data**

**Learning goals:** run import → tidy → transform → visualize → test start-to-finish without a template; diagnose what's wrong with a dataset you've never seen; commit to a final-project dataset you can defend.

**Due this week:** your **final-project dataset selection** (details at the end).

## Real Analysis Is 80% Preparation

The folk wisdom in data work: **80% of the time goes to getting and preparing data, 20% to analyzing it.** Six sessions in, you can now verify this yourself.

::: incremental
- Finding data that can actually answer your question
- Importing it without silently mangling anything
- Discovering what's wrong with it (something always is)
- Building the variables you actually need — the file never has them
- …and then the t-test is **one line of code**
:::

. . .

Every dataset in this course so far arrived clean because *I* cleaned it. That subsidy ends today.

## The Pipeline You Already Own {.smaller}

| Step · learned in | The job | Tools |
|-------------------|---------|-------|
| **Import** · Session 5 | get data into R intact | `read_csv()`, `read_excel()`, `col_types`, `mdy()` |
| **Tidy** · Session 4 | one row = one observation | `pivot_longer()`, `pivot_wider()` |
| **Transform** · Session 3 | filter, derive, summarize | `filter()`, `mutate()`, `summarize()` |
| **Visualize** · Session 2 | see the shape and the pattern | `ggplot()` |
| **Model / test** · Sessions 5–6 | formal inference | `t.test()`, `ttestBF()`, `prop.test()` |

: {tbl-colwidths="[27,27,46]"}

You learned the first four steps in *reverse* order. Today you run the whole thing left to right for the first time.

# One Pass Through the Pipeline {background-color="#73000A"}

## Step 1 · Import

Back to the Alaska crash data — but this time the date is handled *at the door*:

```{r}
#| label: import
crashes <- read_csv(
  "https://smourtgos.github.io/crju705/data/crash-ak.csv",
  col_types = cols(date = col_date(format = "%m/%d/%Y"))
)
# (equivalent Lab 5 habit: import, then mutate(date = mdy(date)))

sum(is.na(crashes$date))   # zero NAs = every date parsed
```

20,000 crashes, 2022–2026. Two roads to a real date column — `col_types` up front, or `mdy()` after. Either way: **verify the parse**.

## Step 2 · Clean & Derive

The file has no "night" or "weekend" column. Files never do — *you* build the variables your question needs:

```{r}
#| label: derive
crashes <- crashes |>
  mutate(
    hour    = hour(time),
    weekend = wday(date, week_start = 1) >= 6,   # Mon = 1 … Sat = 6, Sun = 7
    night   = hour >= 22 | hour < 4              # 10 p.m. – 3:59 a.m.
  )
crashes |> count(night)
```

Night hours hold 25% of all crashes — but 39% of intoxication-involved ones. One `mutate()` made that pattern visible.

## Step 3 · Describe

The Homework 5 question, back for a verdict: **do crashes involving an intoxicated driver injure more people?**

```{r}
#| label: describe
crashes |>
  summarize(
    crashes  = n(),
    mean_inj = mean(injuries),
    sd_inj   = sd(injuries),
    .by = intox_any
  )
```

0.74 vs. 1.13 injured people per crash. Point estimates — you know by now that's not the final word.

## Step 4 · Visualize

```{r}
#| label: viz
#| output-location: slide
#| fig-height: 5
inj_summary <- crashes |>
  mutate(intox_group = factor(intox_any, labels = c("No intox", "Intox involved"))) |>
  summarize(n = n(), mean_inj = mean(injuries), sd_inj = sd(injuries),
            .by = intox_group) |>
  mutate(se = sd_inj / sqrt(n),
         ci_lo = mean_inj - 1.96 * se,
         ci_hi = mean_inj + 1.96 * se)   # Session 5, by hand

ggplot(inj_summary, aes(intox_group, mean_inj)) +
  geom_pointrange(aes(ymin = ci_lo, ymax = ci_hi),
                  color = crju_colors["garnet"], linewidth = 1.2, size = 1) +
  labs(title = "Mean injuries per crash, with 95% confidence intervals",
       subtitle = "Alaska crash data, n = 20,000",
       x = NULL, y = "Mean injuries per crash")
```

## Step 5a · Test It — Frequentist

```{r}
#| label: ttest
t.test(injuries ~ intox_any, data = crashes)
```

Same gotcha as Homework 5: R computes group 0 − group 1, so the negative signs mean the *intoxication* group is higher.

## Step 5b · Test It — Bayesian

`BayesFactor` insists the grouping variable be a **factor** — so we hand it one:

```{r}
#| label: scipen-reset
#| include: false
# setup.R sets scipen = 999; without resetting, the ~2e43 Bayes factor prints
# as a 44-digit integer. Students' consoles (default scipen) show e-notation.
options(scipen = 0)
```

```{r}
#| label: ttest-bf
crashes <- crashes |>
  mutate(intox_group = factor(intox_any, labels = c("No intox", "Intox involved")))

ttestBF(formula = injuries ~ intox_group, data = crashes)
```

BF~10~ ≈ 2 × 10^43^: the data are astronomically better predicted by "the groups differ" than by "no difference."

::: aside
Depending on your R settings, this may print as `2.06e+43` or as a 44-digit integer — same number, different formatting. Don't panic either way.
:::

## Two Grammars, One Verdict

::: incremental
- **Frequentist:** *if* the true difference were zero, data like ours would essentially never occur — t(1733) = −12.5, p < .001
- **Bayesian:** the difference model beats the no-difference model by a factor of ~10^43^
- **The estimate, in English:** intoxication-involved crashes average **0.33 to 0.45 more injured people per crash** (95% CI)
:::

## Two Grammars, One Verdict, continued

::: incremental
- The two frameworks agree *here* because the evidence is overwhelming; with small samples and borderline evidence, they can part ways — that's when running both matters
- **What neither test says:** *why*. Intoxication crashes also differ in hour, speed, and road conditions — an observational gap no test statistic closes
:::

# Now You {background-color="#1B2A4A"}

## Today's Workshop

The rest of the session is **[Lab 7](../labs/lab-07-workshop.qmd)** — the same pipeline, run twice:

1. **With me, on the crash data** — quickly; you've seen every piece before
2. **Alone, on the dataset you're proposing for your final project** — import it, size up its damage, and decide if it can carry a semester

## Workshop Ground Rules

- Work in your own `.R` script from minute one — it becomes Homework 7
- Stuck? Use the Lab 5 ladder: read the error → help page → reprex → me
- Your neighbor's dataset will break in ways yours didn't. **Look over shoulders** — other people's messes are free lessons
- **Exit ticket** before you leave

## Dataset Selection — Due This Week

Your final-project dataset, formally declared in **Homework 7** (before noon Tuesday):

- A **criminal-justice** dataset — policing, crime, courts, corrections, public opinion
- It **imports into R** — you prove it with working code
- It has **numeric variables** that will support correlation and regression (Sessions 10–11)
- You can state your **research question in one sentence**

Torn between two? Submit both — we settle it at your Session 8 checkpoint. Full requirements: **[final project page](../final-project.qmd)**.

## Today's Lab & Homework

- **Lab 7:** the full pipeline — on the crash data, then on *your* data ([lab page](../labs/lab-07-workshop.qmd))
- **Homework 7:** workshop write-up **+ formal dataset selection** (posted on the [Session 7 page](../weeks/week-07.qmd))
- **Read before Session 8:** Stanton Ch. 6 (ANOVA) — and project checkpoints begin
