---
title: "Applied Workshop: Real Questions, Real Tests"
subtitle: "CRJU 705 · Session 7 · Fall 2026"
author: "Scott M. Mourtgos"
format:
  revealjs:
    slide-number: true
    preview-links: auto
    chalkboard: true
    footer: "CRJU 705 · Session 7"
    theme: [default, custom-slides.scss]
    fig-align: center
    include-in-header:
      - text: '<link rel="stylesheet" href="../widgets/gap-explorer.css">'
    include-after-body:
      - text: '<script src="../widgets/gap-explorer.js"></script>'
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(plotly)
source(here::here("R/setup.R"))
source(here::here("R/bayes.R"))
source(here::here("R/gap-explorer.R"))   # "Where Is Zero?" widgets, numbers from the real tests
options(scipen = 0)   # print p-values the way students' consoles do (2.2e-16)
theme_set(theme_minimal(base_size = 18))
set.seed(705)

rossi   <- read_csv(here::here("data/rossi.csv"), show_col_types = FALSE)
arrests <- read_csv(here::here("data/toronto-arrests.csv"), show_col_types = FALSE)
areas   <- read_csv(here::here("data/chicago-areas.csv"), show_col_types = FALSE) |>
  mutate(violence = factor(if_else(high_violence, "High violence", "Other areas"),
                           levels = c("High violence", "Other areas")))
shootings <- read_csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/v2/fatal-police-shootings-data.csv",
                      show_col_types = FALSE) |>
  mutate(period = if_else(year(date) >= 2020, "2020 to 2024", "2015 to 2019"))

# Descriptive animation colors and size (Chicago areas, body cameras by year)
garnet <- "#73000A"; blue <- "#6A8DBE"; gray <- "#BDBDBD"; navy <- "#1B2A4A"
anim_w <- 950; anim_h <- 470
```

## This Week

**Topic:** no new statistics. Tonight we *use* what you already have.

- Four real questions from four corners of criminal justice: reentry, pretrial release, neighborhoods, police use of force
- Every one answered with tools from Sessions 5 and 6: `count()`, `prop.test()`, `bayes_prop_test()`, `t.test()`
- Then: **what data is out there**, so you can choose your final-project dataset this week

**Learning goal:** given a two-group question on a dataset you have never seen, pick the test, run it, and say what it means (and what it does not).

## How Today Works

:::: columns
::: {.column width="33%"}
**Lecture**

Four worked examples, one recipe, and a picture of what each test measures
:::

::: {.column width="33%"}
**Lab 7**

Two more, and you write the code
:::

::: {.column width="33%"}
**What's out there**

A menu of real datasets to explore this week
:::
::::

Everything I run tonight is written up in the [demo walkthrough](../demos/demo-07-workshop.qmd), animations included.

## The Recipe

Every question tonight, the same four steps:

1. **Question.** Name the two groups and the outcome
2. **Count or describe.** `count()` for yes/no; group averages for a number
3. **Test.** Frequentist always; Bayesian too when the outcome is yes/no
4. **Say it.** The honest sentence · how big, in real units · one thing this test **cannot** tell you

## Which Test?

| The outcome is... | Example | Test |
|---|---|---|
| **yes/no** | arrested again? searched? released? | `prop.test()` and `bayes_prop_test()` |
| **a number** | age, income, number of prior arrests | `t.test(outcome ~ group, data = )` |

. . .

That table covers every question in tonight's lecture and lab, and a surprising share of what gets published in criminal justice journals.

# Example 1 · Reentry: Does Money at Release Keep People Out of Jail? {background-color="#73000A"}

## The Study

- **432 men** released from Maryland state prisons in the 1970s, followed for one year
- Half were **randomly assigned** to receive financial aid after release; half were not
- Outcome: **arrested again** within the year (1 = yes)
- One row = one released man

. . .

**Question:** did the aid group get re-arrested less often? Two groups, yes/no outcome: `prop.test()` and `bayes_prop_test()`.

## Count

```{r}
#| label: rossi-count
rossi |> count(fin, arrest)
```

- **Aid** (`fin = yes`): 48 re-arrested out of 48 + 168 = **216** (22.2%)
- **No aid:** 66 out of 66 + 150 = **216** (30.6%)

Aid first all the way through, so every difference reads **Aid minus No aid**.

## Test: `prop.test()`

```{r}
#| label: rossi-prop
prop.test(x = c(48, 66), n = c(216, 216))
```

## Reading It

- Interval: about **17 points lower to 0.4 points higher**. Zero is (just) inside
- **p = 0.063**. Above 0.05: fail to reject the null

. . .

A funder reading only the p-value writes "no significant effect" and cuts the program. Sound familiar? It is Session 6's recidivism trial, except this one is real. Now ask the other question.

## Test: `bayes_prop_test()`

```{r}
#| label: rossi-bayes
bayes_prop_test(x = c(48, 66), n = c(216, 216))
```

## Where Is Zero?

```{r}
#| label: rossi-widget
#| echo: false
#| output: asis
ge_widget(
  ge_prop("Aid vs no aid", A = "Aid", B = "No aid", out = "re-arrested",
          xa = 48, na = 216, xb = 66, nb = 216,
          left = "aid group re-arrested less", right = "aid group re-arrested more",
          rerun = "same program, 432 different men", unit_group = "men"),
  default = 1)
```

## Say It

- **The sentence:** given these data, there is about a **97% probability** that men who got aid were re-arrested at a lower rate
- **How big:** about **22 of every 100** men with aid were re-arrested, against about **31 of every 100** without: roughly 8 fewer re-arrests per 100 people released
- **What this test cannot tell you:** whether the effect lasts past one year, or whether aid would work the same way today

. . .

**"Not significant" is not "no effect."** Here the best estimate is an 8-point drop, and the data say it is very probably real.

# Example 2 · Pretrial Release: Does a Job Get You Released? {background-color="#1B2A4A"}

## The Data

- **5,226 arrests** by Toronto police for possessing small amounts of marijuana, 1997 to 2002
- Outcome: **released** with a summons (instead of held for a bail hearing)
- Group: was the person **employed**?
- One row = one arrest

## Count

```{r}
#| label: tor-count
arrests |> count(employed, released)
```

- **Employed:** 3,568 released out of 3,568 + 543 = **4,111** (86.8%)
- **Not employed:** 766 out of 766 + 349 = **1,115** (68.7%)

## Test: `prop.test()`

```{r}
#| label: tor-prop
#| output-location: slide
prop.test(x = c(3568, 766), n = c(4111, 1115))
```

## Test: `bayes_prop_test()`

```{r}
#| label: tor-bayes
#| output-location: slide
bayes_prop_test(x = c(3568, 766), n = c(4111, 1115))
```

## Reading It

- `prop.test()`: employed minus not employed, about **15 to 21 points**, p below 2.2e-16
- `bayes_prop_test()`: **15.2 to 21.0 points**, and more than a 99% probability the employed rate is higher
- With 5,000 cases, the two traditions land on the same range

. . .

**So a job gets you released?** Not so fast. What *else* is different about these two groups?

## The Groups Differ in Other Ways

`checks` counts how many police databases (prior arrests, convictions, parole, and so on) had the person's name in them.

```{r}
#| label: tor-checks
arrests |> summarize(avg_checks = mean(checks), people = n(), .by = employed)
```

People without a job had **more of a record**, and a record is a reason to hold someone. The 18-point gap could be partly about records, not jobs.

## Compare Like With Like

Keep only people with **no record at all** (`checks == 0`), and ask again.

```{r}
#| label: tor-clean
arrests |> filter(checks == 0) |> count(employed, released)
```

- **Employed:** 1,526 of 1,650 released (92.5%)
- **Not employed:** 166 of 201 (82.6%)

## Test Again: `prop.test()`

```{r}
#| label: tor-clean-prop
#| output-location: slide
prop.test(x = c(1526, 166), n = c(1650, 201))
```

## Test Again: `bayes_prop_test()`

```{r}
#| label: tor-clean-bayes
#| output-location: slide
bayes_prop_test(x = c(1526, 166), n = c(1650, 201))
```

## Where Is Zero?

```{r}
#| label: tor-widget
#| echo: false
#| output: asis
ge_widget(
  ge_prop("Everyone", A = "Employed", B = "Not employed", out = "released",
          xa = 3568, na = 4111, xb = 766, nb = 1115,
          left = "employed released less often", right = "employed released more often",
          rerun = "another 5,226 arrests", unit_group = "arrests"),
  ge_prop("No record only", A = "Employed", B = "Not employed", out = "released",
          xa = 1526, na = 1650, xb = 166, nb = 201,
          left = "employed released less often", right = "employed released more often",
          rerun = "another 1,851 arrests with no record", unit_group = "arrests"),
  default = 2)
```

## Say It

- **Among people with clean records**, the employed were released about **10 points** more often: interval about 4 to 16 points (p = 0.000004), credible interval 5.1 to 15.8 points, and more than a 99% probability the employed rate is higher
- The gap **shrank** from 18 points to 10 when we compared like with like, but it did **not disappear**
- **What this cannot tell you:** whether some *other* difference (age, citizenship, where the arrest happened) explains the rest

. . .

`filter()` is the simplest honest move you have: **make the groups more alike, then ask again.**

# Example 3 · Neighborhoods: Violence and Income {background-color="#73000A"}

## The Data

- Chicago's **77 community areas** (our anchor data, 2025 crimes plus census numbers)
- Group: the 38 areas with violent-crime rates above the city's median vs the other 39
- Outcome: **median household income**, a number in dollars
- One row = one **area**, not one person

. . .

The outcome is a number, so: `t.test()`.

## Describe

```{r}
#| label: chi-describe
areas |>
  summarize(avg_income = mean(median_income), areas = n(), .by = violence) |>
  arrange(violence)
```

## See It

Each dot is one community area. Press Play.

```{r}
#| label: chi-anim
#| echo: false
set.seed(705)
jit <- runif(nrow(areas), -0.18, 0.18)
stage_labels <- c("All 77 areas", "Split by violence", "The two averages")
chi_pts <- map_dfr(1:3, \(s) areas |>
  mutate(x = if (s == 1) 1.5 + jit else if_else(violence == "High violence", 1, 2) + jit,
         stage = stage_labels[s])) |>
  mutate(stage = factor(stage, levels = stage_labels),
         hover = paste0(community, "<br>$", format(round(median_income), big.mark = ",")))
chi_means <- areas |>
  summarize(avg = mean(median_income), .by = violence)
chi_means <- map_dfr(1:3, \(s) chi_means |>
  mutate(x = if (s < 3) -1 else if_else(violence == "High violence", 1, 2),
         stage = stage_labels[s])) |>
  mutate(stage = factor(stage, levels = stage_labels),
         txt = paste0("Average<br>$", format(round(avg), big.mark = ",")))
chi_gap <- tibble(stage = factor(stage_labels, levels = stage_labels),
                  x = c(-1, -1, 1.5), y = 175000, txt = "Gap: about $35,700")

plot_ly(width = anim_w, height = anim_h) |>
  add_markers(data = chi_pts, x = ~x, y = ~median_income, frame = ~stage, ids = ~community,
              color = ~violence, colors = c(garnet, blue), marker = list(size = 11, opacity = 0.8),
              text = ~hover, hoverinfo = "text") |>
  add_markers(data = chi_means, x = ~x, y = ~avg, frame = ~stage, ids = ~violence,
              marker = list(symbol = "line-ew-open", size = 70, color = navy,
                            line = list(width = 4, color = navy)),
              showlegend = FALSE, hoverinfo = "none") |>
  add_text(data = chi_means, x = ~x + 0.33, y = ~avg, text = ~txt, frame = ~stage,
           textfont = list(size = 15, color = navy), showlegend = FALSE, hoverinfo = "none") |>
  add_text(data = chi_gap, x = ~x, y = ~y, text = ~txt, frame = ~stage,
           textfont = list(size = 20, color = navy), showlegend = FALSE, hoverinfo = "none") |>
  layout(xaxis = list(title = "", range = c(0.4, 2.6), tickvals = c(1, 2),
                      ticktext = c("High violence", "Other areas"), tickfont = list(size = 16)),
         yaxis = list(title = "Median household income", tickprefix = "$", range = c(0, 190000)),
         legend = list(orientation = "h", y = 1.08)) |>
  animation_opts(frame = 1800, transition = 900, redraw = FALSE) |>
  animation_slider(currentvalue = list(prefix = "")) |>
  config(displayModeBar = FALSE)
```

## Test

```{r}
#| label: chi-ttest
#| output-location: slide
t.test(median_income ~ violence, data = areas)
```

## Where Is Zero?

```{r}
#| label: chi-widget
#| echo: false
#| output: asis
ge_widget(
  ge_mean("Income", A = "High violence", B = "Other areas",
          a = areas$median_income[areas$violence == "High violence"],
          b = areas$median_income[areas$violence == "Other areas"],
          left = "high-violence areas poorer", right = "high-violence areas richer",
          rerun = "another 77 areas like these", unit_group = "areas"),
  default = 2)
```

## Say It {.smaller}

- High-violence areas minus other areas: about **$25,000 to $47,000 lower** median income (averages $54,355 vs $90,088), p = 0.00000001
- **How big:** the typical household in a high-violence area earns roughly **$36,000 a year less**. The unit is dollars because the column is dollars
- **What this cannot tell you:**
  - which way it runs (does poverty breed violence, or does violence drive out income?) or what else explains both
  - anything about individual *people*. Each row is an area. "People with less income commit more violence" is a claim about people, and these data are not about people

# Example 4 · Use of Force: Body Cameras Over Time {background-color="#1B2A4A"}

## The Data

- The *Washington Post*'s database of **every person shot and killed by on-duty police** in the U.S., 2015 to 2024: **10,430 people**
- `body_camera` is TRUE when news or police reports said a body camera recorded the incident
- One row = one person killed

. . .

**Question:** after 2020's wave of police reform, did body camera footage exist more often? Group: 2015 to 2019 vs 2020 to 2024. Yes/no outcome.

## Count

```{r}
#| label: wapo-count
shootings |> count(period, body_camera)
```

- **2020 to 2024:** 1,205 of 1,205 + 4,302 = **5,507** (21.9%)
- **2015 to 2019:** 594 of 594 + 4,329 = **4,923** (12.1%)

## Year by Year

The share of fatal shootings with a body camera reported, one year at a time. Press Play.

```{r}
#| label: wapo-anim
#| echo: false
yearly <- shootings |>
  mutate(year = year(date)) |>
  summarize(pct = 100 * mean(body_camera), .by = c(year, period)) |>
  arrange(year)
yr_labels <- c(as.character(2015:2024), "The two periods")
bwc_anim <- map_dfr(seq_along(yr_labels), \(s) yearly |>
  mutate(shown = if_else(s <= 10 & year > 2014 + s, 0, pct), step = yr_labels[s])) |>
  mutate(step = factor(step, levels = yr_labels),
         txt = if_else(shown > 0, paste0(round(shown), "%"), ""))
per <- shootings |> summarize(pct = 100 * mean(body_camera), .by = period)
bwc_lines <- map_dfr(seq_along(yr_labels), \(s) per |>
  mutate(x = if (s < 11) 2010 else if_else(period == "2015 to 2019", 2017, 2022),
         step = yr_labels[s])) |>
  mutate(step = factor(step, levels = yr_labels),
         ytxt = if_else(period == "2015 to 2019", 19.5, 35),
         txt = paste0(period, ": ", round(pct, 1), "%"))

plot_ly(width = anim_w, height = anim_h) |>
  add_bars(data = bwc_anim, x = ~year, y = ~shown, frame = ~step, color = ~period,
           colors = c(blue, garnet), text = ~txt, textposition = "outside",
           textfont = list(size = 14), hoverinfo = "none") |>
  add_markers(data = bwc_lines, x = ~x, y = ~pct, frame = ~step, ids = ~period,
              marker = list(symbol = "line-ew-open", size = 230, color = navy,
                            line = list(width = 4, color = navy)),
              showlegend = FALSE, hoverinfo = "none") |>
  add_text(data = bwc_lines, x = ~x, y = ~ytxt, text = ~txt, frame = ~step,
           textfont = list(size = 17, color = navy), showlegend = FALSE, hoverinfo = "none") |>
  layout(xaxis = list(title = "", range = c(2014.4, 2024.6), dtick = 1),
         yaxis = list(title = "Body camera reported (%)", range = c(0, 38)),
         legend = list(orientation = "h", y = 1.08)) |>
  animation_opts(frame = 900, transition = 400, redraw = FALSE) |>
  animation_slider(currentvalue = list(prefix = "")) |>
  config(displayModeBar = FALSE)
```

## What the Picture Says

- The rise is real, but **not smooth**: 2023 hit 30%, and 2024 fell back to 20%
- Comparing two five-year periods averages over those bumps. That is a choice, and an honest write-up says so

. . .

Now the two tests on the two periods.

## Test: `prop.test()`

```{r}
#| label: wapo-prop
#| output-location: slide
prop.test(x = c(1205, 594), n = c(5507, 4923))
```

## Test: `bayes_prop_test()`

```{r}
#| label: wapo-bayes
#| output-location: slide
bayes_prop_test(x = c(1205, 594), n = c(5507, 4923))
```

## Where Is Zero?

```{r}
#| label: wapo-widget
#| echo: false
#| output: asis
ge_widget(
  ge_prop("Body cameras", A = "2020 to 2024", B = "2015 to 2019", out = "with a camera",
          xa = 1205, na = 5507, xb = 594, nb = 4923,
          left = "cameras less common after 2020", right = "cameras more common after 2020",
          rerun = "another 10,430 shootings", unit_group = "shootings"),
  default = 4)
```

## Say It

- `prop.test()`: about **8 to 11 points** higher after 2020, p below 2.2e-16. `bayes_prop_test()`: 8.4 to 11.2 points, more than a 99% probability the later rate is higher
- **How big:** about **12 in 100** fatal shootings had known camera footage before 2020, and about **22 in 100** after. Close to doubling, yet still fewer than 1 in 4
- **What this cannot tell you:** whether cameras changed anything about the shootings, or how many cameras *existed* but were never reported

. . .

With 10,430 cases, the p-value was never in doubt. **The size is the whole story.**

## Four Questions, One Recipe {.smaller}

| | Question | Test | Answer | The lesson |
|---|---|---|---|---|
| **Reentry** | Does aid lower re-arrest? | `prop.test()` + Bayes | 22% vs 31%; p = 0.063; 97% probably lower | "Not significant" is not "no effect" |
| **Pretrial** | Does a job get you released? | `prop.test()` + Bayes, then again after `filter()` | 18 points; 10 among clean records | Groups differ in other ways. Compare like with like |
| **Neighborhoods** | Lower income where violence is high? | `t.test()` | $25k to $47k lower | Size in real units. Areas are not people |
| **Use of force** | More body camera footage after 2020? | `prop.test()` + Bayes | 12% to 22% | Huge n: the p-value adds nothing; the size is the story |

# Now You: Lab 7 {background-color="#73000A"}

## Lab 7

Two more real questions, same recipe, and this time you write the code ([lab page](../labs/lab-07-workshop.qmd)):

- **With me:** a pretrial risk-assessment tool. Do people it labels High risk get re-arrested more often than people it labels Low risk?
- **Your Turn:** Minneapolis police stops. Are people searched more often in suspicious-person stops than in traffic stops? Then a `t.test()` on the *Post* data

Submit your Your Turn script to Canvas before you leave.

# What's Out There {background-color="#1B2A4A"}

## Your Final Project Needs a Dataset

The **[Final-Project Data Options](../project-data.qmd)** page lists real criminal justice datasets that load with **one line of R**:

- Reentry, pretrial release, police stops, police shootings, risk assessment, public opinion about crime and police, state crime rates
- Each one: what a row is, how many rows, the key variables, a starter question, and whether it has the numeric variables regression will need (Sessions 10 and 11)

## What's Eligible

- Everything on that page is eligible, **including the reentry, pretrial, police stops, shootings, and risk-assessment data from tonight**, as long as your question goes beyond what we did in class
- **Not eligible:** our course datasets, which includes the **Chicago** data from Example 3, plus crash, officers, prisoners, and the rest

## Two Ways to Load a Dataset

Our course copy, the way you have loaded everything all semester:

```r
arrests <- read_csv("https://smourtgos.github.io/crju705/data/toronto-arrests.csv")
```

The original, from an R package (install once, then load it each time):

```r
install.packages("carData")   # once, in the console
library(carData)
data(Arrests)
```

Same data either way. Use whichever you like.

## How to Size Up a Dataset in Five Minutes

1. **What is one row?** A person? An arrest? A neighborhood? A year?
2. **How many rows?** Hundreds is plenty; millions is a headache
3. **What are the variables?** `glimpse()` tells you names and types
4. **Anything strange?** 999s, negative ages, lots of `NA`: `summary()` finds them
5. **Could it answer a question you care about?** Name the groups, or the variables

## Homework 7 and Next Week

- **Homework 7:** explore **two** datasets (from the menu, or one you found and cleared with me), then **choose one** for your final project. Due **Sunday by 11:59 pm** (posted on the [Session 7 page](../weeks/week-07.qmd))
- No reading this week
- **Session 8:** more worked examples like tonight's, and review for the midterm
