---
title: "Lab 8 · ANOVA: Three Classrooms, Then Real Crashes"
subtitle: "CRJU 705 · Session 8"
---

**Goal:** run the complete ANOVA workflow from today's lecture — the two variances by hand, `aov()` + Tukey's HSD, and the Bayesian version with `brms` + `emmeans` — first on a small classroom example where you can check every number, then on real Alaska crash data.

The R4DS wrangling thread wrapped up before the break, so there's no new data-skills content today: this lab is pure statistics practice.

**How labs work:** run the Setup chunk as-is, follow the Walkthrough along with me, then complete the **Your Turn** tasks in your own script. The Exit Ticket gets submitted to Canvas before you leave.

## Setup

Open RStudio, start a new R script, and run:

```{r}
#| label: lab-setup
#| message: false
library(tidyverse)
library(brms)
library(emmeans)

crashes <- read_csv("https://smourtgos.github.io/crju705/data/crash-ak.csv")
```

::: {.callout-warning appearance="simple"}
**First time using `brms`?** Install both new packages before anything else: `install.packages(c("emmeans", "brms"))`. And when you run your first `brm()` model of a session, R will sit at `Compiling Stan program...` for a minute before sampling starts. That's normal — Stan is compiling your model into C++. It is not frozen. If instead you get an error mentioning a *compiler* or *toolchain*, flag me down.
:::

Our walkthrough data are the three classrooms from lecture — typed in by hand, because 15 numbers you can see beat 15,000 you can't when you're learning what a method *does*:

```{r}
#| label: classrooms-data
classrooms <- tibble(
  score     = c(75, 80, 78, 74, 77,    # Class 1
                88, 90, 92, 85, 87,    # Class 2
                70, 65, 68, 72, 66),   # Class 3
  classroom = factor(rep(c("Class 1", "Class 2", "Class 3"), each = 5))
)
```

**The question:** did these three classrooms perform differently, or are the differences just the usual student-to-student bounce?

## Walkthrough

### Look first

Always. Before any test:

```{r}
#| label: look-first
class_means <- classrooms |>
  summarize(mean_score = mean(score), .by = classroom)

ggplot(classrooms, aes(classroom, score)) +
  geom_jitter(width = 0.08, height = 0, size = 3, alpha = 0.6) +
  geom_boxplot(alpha = 0.2, width = 0.3) +
  geom_point(data = class_means, aes(y = mean_score),
             shape = 23, size = 5, fill = "firebrick") +
  labs(title = "Test scores by classroom",
       subtitle = "Dots = individual students · Diamonds = class means",
       x = NULL, y = "Test score")
```

Class 2 sits high, Class 3 sits low, and the spread *inside* each class is modest. That's the whole ANOVA story in one picture — the rest of the lab is putting numbers on it.

### The two variances, by hand

**Within-group variance** = the average spread inside a class. **Between-group variance** = the spread of the class means themselves.

```{r}
#| label: hand-variances
class_stats <- classrooms |>
  summarize(class_mean = mean(score), class_var = var(score), .by = classroom)
class_stats

within_var  <- mean(class_stats$class_var)    # average spread inside a class
between_var <- var(class_stats$class_mean)    # spread of the class means

within_var
between_var
between_var / within_var                      # signal-to-noise ratio
```

The class means are about **14.5 times** more spread out than chance bouncing within classes would produce. Strong signal.

::: {.callout-note collapse="true"}
## Why won't this ratio match `aov()`'s F?

Because `aov()` scales the between-group signal by the **group size** before comparing it to the noise: a gap between means built on 5 students each is stronger evidence than the same gap built on 2. With equal groups, F = group size × our hand ratio — here, 5 × 14.5 ≈ 72.7. Watch for exactly that number below. Same logic, different units.
:::

### The frequentist ANOVA

```{r}
#| label: aov-classrooms
class_model <- aov(score ~ classroom, data = classrooms)
summary(class_model)
```

Reading the table against our hand math:

- **Mean Sq Residuals = 7.1** — our within-group variance, exactly
- **Mean Sq classroom = 513.8** — our between-group variance × 5 (the group size): 5 × 102.76
- **F(2, 12) = 72.7, p ≈ 0.0000002** — far beyond anything a null world produces. *Something* differs.

### Which classes differ? Tukey's HSD

The omnibus F never says *which* pairs. Tukey's HSD tests every pair while keeping the family-wise false-alarm rate at 5% (remember the pile-of-t-tests problem from lecture):

```{r}
#| label: tukey-classrooms
TukeyHSD(class_model)
```

All three intervals exclude 0 with tiny adjusted p-values: every class differs from every other class.

### The Bayesian version

Same formula, Bayesian machinery. Fit once, then let `emmeans` do the summarizing:

```{r}
#| label: brm-classrooms
#| output: false
class_bayes <- brm(score ~ classroom, data = classrooms,
                   chains = 2, iter = 2000, seed = 705, refresh = 0)
```

```{r}
#| label: emmeans-classrooms
class_post <- emmeans(class_bayes, ~ classroom)
class_post
```

One row per class: posterior median and 95% HPD interval for that class's mean.

```{r}
#| label: pairs-classrooms
pairs(class_post)
```

```{r}
#| label: plot-pairs-classrooms
#| fig-height: 3
plot(pairs(class_post)) +
  geom_vline(xintercept = 0, linetype = 2)
```

**The interpretation rule for pairwise differences:** if an interval **crosses 0**, zero ("no difference") remains a credible value — no credible difference. If it **excludes 0**, the difference is credible. Here none of the three intervals crosses 0, so the Bayesian analysis agrees with Tukey: all three classes differ.

## Your Turn

Work in your own script. Comment each answer.

**1. Flatten the classrooms.** Erase the group differences but keep the within-class spread:

```{r}
#| label: yt-1
#| eval: false
classrooms_flat <- classrooms |>
  mutate(score = case_when(
    classroom == "Class 2" ~ score - 12,
    classroom == "Class 3" ~ score + 8,
    .default = score
  ))
# Now recompute: class means, within_var, between_var, the ratio,
# and rerun aov() + TukeyHSD() on classrooms_flat.
```

In comments: which numbers changed, which stayed the same, and *why*? What does `aov()` conclude now, and what do the Tukey intervals do?

**2. Real data — do injuries differ by crash cause?** The `crashes` data (20,000 real-world-style Alaska crashes, from Lab 5 and Homework 5) record each crash's `primary_cause`. Keep three causes and look first:

```{r}
#| label: yt-2
#| eval: false
crash3 <- crashes |>
  filter(primary_cause %in% c("speeding", "distracted", "wildlife"))

# a) Compute mean, SD, and n of `injuries` for each cause (.by = primary_cause)
# b) Make the look-first plot (boxplot + mean diamonds, as in the Walkthrough).
#    Injuries are small whole numbers, so the boxes will look blocky —
#    that's the data, not a bug. The diamonds are doing the real work here.
```

**3. Both traditions on `crash3`.** Run `aov()` + `summary()`, then `TukeyHSD()`. In comments: which pairs differ? One pair should *not* — name it, and report the interval that told you so. Then fit the Bayesian version (this one samples on ~8,800 rows — give it a minute or two) and check `pairs()`:

```{r}
#| label: yt-3
#| eval: false
# aov() + summary() + TukeyHSD() on crash3
# then:
# crash_bayes <- brm(injuries ~ primary_cause, data = crash3,
#                    chains = 2, iter = 2000, seed = 705, refresh = 0)
# crash_post  <- emmeans(crash_bayes, ~ primary_cause)
# pairs(crash_post)
```

Do the two traditions agree about which pair shows no real difference? And practically: how many *more* people are injured in the average speeding crash than the average wildlife crash?

**4. *(Stretch)* All nine causes.** Rerun `aov()` and `TukeyHSD()` on the full `crashes` data — 9 groups means 36 pairwise comparisons. How many come out significant? Why would running 36 *unadjusted* t-tests instead be a terrible idea? (Lecture's false-alarm table has the number.)

```{r}
#| label: yt-4
#| eval: false
# your code here
```

## Exit Ticket

Submit a short R script to Canvas before you leave, containing:

1. One sentence each, in comments: what within-group variance measures, what between-group variance measures, and what F compares
2. Your crash-cause conclusion, two ways: one frequentist sentence (F and a Tukey pair) and one Bayesian sentence (an HPD interval) — each including the size of the difference in injuries, not just "significant"
3. A comment naming the pair of crash causes that showed *no* credible difference, and what its interval looked like in each tradition
4. A comment: why is ANOVA-then-Tukey better than running a pile of unadjusted t-tests?
