---
title: "Lab 2 · First Plots with ggplot2"
subtitle: "CRJU 705 · Session 2"
---

**Goal:** make your first plots in R and connect them to today's descriptive statistics. This lab assumes you've read R4DS [Ch. 1 (Data visualization)](https://r4ds.hadley.nz/data-visualize.html) and [Ch. 2 (Workflow: basics)](https://r4ds.hadley.nz/workflow-basics.html).

**How labs work:** run the Setup chunk as-is, follow the Walkthrough along with me, then complete the **Your Turn** tasks in your own script. The Exit Ticket gets submitted to Canvas before you leave.

## Setup

Open RStudio, start a new R script, and run:

```{r}
#| label: lab-setup
#| message: false
library(tidyverse)

prisoners <- read_csv("https://smourtgos.github.io/crju705/data/prisoners.csv")
```

`prisoners` holds records for 300 incarcerated persons (simulated, but realistic): demographics, offense type, sentence, and well-being measures. Look at it first — *always* look at your data first:

```{r}
#| label: glimpse
glimpse(prisoners)
```

## Walkthrough

### A histogram — the shape of a numeric variable

Today's lecture was about summarizing numeric variables. A histogram is the picture version:

```{r}
#| label: hist-age
ggplot(prisoners, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Age distribution of incarcerated persons",
       x = "Age (years)", y = "Count")
```

Anatomy of every ggplot: **data** (`prisoners`), **aesthetics** (`aes(x = age)` — which variables map to which visual channels), and **geometry** (`geom_histogram()` — how to draw them).

How does the picture relate to the numbers?

```{r}
#| label: age-stats
prisoners |>
  summarize(mean_age = mean(age), median_age = median(age), sd_age = sd(age))
```

### A bar chart — counts of a categorical variable

```{r}
#| label: bar-offense
ggplot(prisoners, aes(x = offense_type)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Offense types", x = NULL, y = "Count")
```

### A boxplot — a distribution *split by groups*

The five-number-summary as a picture, one box per offense type:

```{r}
#| label: box-sentence
ggplot(prisoners, aes(x = offense_type, y = sentence_years)) +
  geom_boxplot(fill = "grey90") +
  labs(title = "Sentence length by offense type",
       x = NULL, y = "Sentence (years)")
```

The box is the middle 50% (median = the line inside); the dots are outliers — the same "one 30-year sentence" problem from lecture, visible instantly.

## Your Turn

Work in your own script. Comment each answer.

**1.** Make a histogram of `sentence_years`. Is it symmetric or skewed? Compute the mean and median — do they agree with what the picture shows? *(Hint: today's lecture.)*

```{r}
#| label: yt-1
#| eval: false
# your code here
```

**2.** Make a bar chart of `education`. Which category is most common (the mode)?

```{r}
#| label: yt-2
#| eval: false
# your code here
```

**3.** Make a boxplot of `age` by `offense_type`. Which offense type has the *most spread* in age? Check yourself with `sd()`:

```{r}
#| label: yt-3
#| eval: false
# your code here — for the numeric check:
# prisoners |> summarize(sd_age = sd(age), .by = offense_type)
```

**4.** *(Stretch)* Add `color = "white"` inside your histogram's `geom_histogram()` and change its `binwidth`. What changes? What binwidth seems most honest for `sentence_years`?

## Exit Ticket

Submit a short R script to Canvas before you leave, containing:

1. One plot of your choice from today (code only is fine)
2. A comment (2–3 sentences): what does the plot show, and which descriptive statistic (mean/median/SD) best summarizes it?
3. A comment naming one thing from R4DS Ch. 1–2 or today's lab that is still confusing
