---
title: "Lab 8 · Correlation: Do Two Numbers Move Together?"
subtitle: "CRJU 705 · Session 8"
---

**Goal:** by the end of this lab you can take two numeric variables, look at them, put one number on how tightly they move together, test it, and say what it means (and what it does not). Two new functions tonight, `cor()` and `cor.test()`, and `cor.test()` reads a lot like the `t.test()` output you already know.

**How labs work:** run the Setup chunk as-is, follow the Walkthrough along with me, then complete the **Your Turn** tasks in your own script. That script is what you submit to Canvas before you leave.

## The recipe

Same idea as Session 7, with one change: there are no groups. Both variables are numbers.

1. **Question.** Name the two numeric variables, and what one row is (a person? a neighborhood?).
2. **Look.** A scatterplot, always, before any number.
3. **Number and test.** `cor()` gives r. `cor.test()` gives r, a confidence interval, and a p-value.
4. **Say it.** The honest sentence, how strong in plain words, and one thing a correlation cannot tell you.

## Setup

Open RStudio, start a new R script, and run:

```{r}
#| label: lab-setup
#| message: false
library(tidyverse)

# Walkthrough: Chicago's 77 community areas, 2025
areas <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv",
                  show_col_types = FALSE)

# Your Turn: 500 neighborhoods (the Lab 3 data)
nbhd <- read_csv("https://smourtgos.github.io/crju705/data/neighborhoods.csv",
                 show_col_types = FALSE)

# Your Turn: 200 correctional officers (the Lab 5 and 6 data)
officers <- read_csv("https://smourtgos.github.io/crju705/data/officers.csv",
                     show_col_types = FALSE)
```

Nothing new to install tonight.

## Walkthrough

### The data

`areas` has one row per Chicago **community area** (77 of them), with 2025 crime rates and Census measures ([codebook](../chicago-data.qmd)). One row is one area, **not** one person. Keep that in mind all night: it decides what your sentences are allowed to claim.

### Question

**Do community areas with higher unemployment have higher violent-crime rates?** Two numeric variables: `unemployment_rate` (percent) and `violent_rate` (violent crimes per 1,000 residents).

### Look

```{r}
#| label: wt-look
#| fig-height: 4
ggplot(areas, aes(unemployment_rate, violent_rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

Read the picture before any number: the dots climb from left to right (positive), they hug the line fairly tightly (strong), the pattern is roughly a straight line, and no single wild point is creating it. *Now* one number is a fair summary.

### Number

```{r}
#| label: wt-cor
cor(areas$unemployment_rate, areas$violent_rate)
```

r = 0.84. Positive, and very strong: r always lands between -1 and +1, and 0.84 is close to the top.

### Test

```{r}
#| label: wt-test
cor.test(areas$unemployment_rate, areas$violent_rate)
```

Read it the way you read `t.test()`:

- **`cor`** at the bottom is r, 0.84
- **95 percent confidence interval:** 0.76 to 0.90. It is an interval for the correlation, not for a difference, so it is in r's units (between -1 and +1). Zero is far outside it
- **p-value** below 2.2e-16: reject the null of no correlation

### Say it

**The sentence:** across Chicago's 77 community areas, those with higher unemployment had higher violent-crime rates, r = 0.84 (95% CI 0.76 to 0.90).

**How strong, in plain words:** a very strong association. Knowing an area's unemployment rate tells you a lot about where its violent-crime rate sits.

**What it cannot tell you:**

- **Cause.** Unemployment and violence could both be driven by something else (concentrated poverty, disinvestment). A correlation is a pattern, not a proof.
- **Anything about individuals.** These rows are areas. Nothing here says unemployed *people* commit violence. Making that jump from areas to people is the **ecological fallacy**.

## Your Turn

Work in your own script and comment each answer.

**Question 1:** in the 500 neighborhoods, **do neighborhoods with more poverty have more assaults?** `poverty_rate` is the share of residents in poverty (0.25 means 25%) and `assaults` is the number of assaults recorded there.

**1.** Look. Make a scatterplot with `poverty_rate` on the x-axis and `assaults` on the y-axis, and add `geom_smooth(method = "lm", se = FALSE)`. In a comment: which direction do the dots go, and is the pattern straight enough for r to be a fair summary? (The dots sit in rows because `assaults` is a whole-number count. That is fine.)

```{r}
#| label: yt-1
#| eval: false
# your code here
```

**2.** Number. Run `cor(nbhd$poverty_rate, nbhd$assaults)`. In a comment, write r and describe it in words: positive or negative, and weak, moderate, or strong.

```{r}
#| label: yt-2
#| eval: false
# your code here
```

**3.** Test. Run `cor.test()` on the same two variables. In a comment, report the 95% confidence interval and the p-value. Reject or fail to reject the null of no correlation?

```{r}
#| label: yt-3
#| eval: false
# your code here
```

**4.** Say it. In a comment, write one sentence about these neighborhoods that a city council member would understand. Then name one thing this correlation cannot tell you. Check your sentence: does it accidentally make a claim about *people* rather than neighborhoods?

```{r}
#| label: yt-4
#| eval: false
# a comment-only task
```

**Question 2:** among the 200 correctional officers, **do officers who have been on the job longer have lower (or higher) job satisfaction?** `tenure_years` is years on the job; `job_satisfaction` is a 0 to 100 score.

**5.** Run the recipe again: scatterplot (tenure on the x-axis), then `cor.test(officers$tenure_years, officers$job_satisfaction)`. In a comment: r, the confidence interval, and the p-value. Does the interval include zero? Then write the honest sentence. Careful: failing to reject is not the same as proving there is no relationship.

```{r}
#| label: yt-5
#| eval: false
# your code here
```

**6.** *(Stretch)* Spearman on an ordered category. `disorder_level` is ordered (Low, Moderate, High, Very High), so Pearson's r does not fit it, but a rank correlation does. Run this, then say in a comment what it shows about disorder and assaults:

```{r}
#| label: yt-6
#| eval: false
nbhd |>
  mutate(disorder_rank = as.numeric(factor(disorder_level,
    levels = c("Low", "Moderate", "High", "Very High")))) |>
  with(cor.test(disorder_rank, assaults, method = "spearman"))
```

When you are done, submit your script to Canvas. The in-class exercise is graded on completing it.
