---
title: "Lab 5 · Importing Real Data"
subtitle: "CRJU 705 · Session 5"
---

**Goal:** get real data *into* R correctly — from a CSV on the web and from a messier Excel file — and know what to do when the import goes sideways. This lab assumes you've read R4DS [Ch. 7 (Data import)](https://r4ds.hadley.nz/data-import.html) and [Ch. 8 (Workflow: getting help)](https://r4ds.hadley.nz/workflow-help.html).

**How labs work:** run the Setup chunk as-is, follow the Walkthrough along with me, then complete the **Your Turn** tasks in your own script. The Exit Ticket gets submitted to Canvas before you leave.

## Setup

Open RStudio, start a new R script, and run:

```{r}
#| label: lab-setup
#| message: false
library(tidyverse)
library(readxl)

crashes <- read_csv("https://smourtgos.github.io/crju705/data/crash-ak.csv")
```

`crashes` holds 20,000 Alaska traffic-crash incidents (simulated, but realistic): when and where each crash happened, causes, severity, injuries, driver characteristics, and intoxication flags. As always — look at it first:

```{r}
#| label: glimpse
glimpse(crashes)
```

## Walkthrough

### What did `read_csv()` decide for us?

`read_csv()` *guessed* a type for every column. See its guesses with `spec()`:

```{r}
#| label: spec
spec(crashes)
```

Mostly sensible — but look at `date`: it's `col_character()`. The file stores dates as `"02/20/2023"` (MM/DD/YYYY), and readr only auto-detects ISO-style dates (`2023-02-20`). Right now R thinks our dates are just text — you can't sort them chronologically, subtract them, or extract a year.

**The lesson: never assume the import worked. Check the types.**

### Telling `read_csv()` what you want: `col_types`

You can hand `read_csv()` an explicit specification instead of letting it guess. What happens if we naively declare `date` a date column?

```{r}
#| label: bad-coltypes
#| warning: true
crashes_bad <- read_csv(
  "https://smourtgos.github.io/crju705/data/crash-ak.csv",
  col_types = cols(date = col_date())
)
```

readr warns us — and `problems()` shows exactly what went wrong, row by row:

```{r}
#| label: problems
problems(crashes_bad)
```

Every row "failed to parse": `col_date()` expected ISO format and got `"02/20/2023"`, so the whole column became `NA`. Two fixes:

**Fix 1 — tell readr the format up front:**

```{r}
#| label: coltypes-format
crashes_fixed <- read_csv(
  "https://smourtgos.github.io/crju705/data/crash-ak.csv",
  col_types = cols(date = col_date(format = "%m/%d/%Y"))
)
crashes_fixed |> select(incident_id, date) |> head(3)
```

**Fix 2 — import as text, then parse with lubridate.** This is my everyday habit, because lubridate's helpers are nearly unbreakable: name the function after the *order* of the components — `mdy()` for month/day/year, `dmy()` for day/month/year, `ymd()` for year/month/day:

```{r}
#| label: mdy
crashes <- crashes |>
  mutate(date = mdy(date))

crashes |> select(incident_id, date, time) |> head(3)
```

Now `date` is a real date. Real dates unlock real questions:

```{r}
#| label: dates-work
crashes |>
  mutate(year = year(date)) |>
  count(year)
```

**Always confirm a date parse did what you think:** check `sum(is.na(...))` after parsing. Zero new `NA`s means every value parsed.

```{r}
#| label: parse-check
sum(is.na(crashes$date))
```

### Checking names (no packages required)

Before analyzing, glance at the column names:

```{r}
#| label: names
names(crashes)
```

These are already *clean*: lowercase, no spaces, words separated by underscores — so you never need backticks. When you import data from the wild, look for the troublemakers: spaces (`Driver Age`), capitals, punctuation (`Injuries?`), or the dead giveaway `...1` — readr/readxl's placeholder name for a column with **no header at all**. You're about to meet one.

### The messier cousin: Excel

The same crash data also circulates as an Excel file — as agency data usually does. `read_excel()` can't read from a URL, so download the file first, then read it:

```{r}
#| label: download-xlsx
#| eval: false
download.file(
  "https://smourtgos.github.io/crju705/data/crash-ak.xlsx",
  "crash-ak.xlsx",
  mode = "wb"   # "write binary" — required for Excel files, especially on Windows
)

crashes_xl <- read_excel("crash-ak.xlsx")
```

```{r}
#| label: read-xlsx-local
#| echo: false
crashes_xl <- read_excel(here::here("data/crash-ak.xlsx"))
```

```{r}
#| label: glimpse-xlsx
glimpse(crashes_xl)
```

Same 20,000 crashes, but three problems the CSV didn't have:

1. **A junk first column named `...1`** — someone saved the spreadsheet with Excel's row numbers included, and there's no header for them.
2. **`date` and `time` imported as datetimes.** Excel stores everything as datetimes under the hood; note `time`'s absurd `1899-12-31` dates — that's Excel's "time only" anchor leaking through.
3. **`vehicle2_year` imported as *character*.** Why? Single-vehicle crashes have no second vehicle, and this spreadsheet recorded those cells as the literal *text* `"NA"` — which readxl treats as a word, not as missing:

```{r}
#| label: fake-na
crashes_xl |> count(vehicle2_year == "NA")
```

Fix all three — tell `read_excel()` what "missing" looks like, drop the junk column, and flatten the datetimes:

```{r}
#| label: fix-xlsx
#| eval: false
crashes_xl <- read_excel("crash-ak.xlsx", na = c("", "NA"))
```

```{r}
#| label: fix-xlsx-local
#| echo: false
crashes_xl <- read_excel(here::here("data/crash-ak.xlsx"), na = c("", "NA"))
```

```{r}
#| label: fix-xlsx-clean
crashes_xl <- crashes_xl |>
  select(-`...1`) |>
  mutate(
    date = as_date(date),      # drop the meaningless midnight timestamp
    hour = hour(time)          # keep the useful part of the time column
  )

crashes_xl |> select(incident_id, date, hour, vehicle2_year) |> head(3)
```

`vehicle2_year` is now numeric with honest `NA`s — because `na = c("", "NA")` told readxl both blank cells and the text `"NA"` mean *missing*. This distinction — the two-letter *word* "NA" versus a genuinely empty value — causes an astonishing share of real-world data misery.

### Save your clean version

Once the cleaning works, write the result out so you (and collaborators) never have to repeat it. Keep the raw file untouched and give the clean one its own name:

```{r}
#| label: write-csv
#| eval: false
write_csv(crashes, "crash-ak-clean.csv")
```

`write_csv()` stores dates in ISO format (`2023-02-20`), so the next `read_csv()` parses them automatically — your future self imports with zero fixes.

### When R yells at you: getting help

You will spend more time reading error messages than reading textbooks. A system:

**1. Actually read the message.** Modern tidyverse errors usually tell you the fix:

```{r}
#| label: error-anatomy
#| error: true
crashes |> filter(county = "Anchorage")
```

The error names the problem *and* the solution. Slow down and read before you panic.

**2. Read the help page.** `?read_csv` — skim **Usage** and **Arguments**, then jump to the **Examples** at the bottom, which are runnable.

**3. Make a reprex** (a *repr*oducible *ex*ample) before asking anyone for help: the smallest self-contained chunk of code + data that produces your error. Half the time, building the reprex reveals the bug yourself; the other half, it lets a helper answer in minutes. A question like "my import is broken" is unanswerable; "`read_csv()` turns my date column into `NA` — here are three sample rows and my code" gets solved.

**4. Search the exact error text** (in quotes). Someone has hit it before you.

## Your Turn

Work in your own script. Comment each answer.

**1.** Re-import the CSV from the URL with a `col_types` specification that gets `date` right *on import* (no `mutate()` afterward). Confirm with `problems()` and `spec()` that nothing failed.

```{r}
#| label: yt-1
#| eval: false
# your code here
```

**2.** Using your correctly-imported data, create `year` and `month` columns from `date`, then count crashes per year. Which year in the data has the fewest crashes? (Careful — is that year fully observed, or is it an artifact of when the data ends?)

```{r}
#| label: yt-2
#| eval: false
# your code here
```

**3.** Download and import the Excel version with `read_excel()`, applying all three fixes from the walkthrough. Verify your fix: confirm `vehicle2_year` is numeric and count its `NA`s — then check that the count equals the number of single-vehicle crashes (`num_vehicles == 1`). Does it?

```{r}
#| label: yt-3
#| eval: false
# your code here
```

**4.** Write your cleaned CSV data out with `write_csv()`, then read the new file back in with `read_csv()`. Does `date` come back as a real date column without any help? In a comment, explain why.

```{r}
#| label: yt-4
#| eval: false
# your code here
```

**5.** *(Stretch)* Sabotage on purpose: re-import the CSV forcing `col_types = cols(driver_gender = col_double())`, then inspect `problems()`. In a comment, explain what readr did with the column and why deliberately breaking an import is a useful debugging skill.

```{r}
#| label: yt-5
#| eval: false
# your code here
```

## Exit Ticket

Submit a short R script to Canvas before you leave, containing:

1. One column that `read_csv()` or `read_excel()` guessed wrong today, and the code that fixed it
2. A comment (2–3 sentences): what is the difference between a cell containing the text `"NA"` and a truly empty cell, and how did we handle it?
3. A comment: one error message you hit this week (any assignment), phrased the way you'd ask for help — what you ran, what you expected, what happened instead
