---
title: "Lab 5 · Putting an Interval On It"
subtitle: "CRJU 705 · Session 5"
---

**Goal:** turn today's lecture into two lines of R you will use for the rest of your career — a point estimate with a confidence interval around it, and a sentence about that interval that is actually true. Two functions do all the work: `prop.test()` and `t.test()`. You will read the numbers they need off a table and type them in yourself.

**How labs work:** run the Setup chunk as-is, follow the Walkthrough along with me, then complete the **Your Turn** tasks in your own script. That script, with the four Exit Ticket comments at the bottom, is what you submit to Canvas before you leave.

## Setup

Open RStudio, start a new R script, and run:

```{r}
#| label: lab-setup
#| message: false
library(tidyverse)

officers <- read_csv("https://smourtgos.github.io/crju705/data/officers.csv",
                     show_col_types = FALSE) |>
  filter(!is.na(ptsd_screen)) |>
  mutate(ptsd_screen = factor(ptsd_screen, levels = c("Positive", "Negative")))
```

`officers` holds survey records for correctional officers (simulated, but realistic): shift, tenure, rank, and three wellness measures — `burnout`, `job_satisfaction`, and `sleep_hours` — plus a `ptsd_screen` result.

Two lines in that Setup deserve a sentence each.

The `filter()` line is doing something you should never do silently. Eight of the 200 officers have no PTSD screen on file, and `sum()` returns `NA` the moment a missing value touches it. We drop those eight and **say so out loud**, which leaves 192. What you *should* do about missing data is a real question, and not today's.

The `factor()` line tells R which group comes **first**. R's default is alphabetical, which would put Negative ahead of Positive. Our question is about the officers who screened positive, so we put Positive first, and every comparison below reads **Positive minus Negative**. This is the same line from lecture, and it is what next week's R4DS chapter is about.

## Walkthrough

### Count first

Every interval today starts from a count:

```{r}
#| label: wt-count
officers |> count(ptsd_screen)
```

42 of 192 screened positive:

```{r}
#| label: wt-point
42 / 192
```

About **22%**. Now the honest question: we surveyed 192 officers, not every officer in the state. If someone else surveyed a different 192, they would not get 22%. Session 4 told us exactly how much that number moves around. This week we put that movement in the answer.

### `prop.test()` does it for you

Type the two numbers you just read off the table: how many, out of how many.

```{r}
#| label: wt-prop
prop.test(42, 192)
```

Ignore most of that output. R is testing whether the proportion equals 0.5, which is a question nobody asked. The line you want is the one under **95 percent confidence interval**: roughly **0.16 to 0.29**.

### Saying it out loud without lying

Here is the sentence I want:

> In this sample of 192 correctional officers, about 22% screened positive for PTSD. Accounting for sampling, the plausible range is roughly 16% to 29%.

And here is what not to say:

- ~~"There is a 95% chance the true rate is between 16% and 29%."~~ The true rate is a fixed number. It is not doing anything probabilistic. The 95% describes *the method*, not this one interval.
- ~~"95% of officers fall between 16% and 29%."~~ The interval is about the *rate*, not about individual officers.
- ~~"The rate is 22%."~~ That is the estimate with the uncertainty deleted.

The width of that interval is the honest part. Do not throw it away.

### Width is a choice you pay for

Take a random 50 of the 192 and count again. Run the seed line and the sample line together, so your count matches mine:

```{r}
#| label: wt-small
set.seed(705)
small <- slice_sample(officers, n = 50)

small |> count(ptsd_screen)
```

10 of 50. Same two numbers, typed in:

```{r}
#| label: wt-small-prop
prop.test(10, 50)
```

Same population, a quarter of the sample, and the interval runs from about **0.10 to 0.34**, roughly twice as wide. That is the $\sqrt{n}$ rule from last week, showing up in the thing you actually report.

### Two groups, and an interval that contains zero

Do officers who screen positive sleep less? One line:

```{r}
#| label: wt-ttest
t.test(sleep_hours ~ ptsd_screen, data = officers)
```

Read it from the bottom up. The two **mean in group** lines are the point estimates: about 6.6 hours for Positive, 6.3 for Negative. Positive is listed first because we told R so, which means the interval above them is **Positive minus Negative**.

And here is the whole lesson: that interval runs from about **-0.26 to +0.77 hours**, and **zero is inside it**.

That does not mean the groups are the same. It means this sample cannot tell you which direction the difference runs. The honest sentence:

> Officers screening positive slept about 15 minutes more per night on average, but with 192 officers the data are consistent with anything from roughly 15 minutes *less* to roughly 46 minutes *more*, including no difference at all.

Learning to write that sentence, instead of "no effect" or "they sleep more," is most of what this course is for.

## Your Turn

Same five moves, new columns. Work in your own script and comment each answer.

**1.** Count, then interval. How many of these officers hold a rank above Officer? Run `officers |> count(rank)`, add the two ranks above Officer together, and put that count and 192 into `prop.test()`. In a comment, report the point estimate and the interval.

```{r}
#| label: yt-1
#| eval: false
# your code here
```

**2.** Write the sentence. In a comment, state your Task 1 result the way I did in the walkthrough: the estimate, the interval, and no claim you cannot support. Then write one *wrong* version underneath it and label what is wrong with it.

```{r}
#| label: yt-2
#| eval: false
# a comment-only task
```

**3.** Buy some precision. Take a random 60 officers with `slice_sample(officers, n = 60)`, count `rank` again, and rerun `prop.test()` with the new numbers. Your numbers will differ from your neighbor's, and that is expected. In a comment: which interval is wider, and why is that the right answer rather than a problem?

```{r}
#| label: yt-3
#| eval: false
# your code here
```

**4.** Two groups. Do officers who screen positive report different `job_satisfaction`? Run `t.test(job_satisfaction ~ ptsd_screen, data = officers)`. In a comment: what are the two group means, does the interval contain zero, and what does that let you say?

```{r}
#| label: yt-4
#| eval: false
# your code here
```

**5.** *(Stretch)* Pick any other numeric column, run the same `t.test()` line on it, and write the honest sentence. If your interval excludes zero, say so plainly; if it contains zero, resist the urge to call it "no difference."

```{r}
#| label: yt-5
#| eval: false
# your code here
```

## Exit Ticket

Before you leave, submit your **Your Turn script** to Canvas: the whole thing, Tasks 1 to 4 (and 5 if you got there). The in-class exercise is graded on completing it. At the bottom of that script, add four comment lines:

1. The interval from Task 1, and the two numbers you typed into `prop.test()` to get it
2. Your honest sentence from Task 2, plus the wrong version and what is wrong with it
3. From Task 4: did the interval contain zero, and what did you conclude
4. One thing about confidence intervals that still feels slippery
