---
title: "Lab 6 · Joining Data"
subtitle: "CRJU 705 · Session 6"
---

**Goal:** combine two datasets into one — correctly — and learn to spot the ways joins quietly go wrong. This lab assumes you've read R4DS [Ch. 19 (Joins)](https://r4ds.hadley.nz/joins.html) and skimmed [Ch. 18 (Missing values)](https://r4ds.hadley.nz/missing-values.html).

**How labs work:** run the Setup chunk as-is, follow the Walkthrough along with me, then complete the **Your Turn** tasks in your own script. The Exit Ticket gets submitted to Canvas before you leave.

## Setup

Open RStudio, start a new R script, and run:

```{r}
#| label: lab-setup
#| message: false
library(tidyverse)

crimes <- read_csv("https://smourtgos.github.io/crju705/data/chicago-crimes-2025.csv")
areas  <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv")
```

You know both of these files already — they're the two levels of our Chicago anchor data. Look at what each one knows:

```{r}
#| label: glimpse-crimes
glimpse(crimes)
```

```{r}
#| label: glimpse-areas
areas |> select(community_area, community, median_income, unemployment_rate) |> head(3)
```

## Walkthrough

### A question neither file can answer alone

**Do incidents in lower-income community areas end in arrest at a different rate than incidents in higher-income areas?**

Look at what that question needs: `arrest` lives in `crimes` (one row per *incident*), and `median_income` lives in `areas` (one row per *community area*). No single file has both. This is the everyday reality of CJ data — incidents in one system, census context in another, personnel records in a third. Joins are how you put them together.

### Keys first, join second

A join matches rows using a **key**. Before joining anything, identify:

- the **primary key** — the column(s) that uniquely identify each row of one table
- the **foreign key** — the matching column in the other table

`community_area` should be the primary key of `areas`: one row per area. **Never assume — check:**

```{r}
#| label: pk-check
areas |> count(community_area) |> filter(n > 1)
```

Zero rows means no duplicates — `community_area` really does uniquely identify each of the 77 areas. In `crimes`, `community_area` is a *foreign* key: thousands of incidents point to each area. That's a **many-to-one** join — many incidents, one area each.

### `left_join()` — and a trap on the very first try

```{r}
#| label: join-naive
joined_naive <- crimes |> left_join(areas, by = "community_area")

joined_naive |> select(id, community.x, community.y) |> head(3)
```

Where did `community.x` and `community.y` come from? **Both** tables have a column named `community`, and it wasn't part of the key — so R kept both copies and slapped suffixes on them. Harmless here (they agree), but confusing, and with messier data the two versions might *not* agree.

The cleaner habit: **select just the columns you need from the right-hand table before joining.**

```{r}
#| label: join-clean
joined <- crimes |>
  left_join(
    areas |> select(community_area, median_income, unemployment_rate),
    by = "community_area"
  )

joined |> select(id, community, arrest, median_income) |> head(3)
```

### Row-count discipline

A many-to-one join should return **exactly the rows of the left table**, just wider. Verify it every time:

```{r}
#| label: row-check
nrow(crimes)
nrow(joined)
```

If the joined table had *grown*, that would mean the "one" side secretly had duplicate keys — every duplicate multiplies your rows and silently corrupts every count and mean downstream. `nrow()` before and after is a two-second insurance policy.

### Now answer the question

Split areas at the median of `median_income` and compare arrest rates:

```{r}
#| label: answer
joined |>
  mutate(income_group = if_else(median_income < median(areas$median_income),
                                "Lower-income area", "Higher-income area")) |>
  summarize(arrest_rate = mean(arrest), n = n(), .by = income_group)
```

Incidents in lower-income areas end in arrest about **18%** of the time, versus about **13%** in higher-income areas — a gap we could not have computed from either file alone.

::: {.callout-warning appearance="simple"}
**Interpret with care.** This is an *area-level* pattern attached to incidents. It does not tell us *why* — the mix of offense types differs across areas, police deployment differs, reporting differs. And `median_income` describes the *area*, not the people involved in any incident. Joins expand what you can ask; they don't excuse you from thinking about what the answer means.
:::

### When keys don't match: a small disaster on purpose

Our Chicago files join perfectly, which is lovely and completely unrepresentative of real life. Here's a tiny example with the kind of flaws you'll actually meet — a table of incidents and a lookup table of district commanders:

```{r}
#| label: toy-tables
incidents_toy <- tribble(
  ~incident_id, ~district,
  101, "Central",
  102, "Central",
  103, "Harrison",
  104, "Shakespeare"
)

district_info <- tribble(
  ~district,    ~commander,
  "Central",    "Diaz",
  "Harrison",   "Okafor",
  "Town Hall",  "Reyes"
)
```

Incident 104 is in a district the lookup table doesn't know; Town Hall has a commander but no incidents. Watch what each join does with the mismatches:

```{r}
#| label: toy-left
incidents_toy |> left_join(district_info, by = "district")
```

`left_join()` **keeps every left-hand row** and fills the holes with `NA` — incident 104 survives, commander unknown.

```{r}
#| label: toy-inner
incidents_toy |> inner_join(district_info, by = "district")
```

`inner_join()` **keeps only rows that match** — incident 104 is *silently gone*. Four incidents in, three out, no warning. If you didn't check row counts, you'd never know you dropped data.

### `anti_join()` — find the orphans *before* you commit

`anti_join()` returns the left-hand rows that have **no match** — it's the diagnostic you run before choosing a join:

```{r}
#| label: toy-anti
incidents_toy |> anti_join(district_info, by = "district")   # incidents with no district info
district_info |> anti_join(incidents_toy, by = "district")   # districts with no incidents
```

Run it both directions on our real files:

```{r}
#| label: real-anti
crimes |> anti_join(areas, by = "community_area") |> nrow()
areas  |> anti_join(crimes, by = "community_area") |> nrow()
```

Zero and zero: every incident matches an area, every area has incidents. Our keys are clean — **and the only reason we know that is that we checked.**

### The missing-values moment (the Ch. 18 skim, cashed in)

That `NA` commander `left_join()` created is a genuine missing value, and `NA`s are *contagious by design*:

```{r}
#| label: na-contagion
toy_joined <- incidents_toy |> left_join(district_info, by = "district")

toy_joined |> filter(is.na(commander))   # locate the missing
```

The same discipline applies to missingness that arrives *with* the data. Our crime file has incidents with no coordinates (blocked addresses, data-entry gaps):

```{r}
#| label: na-real
sum(is.na(crimes$latitude))
mean(crimes$latitude)
mean(crimes$latitude, na.rm = TRUE)
```

One missing value poisons the whole mean until you decide — *explicitly* — what to do about it. `na.rm = TRUE` is a decision ("compute on the cases we have"), not a reflex. The worst option is not noticing the `NA`s exist; that's what `sum(is.na(...))` is for.

## Your Turn

Two lab datasets you've met before: correctional **officers** and **prisoners**. Both have a `facility` column.

```r
officers  <- read_csv("https://smourtgos.github.io/crju705/data/officers.csv")
prisoners <- read_csv("https://smourtgos.github.io/crju705/data/prisoners.csv")
```

**1.** Read in both files. What does one row represent in each table? Is `facility` a primary key in *either* table? Check with `count()` — and in a comment, say what kind of join relationship prisoners-to-officers would be (one-to-one? many-to-one? many-to-many?).

```{r}
#| label: yt-1
#| eval: false
# your code here
```

**2.** Before running it: predict how many rows `prisoners |> left_join(officers, by = "facility")` returns. Then run it. Explain (in a comment) where the row count came from and why dplyr printed a warning. *(Hint: every prisoner matches every officer in their facility.)*

```{r}
#| label: yt-2
#| eval: false
# your code here
```

**3.** The many-to-many join in #2 answered no question at all. The right move is to **aggregate first, then join**: summarize mean `burnout` by facility from `officers`, summarize mean `stress` by facility from `prisoners`, and join the two small tables. Do facilities where officers report more burnout also hold prisoners reporting more stress? *(Three facilities is far too few to generalize — just describe what you see.)*

```{r}
#| label: yt-3
#| eval: false
# your code here
```

**4.** Use `anti_join()` in both directions to confirm no prisoner sits in a facility with no officers, and no officer works in a facility with no prisoners. In a comment: which join type in #3 would have *hidden* a mismatch like that, had one existed?

```{r}
#| label: yt-4
#| eval: false
# your code here
```

**5.** *(Stretch)* Staffing ratios: `count()` prisoners per facility, `count()` officers per facility, join the two counts, and compute prisoners-per-officer. Which facility is stretched thinnest?

```{r}
#| label: yt-5
#| eval: false
# your code here
```

## Exit Ticket

Submit a short R script to Canvas before you leave, containing:

1. The code for one join you ran today, plus a comment saying what one row of the *result* represents
2. A comment (2–3 sentences): when would `left_join()` and `inner_join()` give you different answers to the same research question, and which default do you plan to reach for?
3. A comment naming one thing about joins or missing values that is still confusing
