---
title: "Lab 7 · Real Questions, Real Tests"
subtitle: "CRJU 705 · Session 7"
---

**Goal:** by the end of this lab you can take a real criminal justice dataset you have never seen, ask it a two-group question, and answer it with the same four steps every time. No new functions tonight. Everything here is something you already ran in Labs 5 and 6, pointed at data researchers actually use.

**How labs work:** run the Setup chunk as-is, follow the Walkthrough along with me, then complete the **Your Turn** tasks in your own script. That script is what you submit to Canvas before you leave.

## The recipe

Every question tonight, in lecture and here, runs the same four steps:

1. **Question.** Name the two groups and the outcome. If the outcome is yes/no, the test is `prop.test()` (and `bayes_prop_test()`). If the outcome is a number, the test is `t.test()`.
2. **Count or describe.** `count()` for a yes/no outcome; the group averages for a number.
3. **Test.** Type the numbers in, or give `t.test()` the formula.
4. **Say it.** The honest sentence, how big in real units, and one thing this test cannot tell you.

## Setup

Open RStudio, start a new R script, and run:

```{r}
#| label: lab-setup
#| message: false
library(tidyverse)
source("https://smourtgos.github.io/crju705/R/bayes.R")

# Walkthrough: pretrial defendants scored by a risk-assessment tool
compas <- read_csv("https://raw.githubusercontent.com/propublica/compas-analysis/master/compas-scores-two-years.csv",
                   show_col_types = FALSE) |>
  mutate(reoffended = factor(if_else(two_year_recid == 1, "Reoffended", "Did not"),
                             levels = c("Reoffended", "Did not")))

# Your Turn: police stops in Minneapolis, 2017
stops <- read_csv("https://smourtgos.github.io/crju705/data/mpls-stops.csv",
                  show_col_types = FALSE)

# Your Turn: people killed in police shootings, 2015 to 2024
shootings <- read_csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/v2/fatal-police-shootings-data.csv",
                      show_col_types = FALSE) |>
  mutate(crisis = factor(if_else(was_mental_illness_related, "Crisis", "No crisis"),
                         levels = c("Crisis", "No crisis")))
```

Three datasets, three `read_csv()` lines, the same line you have used all semester. Two of them come straight from the people who built them (ProPublica and the *Washington Post*), and one is a copy on our course site. Each `mutate()` line does what you did in Homework 5: it turns a 1/0 or TRUE/FALSE column into two labeled groups, with the group we care about first.

When the first file loads, R prints a message that starts `New names:`. That file has two columns called `decile_score` and two called `priors_count`, and R renamed the second copy of each so it can tell them apart. Nothing is broken, and there is nothing to fix. Real data does this.

## Walkthrough

### The data

In 2013 and 2014, Broward County, Florida, used a commercial tool called **COMPAS** to score people arrested and awaiting trial. Each person got a label: Low, Medium, or High risk. ProPublica then checked who was arrested again within two years. One row is one defendant. `two_year_recid` is 1 if that person was arrested again within two years.

### Question

**Do people the tool labels High risk reoffend more often than people it labels Low risk?** Two groups (High, Low). A yes/no outcome (reoffended or not). So: `prop.test()`, then `bayes_prop_test()`.

### Count

```{r}
#| label: wt-count
compas |> count(score_text, two_year_recid)
```

Four numbers, two per group, High first:

- **High:** 1,001 reoffended, out of 1,001 + 402 = **1,403**
- **Low:** 1,216 reoffended, out of 1,216 + 2,681 = **3,897**

We are comparing only High and Low tonight, so the Medium rows sit out. Say so out loud whenever you leave a group aside.

```{r}
#| label: wt-rates
1001 / 1403
1216 / 3897
```

About **71%** of people labeled High were arrested again, against about **31%** of people labeled Low.

### Test

```{r}
#| label: wt-prop
prop.test(x = c(1001, 1216), n = c(1403, 3897))
```

High minus Low: the interval runs from about **37 to 43 points**, and the p-value is below 2.2e-16. Reject the null.

```{r}
#| label: wt-bayes
bayes_prop_test(x = c(1001, 1216), n = c(1403, 3897))
```

Credible interval about **37 to 43 points**, and more than a 99% probability that the High group's rate is the higher one. With 5,300 people, the two traditions land on the same range.

### Say it

- **The sentence:** given these data, there is a 95% probability that the true gap in two-year re-arrest between people labeled High and Low is between about 37 and 43 percentage points.
- **How big, in real units:** about 71 of every 100 people labeled High were arrested again within two years, compared with about 31 of every 100 labeled Low.
- **What this test cannot tell you:** whether the tool is *good enough* to decide who stays in jail. Almost 3 in 10 people labeled High were *not* arrested again, and "arrested again" is not the same as "committed a crime." A p-value cannot answer whether that error rate is acceptable. That is a question about values, not statistics.

### Same data, a number instead of yes/no

Were the people who reoffended younger? The outcome is now **age**, a number, so the test is `t.test()`. The `reoffended` column was set up in Setup with Reoffended first.

```{r}
#| label: wt-ttest
t.test(age ~ reoffended, data = compas)
```

Reoffended minus Did not: about **4 to 5 years younger** (32.3 vs 36.9 on average), p below 2.2e-16.

In plain words: people who were arrested again were, on average, about four and a half years younger. That is one of the oldest findings in criminology (offending peaks young), showing up again in a dataset built for something else.

## Your Turn

Your question: **in Minneapolis, are people searched more often in stops for a suspicious person than in traffic stops?** `problem` is the reason for the stop (`suspicious` or `traffic`) and `personSearch` is whether the person was searched (`YES` or `NO`). Suspicious goes first all the way through. Work in your own script and comment each answer.

**1.** Count. Run `stops |> count(problem, personSearch)`. In a comment, write the four numbers you need: how many people were searched in suspicious-person stops, out of how many, and the same for traffic stops. Some rows say `NA`: no search was recorded either way. Leave them out of both totals. Then compute the two rates.

```{r}
#| label: yt-1
#| eval: false
# your code here
```

**2.** The frequentist test. Put your four numbers into `prop.test(x = c(__, __), n = c(__, __))`, suspicious first. In a comment, report the confidence interval for the difference in percentage points and the p-value. Reject or fail to reject?

```{r}
#| label: yt-2
#| eval: false
# your code here
```

**3.** The Bayesian test. Same four numbers into `bayes_prop_test()`. In a comment, report the credible interval for the difference and the probability that suspicious-person stops have the higher search rate. Then write the one sentence this output lets you say that Task 2's output does not.

```{r}
#| label: yt-3
#| eval: false
# your code here
```

**4.** Say it. In a comment, state the size of the difference in plain words a city council member would understand (no p-values, no intervals). Then name one thing this comparison cannot tell you.

```{r}
#| label: yt-4
#| eval: false
# a comment-only task
```

**5.** A t-test on a different dataset. Among people killed in police shootings, were those in a mental health crisis older or younger than everyone else? Run `t.test(age ~ crisis, data = shootings)`. In a comment: the two average ages, the confidence interval for the difference in years, the p-value, and one sentence in plain words.

```{r}
#| label: yt-5
#| eval: false
# your code here
```

**6.** *(Stretch)* Suppose a report from another city found about 1 person in 10 searched across all stops, and you trust it about as much as having seen 50 stops: `prior = c(5, 45)`. Rerun Task 3 with it. In a comment: how much did anything move, and why?

```{r}
#| label: yt-6
#| eval: false
# your code here
```

When you are done, submit your script to Canvas. The in-class exercise is graded on completing it.
