---
title: "Lab 10 · Correlation: Chicago Areas, Then a Null Worth Reporting"
subtitle: "CRJU 705 · Session 10"
---

**Goal:** run the complete correlation workflow from today's lecture — scatterplots first, then `cor()` and the correlation matrix, then inference in both traditions with `cor.test()` and the `correlation` package — on the Chicago community areas, and then on two lab datasets you already know. One of them has real associations to find. The other has *nothing*, and learning to say "there's nothing here" out loud is half of this lab's point.

**How labs work:** run the Setup chunk as-is, follow the Walkthrough along with me, then complete the **Your Turn** tasks in your own script. The Exit Ticket gets submitted to Canvas before you leave.

## Setup

Open RStudio, start a new R script, and run:

```{r}
#| label: lab-setup
#| message: false
library(tidyverse)
library(correlation)

areas <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv")
```

::: {.callout-warning appearance="simple"}
**One new package this week:** `install.packages("correlation")`. Good news after Session 8's `brms` marathon — it's a small install, and its Bayesian correlations sample instantly. No "Compiling Stan program..." pause this time.
:::

`areas` is the anchor dataset's second level ([codebook](../chicago-data.qmd)): Chicago's 77 community areas, each with 2025 crime rates and Census-based structural measures. One row = one community area — **not** one person. Keep that in mind all lab; it decides what your sentences are allowed to claim.

## Walkthrough

### Look first

Always. Before any correlation:

```{r}
#| label: look-first
#| fig-height: 4
ggplot(areas, aes(unemployment_rate, violent_rate)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Violent-crime rate vs. unemployment rate — 77 community areas",
       x = "Unemployment rate (%)", y = "Violent crimes per 1,000 residents")
```

A strong, roughly linear, positive association — and no wild outlier manufacturing it. *Now* it's safe to summarize with one number:

```{r}
#| label: single-r
cor(areas$unemployment_rate, areas$violent_rate)
```

r = 0.84: very strong and positive, matching what the plot already told us.

### The correlation matrix

Real analyses look at several candidate variables at once. `cor()` takes every numeric column you feed it:

```{r}
#| label: cor-matrix-lab
areas |>
  select(violent_rate, unemployment_rate, median_income, pct_bachelors) |>
  cor() |>
  round(2)
```

Read it like a mileage chart: symmetric, 1s on the diagonal. Violence correlates strongly with unemployment (0.84), strongly and negatively with income (−0.71), moderately and negatively with college education (−0.45). But this table carries **no uncertainty** — no tests, no intervals.

### The `correlation()` upgrade

The `correlation` package tests every pair and attaches confidence intervals — and it Holm-adjusts the p-values because six tests at once is exactly the pile-of-tests problem from Session 8:

```{r}
#| label: corr-pkg-lab
areas |>
  select(violent_rate, unemployment_rate, median_income, pct_bachelors) |>
  correlation()
```

Same r values as the matrix, now each with a 95% CI and an adjusted p-value.

### Frequentist inference on one pair

When one relationship is the star of your analysis, `cor.test()` gives the full readout:

```{r}
#| label: cor-test-lab
cor.test(areas$unemployment_rate, areas$violent_rate)
```

The reporting sentence, built from that output:

> "Across Chicago's 77 community areas, unemployment was strongly associated with the violent-crime rate, r = .84, t(75) = 13.4, p < .001, 95% CI [.76, .90]."

Note what the sentence does *not* say: nothing about unemployed *people* (ecological fallacy — these are areas), and nothing about unemployment *causing* violence.

### The Bayesian version

Same package, one new argument:

```{r}
#| label: bayes-lab
set.seed(705)   # the Bayesian version samples from the posterior — seed keeps output stable
areas |>
  select(violent_rate, unemployment_rate) |>
  correlation(bayesian = TRUE)
```

Four things to read, all lecture echoes:

- **rho** ≈ 0.82 — the posterior median, pulled *slightly* toward 0 relative to the frequentist 0.84 by the mildly skeptical prior
- **95% CI** — a *credible* interval: given data and prior, 95% probability the correlation lies in this range (the direct statement a confidence interval can't make)
- **pd** = 100% — the posterior probability the correlation is positive
- **BF** ≈ 10¹⁷ — on Session 6's scale, evidence beyond "extreme." Hold this number in mind: in Your Turn 2 you'll meet its opposite.

### "r is blind" — Anscombe's quartet, hands on

From lecture: four datasets that ship with R, statistically identical, visually nothing alike. The pivot gymnastics below just reshape `anscombe`'s eight columns into tidy (set, x, y) rows — copy it as given; the payoff is the summary and the plot:

```{r}
#| label: anscombe-lab
anscombe_long <- anscombe |>
  pivot_longer(everything(),
               names_to = c(".value", "set"), names_pattern = "(.)(.)")

anscombe_long |>
  summarize(mean_x = mean(x), mean_y = mean(y), r = cor(x, y), .by = set)
```

```{r}
#| label: anscombe-plot-lab
#| fig-height: 4
ggplot(anscombe_long, aes(x, y)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  facet_wrap(~set)
```

Four times r = 0.82: one genuine linear trend, one curve, one outlier-hijacked line, one vertical stack with a single point doing all the work. **r summarizes a shape it never checks.** Plot first, every time.

## Your Turn

Work in your own script. Comment each answer.

**1. Neighborhoods — real associations.** Back to the 500 simulated-but-realistic neighborhoods from Lab 3:

```{r}
#| label: yt-1
#| eval: false
nbhd <- read_csv("https://smourtgos.github.io/crju705/data/neighborhoods.csv")

# a) Run correlation() on poverty_rate, unemployment_rate, and assaults.
#    Which pair is strongest? Which is weakest?
# b) Scatterplot the strongest pair (put assaults on the y-axis if it's involved),
#    with a fitted line. Does the shape look linear enough for r to be a fair summary?
# c) cor.test() on poverty_rate and assaults. Write the one-sentence report,
#    lecture style: r, df, p, CI, direction, strength.
# d) These rows are neighborhoods, not residents. Reread your sentence from (c):
#    does it accidentally make a claim about people? Fix it if so.
```

**2. Officers — an honest null.** The 200 correctional officers from Lab 6, and a question the wellness literature cares about: does overtime track burnout?

```{r}
#| label: yt-2
#| eval: false
officers <- read_csv("https://smourtgos.github.io/crju705/data/officers.csv")

# a) Run correlation() on overtime_hours_mo, burnout, sleep_hours, job_satisfaction.
#    Describe what you see. (Expect to be underwhelmed. That's the data, not a bug.)
# b) Now the Bayesian version on just overtime_hours_mo and burnout:
#    officers |>
#      select(overtime_hours_mo, burnout) |>
#      correlation(bayesian = TRUE)
#    Look at the BF. It's BELOW 1 — remember Session 6: a BF < 1 is evidence FOR
#    the null. Compute 1/BF: how many times more consistent are these data with
#    "no correlation" than with a correlation?
# c) In a comment, report this result honestly in both traditions:
#    one frequentist sentence (r, p — what are you allowed to conclude? recall
#    Session 6: can a p-value ever SUPPORT the null?), and one Bayesian sentence
#    (what the BF lets you say that the p-value doesn't).
```

**3. *(Stretch)* Rank correlation on an ordinal variable.** `disorder_level` is ordered categories (Low → Very High) — Pearson's r has no business there, but Spearman's rank correlation is fine:

```{r}
#| label: yt-3
#| eval: false
nbhd |>
  mutate(disorder_rank = as.numeric(factor(disorder_level,
    levels = c("Low", "Moderate", "High", "Very High")))) |>
  with(cor.test(disorder_rank, assaults, method = "spearman"))

# In comments: what does rho say about disorder and assaults?
# And why was Spearman the right tool here when Pearson wasn't?
```

## Exit Ticket

Submit a short R script to Canvas before you leave, containing:

1. A comment, 2–3 sentences: what Pearson's r measures, its possible range, and one thing r = 0 does **not** mean
2. The strongest and the weakest association you found across Your Turn 1 and 2, each with its r value and a plain-English direction/strength label
3. Your two honest sentences about overtime and burnout from Your Turn 2c — the frequentist one and the Bayesian one
4. A comment: what Anscombe's quartet proves about correlation coefficients, and the habit it should build into every analysis you run from now on
