---
title: "Lab 12 · Logistic Regression: Chicago Neighborhoods, Then Alaska Nights"
subtitle: "CRJU 705 · Session 12"
---

**Goal:** run the complete logistic-regression workflow from today's lecture — `glm(family = binomial)`, coefficients → odds ratios, odds ratios → predicted probabilities, and the Bayesian version with `brms` — first on the Chicago community areas, then on your own model of intoxicated crashes in the Alaska data.

This is the last lab with new statistical content: everything after today is your project. Treat this hour as a dress rehearsal — the workflow you practice here is exactly the workflow Checkpoint 4 asks for if your project outcome is binary.

**How labs work:** run the Setup chunk as-is, follow the Walkthrough along with me, then complete the **Your Turn** tasks in your own script. The Exit Ticket gets submitted to Canvas before you leave.

## Setup

Open RStudio, start a new R script, and run:

```{r}
#| label: lab-setup
#| message: false
library(tidyverse)
library(brms)

areas   <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv")
crashes <- read_csv("https://smourtgos.github.io/crju705/data/crash-ak.csv")
```

::: {.callout-warning appearance="simple"}
`brms` is the same package you installed for Lab 8 — nothing new to install. Same reminder as then: the first `brm()` of a session pauses at `Compiling Stan program...` for a minute before sampling. It is not frozen.
:::

Two old friends: the 77 Chicago community areas (Homework 2, and the lecture thread ever since), and the 20,000 Alaska crashes (Labs 5 and 7, Homework 5 and 8's ANOVA cousin). Today the crash data plays the role of *your project dataset*: a binary outcome you care about, predictors you have to build yourself.

## Walkthrough

### Look at the outcome first

Our binary outcome is `high_violence`: is the area's violent-crime rate above the citywide median? It ships in the file, but never trust a variable you haven't checked:

```{r}
#| label: look-outcome
areas |> count(high_violence)

areas |>
  summarize(median_violent = median(violent_rate), .by = high_violence)
```

38 areas above the median, 39 at or below (77 is odd, so the split isn't exactly even). The group medians are wildly different — the flag means what the codebook says it means.

R treats `TRUE` as 1 and `FALSE` as 0, so a logical column drops straight into `glm()` — no conversion needed.

### One predictor, seen

Before any multi-predictor model, look at the shape of the relationship. This is the lecture's curved hill, on real data:

```{r}
#| label: look-curve
#| fig-height: 4
ggplot(areas, aes(unemployment_rate, as.numeric(high_violence))) +
  geom_jitter(height = 0.04, width = 0, size = 2.5, alpha = 0.6) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              se = FALSE, color = "firebrick", linewidth = 1.2) +
  labs(title = "Probability of high-violence status rises with unemployment",
       x = "Unemployment rate (%)", y = "High-violence area (0/1)")
```

Low-unemployment areas pile up at 0, high-unemployment areas at 1, and the fitted curve climbs steeply through the middle — most of the action is between roughly 5% and 15% unemployment.

### Fit the model

`lm()` becomes `glm()` with `family = binomial`. That one argument is the entire syntax difference:

```{r}
#| label: fit-glm
v_model <- glm(high_violence ~ unemployment_rate + pct_vacant,
               data = areas, family = binomial)
summary(v_model)
```

Both coefficients are positive with small p-values — but they're in **log-odds**, and nobody thinks in log-odds.

### Exponentiate: log-odds → odds ratios

```{r}
#| label: odds-ratios
exp(cbind(OR = coef(v_model), confint(v_model)))
```

Reading it out loud (the skill that matters):

- **Unemployment, OR ≈ 1.45:** each additional percentage point of unemployment multiplies the odds of high-violence status by about 1.45 — 45% higher odds — *holding vacancy constant*.
- **Vacancy, OR ≈ 1.33:** each additional point of vacant housing raises the odds by about 33%, *holding unemployment constant*.
- For odds ratios, the no-effect value is **1** (not 0). Both 95% CIs sit entirely above 1.

::: {.callout-important appearance="simple"}
**The classic mistake:** "OR = 1.45 means the *probability* goes up 45%." No — the **odds** go up 45%. Odds = p/(1−p). The two only look alike when p is small. Say "odds" when you mean odds, and use `predict()` when you want probabilities.
:::

### Predicted probabilities: from effects to cases

Odds ratios describe effects; decision-makers ask about *places*. `predict(type = "response")` answers on the probability scale:

```{r}
#| label: predicted-probs
two_areas <- areas |>
  filter(community %in% c("Forest Glen", "West Garfield Park")) |>
  select(community, unemployment_rate, pct_vacant)

two_areas |>
  mutate(p_high_violence = predict(v_model, newdata = two_areas,
                                   type = "response"))
```

About a 3% predicted probability for Forest Glen; over 99% for West Garfield Park. And the "by hand" version, so the machinery isn't a mystery — plug the log-odds prediction into `plogis()` (the inverse of the logit):

```{r}
#| label: by-hand
b <- coef(v_model)
unname(plogis(b[1] + b[2] * 4.2 + b[3] * 2.6))   # Forest Glen's values, by hand
```

### The Bayesian version

Same formula, `bernoulli()` family, the Session 8 settings:

```{r}
#| label: fit-brm
#| output: false
v_bayes <- brm(high_violence ~ unemployment_rate + pct_vacant,
               data = areas, family = bernoulli(),
               chains = 2, iter = 2000, seed = 705, refresh = 0)
```

```{r}
#| label: brm-or
exp(fixef(v_bayes)[, c("Estimate", "Q2.5", "Q97.5")])
```

The posterior odds ratios land near the `glm()` estimates, and both 95% credible intervals exclude 1. The interpretation you're allowed to say out loud: *"Given the data and priors, there's a 95% probability the unemployment odds ratio lies in this interval."*

## Your Turn

Work in your own script. Comment each answer. **The question:** what predicts whether an Alaska crash involves an intoxicated driver (`intox_any`)?

**1. Build your predictors.** No dataset ships with your analysis variables — Lab 7's lubridate moves, one more time:

```{r}
#| label: yt-1
#| eval: false
crashes <- crashes |>
  mutate(
    date    = mdy(date),                         # Lab 5/7 habit
    hour    = hour(time),
    weekend = wday(date, week_start = 1) >= 6,   # Mon = 1 … Sat = 6, Sun = 7
    night   = hour >= 22 | hour < 4              # 10 p.m. – 3:59 a.m.
  )
# Sanity checks before modeling — always:
# a) count(crashes, night) and count(crashes, weekend) — plausible shares?
# b) mean(crashes$intox_any) — what's the baseline rate you're modeling?
```

**2. Fit and translate.** Model `intox_any` from `driver_age`, `night`, and `weekend`; then exponentiate:

```{r}
#| label: yt-2
#| eval: false
# intox_model <- glm(intox_any ~ driver_age + night + weekend,
#                    data = crashes, family = binomial)
# summary(intox_model)
# exp(cbind(OR = coef(intox_model), confint(intox_model)))
```

In comments: (a) interpret the `night` odds ratio in one plain sentence — Lab 7 found intoxication involvement "nearly doubles" at night; does the model agree? (b) interpret `weekend` the same way; (c) what do you conclude about `driver_age`? Its interval covers 1 — practice *saying that no evidence is the finding*.

**3. Predicted probabilities for two drivers.** Contrast a 21-year-old driving on a weekend night with a 45-year-old on a weekday afternoon:

```{r}
#| label: yt-3
#| eval: false
# profiles <- tibble(
#   driver_age = c(21, 45),
#   night      = c(TRUE, FALSE),
#   weekend    = c(TRUE, FALSE)
# )
# profiles |> mutate(p_intox = predict(intox_model, newdata = profiles,
#                                      type = "response"))
```

In a comment: report both probabilities and compare them to the baseline rate from Task 1b. Notice the night OR is ~2 but neither probability is anywhere near "double 50%" — odds ratios multiply *odds*, and when p is small, doubled odds is roughly doubled probability, but it does **not** stay that way as p grows.

**4. Odds-ratio translation drills.** No fitting — just fluency. For each coefficient below (from hypothetical logistic models), write one sentence: the odds ratio, the direction, and the percent change in odds.

```{r}
#| label: yt-4
#| eval: false
# a) beta = 0.30   (body-camera footage available -> case cleared)
# b) beta = -1.00  (defense counsel at first appearance -> pretrial detention)
# c) beta = 0.00   (defendant age -> failure to appear)
# d) A classmate reads (b) as "the probability of detention drops 63%."
#    In a comment: what did they get wrong, and how would you fix the sentence?
```

**5. *(Stretch)* One more predictor.** Add `severity` to your crash model. Do the `night` and `weekend` odds ratios move much? In a comment: why does "holding the other predictors constant" matter when you tell the story?

## Exit Ticket

Submit a short R script to Canvas before you leave, containing:

1. A comment: why `lm()` is the wrong tool for a 0/1 outcome, in one or two sentences
2. Your fitted crash model plus two interpretation sentences: one **odds-ratio** sentence (the `night` effect) and one **predicted-probability** sentence (your two driver profiles)
3. A comment stating the difference between odds and probability — and which one an odds ratio multiplies
4. A comment: which piece of today's workflow you'd use in your final project, and the one thing you most need help with at today's code clinic
