---
title: "Lab 11 · Multiple Regression: Chicago's 77 Areas, Then 500 Neighborhoods"
subtitle: "CRJU 705 · Session 11"
---

**Goal:** run the complete multiple-regression workflow from today's lecture — fit with `lm()`, interpret coefficients in *holding-constant* language, diagnose with a residual plot and VIF, then fit the Bayesian twin with `brms` and read the posteriors — first on the Chicago community-area data from lecture, then on your own model with the Lab 3 neighborhoods.

There's no new data-skills content today: this lab is pure statistics practice. (Homework 11 brings the wrangling back — you'll clean a file *before* regressing on it.)

**How labs work:** run the Setup chunk as-is, follow the Walkthrough along with me, then complete the **Your Turn** tasks in your own script. The Exit Ticket gets submitted to Canvas before you leave.

## Setup

Open RStudio, start a new R script, and run:

```{r}
#| label: lab-setup
#| message: false
library(tidyverse)
library(bayesplot)
library(brms)
library(car)

areas <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv")
```

::: {.callout-warning appearance="simple"}
**One new package today:** `install.packages("car")` — we use it for VIF. (`bayesplot` came along free when you installed `brms` for ANOVA week, so `library(bayesplot)` should just work.) And the usual `brms` reminder: the first `brm()` of a session pauses at `Compiling Stan program...` for a minute. Not frozen. Compiling.
:::

Our walkthrough data are the 77 Chicago community areas — every 2025 incident aggregated to the area level and joined with census measures (codebook on the [Data page](../chicago-data.qmd)). One row per *area*, not per person or per crime. Keep that word — **area** — in every sentence you write today.

**The question:** how do unemployment and vacant housing relate to an area's violent-crime rate — each adjusted for the other?

## Walkthrough

### Look first

Always. Before any model:

```{r}
#| label: look-first
ggplot(areas, aes(unemployment_rate, violent_rate)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  labs(title = "77 community areas, 2025",
       x = "Unemployment rate (%)",
       y = "Violent crimes per 1,000 residents")
```

A strong, roughly linear, upward relationship — with the high-violence areas scattering farther from the line. Remember that scatter; it comes back in diagnostics.

### One predictor

```{r}
#| label: chi-m1
chi_m1 <- lm(violent_rate ~ unemployment_rate, data = areas)
summary(chi_m1)
```

Three numbers to pull out of that wall of output, in order of importance:

- **The slope (~3.3):** each additional percentage point of unemployment is associated with about 3.3 more violent crimes per 1,000 residents.
- **Multiple R-squared (~0.71):** unemployment alone accounts for about 71% of the variance in violent-crime rates *across areas*.
- **The slope's p-value:** far below .05 — a slope this steep essentially never shows up when the true slope is zero.

### Two predictors

Areas with high unemployment also tend to have more vacant housing. Does each carry its own association, or is one riding the other's coattails?

```{r}
#| label: chi-m2
chi_m2 <- lm(violent_rate ~ unemployment_rate + pct_vacant, data = areas)
summary(chi_m2)
```

Compare the unemployment slope across the two models: **~3.3 alone → ~2.1 with vacancy in the model**. That drop is the whole point of multiple regression: part of what the simple model credited to unemployment was really variance it *shares* with vacancy. Each coefficient is now an **adjusted** association — unemployment's relationship with violence *among areas with similar vacancy*, and vice versa. Both stay strongly significant, and R² rises to ~0.82.

**Interpretation sentences worth copying:**

- "Holding vacancy constant, each additional percentage point of unemployment is associated with about 2.1 more violent crimes per 1,000 residents."
- "Holding unemployment constant, each additional percentage point of vacant housing is associated with about 1.9 more violent crimes per 1,000 residents."
- What we may **not** say: anything with "causes," "drives," or "increases" as an action verb — and nothing about individual *people*. These are adjusted associations between *areas* (Session 10's ecological fallacy never left).

### Diagnose: the residual plot

The model assumes its misses are patternless. Look:

```{r}
#| label: chi-resid
chi_diag <- tibble(fitted = fitted(chi_m2), residual = resid(chi_m2))

ggplot(chi_diag, aes(fitted, residual)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point(alpha = 0.8) +
  labs(x = "Fitted violent-crime rate", y = "Residual")
```

What to check, in order: (1) **no curve** — the cloud shouldn't bend (a bend means the straight-plane shape is wrong); (2) **even spread** — here the spread visibly widens at higher fitted values, so the model misses by more in high-violence areas. That fanning doesn't bias the slopes, but it means the error bars are least trustworthy exactly where the stakes are highest — say so when you report.

### Diagnose: VIF

If predictors are near-duplicates of each other, the model can't apportion credit between them and the coefficients wobble. VIF puts a number on the redundancy:

```{r}
#| label: chi-vif
vif(chi_m2)
```

Rules of thumb: **above 5, pay attention; above 10, you have a problem.** Ours sit around 1.8 — unemployment and vacancy are correlated (r ≈ 0.67) but each still brings its own information. If you ever do see a huge VIF: drop one of the offenders, or combine them into an index. More data won't fix redundancy.

### The Bayesian version

Same formula, same `brm()` pattern as ANOVA week:

```{r}
#| label: chi-brm
#| output: false
chi_bayes <- brm(violent_rate ~ unemployment_rate + pct_vacant,
                 data = areas,
                 chains = 2, iter = 2000, seed = 705, refresh = 0)
```

```{r}
#| label: chi-brm-summary
summary(chi_bayes)
```

Reading the `Regression Coefficients` block: `Estimate` is the posterior mean (it lands almost exactly on `lm()`'s estimate — expected with default priors and 77 areas), and `l-95% CI` / `u-95% CI` bound the **credible interval**. Both intervals sit clearly away from zero. Quick health checks: `Rhat` at 1.00 and ESS in the hundreds-plus mean the chains behaved.

```{r}
#| label: chi-mcmc
mcmc_areas(chi_bayes, pars = c("b_unemployment_rate", "b_pct_vacant"),
           prob = 0.95) +
  labs(title = "Posteriors for the two predictors",
       subtitle = "Shaded = central 95% credible interval")
```

**The interpretation rule, carried over from ANOVA week:** if a coefficient's 95% credible interval **includes 0**, zero remains a credible value — no credible association. If it **excludes 0**, the association is credible, and you may say: "given the data and priors, there's a 95% probability the coefficient lies in this range." (The sentence the confidence interval never let you say.)

## Your Turn

Back to the 500 simulated neighborhoods from Lab 3 — poverty, unemployment, disorder, and last year's assault counts:

```{r}
#| label: yt-load
#| eval: false
nbhd <- read_csv("https://smourtgos.github.io/crju705/data/neighborhoods.csv")
```

Work in your own script. Comment each answer.

**1. Make the units speak.** `poverty_rate` and `unemployment_rate` are stored as proportions (0–1), so a raw slope would describe an absurd 0%-to-100% jump. Rescale before fitting — lecture's "watch the units" lesson, applied:

```{r}
#| label: yt-1
#| eval: false
nbhd <- nbhd |>
  mutate(poverty_pct      = poverty_rate * 100,
         unemployment_pct = unemployment_rate * 100)
# Then look first: scatterplot assaults ~ poverty_pct with an lm smooth.
```

**2. Fit and interpret.** Regress `assaults` on `poverty_pct` and `unemployment_pct`. In comments: one holding-constant sentence per coefficient (units: assaults per percentage point), and what R² says. Expect a much humbler R² than Chicago's — real-feeling individual-ish noise, and that's fine; say what the model *does* explain, not more.

```{r}
#| label: yt-2
#| eval: false
# nbhd_m <- lm(assaults ~ poverty_pct + unemployment_pct, data = nbhd)
# summary(nbhd_m)
```

**3. The diagnostics moment.** Residual plot + VIF, exactly as in the Walkthrough. Two things to note in comments: (a) the residuals form diagonal stripes — that's because `assaults` only takes small whole values (0, 1, 2, …), not a bug, but it *is* a hint that a count outcome bends the linear model's assumptions (counts and yes/no outcomes get their own tools — that story starts next session); (b) the VIF values — poverty and unemployment are correlated here (r ≈ 0.48), so is it redundancy trouble or not?

```{r}
#| label: yt-3
#| eval: false
# residual plot on nbhd_m, then vif(nbhd_m)
```

**4. The Bayesian twin.** Fit the `brm()` version (same options as the Walkthrough; ~500 rows samples fast), look at `summary()`, and make the `mcmc_areas()` plot for `b_poverty_pct` and `b_unemployment_pct`. In comments: one frequentist sentence and one Bayesian sentence about the poverty coefficient — each including the **size** of the association (assaults per percentage point, or per 10 points if that reads better), not just "significant"/"credible."

```{r}
#| label: yt-4
#| eval: false
# nbhd_bayes <- brm(assaults ~ poverty_pct + unemployment_pct, data = nbhd,
#                   chains = 2, iter = 2000, seed = 705, refresh = 0)
```

**5. *(Stretch)* Does disorder earn its keep?** Add `disorder_level` to the frequentist model (make it a factor with levels Low → Moderate → High → Very High first). Check the coefficients and the **adjusted** R². Does observed disorder tell you anything about assaults that poverty and unemployment didn't already? Connect your answer to lecture's third-predictor slide.

*(Reading a factor in `lm()`: R drops the first level — Low — as the reference, so each `disorder_level` coefficient is that level's difference from Low, holding the other predictors constant. Same idea as ANOVA's group comparisons, wearing regression clothes.)*

```{r}
#| label: yt-5
#| eval: false
# your code here
```

## Exit Ticket

Submit a short R script to Canvas before you leave, containing:

1. A comment: your holding-constant interpretation of the `poverty_pct` coefficient — one sentence, with units
2. A comment: what R² told you about your neighborhoods model, in signal-and-noise terms — and one thing the model *leaves* unexplained
3. A comment: what your residual plot and VIF each checked, and what each said (one sentence apiece)
4. Your poverty conclusion, two ways: one frequentist sentence (estimate + CI or p) and one Bayesian sentence (posterior + credible interval)
5. A comment: why "holding unemployment constant" is *not* the same claim as "poverty causes assaults" — name one lurker these neighborhoods might be hiding
