---
title: "Lab 6 · One Question, Two Tests"
subtitle: "CRJU 705 · Session 6"
---

**Goal:** by the end of this lab you can take one yes/no question about two groups and answer it twice: with a **p-value** from `prop.test()`, and with a **posterior** from `bayes_prop_test()`. Same four numbers go into both. What changes is the sentence each one lets you say.

**How labs work:** run the Setup chunk as-is, follow the Walkthrough along with me, then complete the **Your Turn** tasks in your own script. That script is what you submit to Canvas before you leave.

## Setup

Open RStudio, start a new R script, and run:

```{r}
#| label: lab-setup
#| message: false
library(tidyverse)
source("https://smourtgos.github.io/crju705/R/bayes.R")

officers <- read_csv("https://smourtgos.github.io/crju705/data/officers.csv",
                     show_col_types = FALSE) |>
  filter(!is.na(ptsd_screen)) |>
  mutate(ptsd_screen = factor(ptsd_screen, levels = c("Positive", "Negative")))
```

These are last week's correctional officers, set up exactly the same way: the eight officers with no PTSD screen on file are dropped (we say so out loud), which leaves 192, and Positive comes first.

One line is new. `source()` runs an R file from the course website, and that file gives you one new function, `bayes_prop_test()`. There is nothing to install. If you ever see `could not find function "bayes_prop_test"`, you skipped this line.

## Walkthrough

### Count first

Our question: **do male and female officers screen positive for PTSD at different rates?** Every test today starts from a count, just like last week:

```{r}
#| label: wt-count
officers |> count(gender, ptsd_screen)
```

Read four numbers off that table, two per group: how many screened positive, out of how many.

- **Male:** 33 positive, out of 33 + 111 = **144**
- **Female:** 8 positive, out of 8 + 37 = **45**

(Three officers are listed as Nonbinary/Other. Three people is too few to compare with anyone, so we leave that row alone and say so.)

```{r}
#| label: wt-rates
33 / 144
8 / 45
```

About **23%** of the men and **18%** of the women. A gap of about 5 percentage points. Male goes first all the way through, so every difference reads **Male minus Female**.

Is a 5-point gap real, or is it the kind of gap you get by luck with only 45 women in the sample?

### The frequentist test: `prop.test()` with two groups

Last week you typed one "how many" and one "out of how many." With two groups you type two of each, wrapped in `c()`. Male first, both times:

```{r}
#| label: wt-prop
prop.test(x = c(33, 8), n = c(144, 45))
```

Read it from the bottom up.

- **`prop 1` and `prop 2`** are the two rates, 0.229 and 0.178. Male is `prop 1` because we typed it first.
- **`95 percent confidence interval`** is for the *difference*, Male minus Female: about **-0.09 to +0.20**, or 9 points lower to 20 points higher. Zero is inside it. That is last week's skill.
- **`p-value = 0.6011`** is tonight's new line. If men and women truly screened positive at the *same* rate, a gap this big or bigger would still turn up about **60%** of the time, just from the luck of who landed in the sample.

A p-value of 0.60 is nowhere near 0.05. The verdict is **"fail to reject the null."**

### Saying it out loud without lying

Here is the sentence I want:

> In this sample, 23% of male officers and 18% of female officers screened positive. A gap that size is well within what chance alone produces (p = 0.60), so these data cannot tell us whether the true rates differ.

And here is what not to say:

- ~~"There is a 60% chance that men and women have the same rate."~~ The p-value is about the *data*, assuming the null is true. It is never the probability that the null is true.
- ~~"We proved there is no difference."~~ Failing to reject is not proving. The interval allows anything from 9 points lower to 20 points higher.
- ~~"Men have higher PTSD rates (23% vs 18%)."~~ That is the estimate with the uncertainty deleted.

### The Bayesian test: same four numbers

Now hand the same four numbers to `bayes_prop_test()`:

```{r}
#| label: wt-bayes
bayes_prop_test(x = c(33, 8), n = c(144, 45))
```

Read it top to bottom.

- The first lines repeat your data back to you, so you can check your typing.
- **`prior`** says "know-nothing." We did not tell it what we believed beforehand, so it started from a blank slate.
- **`estimate`** is the best single guess for each group's true rate, and each group gets its own **credible interval**.
- **`difference (1 minus 2)`** is Male minus Female: about **4 points**, with a 95% credible interval of roughly **-10 to +16 points**.
- The last two lines answer the question directly: the probability that the male rate is higher is about **0.74**.

Here is the sentence this one lets you say, and it is the sentence we were *forbidden* to say about a confidence interval last week:

> Given these data, there is a 95% probability that the true difference is between 10 points lower and 16 points higher for men, and about a 74% probability that the male rate is the higher one.

Notice the two tests do not disagree. Both intervals run from roughly -10 to somewhere between +16 and +20. One stops at "not significant." The other tells you *how probable* each direction is: 74% is better than a coin flip, and a long way from sure.

### Adding what you knew beforehand

A Bayesian can bring in earlier knowledge, as long as they say it out loud. Suppose earlier surveys at these facilities found about **1 officer in 5** screening positive, and you trust that about as much as you would trust having already seen **20 officers: 4 positive, 16 negative**. That belief is two numbers:

```{r}
#| label: wt-prior
bayes_prop_test(x = c(33, 8), n = c(144, 45), prior = c(4, 16))
```

Look at the `prior` line, then at what changed: almost nothing. The probability moved from 0.74 to 0.77. With 189 real officers in the data, 20 imaginary ones do not carry much weight. A prior matters most when you have very little data.

### The p-value was in `t.test()` the whole time

Last week you ran this line and read the interval. Look one line higher:

```{r}
#| label: wt-ttest
t.test(sleep_hours ~ ptsd_screen, data = officers)
```

`p-value = 0.3206`. If officers who screen positive and negative truly slept the same amount, a gap as big as ours (about 15 minutes) would turn up about a third of the time by luck. Fail to reject. It is the same conclusion the interval gave you last week, because zero was inside it.

## Your Turn

Same moves, new question: **do officers at the North Unit and the Central Unit screen positive at different rates?** Work in your own script and comment each answer. North goes first all the way through.

**1.** Count. Run `officers |> count(facility, ptsd_screen)`. In a comment, write the four numbers you need: how many screened positive at North, out of how many, and the same for Central. Then compute the two rates.

```{r}
#| label: yt-1
#| eval: false
# your code here
```

**2.** The frequentist test. Put your four numbers into `prop.test(x = c(__, __), n = c(__, __))`, North first. In a comment, report the confidence interval for the difference and the p-value. Do you reject the null or fail to reject it?

```{r}
#| label: yt-2
#| eval: false
# your code here
```

**3.** Write the sentence. In a comment, state your Task 2 result the way I did in the walkthrough. Then write one *wrong* sentence about your p-value underneath it, and label what is wrong with it.

```{r}
#| label: yt-3
#| eval: false
# a comment-only task
```

**4.** The Bayesian test. Put the same four numbers into `bayes_prop_test()`. In a comment, report the credible interval for the difference and the probability that North's rate is higher. Then write the sentence this output lets you say that Task 2's output does not.

```{r}
#| label: yt-4
#| eval: false
# your code here
```

**5.** The p-value in `t.test()`. Run `t.test(job_satisfaction ~ ptsd_screen, data = officers)`, the line from last week's Your Turn. In a comment: what is the p-value, do you reject or fail to reject, and how does that match what the interval told you last week?

```{r}
#| label: yt-5
#| eval: false
# your code here
```

**6.** *(Stretch)* Rerun Task 4 with the walkthrough's prior, `prior = c(4, 16)`. In a comment: how much did the probability move, and why so little?

```{r}
#| label: yt-6
#| eval: false
# your code here
```

When you are done, submit your script to Canvas. The in-class exercise is graded on completing it.
