---
title: "Demo Walkthrough · Session 5"
subtitle: "Confidence intervals — turning one sample into an honest range"
---

::: {.callout-note appearance="simple"}
**What this page is.** The Session 5 lecture demos, written down step by step. Every chunk below is self-contained — copy any of them into your own R session and they run as-is. Use this page to catch up after a missed class, or to re-run what happened on screen at your own pace.
:::

## What you'll build

1. John Snow's 1854 cholera inference, recomputed from his original data — the classic example of reasoning from a sample to the world
2. A demonstration that point estimates fluctuate: 1,000 samples from a population whose truth we secretly know
3. A 95% confidence interval from `prop.test()`, and **the 100-intervals plot** — the picture that shows what "95% confident" actually means
4. Sentences you can and cannot say when reporting an interval
5. How sample size buys precision, and a CI for a two-group difference with `t.test()`

## Setup

Two packages today: the tidyverse, plus **HistData**, which ships John Snow's original cholera data (`Snow.deaths` and `Snow.pumps`):

```{r}
#| label: setup
library(tidyverse)
library(HistData)   # install.packages("HistData")  — once, if you don't have it
set.seed(705)

crimes <- read_csv("https://smourtgos.github.io/crju705/data/chicago-crimes-2025.csv",
                   show_col_types = FALSE)
```

## 1 · Inference, in 1854

Cholera breaks out in Soho, London. The debate: bad air ("miasma") or contaminated water? John Snow couldn't test every household in London — that's the whole population. So he mapped the **sample of evidence** he could get: deaths, and where people drew their water.

We can rerun his reasoning. For each of the 578 recorded deaths, find the *nearest* pump. Two verbs arrive ahead of schedule here — `crossing()` builds every death–pump pair, and `left_join()` looks up each point's coordinates. Joins get their full treatment in next week's lab; today, read them as "pair up, then look up" and focus on the logic:

```{r}
#| label: snow-nearest
deaths <- as_tibble(Snow.deaths)
pumps  <- as_tibble(Snow.pumps)

pairs <- crossing(case = deaths$case, pump = pumps$pump) |>
  left_join(deaths |> select(case, dx = x, dy = y), by = "case") |>
  left_join(pumps  |> select(pump, label, px = x, py = y), by = "pump") |>
  mutate(dist = sqrt((dx - px)^2 + (dy - py)^2))

nearest <- pairs |>
  slice_min(dist, n = 1, by = case)

nearest |>
  count(label, name = "deaths_nearest") |>
  arrange(desc(deaths_nearest))
```

One pump — **Broad St** — is the nearest pump for 359 of the 578 deaths. Seen on the map:

```{r}
#| label: snow-map
#| fig-height: 5.5
broad <- pumps |> filter(label == "Broad St")

ggplot() +
  geom_point(data = deaths, aes(x, y), color = "firebrick", alpha = 0.5, size = 1.5) +
  geom_point(data = pumps, aes(x, y), color = "navy", size = 3, shape = 17) +
  annotate("label", x = broad$x, y = broad$y - 0.8, label = "Broad St. pump") +
  coord_equal() +
  labs(title = "Cholera deaths cluster around a single pump",
       subtitle = "Red dots = deaths · navy triangles = water pumps",
       x = NULL, y = NULL)
```

Snow wasn't just *describing* this map — he was making a bigger claim about the whole city, and about cholera itself. **That jump — from the data you have to the world you don't — is what we mean by inference.** And the fine print from lecture: you cannot *PROVE* anything from samples. We collect data and analyze it to build up a **weight of evidence** that shifts our certainty toward one conclusion or another.

## 2 · A rare luxury: we know the truth

Our Chicago file holds ~30,000 incidents from 2025. Today, *pretend the file is the whole population* — then we can check every inference against the truth:

```{r}
#| label: pop-truth
crimes |> summarize(pop_arrest_rate = mean(arrest), n = n())
```

The "true" citywide arrest proportion is **16.1%**. Lock that in — in real research you never get to see this number.

Now act like a researcher who can only afford to pull 250 case files:

```{r}
#| label: one-sample
sample_250 <- crimes |> slice_sample(n = 250)

sample_250 |> summarize(sample_arrest_rate = mean(arrest))
```

This is a **point estimate** — the sample's best single guess at the population value. About 14.8% versus the true 16.1%. Close-ish. But how good is this estimate? How *certain* are we?

## 3 · Caution! Samples fluctuate

Draw a fresh sample of 250 and the estimate moves. Do it 1,000 times:

```{r}
#| label: many-samples
#| fig-height: 4.5
many_estimates <- map_dbl(1:1000, \(i) {
  crimes |> slice_sample(n = 250) |> summarize(p = mean(arrest)) |> pull(p)
})

ggplot(tibble(p = many_estimates), aes(p)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = mean(crimes$arrest), color = "firebrick", linewidth = 1.2) +
  labs(title = "1,000 sample estimates of the arrest rate (n = 250 each)",
       subtitle = "Red line = the true population value",
       x = "Sample arrest proportion", y = "Number of samples")
```

That's Session 4's sampling distribution, live. Where did the middle 95% of estimates land?

```{r}
#| label: quantile-bridge
quantile(many_estimates, c(0.025, 0.975))
```

Roughly **12% to 21%** — any *single* sample's estimate could easily be a couple points off the truth. So reporting "14.8%" alone is **overconfident**; we owe our reader a range that reflects the fluctuation. Problem: this exercise required 1,000 samples and a known population. A real researcher has **one sample**. Now what?

## 4 · The confidence interval

The confidence interval turns *one* sample into an honest range, using what Session 4 taught us about how sample statistics behave:

$$\text{estimate} \; \pm \; \underbrace{1.96 \times SE}_{\text{margin of error}}$$

The **SE** (standard error) is the standard deviation of the sampling distribution, *estimated from the one sample you have*; **1.96** comes from the normal distribution — ±1.96 SDs covers the middle 95%. R does the whole thing for us:

```{r}
#| label: prop-test
prop.test(sum(sample_250$arrest), n = 250)
```

Our one sample of 250 says: point estimate **14.8%**, 95% CI **10.8% to 20.0%**. The true value (16.1%) is inside — *this time*. But "95% confident" of what, exactly?

## 5 · The classic picture: 100 samples, 100 intervals

Draw 100 samples; build a 95% CI from each. (`map()` runs the procedure 100 times and `list_rbind()` stacks the 100 one-row results into one tibble — Session 3's `replicate()`, tibble-flavored; see the [simulation verbs reference](../resources.qmd#simulation-verbs).)

```{r}
#| label: ci-sim
ci_100 <- map(1:100, \(i) {
  s  <- crimes |> slice_sample(n = 250)
  ci <- prop.test(sum(s$arrest), n = 250)$conf.int
  tibble(sample = i, estimate = mean(s$arrest), lo = ci[1], hi = ci[2])
}) |>
  list_rbind() |>
  mutate(captured = lo <= mean(crimes$arrest) & mean(crimes$arrest) <= hi)

sum(ci_100$captured)   # how many of the 100 intervals caught the truth?
```

```{r}
#| label: ci-sim-plot
#| fig-height: 5.5
ggplot(ci_100, aes(sample, estimate, color = captured)) +
  geom_hline(yintercept = mean(crimes$arrest), color = "grey30", linewidth = 1) +
  geom_pointrange(aes(ymin = lo, ymax = hi), linewidth = 0.7, size = 0.25) +
  scale_color_manual(values = c("TRUE" = "grey65", "FALSE" = "firebrick")) +
  labs(title = paste0("100 samples, 100 confidence intervals — ",
                      sum(ci_100$captured), " capture the truth"),
       subtitle = "Horizontal line = true population arrest rate; red = intervals that missed",
       x = "Sample number", y = "Arrest proportion", color = "Captured?")
```

In plain language: each vertical bar is one research team's interval, the truth never moves, and about 95 of 100 intervals catch it. Each sample produced a *different* interval — the interval is a **random quantity**; the population value is **fixed**. In our run, 97 of 100 captured the truth. The teams who drew the red samples did nothing wrong — and their data give **no warning** that theirs are the unlucky intervals.

That long-run capture rate is *the entire meaning* of "95% confident": if we reran the whole study many times, in 95% of those replications the interval would contain the true value. Three things follow:

1. Our 95% CI does **not** necessarily contain the true population value
2. Our 95% CI is **not** a statement about the probability that *this particular* interval is correct
3. It is a **long-run prediction** based on numerous replications

(Everyone reacts to this with some version of "WTF?!?" — stare at the plot until it clicks.)

## 6 · Saying it out loud

Defensible sentences:

> "We estimate the citywide arrest rate at 14.8% (95% CI: 10.8% to 20.0%). Values in this range are consistent with our data."

> "The method we used captures the true value 95% of the time."

Not defensible:

> "There is a 95% probability the true arrest rate is between 10.8% and 20.0%." — the truth is fixed; it's either in there or it isn't.

> "95% of arrest rates fall in this interval." — the CI is about the population parameter, not individual observations.

The forbidden first sentence is forbidden only under **frequentist rules**, where probability attaches to the *procedure*, never to the parameter. Attaching probability to a claim itself is exactly what Session 3's Bayes' theorem does — and next session we compute an interval you *can* say that sentence about: the **credible interval**. If the forbidden sentence felt like the natural reading all along, your intuitions were Bayesian. You're in good company.

## 7 · Sample size buys precision

Same procedure, four sample sizes, one draw each:

```{r}
#| label: width-by-n
map(c(100, 250, 1000, 4000), \(n) {
  s  <- crimes |> slice_sample(n = n)
  ci <- prop.test(sum(s$arrest), n = n)$conf.int
  tibble(n = n, estimate = mean(s$arrest),
         lo = ci[1], hi = ci[2], width = ci[2] - ci[1])
}) |>
  list_rbind()
```

Quadrupling *n* roughly **halves** the width — precision grows with $\sqrt{n}$, so it gets expensive fast.

## 8 · Comparing two groups: do violent crimes happen later in the day?

Street lore says violence is a nighttime phenomenon; property crime, the daytime hustle. Our data:

```{r}
#| label: two-group-est
vp <- crimes |> filter(crime_category %in% c("Violent", "Property"))

vp |> summarize(mean_hour = mean(hour), sd_hour = sd(hour),
                n = n(), .by = crime_category)
```

Two point estimates: violent crimes average hour 12.8, property 12.5. Is that ~15-minute gap real, or sample fluctuation?

```{r}
#| label: two-group-box
#| fig-height: 4.5
ggplot(vp, aes(crime_category, hour, fill = crime_category)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(x = NULL, y = "Hour of day (0–23)")
```

The boxes nearly coincide — the eye can't settle this one. So resample: pull a bootstrap sample from each group, compute the difference in means, repeat 1,000 times:

```{r}
#| label: mean-diffs
violent_hours  <- vp$hour[vp$crime_category == "Violent"]
property_hours <- vp$hour[vp$crime_category == "Property"]

mean_diffs <- replicate(1000, {
  mean(sample(violent_hours,  length(violent_hours),  replace = TRUE)) -
  mean(sample(property_hours, length(property_hours), replace = TRUE))
})

quantile(mean_diffs, c(0.025, 0.975))
```

The middle 95% of resampled differences (Violent − Property) runs from about **0.06 to 0.44 hours** — all above zero. The two-sample t-test gives the CI directly, no loop required:

```{r}
#| label: t-test
t.test(hour ~ crime_category, data = vp)
```

Watch the direction: R orders groups alphabetically, so this tests **Property − Violent** — the mirror image of our resampled differences. Same finding, flipped sign:

```{r}
#| label: t-test-ci
t.test(hour ~ crime_category, data = vp)$conf.int
```

Reading it honestly: the 95% CI for the difference runs from **−0.44 to −0.06 hours** — it excludes zero, so the data are consistent with violent crimes genuinely averaging later. But look at the *magnitude*: somewhere between about **4 and 27 minutes**. Detectable ≠ dramatic. With ~19,500 incidents we can detect tiny differences; the CI hands you both facts at once — *direction* (probably real) and *size* (small). **Never report the first without the second.** This is your vaccination against next session's topic: a "significant" result can be substantively trivial.

## Try a variation

1. Compute a 95% CI for the proportion of incidents flagged as **domestic**: `prop.test(sum(crimes$domestic), nrow(crimes))`. Then write one *defensible* sentence reporting it — and audit yourself: did you claim a probability about *this* interval?
2. Add `conf.level = 0.80` to a `prop.test()` call. Does the interval get wider or narrower — and what did you trade to get it?
3. In the 100-intervals simulation, change `n = 250` to `n = 1000` in both places. What changes: the capture rate, or the widths?
4. Remove the Broad Street pump before the `slice_min()` step (`filter(pump != 7)` on the pairs). Which pump inherits the deaths — and why should that make you *more* confident in Snow's inference, not less?
