---
title: "Demo Walkthrough · Session 8"
subtitle: "Correlation: covariance, Pearson's r, what r cannot see, and cor.test()"
---

::: {.callout-note appearance="simple"}
**What this page is.** The Session 8 correlation lecture, written down step by step. Every chunk below is self-contained: copy any of them into your own R session and they run as-is. Use this page to catch up after a missed class, or to re-run what happened on screen at your own pace. (The second half of Session 8, midterm review, is at the end of the [slides](../slides/week-10-correlation.qmd).)
:::

## What you'll build

1. Covariance on real Chicago data, and the units problem that makes its size hard to read
2. Pearson's r as rescaled covariance, computed by hand before `cor()` does it for you
3. A calibration grid: what r = 0.1 through 0.9 actually *look* like
4. Three demonstrations of what r cannot see: identical summaries, curves, and one point that flips a sign
5. Spearman's rho as the rank-based backup
6. The real question: which features of Chicago's 77 community areas track violence, tested with `cor.test()`, with the ecological-fallacy guardrail on every claim

## Setup

Everything below needs the tidyverse and the *second level* of the course's anchor data: 2025's incidents added up to the 77 community areas and joined with Census measures.

```{r}
#| label: setup
library(tidyverse)
set.seed(705)

areas <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv",
                  show_col_types = FALSE)
```

## 1 · Covariance: moving together, in raw units

Row by row: do x and y sit on the *same side* of their averages?

$$\text{cov}(x, y) \;=\; \frac{\sum_i (x_i - \bar{x})(y_i - \bar{y})}{n - 1}$$

Both above average (or both below) gives positive products and a positive covariance; opposite sides give a negative one. The **sign** is trustworthy. The **size** is not. Watch what happens on real data:

```{r}
#| label: cov-units
cov(areas$unemployment_rate, areas$violent_rate)
cov(areas$median_income,     areas$violent_rate)
```

Is `r format(round(cov(areas$median_income, areas$violent_rate)), big.mark = ",")` a "stronger" relationship than `r round(cov(areas$unemployment_rate, areas$violent_rate))`? No idea. The first number is in *dollars times crimes per 1,000*, the second in *percentage points times crimes per 1,000*. **Covariance is glued to its units**, so you can't compare across pairs.

## 2 · Correlation rescales it

Divide the covariance by both standard deviations and the units cancel:

$$r \;=\; \frac{\text{cov}(x, y)}{s_x \, s_y}$$

```{r}
#| label: r-standardize
cov(areas$unemployment_rate, areas$violent_rate) /
  (sd(areas$unemployment_rate) * sd(areas$violent_rate))

cor(areas$unemployment_rate, areas$violent_rate)   # same thing, one function
```

Always between -1 and +1, no units, comparable: income runs r = `r sprintf("%.2f", cor(areas$median_income, areas$violent_rate))` with violence, so unemployment (`r sprintf("%.2f", cor(areas$unemployment_rate, areas$violent_rate))`) tracks violence more tightly than income does, direction aside. The usual reading bands (about 0.8 and up "very strong," 0.4 to 0.7 "moderate," 0.1 to 0.3 "weak," either sign) are **conventions, not laws**: in messy social-science data, a "moderate" r can be a major finding.

## 3 · Calibrate your eye

Before trusting the bands, see them. Simulate five datasets with known correlations and look:

```{r}
#| label: r-grid
#| fig-height: 3.5
set.seed(705)   # inline seed: keeps these particular clouds (and their r labels) stable
r_targets <- c(0.1, 0.3, 0.5, 0.7, 0.9)

r_grid <- map(r_targets, \(r) {
  x <- rnorm(150)
  y <- r * x + sqrt(1 - r^2) * rnorm(150)
  tibble(label = paste0("r = ", sprintf("%.2f", cor(x, y))), x = x, y = y)
}) |>
  list_rbind() |>
  mutate(label = fct_inorder(label))

ggplot(r_grid, aes(x, y)) +
  geom_point(alpha = 0.5, size = 1.4, color = "navy") +
  facet_wrap(~label, nrow = 1) +
  labs(title = "Five simulated datasets, 150 points each",
       x = NULL, y = NULL) +
  theme(axis.text = element_blank())
```

(The panel labels show each sample's *actual* r: 150 draws don't land exactly on the target, which is Session 4's sampling variability all over again.) By about 0.5 the trend is visible; below about 0.2 your eye mostly sees noise, and it's right. Negative correlations look the same with the tilt reversed.

## 4 · What r cannot see

### Same summaries, four different stories

Anscombe's quartet ships with R (Anscombe, 1973): four little datasets built to share their summary statistics.

```{r}
#| label: anscombe
#| fig-height: 4.5
anscombe_long <- anscombe |>
  pivot_longer(everything(), names_to = c(".value", "set"), names_pattern = "(.)(.)")

anscombe_long |>
  summarize(mean_x = mean(x), mean_y = mean(y), r = cor(x, y), .by = set)
ggplot(anscombe_long, aes(x, y)) +
  geom_point(size = 2.5, color = "navy") +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick", linewidth = 0.8) +
  facet_wrap(~set, labeller = labeller(set = \(s) paste0("Set ", s, ": r = 0.82"))) +
  labs(title = "Same r, four completely different stories", x = NULL, y = NULL)
```

Same averages, same r = 0.82, and only Set 1 is what you picture when someone says "r = 0.82." **Always plot before you correlate.**

### Limitation 1: r only sees straight lines

```{r}
#| label: u-shape
#| fig-height: 4
set.seed(705)   # inline seed: stable curve
u_data <- tibble(
  x = seq(-4, 4, length.out = 50),
  y = x^2 + rnorm(50, 0, 1)
)

ggplot(u_data, aes(x, y)) +
  geom_point(size = 2.5, color = "chocolate3") +
  geom_smooth(method = "loess", se = FALSE, color = "chocolate3") +
  labs(title = "A clear relationship that r cannot see",
       subtitle = paste0("Pearson's r = ", sprintf("%.2f", cor(u_data$x, u_data$y)),
                         ": the data follow a U-shaped curve"),
       x = "Variable X", y = "Variable Y")
```

Think stress and performance: an upside-down U. r near 0 means no **straight-line** relationship, not no relationship.

### Limitation 2: one point can manufacture a correlation

Twenty points of pure noise, then the same twenty plus a single wild point:

```{r}
#| label: outlier-demo
#| fig-height: 4
set.seed(705)   # inline seed: stable demo; the numbers below depend on it
base_data    <- tibble(x = 1:20, y = rnorm(20, 10, 4))   # pure noise
outlier_data <- bind_rows(base_data, tibble(x = 25, y = 100))

bind_rows(
  base_data    |> mutate(version = paste0("20 points of pure noise: r = ",
                                          sprintf("%.2f", cor(base_data$x, base_data$y)))),
  outlier_data |> mutate(version = paste0("Same noise + ONE outlier: r = ",
                                          sprintf("%.2f", cor(outlier_data$x, outlier_data$y))))
) |>
  mutate(version = fct_inorder(version)) |>
  ggplot(aes(x, y)) +
  geom_point(size = 2.5, color = "grey35") +
  geom_point(data = \(d) filter(d, y > 50), color = "firebrick", size = 4.5) +
  geom_smooth(method = "lm", se = FALSE, color = "navy") +
  facet_wrap(~version) +
  labs(x = "Variable X", y = "Variable Y")
```

Two lessons in one picture. First, the noise-only panel gave r = `r sprintf("%.2f", cor(base_data$x, base_data$y))`: **small samples produce correlations out of nothing**, which is why we test rather than just compute. Second, adding **one** extreme point swung r to `r sprintf("%+.2f", cor(outlier_data$x, outlier_data$y))`: a single point flipped the sign *and* manufactured a "moderate positive" correlation. Neither number describes a real relationship. (Outliers can also work the other way, *hiding* a real correlation.) Defense one is plotting first; you'd catch this instantly. Defense two is a correlation that doesn't care about wild values.

### The rank-based backup: Spearman's rho

Replace each value with its **rank** (1st, 2nd, 3rd...), then correlate the ranks. Spearman asks only: *is the relationship consistently up, or consistently down?*

```{r}
#| label: spearman-outlier
c(pearson  = cor(outlier_data$x, outlier_data$y),
  spearman = cor(outlier_data$x, outlier_data$y, method = "spearman"))
```

To Spearman, the wild point is just "rank 21": rho drifts from `r sprintf("%.2f", cor(base_data$x, base_data$y, method = "spearman"))` to `r sprintf("%.2f", cor(outlier_data$x, outlier_data$y, method = "spearman"))`, still weak, while Pearson flipped sign. When data are skewed or have wild values, check Spearman alongside Pearson.

## 5 · Real data: structure and violence across Chicago

A classic question about places: **which features of a neighborhood track its violence?**

```{r}
#| label: areas-peek
areas |>
  select(community, violent_rate, unemployment_rate, median_income, pct_vacant) |>
  head(3)
```

Rule from Section 4: plot first. One dot = one community area:

```{r}
#| label: unemp-scatter
#| fig-height: 4.5
ggplot(areas, aes(unemployment_rate, violent_rate)) +
  geom_point(size = 3, alpha = 0.7, color = "firebrick") +
  geom_smooth(method = "lm", se = FALSE, color = "navy") +
  labs(title = "Violent-crime rate and unemployment rate, 77 community areas",
       subtitle = paste0("r = ", sprintf("%.2f", cor(areas$unemployment_rate, areas$violent_rate))),
       x = "Unemployment rate (%)", y = "Violent crimes per 1,000 residents")
```

About as strong as area-level associations get in criminology. And the full menu, in one dataset:

```{r}
#| label: four-covariates
#| fig-height: 5
areas |>
  select(violent_rate, unemployment_rate, pct_renter, median_income, median_age) |>
  pivot_longer(-violent_rate, names_to = "covariate", values_to = "value") |>
  mutate(covariate = case_match(covariate,
    "unemployment_rate" ~ paste0("Unemployment rate: r = ",
      sprintf("%.2f", cor(areas$unemployment_rate, areas$violent_rate))),
    "pct_renter" ~ paste0("% renter-occupied: r = ",
      sprintf("%.2f", cor(areas$pct_renter, areas$violent_rate))),
    "median_income" ~ paste0("Median income: r = ",
      sprintf("%.2f", cor(areas$median_income, areas$violent_rate))),
    "median_age" ~ paste0("Median age: r = ",
      sprintf("%.2f", cor(areas$median_age, areas$violent_rate))))) |>
  ggplot(aes(value, violent_rate)) +
  geom_point(alpha = 0.6, size = 2, color = "navy") +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick", linewidth = 0.8) +
  facet_wrap(~covariate, scales = "free_x") +
  labs(title = "Violent-crime rate and four neighborhood measures",
       x = NULL, y = "Violent crimes per 1,000")
```

Strong positive (unemployment), moderate positive (renters), strong negative (income), and one genuine *nothing* (median age). Real data span the whole menu. Income is also where Pearson and Spearman disagree: the decline is real but **bent** (steep across lower-income areas, nearly flat among the richest), and income is right-skewed, so the rank-based Spearman runs stronger than Pearson because it only asks "downhill?", not "straight line?":

```{r}
#| label: income-spearman
c(pearson  = cor(areas$median_income, areas$violent_rate),
  spearman = cor(areas$median_income, areas$violent_rate, method = "spearman"))
```

## 6 · These are areas, not people

Our r = `r sprintf("%.2f", cor(areas$unemployment_rate, areas$violent_rate))` says: **community areas** with higher unemployment rates record more violent crime per 1,000 residents.

::: {.callout-important appearance="simple"}
**The ecological fallacy:** using an association between *groups* (areas, states, cities) to make claims about *individuals*. Our data contain **no individuals**. Nothing here says *unemployed people* commit violence.
:::

Saying it out loud, with area data. ✅ Defensible: "Community areas with higher unemployment rates recorded higher violent-crime rates." ❌ Not defensible: "Unemployed residents commit more violent crime," a claim about individuals these data can't see. ❌ Also not defensible: "Unemployment drives up violence," a claim about cause; concentrated poverty, disinvestment, and policing patterns all lurk behind both variables. Correlation is a pattern, not proof of cause.

The classic example (Robinson, 1950): across U.S. **states**, more immigrants went with *higher* literacy; across **individuals**, immigrants had *lower* literacy. Immigrants had settled in states with high literacy. Adding people up into areas can weaken, exaggerate, or even **flip** a relationship.

## 7 · Testing r: `cor.test()`

Same question as every test since Session 6. **The null:** the true correlation is 0. *If* it were, how surprising would an r like ours be, with 77 areas? One function does the whole thing:

```{r}
#| label: cor-test
unemp <- cor.test(areas$unemployment_rate, areas$violent_rate)
unemp
```

Reading it: **r = `r sprintf("%.2f", unemp$estimate)`**, with a 95% confidence interval of **`r sprintf("%.2f", unemp$conf.int[1])` to `r sprintf("%.2f", unemp$conf.int[2])`**. It is an interval for the *correlation*, so it lives between -1 and +1, and zero is far outside it. The p-value is below 2.2e-16: reject the null. Two perspective checks. First, with 77 areas, any r above about 0.22 (either sign) clears p < 0.05, so passing the test is a low bar and **the strength is the story**. Second, these 77 areas are *all of Chicago*, not a sample; read the null as a claim about the *process* that produces neighborhood outcomes.

### When the answer is "nothing here"

```{r}
#| label: cor-test-null
age <- cor.test(areas$median_age, areas$violent_rate)
age
```

r = `r sprintf("%.2f", age$estimate)`, and the interval runs from `r sprintf("%.2f", age$conf.int[1])` to `r sprintf("%.2f", age$conf.int[2])`: **zero is inside**, so we fail to reject. The honest sentence: *these data show no evidence of a straight-line association between median age and violence*. That is **not** the same as proving there is none.

### Several pairs at once

`cor()` takes every numeric column you give it:

```{r}
#| label: cor-matrix
areas |>
  select(violent_rate, unemployment_rate, median_income, pct_vacant) |>
  cor() |>
  round(2)
```

Read it like a mileage chart: each pair appears twice, and the 1s down the diagonal are each variable with itself. A good first look at a project's variables; then plot and `cor.test()` the pairs that matter.

### How to report it

> "Across Chicago's 77 community areas, unemployment was strongly associated with the violent-crime rate, r = `r sprintf("%.2f", unemp$estimate)`, 95% CI `r sprintf("%.2f", unemp$conf.int[1])` to `r sprintf("%.2f", unemp$conf.int[2])`, p < 0.001. This is an area-level association; it does not speak to individuals."

Direction, strength, uncertainty, and the level of analysis. Every time.

## Try a variation

1. `pct_no_hs_diploma` is the share of an area's adults without a high-school diploma. *Before computing anything*, write down your guess for the sign and rough size of its correlation with `violent_rate`. Then plot it and run `cor.test()`. Most people guess a strong positive; check what the data actually say, then try `pct_bachelors`.
2. In the outlier demo, change the wild point from `y = 100` to `y = 40`, then `y = 15`. At what point does Pearson stop being fooled, and does Spearman ever care?
3. Take a random 15 areas (`areas |> slice_sample(n = 15)`) and rerun the unemployment `cor.test()`. How much wider is the confidence interval with 15 areas instead of 77? Run it a few times: does r bounce around?
