---
title: "Demo Walkthrough · Session 7"
subtitle: "The full pipeline, run once — Alaska crash data, import to verdict"
---

::: {.callout-note appearance="simple"}
**What this page is.** The Session 7 worked example, written down step by step. No new statistics — this is the import → tidy → describe → visualize → test pipeline you already own, run start-to-finish on one dataset. Every chunk is self-contained; copy any of them into your own R session and they run as-is.
:::

## What you'll build

1. An import where the dates are parsed *at the door* — and verified
2. The variables your question needs that the file doesn't have (`hour`, `weekend`, `night`)
3. Honest description and visualization of the intoxication–injury gap
4. `t.test()` **and** `ttestBF()` on the same question — two grammars, one verdict

The folk wisdom holds: 80% of data work is getting and preparing the data — and then the t-test is one line. Every dataset in this course so far arrived clean because your instructor cleaned it. That subsidy ends here.

## Setup

```{r}
#| label: setup
library(tidyverse)
library(BayesFactor)   # install.packages("BayesFactor")  — once, if you don't have it
set.seed(705)
```

## Step 1 · Import

The Alaska crash data, with the date column handled up front via `col_types` (the equivalent Lab 5 habit: import first, then `mutate(date = mdy(date))` — two roads to the same real date column):

```{r}
#| label: import
crashes <- read_csv(
  "https://smourtgos.github.io/crju705/data/crash-ak.csv",
  col_types = cols(date = col_date(format = "%m/%d/%Y"))
)

sum(is.na(crashes$date))   # zero NAs = every date parsed
```

20,000 crashes, 2022–2026. Whichever road you take: **verify the parse**. A silent `NA` here poisons everything downstream.

## Step 2 · Clean and derive

The file has no "night" or "weekend" column. Files never do — *you* build the variables your question needs:

```{r}
#| label: derive
crashes <- crashes |>
  mutate(
    hour    = hour(time),
    weekend = wday(date, week_start = 1) >= 6,   # Mon = 1 … Sat = 6, Sun = 7
    night   = hour >= 22 | hour < 4              # 10 p.m. – 3:59 a.m.
  )
crashes |> count(night)
```

Night hours hold about 25% of all crashes — but look what happens inside the intoxication-involved ones:

```{r}
#| label: night-by-intox
crashes |> summarize(p_night = mean(night), .by = intox_any)
```

24% of sober-driver crashes versus 39% of intoxication-involved ones. One `mutate()` made that pattern visible.

## Step 3 · Describe

The Homework 5 question, back for a verdict: **do crashes involving an intoxicated driver injure more people?**

```{r}
#| label: describe
crashes |>
  summarize(
    crashes  = n(),
    mean_inj = mean(injuries),
    sd_inj   = sd(injuries),
    .by = intox_any
  )
```

0.74 vs. 1.13 injured people per crash. Point estimates — you know by now that's not the final word.

## Step 4 · Visualize

Session 5's confidence interval, built by hand and drawn with an honest caption:

```{r}
#| label: viz
#| fig-height: 4.5
inj_summary <- crashes |>
  mutate(intox_group = factor(intox_any, labels = c("No intox", "Intox involved"))) |>
  summarize(n = n(), mean_inj = mean(injuries), sd_inj = sd(injuries),
            .by = intox_group) |>
  mutate(se = sd_inj / sqrt(n),
         ci_lo = mean_inj - 1.96 * se,
         ci_hi = mean_inj + 1.96 * se)

ggplot(inj_summary, aes(intox_group, mean_inj)) +
  geom_pointrange(aes(ymin = ci_lo, ymax = ci_hi),
                  color = "firebrick", linewidth = 1.2, size = 1) +
  labs(title = "Mean injuries per crash, with 95% confidence intervals",
       subtitle = "Alaska crash data, n = 20,000",
       x = NULL, y = "Mean injuries per crash")
```

The intervals don't come close to overlapping. Formal tests next — both traditions.

## Step 5a · Test it — frequentist

```{r}
#| label: ttest
t.test(injuries ~ intox_any, data = crashes)
```

Same gotcha as Homework 5: R computes group 0 − group 1, so the negative signs mean the *intoxication* group is higher. t(1733) = −12.5, p < .001.

## Step 5b · Test it — Bayesian

`BayesFactor` insists the grouping variable be a **factor** — so we hand it one:

```{r}
#| label: ttest-bf
crashes <- crashes |>
  mutate(intox_group = factor(intox_any, labels = c("No intox", "Intox involved")))

ttestBF(formula = injuries ~ intox_group, data = crashes)
```

BF~10~ ≈ 2 × 10^43^: the data are astronomically better predicted by "the groups differ" than by "no difference."

A formatting aside: depending on your R settings, that number may print as `2.13e+43` (scientific notation) or as a 44-digit integer (if you've set `options(scipen = 999)`). Same number, different formatting — don't panic either way.

## Two grammars, one verdict

- **Frequentist:** *if* the true difference were zero, data like ours would essentially never occur — t(1733) = −12.5, p < .001
- **Bayesian:** the difference model beats the no-difference model by a factor of ~10^43^
- **The estimate, in English:** intoxication-involved crashes average **0.33 to 0.45 more injured people per crash** (95% CI)

The two frameworks agree *here* because the evidence is overwhelming; with small samples and borderline evidence they can part ways — Session 6's collision — and that's exactly when running both matters. **What neither test says:** *why*. Intoxication crashes also differ in hour, speed, and road conditions — an observational gap no test statistic closes.

That's the whole pipeline: import, derive, describe, visualize, test — and the tests were the shortest part. The rest of the workshop turns this loose on the dataset *you're* proposing for your final project ([Lab 7](../labs/lab-07-workshop.qmd), and the [final project page](../final-project.qmd) for the dataset-selection requirements).

## Try a variation

1. Swap the grouping variable: do **weekend** crashes injure more people than weekday ones? Run all of Steps 3–5 with `weekend` in place of `intox_any`.
2. Ask the same intoxication question about `fatalities` instead of `injuries`. Do both grammars still agree — and does the *size* of the difference still matter to a policy audience?
3. Filter to a single county (`filter(county == "Anchorage")`) and re-run Step 5. What happens to the CI width and the Bayes factor when n shrinks?
4. Redefine `night` as 9 p.m.–4:59 a.m. Does the 25%-vs-39% intoxication pattern from Step 2 survive your definition? *(Robustness to arbitrary cutoffs is an analysis habit, not a luxury.)*
