---
title: "Demo Walkthrough · Session 8"
subtitle: "ANOVA — comparing three or more means, in both traditions"
---

::: {.callout-note appearance="simple"}
**What this page is.** The Session 8 lecture demos, written down step by step. Every chunk below is self-contained — copy any of them into your own R session and they run as-is. Use this page to catch up after a missed class, or to re-run what happened on screen at your own pace.
:::

## What you'll build

1. A demonstration of why "just run a pile of t-tests" backfires — the false-alarm rate, computed
2. ANOVA's two kinds of variance, built by hand on three classrooms before any function does it for you
3. A null world *by construction* — three random groups from the same 70 cities — and the F distribution it produces
4. `aov()` and Tukey's HSD, with the ANOVA table reconciled against your hand computations
5. The Bayesian twin: `brm()` + `emmeans()` + `pairs()`, read through HPD intervals
6. The same workflow on 30,000 real Chicago incidents — where "significant" and "dramatic" part ways

## Setup

Everything below needs the tidyverse, two new packages (`brms` and `emmeans` — install once with `install.packages(c("brms", "emmeans"))`), and the course's anchor dataset:

```{r}
#| label: setup
library(tidyverse)
library(brms)
library(emmeans)
set.seed(705)

crimes <- read_csv("https://smourtgos.github.io/crju705/data/chicago-crimes-2025.csv",
                   show_col_types = FALSE)
```

## 1 · Why not just run a pile of t-tests?

Three groups means 3 pairwise comparisons. Six groups means 15. Each test carries its own 5% false-alarm risk, and the risks pile up:

```{r}
#| label: alpha-inflation
tibble(groups = c(2, 3, 4, 6, 10)) |>
  mutate(
    pairwise_tests = choose(groups, 2),
    p_false_alarm  = round(1 - 0.95^pairwise_tests, 2)
  )
```

By six groups you have a coin-flip's chance of at least one bogus "significant" difference; by ten it's near-certain. ANOVA's answer: ask **one omnibus question** — *are all the group means plausibly equal?* — at a single 5% risk. (When we go pairwise afterward, Tukey will restore the family-wise rate. Hold that thought.)

## 2 · Two kinds of variance, built by hand

The name is strange — **AN**alysis **O**f **VA**riance for a question about *means* — but the trick is the whole method. Start with three imaginary classrooms:

```{r}
#| label: classrooms
classrooms <- tibble(
  score     = c(75, 80, 78, 74, 77,    # Class 1
                88, 90, 92, 85, 87,    # Class 2
                70, 65, 68, 72, 66),   # Class 3
  classroom = factor(rep(c("Class 1", "Class 2", "Class 3"), each = 5))
)

class_stats <- classrooms |>
  summarize(class_mean = mean(score), class_var = var(score), .by = classroom)
class_stats
```

Two different kinds of "spread" are hiding in that table. **Within-group variance** is how much scores bounce around *inside* each classroom; **between-group variance** is how much the classroom *averages* differ from each other:

```{r}
#| label: classrooms-two-vars
mean(class_stats$class_var)   # WITHIN: average spread inside a class
var(class_stats$class_mean)   # BETWEEN: spread of the class means
```

Between (102.8) dwarfs within (7.1) — the class averages are far more spread out than the chance bouncing inside classes would suggest. Signal-to-noise ratio ≈ 14.5. These classes look genuinely different.

```{r}
#| label: classrooms-plot
#| fig-height: 4.5
ggplot(classrooms, aes(classroom, score)) +
  geom_boxplot(alpha = 0.2, width = 0.3) +
  geom_jitter(width = 0.08, size = 3, alpha = 0.6, color = "navy") +
  geom_point(data = class_stats, aes(y = class_mean),
             shape = 23, size = 5, fill = "firebrick", color = "white") +
  labs(title = "Within vs. between group variance",
       subtitle = "Within = spread inside each box · Between = spread OF the diamonds (class means)",
       x = NULL, y = "Test score")
```

## 3 · A null world by construction: Stanton's rain example

Now flip the situation: three groups where **nothing differs** — by construction. Base R's `precip` holds annual precipitation for 70 U.S. cities. Draw three random groups of 20 from the *same* 70 cities:

```{r}
#| label: rain-groups
data("precip")
set.seed(705)   # so you draw the same "random" groups the lecture did
group1 <- sample(precip, 20)
group2 <- sample(precip, 20)
group3 <- sample(precip, 20)

c(mean(group1), mean(group2), mean(group3))
```

The means differ a little — but any differences here are **pure sampling luck**. The null hypothesis is *true by construction*. Now the same two variances:

```{r}
#| label: rain-two-vars
mean(c(var(group1), var(group2), var(group3)))    # WITHIN
var(c(mean(group1), mean(group2), mean(group3)))  # BETWEEN
```

Within ≈ 133, between ≈ 2.3 — the classrooms' pattern, completely flipped. Why so lopsided? City-to-city rainfall varies a lot (Phoenix vs. Mobile), so within is big; but each group's *mean* averages 20 cities, and Session 4 taught you means fluctuate far less than raw values ($SE = \sigma/\sqrt{n}$). With no real group differences, between-group variance is nothing but the tiny sampling noise of three means. That's the diagnostic: **between ≫ within → something real separates the groups; between ≪ within → nothing but noise.** All we need is a ruler for "how much bigger is big enough."

### The ruler: F, and what it does in a null world

$$F \;=\; \frac{\text{between-group variance}}{\text{within-group variance}} \;=\; \frac{\text{signal}}{\text{noise}}$$

When the null is true, F bounces around low values by luck of the draw. Don't take that on faith — draw 10,000 F values from a null world and look:

```{r}
#| label: f-dist
#| fig-height: 5
set.seed(705)   # inline seed: keeps this particular histogram (and its cutoff) stable
f_null <- tibble(f = rf(10000, df1 = 2, df2 = 57))   # our design: 3 groups, N = 60
cutoff <- quantile(f_null$f, 0.95)

ggplot(f_null, aes(f)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  geom_vline(xintercept = cutoff, color = "firebrick", linewidth = 1.1) +
  labs(title = "10,000 F values from a null world — F(2, 57)",
       subtitle = "Red line = 95th percentile: F values beyond it are rare when nothing differs",
       x = "F value", y = "Count")
cutoff
```

The 95% cutoff for this design lands near ≈ 3.1 (theory says `qf(0.95, 2, 57)` ≈ 3.16 — simulation is how we watch the theory). An F far beyond it means our data are *unlikely under the null* — the same logic as every test we've run.

## 4 · Running the ANOVA with `aov()`

`aov()` wants long format — one row per observation, one column saying which group it belongs to:

```{r}
#| label: rain-aov
rain <- tibble(
  rainfall = c(group1, group2, group3),
  group    = factor(rep(c("Group 1", "Group 2", "Group 3"), each = 20))
)

rain_model <- aov(rainfall ~ group, data = rain)
summary(rain_model)
```

**F = 0.34, p = 0.71.** That F sits squarely in the fat part of the histogram above — nowhere near the cutoff. No evidence the groups differ. *Correct answer* — they're the same population by construction.

### The table is our two variances, dressed up

Reconcile the ANOVA table against the numbers you computed by hand:

- **Mean Sq Residuals = 132.7** — exactly the within-group variance from Section 3.
- **Mean Sq group = 45.2 = 20 × 2.26** — the between-group variance, *scaled by the group size*. Why × 20? A 2-inch gap between means built on 20 cities each is far stronger evidence than the same gap built on 2 — `aov()` bakes the sample size into the signal.
- **F = 45.2 / 132.7 = 0.34.** So when your hand-computed ratio and `aov()`'s F disagree, this scaling is why — same logic, different units.

### Now make a difference artificially

Shift Group 3 down 15 inches — now there *is* a real difference, and we know exactly where it lives:

```{r}
#| label: rain-aov-shifted
rain_shifted <- rain |>
  mutate(rainfall = if_else(group == "Group 3", rainfall - 15, rainfall))

shifted_model <- aov(rainfall ~ group, data = rain_shifted)
summary(shifted_model)
```

The within-group noise didn't move (132.7 — shifting a whole group doesn't change the spread *inside* it). The between-group signal exploded. **F = 14.2, p ≈ 0.00001** — deep in the tail.

### A significant F only says "*something* differs" — Tukey says what

```{r}
#| label: rain-tukey
TukeyHSD(shifted_model)
```

Group 3 differs from both others — intervals exclude 0, adjusted p-values tiny, exactly where we put the difference. Group 1 vs. Group 2: interval spans 0, p = 0.81 — no evidence, *as designed*. And these p-values are already adjusted for making all three comparisons — Tukey is the fix for the t-test pile-up problem from Section 1.

## 5 · The Bayesian twin: `brm()` + `emmeans()`

::: {.callout-warning appearance="simple"}
`brms` compiles models with your computer's C++ toolchain. Expect the *first* model of each R session to pause ~a minute at "Compiling Stan program..." — that's normal, not a crash.
:::

Identical formula, Bayesian machinery:

```{r}
#| label: brm-rain
#| output: false
bayes_rain <- brm(rainfall ~ group, data = rain,
                  chains = 2, iter = 2000, seed = 705, refresh = 0)
```

Two independent MCMC chains, 2,000 draws each → a full **posterior distribution** for every group mean. `emmeans()` turns the fitted model into group summaries — one row per group, with the posterior median and a 95% **HPD interval**, the range holding the most credible values for that group's mean:

```{r}
#| label: emmeans-rain
rain_post <- emmeans(bayes_rain, ~ group)
rain_post
```

What we really care about are *differences*:

```{r}
#| label: pairs-rain
pairs(rain_post)
```

```{r}
#| label: plot-pairs-rain
#| fig-height: 4
plot(pairs(rain_post)) +
  geom_vline(xintercept = 0, linetype = 2, color = "firebrick")
```

Every 95% HPD interval for a pairwise difference **crosses 0**: zero — "no difference" — remains a credible value for every contrast. Same verdict as F = 0.34, p = 0.71. Two traditions, one conclusion. When a difference is *real*, its interval abandons 0 — you'll see that on the Chicago data below.

**The prior we just used.** We didn't state a prior, so `brms` fell back on its **defaults**: essentially flat on the group coefficients — every difference equally believable before the data — and weakly informative on the intercept and spread. Flat priors mean "the data do the talking," which is fine at this *n*, but *say so*: "we used brms default priors" belongs in any writeup that reports a posterior. And notice the pattern across the course — `ttestBF()` in Session 6, `brm()` today, the `correlation` package in Session 10 are **different interfaces to the same move**: prior + likelihood → posterior. The interface changes; the reasoning never does. Session 10 shows what happens when you deliberately *change* the prior — and when it matters.

## 6 · Real data: does the clock differ by crime category?

In Session 5 we found violent crimes average slightly *later* in the day than property crimes — two groups. Chicago's data sort every incident into **three** categories, and ANOVA takes all three in one shot:

```{r}
#| label: chicago-means
crimes |> summarize(mean_hour = mean(hour), sd_hour = sd(hour),
                    n = n(), .by = crime_category)
```

Violent runs latest (12.8), Other earliest (12.1). Gaps of fractions of an hour, with SDs near 7 — real structure, or noise? And remember: n = 30,000, and big samples detect *tiny* things.

```{r}
#| label: chicago-aov
chicago_model <- aov(hour ~ crime_category, data = crimes)
summary(chicago_model)
```

**F(2, 29997) = 25.3, p ≈ 10⁻¹¹.** *Something* differs, emphatically — but that's all the omnibus F will ever tell us. Which pairs, and by how much?

```{r}
#| label: chicago-tukey
TukeyHSD(chicago_model)
```

All three intervals exclude 0 — every category differs from every other. Now look at the **sizes**: 0.25 to 0.68 hours. That's **15 to 41 minutes**, on a 24-hour clock, against SDs near 7 hours. The Bayesian version is the exact three lines from Section 5 with `hour ~ crime_category` swapped in — nothing new to learn — and it agrees: every HPD interval for a pairwise difference excludes zero (the lecture runs it in full; only the compile-and-sample wait argues against re-fitting it here).

**Detectable ≠ dramatic.** Both traditions agree the mean incident hour genuinely differs across categories — and the biggest gap is 41 minutes. Detecting a difference is not the same as its *mattering*. Report **direction** *and* **size**, every time: "Mean incident hour differed across crime categories, F(2, 29997) = 25.3, p < .001; violent crimes averaged ≈ 15 minutes later than property crimes, 95% CI [0.02, 0.48] hours." "Significant" without magnitude is malpractice — Session 6's lesson, now with three groups.

## Try a variation

1. In the shift demo, change `- 15` to `- 5` inches. Does F still clear the cutoff? Which Tukey comparisons survive, and which lose significance?
2. Add `groups = 15` to the alpha-inflation table. How many pairwise tests would that demand, and what's the false-alarm probability?
3. Delete the `set.seed(705)` line in the rain-groups chunk and re-run the draw plus `aov()` a few times. How much do F and p bounce around when the null is true? *(You're watching the left half of the F histogram, one draw at a time.)*
4. Filter `crimes` to the four most common `primary_type` values (`count()` will find them) and run `aov(hour ~ primary_type)` plus Tukey. With six pairwise comparisons, which specific crime types actually differ on the clock?
