---
title: "Demo Walkthrough · Session 2"
subtitle: "Central tendency, dispersion, and distribution shapes — the hard way first, then the easy way"
---

::: {.callout-note appearance="simple"}
**What this page is.** The Session 2 lecture demos, written down step by step. Every chunk below is self-contained — copy any of them into your own R session and they run as-is. Use this page to catch up after a missed class, or to re-run what happened on screen at your own pace.
:::

## What you'll build

1. Mean, median, and mode — computed *by hand* first, so the built-ins never feel like magic
2. The sentencing demo: one 30-year outlier, three very different answers to "what's typical?" — then the same lesson live in Chicago's 77 community areas
3. The dispersion machinery, one step at a time: raw values → deviations → squared deviations → variance → SD
4. The empirical rule (68–95–99.7), checked against real daily crime counts
5. The Poisson distribution — and proof that real robbery counts actually follow it

## Setup

Everything below needs the tidyverse, plus **modeest** for the mode (`install.packages("modeest")` once — then `library()` every session):

```{r}
#| label: setup
library(tidyverse)
library(modeest)
set.seed(705)
crimes <- read_csv("https://smourtgos.github.io/crju705/data/chicago-crimes-2025.csv",
                   show_col_types = FALSE)
areas  <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv",
                   show_col_types = FALSE)
```

## 1 · Three ways to say "typical"

Seven days of rainfall. The **mean** is just "add them up, divide by how many" — do it the hard way once so you know exactly what `mean()` does:

```{r}
#| label: mean-hard-easy
rainfall <- c(0, 0, 0, 2, 1, 1, 4)   # rain each day this week

# The hard way:
sum(rainfall) / length(rainfall)

# The easy way:
mean(rainfall)
```

The **median** is the middle value of the *sorted* data, and the **mode** is the most frequent value. Base R has no mode function (`mode()` does something else entirely!), so we use `mfv()` — "most frequent value" — from modeest:

```{r}
#| label: median-mode
sort(rainfall)   # seven values, so the 4th one is the middle
median(rainfall)
mfv(rainfall)
```

Three answers — mean ≈ 1.14, median 1, mode 0 — from the same seven numbers. When could they disagree by enough to matter? Glad you asked.

## 2 · When "the average" misleads: sentencing

Ten people convicted of robbery, and one judge in a bad mood:

```{r}
#| label: sentences
sentences <- c(2, 2, 3, 3, 3, 4, 4, 5, 5, 30)   # note the outlier!

mean(sentences)
median(sentences)
mfv(sentences)
```

One 30-year sentence drags the mean to **6.1 years** — yet *nobody* in the data actually received a sentence near 6 years. The median (3.5) and mode (3) describe the group far better:

```{r}
#| label: sentences-plot
#| fig-height: 4.5
sent_stats <- tibble(
  measure = c("Mean", "Median", "Mode"),
  value = c(mean(sentences), median(sentences), mfv(sentences))
)

ggplot(tibble(sentences), aes(sentences)) +
  geom_histogram(binwidth = 1, fill = "grey85", color = "black") +
  geom_vline(data = sent_stats, aes(xintercept = value, color = measure),
             linewidth = 1.2, linetype = "dashed") +
  labs(title = "Sentence Lengths (in Years)",
       x = "Years in Prison", y = "Number of People", color = NULL)
```

Worth sitting with: which of the three would be most honest in a press release about sentencing policy — and which would an advocate quote?

### The same thing in live data

Crime rates per 1,000 residents across Chicago's 77 community areas, 2025:

```{r}
#| label: chicago-skew
areas |>
  summarize(
    mean_rate   = mean(crime_rate),
    median_rate = median(crime_rate),
    max_rate    = max(crime_rate)
  )
```

Mean about 97, median 82 — the mean sits well above the median, the signature of a **right-skewed** distribution. Who's dragging it up?

```{r}
#| label: chicago-worst
areas |> slice_max(crime_rate, n = 3) |> select(community, crime_rate)
```

Fuller Park tops the list at roughly 292 crimes per 1,000 residents — partly because only about 2,300 people live there, so even a modest crime count divided by a tiny population produces a huge *rate*. Same lesson as the 30-year sentence, in live data. (Want to see it? Reuse the sentencing-histogram pattern on `areas$crime_rate` — the mean line lands visibly right of the median.)

## 3 · Dispersion: the machinery, step by step

Two three-candidate elections, identical mean support, completely different situations:

```{r}
#| label: votes-setup
votes1 <- c(200, 300, 400)   # wide spread
votes2 <- c(299, 300, 301)   # tight cluster

mean(votes1); mean(votes2)       # identical means...
diff(range(votes1)); diff(range(votes2))   # ...very different ranges
```

**Step 1 — deviations from the mean.** The range only uses two values; the real machinery asks about every value. How far is each one from 300?

```{r}
#| label: votes-dev
votes1 - mean(votes1)
votes2 - mean(votes2)
```

Why not just average these? Because they *always* cancel:

```{r}
#| label: votes-sum-dev
sum(votes1 - mean(votes1))   # the sum of deviations is always zero
```

**Step 2 — square them.** Squaring makes every deviation positive (and makes big misses count extra):

```{r}
#| label: votes-ss
sum((votes1 - mean(votes1))^2)
sum((votes2 - mean(votes2))^2)
```

20,000 versus 2 — *now* the two elections look as different as they are.

**Step 3 — average, then un-square.** The mean squared deviation is the **variance**; its square root is the **standard deviation**, which puts you back in vote units:

```{r}
#| label: votes-var-hand
sum((votes1 - mean(votes1))^2) / length(votes1)          # variance (population formula)
sqrt(sum((votes1 - mean(votes1))^2) / length(votes1))    # SD — back to vote units
```

```{r}
#| label: votes-var-fn
var(votes1); sd(votes1)
var(votes2); sd(votes2)
```

R's built-ins give slightly different answers *on purpose*: `var()` and `sd()` divide by **n − 1**, not n (10,000 vs. our hand-computed 6,666.7). *Why* is Session 4's simulation payoff — for now, file the IOU.

### Seeing the machinery

The same three steps, drawn. First build one tidy table holding both scenarios plus their deviations and squared deviations:

```{r}
#| label: votes-df
votes_df <- tibble(
  scenario = rep(c("Wide Spread", "Tight Cluster"), each = 3),
  candidate = factor(rep(1:3, times = 2)),
  votes = c(votes1, votes2)
) |>
  mutate(dev = votes - mean(votes), sq_dev = dev^2, .by = scenario)
votes_df
```

Raw values — same dashed mean line, very different stories:

```{r}
#| label: machinery-raw
#| fig-height: 4
ggplot(votes_df, aes(candidate, votes)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 300, color = "firebrick", linetype = "dashed", linewidth = 1) +
  facet_wrap(~scenario) +
  labs(title = "Same mean (dashed line) — very different stories",
       x = "Candidate", y = "Votes")
```

Deviations — notice the Tight Cluster bars are barely visible at this scale:

```{r}
#| label: machinery-dev
#| fig-height: 4
ggplot(votes_df, aes(candidate, dev)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 0, color = "grey40") +
  facet_wrap(~scenario) +
  labs(title = "How far is each value from the mean?",
       x = "Candidate", y = "Deviation from mean")
```

Squared deviations — all positive, and the big misses dominate:

```{r}
#| label: machinery-sq
#| fig-height: 4
ggplot(votes_df, aes(candidate, sq_dev)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~scenario) +
  labs(title = "Squaring: all positive, and big misses dominate",
       x = "Candidate", y = "Squared deviation")
```

Average those bars and you have the variance; take the square root and you're back in the original units — the SD. In words: **"on average, how far is each value from the mean?"**

## 4 · Distributions: the normal and the empirical rule

A **distribution** is the shape of the data — how many observations fall at each value. The most famous shape is the normal, defined by just two numbers (mean and SD). Simulate one and look:

```{r}
#| label: normal-hist
#| fig-height: 4
sims <- tibble(x = rnorm(n = 1000, mean = 100, sd = 10))
ggplot(sims, aes(x)) + geom_histogram(bins = 30, fill = "grey80", color = "white")
```

For normal data, the SD comes with a promise — the **empirical rule**: about 68% of values within ±1 SD of the mean, 95% within ±2, 99.7% within ±3. The exact theoretical values:

```{r}
#| label: pct-within
pct_within <- function(k) 100 * (pnorm(k) - pnorm(-k))
tibble(k = 1:3, pct_within = round(pct_within(1:3), 2))
```

### The empirical rule, on real data

Does the promise hold outside a textbook? Count total crimes per day in the real Chicago data (365 days), then check what share of days land within 1, 2, and 3 SDs of the mean:

```{r}
#| label: empirical-rule-real
daily_crimes <- crimes |> count(date)

m <- mean(daily_crimes$n)
s <- sd(daily_crimes$n)
c(mean = m, sd = s)

daily_crimes |>
  summarize(
    within_1_sd = mean(n > m - 1*s & n < m + 1*s),
    within_2_sd = mean(n > m - 2*s & n < m + 2*s),
    within_3_sd = mean(n > m - 3*s & n < m + 3*s)
  )
```

About 68% / 96% / 99% — on a variable nobody simulated, the rule practically nails it. Daily *totals* of many small events pile up into a roughly bell-shaped heap (a preview of Session 4's central limit theorem), and once a variable is roughly normal, its mean and SD tell you almost everything: about 82 crimes on a typical day, and only about one day in twenty lands outside the 57–107 band. (Histogram `daily_crimes$n` yourself and you'll see the bell.)

## 5 · The Poisson — and proof it's not hypothetical

Counts of *rare* events (crimes per day, 911 calls per hour) follow a different, often skewed shape: the **Poisson**, with one parameter, λ = the expected number of events. Simulate a rare event (λ = 1):

```{r}
#| label: poisson-hist
#| fig-height: 4
pois_sims <- tibble(x = rpois(n = 1000, lambda = 1))
ggplot(pois_sims, aes(x)) + geom_bar(fill = "navy")
```

Zero and one dominate, with a right tail — nothing like a bell. Now the real thing — robberies per day in our Chicago sample:

```{r}
#| label: chicago-poisson
#| fig-height: 4.5
daily_robberies <- crimes |>
  filter(primary_type == "ROBBERY") |>
  count(date)

ggplot(daily_robberies, aes(n)) +
  geom_bar(fill = "firebrick") +
  labs(x = "Robberies per day", y = "Number of days",
       title = "Real crime counts look Poisson")
```

Most days see one or two robberies, a handful see six or more — that skewed staircase is the Poisson shape, live in real data. (One honest footnote: `count(date)` only lists days with *at least one* robbery, so the zero-robbery days are missing from this plot — see the last variation prompt.)

Recognizing shape is the point: bell-shaped variables and count variables call for different tools, and matching tool to shape is a theme for the whole semester.

## Try a variation

1. In the rainfall data, change one day to `10` inches. Which of the three "typical" measures move, and which barely budge? *(This is why we say the median is robust.)*
2. In the votes demo, change `400` to `1000` and re-run the sum-of-squares chunk. Squaring emphasizes big misses — by how much did that one change inflate the total?
3. Run the `summarize()` pattern on `areas$violent_rate`: mean, SD, min, max. The SD comes out roughly three-quarters the size of the mean — what does that say about how *unequally* violence is distributed across Chicago's communities?
4. The robbery plot is missing its zero days: 2025 has 365 days, but `nrow(daily_robberies)` is smaller. How many zero-robbery days went missing, and where would that bar sit? Then swap in `primary_type == "THEFT"` — with a much higher λ, does the shape still look skewed?
