---
title: "Demo Walkthrough · Session 6"
subtitle: "Hypothesis testing, twice — t.test(), ttestBF(), and the collision between them"
---

::: {.callout-note appearance="simple"}
**What this page is.** The Session 6 lecture demos, written down step by step. Every chunk below is self-contained — copy any of them into your own R session and they run as-is. Use this page to catch up after a missed class, or to re-run what happened on screen at your own pace.
:::

## What you'll build

1. A bootstrap of the violent-vs-property time gap — 10,000 draws — and the same machinery run after *erasing* the gap
2. The NHST verdict from `t.test()`: p ≈ .009, "statistically significant"
3. The Bayesian verdict from `ttestBF()` on the *same data*: mild evidence **for** the null — and why both are right
4. The posterior distribution for the difference, a credible interval, and the sentence Session 5 forbade
5. The recidivism trial: p = 0.20 vs. a ~91% posterior probability the program works

## Setup

The tidyverse plus **BayesFactor**, and the two groups from Session 5:

```{r}
#| label: setup
library(tidyverse)
library(BayesFactor)   # install.packages("BayesFactor")  — once, if you don't have it
set.seed(705)

crimes <- read_csv("https://smourtgos.github.io/crju705/data/chicago-crimes-2025.csv",
                   show_col_types = FALSE)

vp <- crimes |> filter(crime_category %in% c("Violent", "Property"))
violent_hours  <- vp$hour[vp$crime_category == "Violent"]
property_hours <- vp$hour[vp$crime_category == "Property"]
```

## 1 · The question, again

Session 5 asked whether violent crimes happen later in the day than property crimes:

```{r}
#| label: recap-means
vp |> summarize(mean_hour = mean(hour), sd_hour = sd(hour),
                n = n(), .by = crime_category)
```

Violent incidents average about **15 minutes later**. Session 5 gave us an interval for that gap. Today we get the *test* — a yes/no verdict machine — and then a second tradition that answers the question you probably *meant* to ask.

## 2 · Pretend we sampled repeatedly

Bootstrap each group's mean 10,000 times — Session 4's logic, applied to both groups at once:

```{r}
#| label: boot-means
#| fig-height: 4.5
boot <- tibble(
  Violent  = replicate(10000, mean(sample(violent_hours,  replace = TRUE))),
  Property = replicate(10000, mean(sample(property_hours, replace = TRUE)))
)

boot |>
  pivot_longer(everything(), names_to = "group", values_to = "boot_mean") |>
  ggplot(aes(boot_mean, fill = group)) +
  geom_histogram(bins = 60, alpha = 0.6, position = "identity", color = NA) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(title = "10,000 bootstrap means per group",
       x = "Mean hour of day", y = "Count", fill = NULL)
```

Subtract, pair by pair, and the *difference* gets its own distribution:

```{r}
#| label: boot-diff
#| fig-height: 4.5
boot_diff <- boot$Violent - boot$Property

quantile(boot_diff, c(0.025, 0.975))

ci_boot <- quantile(boot_diff, c(0.025, 0.975))
ggplot(tibble(boot_diff), aes(boot_diff)) +
  geom_histogram(bins = 60, fill = "navy", color = "white") +
  geom_vline(xintercept = 0, color = "firebrick", linewidth = 1.2) +
  geom_vline(xintercept = ci_boot, color = "goldenrod", linetype = 2, linewidth = 1) +
  labs(title = "Bootstrap distribution of the difference (Violent − Property)",
       subtitle = "Red line = zero (no difference); gold dashes = 95% interval",
       x = "Difference in mean hour", y = "Count")
```

The middle 95% of plausible differences runs from about **0.06 to 0.44 hours** — zero is not among the plausible values. The data lean toward a real (if small) gap.

### What if the interval crossed zero?

Erase the real gap on purpose — shift every violent incident earlier by the observed difference — and re-run the exact same machinery:

```{r}
#| label: boot-null
violent_shifted <- violent_hours - (mean(violent_hours) - mean(property_hours))

boot_diff_null <- replicate(10000,
  mean(sample(violent_shifted, replace = TRUE)) -
  mean(sample(property_hours,  replace = TRUE)))

quantile(boot_diff_null, c(0.025, 0.975))
```

Now the interval runs from about **−0.19 to +0.19** — zero sits comfortably inside, and the data are *consistent with no difference*. Formalize this intuition and you have invented the NHST.

## 3 · The Null Hypothesis Significance Test, in five steps

1. **State the null**, $H_0$: no difference. The boring world.
2. **State the alternative**, $H_1$: *some* difference — note how little we commit to.
3. **Compute a test statistic**: $t = \text{observed difference} / SE$ — the gap, in SE units.
4. **Ask:** how often would $|t|$ this large appear by luck in the boring world? That's the **p-value**.
5. **Verdict:** $p < \alpha$ (usually .05) → "reject $H_0$." Otherwise → "fail to reject."

Notice the machine only weighs evidence **against the null** — $H_1$ never gets scored. And be precise about the p-value. It **is** the probability of data *this extreme or more, if the null is true and the model is right*. It is **not** the probability the null is true (that's $P(H_0 \mid \text{data})$, which frequentism never computes), **not** the probability the result "is due to chance," and **not** a measure of how *big* the effect is — with enough data, microscopic effects earn microscopic p-values. Hold that thought.

```{r}
#| label: ttest-real
t.test(hour ~ crime_category, data = vp)
```

Reading the verdict: R orders groups alphabetically, so this tests **Property − Violent** (hence the negative signs — same finding as our bootstrap, mirrored). $t = -2.61$, $p = .009$. Under .05 — **"statistically significant."** Reject the null. Publish the press release?

One caution before we do: everything above was scored against *one* curve — the null — and that tidy t-distribution itself assumes normal-enough data, independent observations, and well-behaved variances. NHST never asks what the alternative world looks like, and researchers rarely test the assumptions before trusting the p-value. (The lecture's sketch figures make this vivid — see the [Session 6 slides](../slides/week-06-hypothesis-testing.qmd).)

## 4 · The Bayesian t-test: `ttestBF()`

Session 3's Bayes' theorem, now pointed at hypotheses: the p-value lives entirely in the **likelihood**, $P(\text{data} \mid H_0)$, and never becomes $P(H_0 \mid \text{data})$. Bayesian testing finishes the flip — at the price of a stated prior. Same question, second tradition. We supply the groups ourselves, so the direction is ours: **x − y = Violent − Property**.

```{r}
#| label: bf-run
bf <- ttestBF(x = violent_hours, y = property_hours)
bf
```

```{r}
#| label: bf-extract
extractBF(bf)$bf        # BF10: evidence for a difference vs. no difference
1 / extractBF(bf)$bf    # BF01: the same evidence, read the other way
```

A Bayes factor is a **ratio of predictions**: how well did "some difference" predict these data versus "no difference"? Here $BF_{10} \approx 0.49$ — *below* 1. The data are about **2 times more consistent with the null**.

Rules of thumb for reading a BF: **1–3** barely worth mentioning · **3–10** moderate · **10–30** strong · **30–100** very strong · **100+** extreme — with the mirror image below 1 (so a BF₁₀ of 0.49, i.e. BF₀₁ ≈ 2, is "barely worth mentioning" evidence *for* the null).

### Hold on — the p-value said "significant"

| | Frequentist | Bayesian |
|---|---|---|
| Result | $p = .009$ → reject $H_0$ | $BF_{10} \approx 0.49$ → mild evidence *for* $H_0$ |

Both machines ran on the same 19,500 cases. With that much data, the t-test can detect *microscopic* gaps — and 15 minutes is microscopic. The BF asks a harder question: did "a difference of plausible size" predict the data better than "exactly zero"? This close to zero, **the null predicted slightly better**. Neither is wrong. They answer **different questions** — and only one is impressed by sheer sample size.

## 5 · The posterior for the difference

The Bayes factor compares models. The **posterior** describes the difference itself, drawn by a sampler. (The machinery behind that sampler — MCMC — is in the optional appendix of the [Session 6 slides](../slides/week-06-hypothesis-testing.qmd), if you want to see how Bayes became computable.)

```{r}
#| label: bf-posterior
post <- suppressMessages(posterior(bf, iterations = 10000))
diff_post <- as.numeric(post[, "beta (x - y)"])

quantile(diff_post, c(0.025, 0.975))
mean(diff_post > 0)
```

```{r}
#| label: posterior-plot
#| fig-height: 4.5
cred <- quantile(diff_post, c(0.025, 0.975))

ggplot(tibble(diff_post), aes(diff_post)) +
  geom_histogram(bins = 60, fill = "darkgreen", color = "white") +
  geom_vline(xintercept = 0, color = "firebrick", linewidth = 1.2) +
  geom_vline(xintercept = cred, color = "goldenrod", linetype = 2, linewidth = 1) +
  labs(title = "Posterior distribution of the difference (Violent − Property, hours)",
       subtitle = paste0("95% credible interval: ", round(cred[1], 2), " to ",
                         round(cred[2], 2), " hours"),
       x = "Difference in mean hour", y = "Posterior samples")
```

Now say the sentence Session 5 forbade — **legally**: *"There is a 95% probability the true difference lies between about 4 and 26 minutes."* Under the Bayesian license (given the model and prior), that is exactly what this **credible interval** means. And about 99.5% of the posterior sits above zero — the difference is almost certainly *real*. Real, but…

## 6 · Detectable, probable — and tiny

One more number every report owes its reader: a standardized **effect size**, Cohen's *d*:

```{r}
#| label: cohens-d
sd_pooled <- sqrt((var(violent_hours) + var(property_hours)) / 2)
(mean(violent_hours) - mean(property_hours)) / sd_pooled
```

Benchmarks: 0.2 "small" · 0.5 "medium" · 0.8 "large." We measured **~0.04** — a ~15-minute shift against SDs of nearly 7 *hours*. Would any commander re-draw a patrol schedule over this? The full picture: the difference is *probably real* (posterior), *utterly trivial* (d), and "significant" (p) only because n is enormous. **Report all three; never let the star do the talking.**

## 7 · A cleaner collision: the recidivism trial

The setup: typical recidivism runs about **60%** — that's our prior belief. A randomized trial of a reentry program enrolls **100 treated, 100 control**. Outcomes: control **50/100** reoffend, treatment **40/100**. Ten percentage points — exactly the kind of effect a city would fund. Both traditions, go.

**Frequentist:**

```{r}
#| label: recid-freq
prop.test(x = c(40, 50), n = c(100, 100))
```

$p = 0.20$ — **not statistically significant**. Honest reading: *a 10-point gap like this could plausibly occur even if the program did nothing.* The grant report writes itself — "no significant effect" — and the program probably dies there.

**Bayesian:** prior Beta(6, 4) — mean 0.6, worth about 10 observations. Update each group with its data, then ask the question we care about directly:

```{r}
#| label: recid-bayes
set.seed(705)   # re-seeded so these exact draws reproduce on their own
theta_treat   <- rbeta(10000, 6 + 40, 4 + 60)   # posterior, treatment
theta_control <- rbeta(10000, 6 + 50, 4 + 50)   # posterior, control

mean(theta_treat < theta_control)   # P(program reduces recidivism)
```

```{r}
#| label: recid-plot
#| fig-height: 4.5
tibble(theta = c(theta_treat, theta_control),
       group = rep(c("Treatment", "Control"), each = 10000)) |>
  ggplot(aes(theta, fill = group)) +
  geom_density(alpha = 0.5, color = NA) +
  scale_fill_manual(values = c("gray50", "firebrick")) +
  labs(title = "Posterior beliefs about each group's recidivism rate",
       subtitle = "Overlap is real — and still, treatment sits lower in ~91% of draws",
       x = "Recidivism probability (θ)", y = "Posterior density", fill = NULL)
```

Given the prior and the data, there is about a **91% probability the program reduces recidivism**. Same data, different questions:

| | Frequentist | Bayesian |
|---|---|---|
| Verdict | $p = 0.20$: cannot reject "no effect" | $P(\text{treatment} < \text{control}) \approx 0.91$ |
| Plain English | "This gap could plausibly be luck" | "9-to-1 odds the program works" |

Both statements are **true of the same data**. Neither is a trick. A policymaker deciding whether to fund the program next year: which sentence actually helps?

## 8 · Sample size changes everything

The lecture closed with two animations you should watch in the [Session 6 slides](../slides/week-06-hypothesis-testing.qmd): fix a *trivial* true effect and just keep collecting data, and *p* still dives below .05 — **given enough data, everything is "significant."** The same growing dataset through Bayesian glasses doesn't flip a verdict; it **sharpens the estimate**, which visibly converges on "almost nothing." The tininess stays in view. That's Act III's collision, run in slow motion — and it's why, whatever tradition you use, you **report the effect size, in units a human cares about**.

## Try a variation

1. Run both tests on a new question: does `hour` differ for **domestic vs. non-domestic** incidents? (`t.test(hour ~ domestic, data = crimes)` and `ttestBF(formula = hour ~ domestic, data = mutate(crimes, domestic = factor(domestic)))`.) Do the two traditions agree this time?
2. In the recidivism trial, flatten the prior to `Beta(1, 1)` (no prior information). How does P(program works) move? *(This is the prior's weight, measured in one edit.)*
3. Make the trial's evidence stronger — treatment 35/100 instead of 40/100 — and re-run both. Where does *p* land now, and where does the posterior probability?
4. Rerun the bootstrap with `replicate(100, …)` instead of 10,000. How stable are the interval endpoints across reruns? *(Monte Carlo error is a sample-size problem too.)*
