---
title: "Demo Walkthrough · Session 7"
subtitle: "Four real criminal justice questions, one recipe: count, test, say it"
---

::: {.callout-note appearance="simple"}
**What this page is.** The Session 7 lecture, written down step by step. Every chunk below is self-contained: copy any of them into your own R session and it runs as-is. Use this page to catch up after a missed class, or to re-run the examples at your own pace.
:::

## What you'll build

Four answers to four real questions, each with the same recipe:

1. **Question.** Name the two groups and the outcome. A yes/no outcome means `prop.test()` and `bayes_prop_test()`; a number means `t.test()`.
2. **Count or describe.**
3. **Test.**
4. **Say it.** The honest sentence, how big in real units, and one thing the test cannot tell you.

The questions: did financial aid at release lower re-arrest? Were employed arrestees released more often? How much lower is income in Chicago's high-violence areas? Was body camera footage reported more often in fatal police shootings after 2020?

Each example also has the **animation** from class. Press **Play** under each one, or drag the slider to step through it at your own pace. The code that draws them is folded away (click "Code" to see it). You do not need to write code like that; it is there so nothing is hidden.

## Setup

The tidyverse, the course's Bayesian function, and four datasets. Two of them are new to you and live on the course site; one is our Chicago anchor data; one comes straight from the *Washington Post*. `plotly` draws the animations; if you want to run those chunks yourself, install it once with `install.packages("plotly")`.

```{r}
#| label: setup
library(tidyverse)
library(plotly)   # only for the animations
source("https://smourtgos.github.io/crju705/R/bayes.R")   # gives you bayes_prop_test()
set.seed(705)

rossi   <- read_csv("https://smourtgos.github.io/crju705/data/rossi.csv",
                    show_col_types = FALSE)
arrests <- read_csv("https://smourtgos.github.io/crju705/data/toronto-arrests.csv",
                    show_col_types = FALSE)
areas   <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv",
                    show_col_types = FALSE)
shootings <- read_csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/v2/fatal-police-shootings-data.csv",
                      show_col_types = FALSE)
```

More about each dataset, and four more like them, is on the [Final-Project Data Options](../project-data.qmd) page.

## 1 · Reentry: does money at release keep people out of jail?

432 men released from Maryland state prisons in the 1970s were followed for one year. Half were **randomly assigned** to receive financial aid (`fin`). The outcome is `arrest`: 1 if arrested again within the year. One row is one man.

```{r}
#| label: rossi-count
rossi |> count(fin, arrest)
```

Aid: 48 re-arrested out of 48 + 168 = **216** (22.2%). No aid: 66 out of 66 + 150 = **216** (30.6%). Aid goes first, so every difference reads **Aid minus No aid**.

```{r}
#| label: rossi-prop
prop.test(x = c(48, 66), n = c(216, 216))
```

The interval runs from about 17 points lower to 0.4 points higher, and **p = 0.063**. Zero is just inside, so: fail to reject the null. A funder who reads only that line cuts the program.

**What is 0.063 measuring? Watch it.** Pretend aid did nothing: everyone has the same 26% chance of re-arrest. Run the study again and again, and pile up the gaps that luck alone produces:

```{r}
#| label: rossi-anim
#| code-fold: true
set.seed(705)
pooled <- (48 + 66) / 432
null_gaps <- tibble(gap = 100 * (rbinom(10000, 216, pooled) - rbinom(10000, 216, pooled)) / 216)
step_size <- 100 / 216
steps <- c(20, 100, 500, 2000, 10000)
rossi_anim <- map_dfr(steps, \(k) {
  first_k <- slice_head(null_gaps, n = k)
  first_k |>
    mutate(bin = round(gap / step_size)) |>
    count(bin) |>
    complete(bin = -45:45, fill = list(n = 0)) |>
    mutate(share = 100 * n / k, bin = bin * step_size,
           as_big = if_else(abs(bin) >= 8.3, "At least as big as ours", "Smaller than ours"),
           label = paste(format(k, big.mark = ","), "pretend studies"),
           pct_big = 100 * mean(abs(first_k$gap) >= 8.3))
}) |>
  mutate(label = factor(label, levels = unique(label)))
rossi_text <- rossi_anim |>
  distinct(label, pct_big) |>
  mutate(x = 13.5, y = 7.2,
         txt = paste0(sprintf("%.1f", pct_big), "% had a gap<br>at least as big as ours"))

plot_ly(height = 470) |>
  add_bars(data = rossi_anim, x = ~bin, y = ~share, frame = ~label, color = ~as_big,
           colors = c("firebrick", "gray75"), hoverinfo = "none") |>
  add_text(data = rossi_text, x = ~x, y = ~y, text = ~txt, frame = ~label,
           textfont = list(size = 16, color = "firebrick"), showlegend = FALSE, hoverinfo = "none") |>
  layout(barmode = "overlay",
         xaxis = list(title = "Gap in re-arrest rate, aid minus no aid (points)", range = c(-20.5, 20.5)),
         yaxis = list(title = "Share of pretend studies (%)", range = c(0, 8)),
         shapes = list(
           list(type = "line", x0 = -8.3, x1 = -8.3, y0 = 0, y1 = 8, line = list(color = "navy", width = 3)),
           list(type = "line", x0 = 8.3, x1 = 8.3, y0 = 0, y1 = 8, line = list(color = "navy", width = 3, dash = "dash"))),
         annotations = list(list(x = -8.6, y = 7.8, text = "Our study: -8.3", showarrow = FALSE,
                                 xanchor = "right", font = list(size = 14, color = "navy"))),
         legend = list(orientation = "h", y = 1.08)) |>
  animation_opts(frame = 1500, transition = 700, redraw = FALSE) |>
  animation_slider(currentvalue = list(prefix = "")) |>
  config(displayModeBar = FALSE)
```

In a world where aid did nothing, about **5 studies in 100** still show a gap of 8.3 points or more, in one direction or the other. That is what p = 0.063 is measuring (`prop.test()` adds a small correction, so its number is a little bigger). Not rare enough to pass the 0.05 bar. Now ask the other question:

```{r}
#| label: rossi-bayes
bayes_prop_test(x = c(48, 66), n = c(216, 216))
```

**Say it.** Given these data, there is about a **97% probability** that men who got aid were re-arrested at a lower rate. About 22 of every 100 men with aid were re-arrested, against about 31 of every 100 without: roughly 8 fewer re-arrests per 100 people released. What this cannot tell you: whether the effect lasts past a year, or would hold today. "Not significant" is not "no effect."

## 2 · Pretrial release: does a job get you released?

5,226 Toronto arrests for possessing small amounts of marijuana, 1997 to 2002. `released` is whether the person was released with a summons instead of held for a bail hearing. The group is `employed`. One row is one arrest.

```{r}
#| label: tor-count
arrests |> count(employed, released)
```

Employed: 3,568 released out of 3,568 + 543 = **4,111** (86.8%). Not employed: 766 out of 766 + 349 = **1,115** (68.7%).

```{r}
#| label: tor-prop
prop.test(x = c(3568, 766), n = c(4111, 1115))
```

```{r}
#| label: tor-bayes
bayes_prop_test(x = c(3568, 766), n = c(4111, 1115))
```

About 15 to 21 points more often for the employed, by either test, and more than a 99% probability the employed rate is higher. But the two groups differ in other ways. `checks` counts how many of six police databases (prior arrests, convictions, parole, and so on) had the person's name:

```{r}
#| label: tor-checks
arrests |> summarize(avg_checks = mean(checks), people = n(), .by = employed)
```

People without a job had more of a record, and a record is a reason to hold someone. **Watch the gap shrink** as we keep only people with less and less of a record:

```{r}
#| label: tor-anim
#| code-fold: true
cutoffs <- tibble(k = c(6, 3, 2, 1, 0),
                  label = c("Everyone", "In 3 or fewer databases", "In 2 or fewer",
                            "In 1 or fewer", "No record at all"))
tor_anim <- cutoffs |>
  mutate(d = map(k, \(kk) arrests |> filter(checks <= kk) |>
                   summarize(rate = 100 * mean(released == "Yes"), people = n(), .by = employed))) |>
  unnest(d) |>
  mutate(employed = if_else(employed == "Yes", "Employed", "Not employed"),
         xpos = if_else(employed == "Employed", 1, 2),
         label = factor(label, levels = cutoffs$label),
         txt = paste0(round(rate, 1), "%<br>(", format(people, big.mark = ",", trim = TRUE), " people)"))
tor_gap <- tor_anim |>
  summarize(gap = rate[employed == "Employed"] - rate[employed == "Not employed"], .by = label) |>
  mutate(x = 1.5, y = 104, txt = paste0("Gap: ", round(gap, 1), " points"))

plot_ly(height = 470) |>
  add_bars(data = tor_anim, x = ~xpos, y = ~rate, frame = ~label, color = ~employed,
           colors = c("firebrick", "steelblue"), text = ~txt, textposition = "inside",
           insidetextfont = list(size = 15, color = "white"), hoverinfo = "none", showlegend = FALSE) |>
  add_text(data = tor_gap, x = ~x, y = ~y, text = ~txt, frame = ~label,
           textfont = list(size = 18, color = "navy"), showlegend = FALSE, hoverinfo = "none") |>
  layout(xaxis = list(title = "", tickvals = c(1, 2), ticktext = c("Employed", "Not employed"),
                      range = c(0.4, 2.6)),
         yaxis = list(title = "Released with a summons (%)", range = c(0, 110))) |>
  animation_opts(frame = 1600, transition = 800, redraw = FALSE) |>
  animation_slider(currentvalue = list(prefix = "Who is included: ")) |>
  config(displayModeBar = FALSE)
```

Now do the last frame properly: keep only people with **no record at all**, and ask again.

```{r}
#| label: tor-clean
arrests |> filter(checks == 0) |> count(employed, released)
prop.test(x = c(1526, 166), n = c(1650, 201))
bayes_prop_test(x = c(1526, 166), n = c(1650, 201))
```

**Say it.** Among people with clean records, the employed were released about 10 points more often (92.5% vs 82.6%; interval about 4 to 16 points, credible interval 5.1 to 15.8 points, more than a 99% probability the employed rate is higher). The gap shrank from 18 points to 10 when we compared like with like, but it did not disappear. What this cannot tell you: whether some other difference (age, citizenship, where the arrest happened) explains the rest.

## 3 · Neighborhoods: violence and income

Chicago's 77 community areas. `high_violence` marks the 38 areas whose violent-crime rate is above the city's median. The outcome is `median_income`, a number in dollars, so the test is `t.test()`. One row is one **area**, not one person. First, name the groups and put High violence first:

```{r}
#| label: chi-groups
areas <- areas |>
  mutate(violence = factor(if_else(high_violence, "High violence", "Other areas"),
                           levels = c("High violence", "Other areas")))

areas |>
  summarize(avg_income = mean(median_income), areas = n(), .by = violence) |>
  arrange(violence)
```

**See it.** Each dot is one community area:

```{r}
#| label: chi-anim
#| code-fold: true
set.seed(705)
jit <- runif(nrow(areas), -0.18, 0.18)
stage_labels <- c("All 77 areas", "Split by violence", "The two averages")
chi_pts <- map_dfr(1:3, \(s) areas |>
  mutate(x = if (s == 1) 1.5 + jit else if_else(violence == "High violence", 1, 2) + jit,
         stage = stage_labels[s])) |>
  mutate(stage = factor(stage, levels = stage_labels),
         hover = paste0(community, "<br>$", format(round(median_income), big.mark = ",")))
chi_means <- areas |>
  summarize(avg = mean(median_income), .by = violence)
chi_means <- map_dfr(1:3, \(s) chi_means |>
  mutate(x = if (s < 3) -1 else if_else(violence == "High violence", 1, 2),
         stage = stage_labels[s])) |>
  mutate(stage = factor(stage, levels = stage_labels),
         txt = paste0("Average<br>$", format(round(avg), big.mark = ",")))
chi_gap <- tibble(stage = factor(stage_labels, levels = stage_labels),
                  x = c(-1, -1, 1.5), y = 175000, txt = "Gap: about $35,700")

plot_ly(height = 470) |>
  add_markers(data = chi_pts, x = ~x, y = ~median_income, frame = ~stage, ids = ~community,
              color = ~violence, colors = c("firebrick", "steelblue"),
              marker = list(size = 10, opacity = 0.8), text = ~hover, hoverinfo = "text") |>
  add_markers(data = chi_means, x = ~x, y = ~avg, frame = ~stage, ids = ~violence,
              marker = list(symbol = "line-ew-open", size = 60, color = "navy",
                            line = list(width = 4, color = "navy")),
              showlegend = FALSE, hoverinfo = "none") |>
  add_text(data = chi_means, x = ~x + 0.33, y = ~avg, text = ~txt, frame = ~stage,
           textfont = list(size = 14, color = "navy"), showlegend = FALSE, hoverinfo = "none") |>
  add_text(data = chi_gap, x = ~x, y = ~y, text = ~txt, frame = ~stage,
           textfont = list(size = 18, color = "navy"), showlegend = FALSE, hoverinfo = "none") |>
  layout(xaxis = list(title = "", range = c(0.4, 2.6), tickvals = c(1, 2),
                      ticktext = c("High violence", "Other areas")),
         yaxis = list(title = "Median household income", tickprefix = "$", range = c(0, 190000)),
         legend = list(orientation = "h", y = 1.08)) |>
  animation_opts(frame = 1800, transition = 900, redraw = FALSE) |>
  animation_slider(currentvalue = list(prefix = "")) |>
  config(displayModeBar = FALSE)
```

```{r}
#| label: chi-ttest
t.test(median_income ~ violence, data = areas)
```

**Say it.** High-violence areas minus other areas: about **$25,000 to $47,000 lower** median household income ($54,355 vs $90,088 on average), p = 0.00000001. The unit is dollars because the column is dollars. What this cannot tell you: which way it runs (does poverty breed violence, or does violence drive income out?), and anything about individual people. Each row is an area.

## 4 · Use of force: body cameras over time

The *Washington Post*'s database of every person shot and killed by on-duty police in the U.S., 2015 to 2024. `body_camera` is TRUE when news or police reports said a camera recorded the incident. One row is one person killed. Group: before and after 2020.

```{r}
#| label: wapo-period
shootings <- shootings |>
  mutate(period = if_else(year(date) >= 2020, "2020 to 2024", "2015 to 2019"))

shootings |> count(period, body_camera)
```

2020 to 2024: 1,205 of 1,205 + 4,302 = **5,507** (21.9%). 2015 to 2019: 594 of 594 + 4,329 = **4,923** (12.1%).

**Year by year** first, because two five-year averages can hide a lot:

```{r}
#| label: wapo-anim
#| code-fold: true
yearly <- shootings |>
  mutate(year = year(date)) |>
  summarize(pct = 100 * mean(body_camera), .by = c(year, period)) |>
  arrange(year)
yr_labels <- c(as.character(2015:2024), "The two periods")
bwc_anim <- map_dfr(seq_along(yr_labels), \(s) yearly |>
  mutate(shown = if_else(s <= 10 & year > 2014 + s, 0, pct), step = yr_labels[s])) |>
  mutate(step = factor(step, levels = yr_labels),
         txt = if_else(shown > 0, paste0(round(shown), "%"), ""))
per <- shootings |> summarize(pct = 100 * mean(body_camera), .by = period)
bwc_lines <- map_dfr(seq_along(yr_labels), \(s) per |>
  mutate(x = if (s < 11) 2010 else if_else(period == "2015 to 2019", 2017, 2022),
         step = yr_labels[s])) |>
  mutate(step = factor(step, levels = yr_labels),
         ytxt = if_else(period == "2015 to 2019", 19.5, 35),
         txt = paste0(period, ": ", round(pct, 1), "%"))

plot_ly(height = 470) |>
  add_bars(data = bwc_anim, x = ~year, y = ~shown, frame = ~step, color = ~period,
           colors = c("steelblue", "firebrick"), text = ~txt, textposition = "outside",
           hoverinfo = "none") |>
  add_markers(data = bwc_lines, x = ~x, y = ~pct, frame = ~step, ids = ~period,
              marker = list(symbol = "line-ew-open", size = 200, color = "navy",
                            line = list(width = 4, color = "navy")),
              showlegend = FALSE, hoverinfo = "none") |>
  add_text(data = bwc_lines, x = ~x, y = ~ytxt, text = ~txt, frame = ~step,
           textfont = list(size = 15, color = "navy"), showlegend = FALSE, hoverinfo = "none") |>
  layout(xaxis = list(title = "", range = c(2014.4, 2024.6), dtick = 1),
         yaxis = list(title = "Body camera reported (%)", range = c(0, 38)),
         legend = list(orientation = "h", y = 1.08)) |>
  animation_opts(frame = 900, transition = 400, redraw = FALSE) |>
  animation_slider(currentvalue = list(prefix = "")) |>
  config(displayModeBar = FALSE)
```

The rise is real but not smooth: 2023 hit 30%, and 2024 fell back to 20%. Comparing two five-year periods averages over those bumps; that is a choice, and an honest write-up says so. Now both tests on the two periods:

```{r}
#| label: wapo-tests
prop.test(x = c(1205, 594), n = c(5507, 4923))
bayes_prop_test(x = c(1205, 594), n = c(5507, 4923))
```

**Say it.** About 12 in 100 fatal shootings had known camera footage before 2020, and about 22 in 100 after (8 to 11 points higher, by either test). Close to doubling, and still fewer than 1 in 4. What this cannot tell you: whether cameras changed anything about the shootings, or how many cameras existed but went unreported. With 10,430 cases, the p-value was never in doubt. The size is the whole story.

## Four questions, one recipe

| | Test | Answer | The lesson |
|---|---|---|---|
| Reentry | `prop.test()` + Bayes | 22% vs 31%; p = 0.063; 97% probably lower | "Not significant" is not "no effect" |
| Pretrial | `prop.test()` + Bayes, then again after `filter()` | 18 points; 10 among clean records | Groups differ in other ways. Compare like with like |
| Neighborhoods | `t.test()` | $25k to $47k lower | Say the size in real units. Areas are not people |
| Use of force | `prop.test()` + Bayes | 12% to 22% | With a huge n, the p-value adds nothing; the size is the story |

## Try a variation

1. **Rossi:** were men released on **parole** (`paro`) re-arrested less often? Count, then both tests, then say it.
2. **Toronto:** were **citizens** (`citizen`) released more often than non-citizens? Then ask again among people with `checks == 0`. Does the gap shrink the way the employment gap did?
3. **Chicago areas:** run `t.test(pct_vacant ~ violence, data = areas)`. How much higher is the share of vacant housing in high-violence areas, in percentage points?
4. **Shootings:** was a body camera reported more often when the person was in a mental health crisis (`was_mental_illness_related`)? Count, then both tests, and say the size in plain words.
