---
title: "Demo Walkthrough · Session 12"
subtitle: "Logistic regression — odds, odds ratios, predicted probabilities, and the Bayesian version"
---

::: {.callout-note appearance="simple"}
**What this page is.** The Session 12 lecture demos, written down step by step. Run the setup chunk once, then work top to bottom — every demo reproduces what happened on screen. Use this page to catch up after a missed class, or to re-run the lecture at your own pace.
:::

## What you'll build

1. A demonstration of *why* `lm()` breaks on a yes/no outcome — impossible predictions, caught in the act
2. The ladder that fixes it: probability → odds → log-odds, plus the odds-ratio Rosetta Stone
3. `glm(family = binomial)` on 32 cars: exponentiate the coefficients, then predict actual cases
4. The same workflow on real data — what separates Chicago's high-violence community areas — plus a by-hand `plogis()` check that demystifies `predict()`
5. The Bayesian version with `brms`: posterior odds ratios, and the reporting sentence each tradition licenses

## Setup

```{r}
#| label: setup
library(tidyverse)
library(brms)
set.seed(705)

areas <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv",
                  show_col_types = FALSE)
```

::: {.callout-warning appearance="simple"}
`brms` is the same package you installed for Lab 8 — nothing new to install. Same reminder as always: the first `brm()` of a session pauses at `Compiling Stan program...` for a minute before sampling starts. It is not frozen.
:::

Was an **arrest** made? Does the defendant **fail to appear**? Is the parolee **re-arrested** within three years? Is this a **high-violence** neighborhood? Every model so far predicted a *number* — a crime rate, an incident hour. Logistic regression predicts a **yes/no**, and its currency is probability (0 to 1), expressed along the way as odds and log-odds. This session also pays off a Session 1 promise: the opening-week bank-robbery model — patrol cars sent to the West Zone off a coefficient table nobody could read yet — was logistic regression all along, its mysterious `estimate` column just log-odds, one `exp()` away from sense. The [Session 12 slides](../slides/week-12-logistic.qmd) reopen that exact model with today's eyes ("The Machinery, Reopened").

## 1 · Watch linear regression fail

Predicting *graduated the academy (yes/no)* from study hours:

```{r}
#| label: recruits-sim
set.seed(42)   # kept from the lecture — same simulated recruits as the slides
recruits <- tibble(
  study_hours = runif(100, 0, 10),
  prob_grad   = 1 / (1 + exp(-(-4 + 0.8 * study_hours))),
  graduated   = rbinom(100, 1, prob_grad)
)

model_linear <- lm(graduated ~ study_hours, data = recruits)
model_logit  <- glm(graduated ~ study_hours, data = recruits, family = binomial)
```

Ask the linear model about recruits at the edges:

```{r}
#| label: impossible-preds
predict(model_linear, newdata = tibble(study_hours = c(-5, 15)))
```

"Probabilities" of −0.75 and 1.66. Those cannot exist. See the whole picture — linear regression is a straight road that runs forever; logistic regression is a curved hill that flattens at 0 and 1:

```{r}
#| label: road-and-hill
#| fig-height: 4.5
grid_hours <- tibble(study_hours = seq(-5, 15, length.out = 400)) |>
  mutate(
    Linear   = predict(model_linear, newdata = pick(study_hours)),
    Logistic = predict(model_logit, newdata = pick(study_hours),
                       type = "response")
  ) |>
  pivot_longer(c(Linear, Logistic), names_to = "model", values_to = "pred")

ggplot(recruits, aes(study_hours, graduated)) +
  geom_jitter(height = 0.05, width = 0, alpha = 0.5, color = "grey40") +
  geom_hline(yintercept = c(0, 1), linetype = 2, color = "grey60") +
  geom_line(data = grid_hours, aes(y = pred, color = model), linewidth = 1.2) +
  scale_color_manual(values = c("firebrick", "steelblue")) +
  scale_y_continuous(limits = c(-1.5, 2.5), breaks = seq(0, 1, 0.5)) +
  annotate("text", x = -3, y = 2.1, label = "Impossible (>1)",
           color = "firebrick", fontface = "bold", size = 5) +
  annotate("text", x = 13, y = -1.1, label = "Impossible (<0)",
           color = "firebrick", fontface = "bold", size = 5) +
  labs(x = "Study hours", y = "Predicted probability of graduating", color = NULL)
```

Inside the middle of the data the two lines nearly agree; the damage is at the edges — exactly where predictions get used ("what about a recruit who studies *a lot*?"). The fix: don't model the probability directly; model something built from it that a straight line *can* live on.

## 2 · The ladder: probability → odds → log-odds

Suppose 3 out of 4 parolees complete supervision: p = 0.75. Then odds = p / (1 − p) = 0.75 / 0.25 = **3** — "three to one." Odds run 0 to ∞; *log*-odds run −∞ to +∞. That's a scale a straight line can live on. Climb the ladder for a few probabilities:

```{r}
#| label: odds-ladder
tibble(probability = c(0.25, 0.50, 0.75)) |>
  mutate(
    odds     = probability / (1 - probability),
    log_odds = log(odds)
  )
```

The logistic model treats the log-odds as its outcome:

$$\text{logit}(p) = \ln\!\left(\frac{p}{1-p}\right) = b_0 + b_1 X_1 + b_2 X_2$$

Same intercept, same slopes, same "holding the others constant" as Session 11. The line predicts log-odds, which convert back to a probability that politely stays between 0 and 1 — the curved hill.

But the coefficients are log-odds, and nobody thinks in log-odds. So what do we do? **EXPONENTIATE.** `exp(coefficient)` is the **odds ratio**: *multiply the odds of the outcome by this much for each one-unit increase in X, holding the other predictors constant.*

The Rosetta Stone, shrink side — negative log-odds → ratio below 1 → odds shrink:

| Log-odds coefficient (β) | Odds ratio (e^β^) | Effect on odds | Intuitive interpretation |
|:---:|:---:|:---:|:---|
| −2.0 | 0.14 | ↓ 86% | Odds are 86% lower than baseline |
| −1.0 | 0.37 | ↓ 63% | Odds are 63% lower than baseline |
| −0.5 | 0.61 | ↓ 39% | Odds are 39% lower than baseline |
| 0.0 | 1.00 | No change | Odds unchanged |

And the grow side — positive log-odds → ratio above 1 → odds grow:

| Log-odds coefficient (β) | Odds ratio (e^β^) | Effect on odds | Intuitive interpretation |
|:---:|:---:|:---:|:---|
| 0.0 | 1.00 | No change | Odds unchanged |
| 0.3 | 1.35 | ↑ 35% | Odds are 35% higher than baseline |
| 0.7 | 2.01 | ↑ 101% | Odds are about twice as high |
| 1.0 | 2.72 | ↑ 172% | Odds are 2.7× higher |
| 2.0 | 7.39 | ↑ 639% | Odds are over seven times higher |

## 3 · Fitting it: glm(family = binomial)

Toy data first — 32 cars from a 1974 *Motor Trend* issue (`mtcars`, built into R), predicting **manual transmission (1) vs. automatic (0)** from horsepower and weight. Small enough to check every number; real CJ data below.

```{r}
#| label: mtcars-glm
cars_df <- mtcars |>
  rownames_to_column("car") |>
  mutate(am = as.factor(am))   # 0 = automatic, 1 = manual

model_freq <- glm(am ~ hp + wt, data = cars_df, family = binomial)
summary(model_freq)
```

`glm()` is the *generalized* linear model — regression's family business — and `family = binomial` is the switch that makes it logistic. The estimates are log-odds. Exponentiate:

```{r}
#| label: mtcars-or
exp(coef(model_freq))
```

- `hp`: log-odds 0.036 → **odds ratio 1.04** — each additional horsepower multiplies the odds of a manual transmission by about 1.04 (+3.7%), *holding weight constant*
- `wt`: log-odds −8.08 → **odds ratio 0.0003** — each additional 1,000 lbs cuts the odds by about 99.97%, *holding horsepower constant*
- Wildly different sizes partly because the **units** differ: one horsepower is a nudge; a thousand pounds is a different car (Session 11's units lesson, again)

### From effects to cases: predict()

Odds ratios describe *effects*. For an actual *case*, ask for the probability directly:

```{r}
#| label: mtcars-pred
two_cars <- cars_df |>
  filter(car %in% c("Honda Civic", "Cadillac Fleetwood")) |>
  select(car, hp, wt)

two_cars |>
  mutate(p_manual = predict(model_freq, newdata = two_cars, type = "response"))
```

`type = "response"` returns probabilities: the light Civic is a near-certain manual (99.95%); the 5,250-lb Fleetwood's probability of a manual has seven zeros after the decimal. **Report both: odds ratios for effects, predicted probabilities for cases.**

### The curved surface, seen from above

Session 11's flat plane, bent: with two predictors, the logistic model is a **curved sheet** — a plateau near probability 1, a floor near 0, and an S-shaped cliff between them. Ask the model for a prediction at every (hp, wt) combination and look straight down at the surface:

```{r}
#| label: surface-map
#| fig-height: 4.5
grid <- expand_grid(
  hp = seq(min(cars_df$hp), max(cars_df$hp), length.out = 40),
  wt = seq(min(cars_df$wt), max(cars_df$wt), length.out = 40)
)
grid$prob <- predict(model_freq, newdata = grid, type = "response")

ggplot(grid, aes(hp, wt)) +
  geom_raster(aes(fill = prob)) +
  geom_contour(aes(z = prob), breaks = 0.5, color = "white", linewidth = 1) +
  geom_point(data = cars_df, aes(shape = am), color = "black", size = 2.5) +
  scale_fill_gradient(low = "lightblue", high = "firebrick") +
  labs(title = "Predicted probability of a manual transmission",
       subtitle = "The white line is p = 0.5 — the S-curve's cliff edge, seen from above",
       x = "Horsepower", y = "Weight (1,000 lbs)",
       fill = "P(manual)", shape = "Manual?")
```

Color is the surface's height; the tight color transition is where the cliff lives. For the spinnable 3-D version — and the recorded guided tour — open the [Session 12 slides](../slides/week-12-logistic.qmd).

## 4 · Real data: Chicago's high-violence areas

What separates Chicago's **high-violence** community areas from the rest? The outcome ships in the file: is the area's violent-crime rate above the citywide median?

```{r}
#| label: chicago-outcome
areas |> count(high_violence)
```

The line that built it — `areas |> mutate(high_violence = violent_rate > median(violent_rate))` — already ran for you. R treats `TRUE` as 1 and `FALSE` as 0, so a logical column drops straight into `glm()`; if your project outcome isn't binary yet, it may be one `mutate()` away too. One caution: cutting a continuous variable in half **throws away information** — defensible when the yes/no version is the question you actually care about.

Look first (always) — one predictor against the outcome, with the fitted logistic curve:

```{r}
#| label: chicago-s-curve
#| fig-height: 4.5
ggplot(areas, aes(unemployment_rate, as.numeric(high_violence))) +
  geom_jitter(height = 0.04, width = 0, size = 2.5, alpha = 0.6,
              color = "firebrick") +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              se = FALSE, color = "navy", linewidth = 1.2) +
  labs(title = "The S-curve, on real data",
       subtitle = "77 community areas; curve = single-predictor logistic fit",
       x = "Unemployment rate (%)", y = "High-violence area (0/1)")
```

Now the two-predictor model — unemployment and vacant housing, both plausible markers of concentrated disadvantage:

```{r}
#| label: chicago-glm
v_model <- glm(high_violence ~ unemployment_rate + pct_vacant,
               data = areas, family = binomial)
summary(v_model)
```

```{r}
#| label: chicago-or
exp(cbind(OR = coef(v_model), confint(v_model)))   # translate: log-odds -> ORs
```

- **Unemployment: OR 1.45** [1.19, 1.85] — each additional point of unemployment multiplies an area's odds of high-violence status by about 1.45 (+45%), holding vacancy constant
- **Vacancy: OR 1.33** [1.04, 1.84] — +33% odds per point, holding unemployment constant
- For a *ratio*, no effect = **1**, not 0 — both intervals exclude it. And these are associations, **not causes**: nobody randomly assigned unemployment to neighborhoods

### Coefficients into cases

```{r}
#| label: chicago-pred
two_areas <- areas |>
  filter(community %in% c("Forest Glen", "West Garfield Park")) |>
  select(community, unemployment_rate, pct_vacant)

two_areas |>
  mutate(p_high_violence = predict(v_model, newdata = two_areas,
                                   type = "response"))
```

In plain language: for an area like Forest Glen (4.2% unemployment, 2.6% vacant), the model puts the probability of high-violence status at about **3%**. For one like West Garfield Park (18.9%, 20.0%), it's over **99%**. That's a sentence a chief — or a city council — can actually use.

### No magic: the plogis() check

`predict()` is just the ladder run in reverse — plug the predictors into the log-odds line, then squash the result back into [0, 1]. Do Forest Glen by hand:

```{r}
#| label: plogis-check
b <- coef(v_model)
log_odds_fg <- b[1] + b[2] * 4.2 + b[3] * 2.6   # Forest Glen's log-odds
plogis(log_odds_fg)                             # back to probability
```

Same 0.029 as `predict()`. `plogis()` is R's name for the inverse logit — the function that bends the straight line into the hill.

## 5 · The Bayesian version

Same formula; `family = bernoulli()` is brms's name for a 0/1 outcome — the switch that `family = binomial` flips in `glm()`. Everything else is the Session 8 pattern. (In lecture we first ran this machinery on the 32 cars, where the honest posterior intervals came out enormous — with 77 areas they tighten to something usable.) This is the chunk with the compile pause.

```{r}
#| label: chicago-brm
#| output: false
v_bayes <- brm(high_violence ~ unemployment_rate + pct_vacant,
               data = areas, family = bernoulli(),
               chains = 2, iter = 2000, seed = 705, refresh = 0)
```

```{r}
#| label: chicago-brm-or
exp(fixef(v_bayes)[, c("Estimate", "Q2.5", "Q97.5")])
```

Both traditions land in the same place: posterior odds ratios near the `glm()` estimates, credible intervals excluding 1. The prior: we stated none, so these are `brms` defaults — flat on the coefficients, weakly informative on the intercept. Your project's Bayesian model will use these too — **say so in your methods**, one sentence: "brms default priors, 2 chains × 2,000 iterations, seed 705."

How to report it, in each tradition:

> **Frequentist:** "Each one-point increase in the unemployment rate was associated with 45% higher odds of being a high-violence area (OR = 1.45, 95% CI [1.19, 1.85]), holding vacancy constant. Predicted probabilities ranged from about 3% for the least-disadvantaged areas to over 99% for the most."
>
> **Bayesian:** "Given the data and priors, each one-point increase in unemployment multiplied the odds of high-violence status by about 1.5 (posterior mean OR ≈ 1.5, 95% credible interval ≈ [1.2, 1.9])."

Same three ingredients every time: the effect (as an odds ratio), the uncertainty, and a **probability a reader can feel**. And say ***odds*** when you mean odds — an odds ratio of 1.45 is not "45% more likely."

## Try a variation

1. A reentry study reports two log-odds coefficients for re-arrest within 12 months: risk score **0.7** per point, completed programming **−0.5**. Convert both with `exp()` and say each in a sentence a police chief would understand. (Check yourself: one roughly doubles the odds; the other cuts them by about 39%.)
2. Rebuild the outcome with a harsher cut — `areas |> mutate(top_quartile = violent_rate > quantile(violent_rate, 0.75))` — and refit `v_model` on it. What happens to the odds ratios and their intervals when only about 19 areas are a "yes"?
3. Swap `pct_vacant` for `pct_no_hs_diploma` in `v_model`. Does the unemployment odds ratio move? Connect what you see to Session 11's lurker lesson.
4. Pick any other Chicago community area and compute its predicted probability by hand with `plogis()`, then check yourself with `predict()`.
