---
title: "Demo Walkthrough · Session 11"
subtitle: "Multiple linear regression — least squares, R², 'holding constant,' and the Bayesian version"
---

::: {.callout-note appearance="simple"}
**What this page is.** The Session 11 lecture demos, written down step by step. Run the setup chunk once, then work top to bottom — every demo reproduces exactly what happened on screen. Use this page to catch up after a missed class, or to re-run the lecture at your own pace.
:::

## What you'll build

1. A hand-rolled ruler for "best line" — `SSR()` — tested on three candidate lines, the proof that `lm()` beats every hand-picked guess, and the check that one-predictor R² is just Session 10's correlation, squared
2. The best-fitting *plane* for two predictors (static slices here — the slides have the spinnable 3D version)
3. The ice-cream/drownings lurker: a "significant" coefficient that evaporates when the third variable enters
4. A real two-predictor model of violent crime across Chicago's 77 community areas — read honestly, then diagnosed (residual plot, VIF)
5. The same model in `brms`: posterior distributions, and the reporting sentence each tradition licenses

## Setup

```{r}
#| label: setup
library(tidyverse)
library(bayesplot)
library(brms)
library(car)
set.seed(705)

areas <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv",
                  show_col_types = FALSE)
```

::: {.callout-warning appearance="simple"}
Two non-tidyverse packages today: `car` (for VIF — `install.packages("car")` if you don't have it) and `brms` from ANOVA week (`bayesplot` came along free when you installed it). The usual `brms` reminder: the first `brm()` of a session pauses at `Compiling Stan program...` for a minute before sampling starts. It is not frozen. Compiling.
:::

## 1 · Least squares: build the ruler by hand

Ten students, hours studied and exam scores. Some line through this cloud is the "best" one — but best by what ruler? Start with three candidates. (Seed 1 is kept from the lecture so every number matches the slides.)

```{r}
#| label: candidate-lines
#| fig-height: 4.5
set.seed(1)
students <- tibble(
  hours = 1:10,
  score = 40 + 5 * hours + rnorm(10, 0, 5)
)

candidates <- tibble(
  a = c(20, 50, 37),
  b = c(5, 3, 5.2),
  line = c("a = 20, b = 5", "a = 50, b = 3", "a = 37, b = 5.2")
)

ggplot(students, aes(hours, score)) +
  geom_abline(data = candidates, aes(intercept = a, slope = b, color = line),
              linewidth = 1.1) +
  geom_point(size = 4, color = "navy") +
  scale_color_manual(values = c("firebrick", "seagreen", "steelblue")) +
  labs(x = "Hours studied", y = "Exam score", color = NULL,
       title = "Which line is best? We need a ruler, not a vibe.")
```

The ruler: for any candidate line, measure every vertical miss, square it (so misses above and below don't cancel), and add them up. That total is the **sum of squared residuals**:

```{r}
#| label: ssr-function
SSR <- function(a, b) {
  predicted <- a + b * students$hours
  sum((students$score - predicted)^2)
}

SSR(20, 5)      # too low
SSR(50, 3)      # too flat
SSR(37, 5.2)    # roughly right
```

Smaller total = better line. **Least squares** = find the line with the smallest total of all. No trial and error required — `lm()` solves for it directly:

```{r}
#| label: lm-wins
fit_hours <- lm(score ~ hours, data = students)
coef(fit_hours)

SSR(coef(fit_hours)[1], coef(fit_hours)[2])
```

Our best hand-picked guess left about 197 in squared misses; `lm()` gets it down to about 131 — and **no other line does better**. Read the slope: each additional hour of studying predicts about 5.3 more points, in this sample. The vertical gap between each point and the winning line is that student's **residual** — the miss the model owns. (The slides animate the search: the line tilts until the SSR stops shrinking.) So how much did the winning line explain? That's **R²**:

```{r}
#| label: r2-check
summary(fit_hours)$r.squared
cor(students$hours, students$score)^2
```

**R² is the proportion of the outcome's variance the model accounts for** — here about 95% (it's simulated data; real life humbles us below). And notice the second line: with one predictor, R² is just Session 10's correlation, squared. With multiple predictors, R² keeps the same meaning — variance explained by the whole team — but no single correlation owns it anymore.

## 2 · Two predictors: from line to plane

With one predictor the model is a line through a 2-D cloud. With two, it's a **plane** floating in a 3-D cloud — one slope along each predictor's axis, chosen by the same ruler: minimize the total squared vertical distance from every point to the surface. Simulate data where we *know* the truth — intercept 10, slopes 2 and 3 — and check that `lm()` finds it:

```{r}
#| label: plane-fit
set.seed(42)   # kept from the lecture — same cloud as the slides
plane_data <- tibble(
  x1 = runif(100, 0, 10),
  x2 = runif(100, 0, 10),
  y  = 10 + 2 * x1 + 3 * x2 + rnorm(100, 0, 2)
)

plane_fit <- lm(y ~ x1 + x2, data = plane_data)
coef(plane_fit) |> round(2)
```

Close to (10, 2, 3) — least squares recovered the plane. A flat page can't spin a 3-D plot, but slicing the plane says the same thing: hold `x2` fixed at a few values and plot predicted `y` against `x1`:

```{r}
#| label: plane-slices
#| fig-height: 4.5
slices <- expand_grid(x1 = seq(0, 10, length.out = 50), x2 = c(2, 5, 8))
slices$predicted <- predict(plane_fit, newdata = slices)

ggplot(slices, aes(x1, predicted, color = factor(x2))) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("steelblue", "navy", "firebrick")) +
  labs(title = "The plane, sliced: the x1 slope is the same at every fixed x2",
       subtitle = "Parallel lines are what 'holding x2 constant' looks like",
       x = "x1", y = "Predicted y", color = "x2 held at")
```

Every slice climbs at slope ≈ 2 (the `x1` coefficient); jumping *between* slices is the `x2` coefficient's job. For the grab-and-spin 3-D version — points, plane, and residuals — open the [Session 11 slides](../slides/week-11-regression.qmd) ("The Best-Fitting Plane, Live").

## 3 · The lurking third variable

Ice cream sales and drowning deaths are strongly correlated. Every summer. Reliably. Build the world where temperature secretly drives both:

```{r}
#| label: lurker-data
#| fig-height: 4.5
set.seed(705)
lurk <- tibble(
  temperature    = runif(100, 60, 100),
  icecream_sales = 20 + 5 * temperature + rnorm(100, 0, 30),
  drownings      = -10 + 0.8 * temperature + rnorm(100, 0, 10)
)

ggplot(lurk, aes(icecream_sales, drownings)) +
  geom_point(color = "navy", size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  labs(title = "r = 0.6 — ban ice cream?",
       x = "Ice cream sales", y = "Drownings")
```

Model 1 fits what the scatterplot suggests — ice cream "predicts" drownings:

```{r}
#| label: lurker-m1
lurk_m1 <- lm(drownings ~ icecream_sales, data = lurk)
summary(lurk_m1)$coefficients |> round(4)
```

The coefficient is positive and the p-value has ten zeros in it. Statistically significant; the model says it's real. Do we believe it? No — but notice that *nothing in this output warns us*. The model can only see the variables we gave it. So give it the lurker:

```{r}
#| label: lurker-m2
lurk_m2 <- lm(drownings ~ icecream_sales + temperature, data = lurk)
summary(lurk_m2)$coefficients |> round(4)
```

Holding temperature constant, ice cream's coefficient collapses from 0.13 to about −0.03 and goes nonsignificant (p ≈ 0.26), while R² jumps from 0.37 to 0.55. Temperature was driving both variables the whole time. This is *holding constant* doing real work: comparing days with the **same temperature**, ice cream tells you nothing about drownings. An **added-variable plot** (from `car`) shows exactly that — the ice cream–drownings relationship after temperature's contribution is removed from both:

```{r}
#| label: lurker-avplot
#| fig-height: 4.5
avPlots(lurk_m2, terms = ~ icecream_sales,
        main = "Ice cream vs. drownings, with temperature partialed out")
```

Flat. That flat line *is* the multiple-regression coefficient for ice cream. The moral: multiple regression holds constant **only the variables you measured and included**. Every observational model has potential lurkers you didn't measure — so coefficients are **adjusted associations, not causal effects**. Keep that loaded; the Chicago models below will tempt you to talk causally.

## 4 · Real data: violence across Chicago's 77 areas

You plotted violent-crime rate against unemployment across the 77 community areas in Session 10 — a strong upward cloud, each dot an *area*, not a person (say it with me: ecological). Look at that scatter again first if it isn't fresh, then fit one predictor:

```{r}
#| label: chicago-m1
chi_m1 <- lm(violent_rate ~ unemployment_rate, data = areas)
summary(chi_m1)
```

Each additional percentage point of unemployment is associated with **3.3 more violent crimes per 1,000 residents**, and unemployment alone accounts for about 70% of the between-area variance. Strong — but is unemployment getting credit for other disadvantage it travels with? Add vacant housing:

```{r}
#| label: chicago-m2
chi_m2 <- lm(violent_rate ~ unemployment_rate + pct_vacant, data = areas)
summary(chi_m2)
```

Both predictors earn significant coefficients, and R² climbs from 0.70 to **0.82**. But watch unemployment's slope: **3.3 → 2.1**. Some of what looked like "unemployment" in the simple model was shared with vacancy. That shrinkage *is* "holding constant" — the lurker lesson, on real data. Reading the coefficients out loud:

- **Intercept (−9.3):** predicted rate when both predictors are 0 — no Chicago area looks like that; it anchors the plane, nothing more
- **`unemployment_rate` (2.12):** +1 point of unemployment ↔ about 2.1 more violent crimes per 1,000 residents, **vacancy held fixed**
- **`pct_vacant` (1.94):** +1 point of vacancy ↔ about 1.9 more violent crimes per 1,000 residents, **unemployment held fixed**

And what they are *not*: not causal effects (these are adjusted associations between *areas*, with every unmeasured lurker still at large); not people-level claims (an unemployed *person* is not "2.1 crimes more dangerous" — Session 10's ecological fallacy rides along with every coefficient today); and not the full story (the model holds constant *only what's in the model*).

### Watch the units

Coefficient *size* is meaningless until you check the predictor's units. Median income, in dollars:

```{r}
#| label: chicago-units
coef(lm(violent_rate ~ median_income, data = areas))

areas |>
  mutate(income_10k = median_income / 10000) |>
  lm(violent_rate ~ income_10k, data = _) |>
  coef()
```

Same model, same fit. But −0.0005 "looks like" nothing; the honest reading is **about 5 *fewer* violent crimes per 1,000 residents per $10,000 of median income**. Rescale before you interpret.

### Does a third predictor earn its keep?

```{r}
#| label: chicago-m3
chi_m3 <- lm(violent_rate ~ unemployment_rate + pct_vacant + pct_no_hs_diploma,
             data = areas)
summary(chi_m3)$coefficients |> round(3)
```

Coefficient ≈ 0, p ≈ 0.77 — with unemployment and vacancy already in, the diploma share adds nothing detectable. And **adjusted R²** (R² with a per-predictor penalty) actually *falls*, 0.813 → 0.811 (compare `summary(chi_m2)$adj.r.squared` with `summary(chi_m3)$adj.r.squared`). More predictors ≠ more truth. Every variable argues its way in — theory first, fit second.

### Diagnostics: the residual plot and VIF

Regression assumes the misses are patternless. So look at the misses:

```{r}
#| label: residual-plot
#| fig-height: 4.5
tibble(fitted = fitted(chi_m2), residual = resid(chi_m2)) |>
  ggplot(aes(fitted, residual)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point(color = "navy", size = 2.5, alpha = 0.8) +
  labs(title = "Residuals vs. fitted values — the all-purpose diagnostic",
       x = "Fitted violent-crime rate", y = "Residual")
```

Read it honestly. What we want: a patternless horizontal band around zero. What we have: centered on zero with no obvious curve (good — the straight-plane assumption isn't visibly wrong), but the **spread widens at higher fitted values** — the model misses by more in high-violence areas. That fanning (**heteroscedasticity**) is common with rate outcomes; it doesn't bias the coefficients, but it makes the standard errors less trustworthy at the extremes — worth a sentence when you report. A *curve* in this plot would be worse news: it would say the model's *shape* is wrong, not just its noise.

**Multicollinearity** is the other standard check: if two predictors move in near-lockstep, the model can't tell whose coefficient the shared variance belongs to — estimates get wobbly and standard errors balloon. **VIF** measures how redundant each predictor is with the others (rules of thumb: > 5, pay attention; > 10, you have a problem):

```{r}
#| label: chicago-vif
vif(chi_m2)
```

About 1.8: unemployment and vacancy are correlated (r ≈ 0.67), but each still carries plenty of its own information. If VIF *were* high: drop one predictor, combine them into an index, or accept wide intervals — you cannot p-value your way out.

## 5 · The Bayesian version

Identical formula, Bayesian machinery — the same `brms` workflow you've run since ANOVA week. (This is the chunk with the compile pause. Go refill your coffee.)

```{r}
#| label: chicago-brm
#| output: false
chi_bayes <- brm(violent_rate ~ unemployment_rate + pct_vacant,
                 data = areas,
                 chains = 2, iter = 2000, seed = 705, refresh = 0)
```

```{r}
#| label: chicago-brm-fixef
fixef(chi_bayes) |> round(2)
```

Posterior means land almost exactly on the least-squares estimates — with `brms`'s default priors (flat on the slopes, weakly informative on the intercept and spread) and a decent *n*, they should. The Session 10 rule travels: defaults are fine, **but say which prior you used**. One sentence in your methods — "brms default priors, 2 chains × 2,000 iterations, seed 705" — covers it. Below the posterior plot: the sentence each tradition licenses.

```{r}
#| label: chicago-mcmc-areas
#| fig-height: 4
mcmc_areas(chi_bayes, pars = c("b_unemployment_rate", "b_pct_vacant"),
           prob = 0.95) +
  labs(title = "Posterior distributions: predictors of violent-crime rate",
       subtitle = "Shaded = central 95% credible interval — both sit clear of zero")
```

> **Frequentist:** "Holding vacancy constant, each percentage point of unemployment was associated with about 2.1 more violent crimes per 1,000 residents (b = 2.12, 95% CI [1.59, 2.64], p < .001)."
>
> **Bayesian:** "Given the data and priors, there is a 95% probability the unemployment coefficient lies between ≈ 1.6 and ≈ 2.6 — zero is not a credible value."

Both are area-level, adjusted **associations** — neither sentence licenses "unemployment causes violence."

## Try a variation

1. Beat the machine: using `SSR()`, hunt by hand for an (a, b) pair that gets below 131. (Spoiler: you can tie `lm()` to more decimal places, but never beat it.)
2. In the lurker simulation, give ice cream a *real* effect — change the `drownings` line to `-10 + 0.8 * temperature + 0.05 * icecream_sales + rnorm(100, 0, 10)`. Does `lurk_m2` find it?
3. In the third-predictor test, swap `pct_no_hs_diploma` for a predictor of your choice (`pct_renter`? `median_rent`?). Does anything earn its keep next to unemployment and vacancy — by its coefficient *and* by adjusted R²?
4. Run `predict(chi_m2, newdata = tibble(unemployment_rate = 23, pct_vacant = 25))` — roughly Englewood — and then again with 10 and 10. Which prediction does the residual plot say to trust more, and why?
