---
title: "Demo Walkthrough · Session 10"
subtitle: "Correlation — covariance, Pearson's r, and what r cannot see"
---

::: {.callout-note appearance="simple"}
**What this page is.** The Session 10 lecture demos, written down step by step. Every chunk below is self-contained — copy any of them into your own R session and they run as-is. Use this page to catch up after a missed class, or to re-run what happened on screen at your own pace.
:::

## What you'll build

1. Covariance on real Chicago data — and the units problem that makes it uninterpretable
2. Pearson's r as rescaled covariance, computed by hand before `cor()` does it for you
3. A calibration grid: what r = 0.1 through 0.9 actually *look* like
4. Three demonstrations of what r cannot see — identical summaries, curves, and one point that flips a sign
5. Spearman's rho as the rank-based backup
6. The real question: which structural features of Chicago's 77 community areas track violence — with inference in both traditions, and the ecological-fallacy guardrail on every claim

## Setup

Everything below needs the tidyverse, the `correlation` package (easystats — install once with `install.packages("correlation")`), and the *second level* of the course's anchor data: 2025's incidents aggregated to the 77 community areas and joined with Census measures.

```{r}
#| label: setup
library(tidyverse)
library(correlation)
set.seed(705)

areas <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv",
                  show_col_types = FALSE)
```

## 1 · Covariance: moving together, in raw units

Observation by observation: do x and y sit on the *same side* of their means?

$$\text{cov}(x, y) \;=\; \frac{\sum_i (x_i - \bar{x})(y_i - \bar{y})}{n - 1}$$

Both above average (or both below) → positive products → positive covariance; opposite sides → negative. The **sign** is trustworthy. The **size** is not — watch what happens on real data:

```{r}
#| label: cov-units
cov(areas$unemployment_rate, areas$violent_rate)
cov(areas$median_income,     areas$violent_rate)
```

Is −482,660 a "stronger" relationship than 108? No idea — the first number lives in *dollars × crimes per 1,000*, the second in *percentage points × crimes per 1,000*. **Covariance is glued to its units.** You can't compare across pairs.

## 2 · Correlation rescales it

Divide the covariance by both standard deviations and the units cancel:

$$r \;=\; \frac{\text{cov}(x, y)}{s_x \, s_y}$$

```{r}
#| label: r-standardize
cov(areas$unemployment_rate, areas$violent_rate) /
  (sd(areas$unemployment_rate) * sd(areas$violent_rate))

cor(areas$unemployment_rate, areas$violent_rate)   # same thing, one function
```

Always between −1 and +1, unit-free, comparable: income runs r = −0.71 with violence — so unemployment (0.84) tracks violence more tightly than income does, direction aside. The conventional reading bands (|r| beyond ≈ 0.7 "strong," 0.4–0.7 "moderate," under 0.3 "weak") are **conventions, not laws** — in messy social-science data, a "moderate" r can be a major finding.

## 3 · Calibrate your eye

Before trusting the bands, see them. Simulate five datasets with known correlations and look:

```{r}
#| label: r-grid
#| fig-height: 3.5
set.seed(705)   # inline seed: keeps these particular clouds (and their r labels) stable
r_targets <- c(0.1, 0.3, 0.5, 0.7, 0.9)

r_grid <- map(r_targets, \(r) {
  x <- rnorm(150)
  y <- r * x + sqrt(1 - r^2) * rnorm(150)
  tibble(label = paste0("r = ", sprintf("%.2f", cor(x, y))), x = x, y = y)
}) |>
  list_rbind() |>
  mutate(label = fct_inorder(label))

ggplot(r_grid, aes(x, y)) +
  geom_point(alpha = 0.5, size = 1.4, color = "navy") +
  facet_wrap(~label, nrow = 1) +
  labs(title = "Five simulated datasets, 150 points each",
       x = NULL, y = NULL) +
  theme(axis.text = element_blank())
```

(The panel labels show each sample's *achieved* r — 150 draws don't land exactly on the target, which is Session 4's sampling variability all over again.) By ≈ 0.5 the trend is visible; below ≈ 0.2 your eye mostly sees noise — and it's right. Negative correlations look identical with the tilt reversed.

## 4 · What r cannot see

### Same summaries, four different stories

Anscombe's quartet ships with R (Anscombe, 1973) — four little datasets engineered to share their summary statistics:

```{r}
#| label: anscombe
#| fig-height: 4.5
anscombe_long <- anscombe |>
  pivot_longer(everything(), names_to = c(".value", "set"), names_pattern = "(.)(.)")

anscombe_long |>
  summarize(mean_x = mean(x), mean_y = mean(y), r = cor(x, y), .by = set)
ggplot(anscombe_long, aes(x, y)) +
  geom_point(size = 2.5, color = "navy") +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick", linewidth = 0.8) +
  facet_wrap(~set, labeller = labeller(set = \(s) paste0("Set ", s, " — r = 0.82"))) +
  labs(title = "Same r, four completely different stories", x = NULL, y = NULL)
```

Same means, same r = 0.82 — statistically interchangeable, and only Set 1 is what you picture when someone says "r = 0.82." **Always plot before you correlate.**

### Limitation 1: r only sees straight lines

```{r}
#| label: u-shape
#| fig-height: 4
set.seed(705)   # inline seed: stable curve
u_data <- tibble(
  x = seq(-4, 4, length.out = 50),
  y = x^2 + rnorm(50, 0, 1)
)

ggplot(u_data, aes(x, y)) +
  geom_point(size = 2.5, color = "chocolate3") +
  geom_smooth(method = "loess", se = FALSE, color = "chocolate3") +
  labs(title = "A clear relationship that r cannot see",
       subtitle = paste0("Pearson's r = ", sprintf("%.2f", cor(u_data$x, u_data$y)),
                         " — the data follow a curved (U-shaped) pattern"),
       x = "Variable X", y = "Variable Y")
```

Think stress vs. performance: an inverted U. r ≈ 0 means no **linear** relationship — not no relationship.

### Limitation 2: one point can manufacture a correlation

Twenty points of pure noise, then the same twenty plus a single wild pair:

```{r}
#| label: outlier-demo
#| fig-height: 4
set.seed(705)   # inline seed: stable demo — the numbers below depend on it
base_data    <- tibble(x = 1:20, y = rnorm(20, 10, 4))   # pure noise
outlier_data <- bind_rows(base_data, tibble(x = 25, y = 100))

bind_rows(
  base_data    |> mutate(version = paste0("20 points of pure noise — r = ",
                                          sprintf("%.2f", cor(base_data$x, base_data$y)))),
  outlier_data |> mutate(version = paste0("Same noise + ONE outlier — r = ",
                                          sprintf("%.2f", cor(outlier_data$x, outlier_data$y))))
) |>
  mutate(version = fct_inorder(version)) |>
  ggplot(aes(x, y)) +
  geom_point(size = 2.5, color = "grey35") +
  geom_point(data = \(d) filter(d, y > 50), color = "firebrick", size = 4.5) +
  geom_smooth(method = "lm", se = FALSE, color = "navy") +
  facet_wrap(~version) +
  labs(x = "Variable X", y = "Variable Y")
```

Two lessons in one picture. First, the noise-only panel gave r = −0.28 — **small samples produce correlations out of nothing**; that's why we test rather than just compute. Second, adding **one** extreme pair swung r from −0.28 to +0.42 — a single point flipped the sign *and* manufactured a "moderate positive" correlation. Neither number describes a real relationship. (Outliers can also work the other way, *hiding* a real correlation.) Defense one is plotting first — you'd catch this instantly. Defense two is a correlation that doesn't care about wild values:

### The rank-based backup: Spearman's rho

Replace each value with its **rank** (1st, 2nd, 3rd…), then correlate the ranks. Spearman asks only: *is the relationship consistently up or consistently down?*

```{r}
#| label: spearman-outlier
c(pearson  = cor(outlier_data$x, outlier_data$y),
  spearman = cor(outlier_data$x, outlier_data$y, method = "spearman"))
```

To Spearman, the wild point is just "rank 21": it drifts from −0.29 to −0.11 — sign intact, still weak — while Pearson swung 0.70 and flipped sign. When data are skewed or outlier-prone, check Spearman alongside Pearson.

## 5 · Real data: structure and violence across Chicago

A classic criminology-of-place question: **which structural features of a neighborhood track its violence?**

```{r}
#| label: areas-peek
areas |>
  select(community, violent_rate, unemployment_rate, median_income, pct_vacant) |>
  head(3)
```

Rule from Section 4: plot first. One dot = one community area:

```{r}
#| label: unemp-scatter
#| fig-height: 4.5
ggplot(areas, aes(unemployment_rate, violent_rate)) +
  geom_point(size = 3, alpha = 0.7, color = "firebrick") +
  geom_smooth(method = "lm", se = FALSE, color = "navy") +
  labs(title = "Violent-crime rate vs. unemployment rate — 77 community areas",
       subtitle = paste0("r = ", sprintf("%.2f", cor(areas$unemployment_rate, areas$violent_rate))),
       x = "Unemployment rate (%)", y = "Violent crimes per 1,000 residents")
```

r = 0.84 — about as strong as area-level associations get in criminology. And the full menu, in one dataset:

```{r}
#| label: four-covariates
#| fig-height: 5
areas |>
  select(violent_rate, unemployment_rate, pct_renter, median_income, median_age) |>
  pivot_longer(-violent_rate, names_to = "covariate", values_to = "value") |>
  mutate(covariate = case_match(covariate,
    "unemployment_rate" ~ paste0("Unemployment rate — r = ",
      sprintf("%.2f", cor(areas$unemployment_rate, areas$violent_rate))),
    "pct_renter" ~ paste0("% renter-occupied — r = ",
      sprintf("%.2f", cor(areas$pct_renter, areas$violent_rate))),
    "median_income" ~ paste0("Median income — r = ",
      sprintf("%.2f", cor(areas$median_income, areas$violent_rate))),
    "median_age" ~ paste0("Median age — r = ",
      sprintf("%.2f", cor(areas$median_age, areas$violent_rate))))) |>
  ggplot(aes(value, violent_rate)) +
  geom_point(alpha = 0.6, size = 2, color = "navy") +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick", linewidth = 0.8) +
  facet_wrap(~covariate, scales = "free_x") +
  labs(title = "Violent-crime rate vs. four structural covariates",
       x = NULL, y = "Violent crimes per 1,000")
```

Strong positive (unemployment, 0.84), moderate positive (renters, 0.55), strong negative (income, −0.71) — and one genuine *nothing* (median age, −0.02). Real data span the whole menu. Income is also where Pearson and Spearman disagree: the decline is real but **bent** — steep across lower-income areas, nearly flat past ≈ $90k — and income is right-skewed, so the rank-based Spearman (−0.82) runs stronger than Pearson (−0.71) because it only asks "downhill?", not "straight line?":

```{r}
#| label: income-spearman
c(pearson  = cor(areas$median_income, areas$violent_rate),
  spearman = cor(areas$median_income, areas$violent_rate, method = "spearman"))
```

## 6 · These are areas, not people

Our r = 0.84 says: **community areas** with higher unemployment rates record more violent crime per 1,000 residents.

::: {.callout-important appearance="simple"}
**The ecological fallacy:** using an aggregate-level association to make claims about individuals. Our data contain **no individuals** — every resident of an area gets the same 77 values. Nothing here says *unemployed people* commit violence.
:::

Saying it out loud, area-data edition. ✅ Defensible: "Community areas with higher unemployment rates recorded higher violent-crime rates (r = 0.84)." ❌ Not defensible: "Unemployed residents commit more violent crime" — an individual-level claim these data can't see. ❌ Also not defensible: "Unemployment drives up violence" — a causal claim, and concentrated disadvantage, disinvestment, and policing patterns all lurk behind both variables. Correlation is a pattern, not proof of cause.

The classic demonstration (Robinson, 1950): across U.S. **states**, more foreign-born residents went with *higher* literacy; across **individuals**, foreign-born residents had *lower* literacy. Same data, opposite signs — immigrants had settled in high-literacy states. Aggregation can weaken, exaggerate, or even **flip** an individual-level relationship.

## 7 · Putting uncertainty on r

We can *see* r = 0.84. The frequentist question: if nothing truly connected these variables, how often would 77 paired observations line up this tightly by luck? One function does the whole thing:

```{r}
#| label: cor-test
cor.test(areas$unemployment_rate, areas$violent_rate)
```

Reading it honestly: **r = 0.84, t(75) = 13.4, p < .001** — a null world essentially never produces this — with **95% CI [0.76, 0.90]**, read exactly as Session 5 taught (the *procedure* captures the truth in 95% of replications). Two perspective checks. First, with n = 77, any |r| above about 0.22 clears p < .05 — significance is a low bar, and **the strength is the story**. Second, these 77 areas are *all of Chicago*, not a sample from a bag of Chicagos — read the null as a claim about the *process generating neighborhood outcomes*, not a larger population of areas.

Real analyses rarely stop at one pair. The `correlation` package tests every pair, attaches CIs, and **Holm-adjusts the p-values** for the number of pairs — six tests at once is exactly Session 8's pile-of-tests problem, handled by default:

```{r}
#| label: corr-pkg
areas |>
  select(violent_rate, unemployment_rate, median_income, pct_vacant) |>
  correlation()
```

## 8 · The Bayesian version

Same package, one new argument — and unlike `brms` in Session 8, no compilation wait; this samples in a blink. Instead of a single r and a p-value, you get a **posterior distribution** of plausible correlation values:

```{r}
#| label: bayes-pair
set.seed(705)   # inline seed: posterior sampling, keeps printed values stable
areas |>
  select(violent_rate, unemployment_rate) |>
  correlation(bayesian = TRUE)
```

Four read-outs, one line each. **rho ≈ 0.82** is the posterior *median* — a touch below the frequentist 0.84, which is the prior's gentle pull toward zero. The **95% CI [≈ 0.73, 0.87]** is a **credible interval**: given data and prior, 95% probability ρ lies in this range — the direct statement Session 5 taught you the confidence interval can't make. **pd = 100%** is the posterior probability the correlation is positive, and **0% in ROPE** says none of the posterior sits in a "practically zero" band. The **BF ≈ 10¹⁷** means the data are astronomically more likely under "correlated" than "not" — on Session 6's scale, anything past 100 is extreme evidence.

### But the priors…

The prior here encodes mild skepticism — real-world correlations are rarely extreme — gently pulling estimates toward 0. The default is `"medium"`; loosen it and see what moves:

```{r}
#| label: bayes-wide
set.seed(705)   # inline seed: posterior sampling
areas |>
  select(violent_rate, unemployment_rate) |>
  correlation(bayesian = TRUE, bayesian_prior = "wide")
```

Almost nothing — *here*. The wide prior is less skeptical of big correlations, so it shrinks rho toward 0 slightly less, and the estimate creeps toward the raw 0.84. With 77 areas of strongly correlated data, **the data dominate**: the prior choice moves the second decimal. With 15 areas and a modest correlation, the same choice could visibly move the estimate — priors matter most when data are thin. Practical rule for this course: **defaults are fine — but say which prior you used.** If your conclusion survives medium *and* wide, say that too; it's a free robustness check. (Same reporting habit as Session 8's brms defaults — the badge changes, the rule doesn't.)

## Try a variation

1. `pct_no_hs_diploma` is the share of an area's adults without a high-school diploma. *Before computing anything*, write down your predicted sign and rough size for its correlation with `violent_rate` — then plot it and run `cor.test()`. Most people predict a strong positive; check what the data actually say, then try `pct_bachelors`.
2. In the outlier demo, change the wild point from `y = 100` to `y = 40`, then `y = 15`. Where's the threshold at which Pearson stops being fooled — and does Spearman ever care?
3. Add `pct_vacant` and `median_rent` to the `correlation()` call. With more pairs, what happens to the Holm-adjusted p-values of the correlations you already had?
4. Re-run the Bayesian pair on a thinner dataset: `areas |> slice_sample(n = 15)` first. Now compare `bayesian_prior = "medium"` vs. `"wide"` — this is the "priors matter when data are thin" lesson, live.
