---
title: "Demo Walkthrough · Session 7"
subtitle: "Four real criminal justice questions, one recipe: count, test, say it"
include-in-header:
  - text: '<link rel="stylesheet" href="../widgets/gap-explorer.css">'
include-after-body:
  - text: '<script src="../widgets/gap-explorer.js"></script>'
---

::: {.callout-note appearance="simple"}
**What this page is.** The Session 7 lecture, written down step by step. Every chunk below is self-contained: copy any of them into your own R session and it runs as-is. Use this page to catch up after a missed class, or to re-run the examples at your own pace.
:::

## What you'll build

Four answers to four real questions, each with the same recipe:

1. **Question.** Name the two groups and the outcome. A yes/no outcome means `prop.test()` and `bayes_prop_test()`; a number means `t.test()`.
2. **Count or describe.**
3. **Test.**
4. **Say it.** The honest sentence, how big in real units, and one thing the test cannot tell you.

The questions: did financial aid at release lower re-arrest? Were employed arrestees released more often? How much lower is income in Chicago's high-violence areas? Was body camera footage reported more often in fatal police shootings after 2020?

Each example also has the **"Where is zero?" explorer** from class: four ways to see what the test is measuring, built from that example's own numbers. Pick one of the four buttons, then press **Play** (or **Step** to go one at a time). Every one asks the same question: the gap between the two groups is uncertain, so where could it really be, and is zero (no difference) one of the places?

- **Rerun the study:** run the same study again and again; each rerun's gap drops onto the number line. The middle 95% of reruns is the confidence interval.
- **Arrows between groups:** pick a plausible value for each group; the arrow between them is the gap. Arrows pointing the other way cross zero.
- **Is zero plausible?:** imagine a world where the two groups are truly the same. Luck alone would still spread studies like ours around zero; 95% of them land in the green. Then drop our actual result on it: inside the green means zero is plausible, far outside means it is ruled out.
- **More-data dial:** keep the gap, change the amount of data, and watch the interval shrink or grow around it.

Chicago and the body-camera example also have a descriptive animation from class (Chicago's areas splitting into two groups, and camera footage year by year). Their code is folded away (click "Code" to see it); you do not need to write code like that.

## Setup

The tidyverse, the course's Bayesian function, and four datasets. Two of them are new to you and live on the course site; one is our Chicago anchor data; one comes straight from the *Washington Post*. `plotly` draws the two descriptive animations; if you want to run those chunks yourself, install it once with `install.packages("plotly")`.

```{r}
#| label: setup
library(tidyverse)
library(plotly)   # only for the animations
source("https://smourtgos.github.io/crju705/R/bayes.R")   # gives you bayes_prop_test()
set.seed(705)

rossi   <- read_csv("https://smourtgos.github.io/crju705/data/rossi.csv",
                    show_col_types = FALSE)
arrests <- read_csv("https://smourtgos.github.io/crju705/data/toronto-arrests.csv",
                    show_col_types = FALSE)
areas   <- read_csv("https://smourtgos.github.io/crju705/data/chicago-areas.csv",
                    show_col_types = FALSE)
shootings <- read_csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/v2/fatal-police-shootings-data.csv",
                      show_col_types = FALSE)
```

```{r}
#| label: explorer-helper
#| include: false
source("../R/gap-explorer.R")   # builds each "Where is zero?" explorer from the real test output
```

More about each dataset, and four more like them, is on the [Final-Project Data Options](../project-data.qmd) page.

## 1 · Reentry: does money at release keep people out of jail?

432 men released from Maryland state prisons in the 1970s were followed for one year. Half were **randomly assigned** to receive financial aid (`fin`). The outcome is `arrest`: 1 if arrested again within the year. One row is one man.

```{r}
#| label: rossi-count
rossi |> count(fin, arrest)
```

Aid: 48 re-arrested out of 48 + 168 = **216** (22.2%). No aid: 66 out of 66 + 150 = **216** (30.6%). Aid goes first, so every difference reads **Aid minus No aid**.

```{r}
#| label: rossi-prop
prop.test(x = c(48, 66), n = c(216, 216))
```

The interval runs from about 17 points lower to 0.4 points higher, and **p = 0.063**. Zero is just inside, so: fail to reject the null. A funder who reads only that line cuts the program.

Now ask the other question:

```{r}
#| label: rossi-bayes
bayes_prop_test(x = c(48, 66), n = c(216, 216))
```

**Where is zero?** Start with *Rerun the study*, then try *Is zero plausible?*: our gap of -8.3 lands just inside the green of the no-difference world, which is exactly why p = 0.063.

```{r}
#| label: rossi-widget
#| echo: false
#| output: asis
ge_widget(
  ge_prop("Aid vs no aid", A = "Aid", B = "No aid", out = "re-arrested",
          xa = 48, na = 216, xb = 66, nb = 216,
          left = "aid group re-arrested less", right = "aid group re-arrested more",
          rerun = "same program, 432 different men", unit_group = "men"),
  default = 1)
```

**Say it.** Given these data, there is about a **97% probability** that men who got aid were re-arrested at a lower rate. About 22 of every 100 men with aid were re-arrested, against about 31 of every 100 without: roughly 8 fewer re-arrests per 100 people released. What this cannot tell you: whether the effect lasts past a year, or would hold today. "Not significant" is not "no effect."

## 2 · Pretrial release: does a job get you released?

5,226 Toronto arrests for possessing small amounts of marijuana, 1997 to 2002. `released` is whether the person was released with a summons instead of held for a bail hearing. The group is `employed`. One row is one arrest.

```{r}
#| label: tor-count
arrests |> count(employed, released)
```

Employed: 3,568 released out of 3,568 + 543 = **4,111** (86.8%). Not employed: 766 out of 766 + 349 = **1,115** (68.7%).

```{r}
#| label: tor-prop
prop.test(x = c(3568, 766), n = c(4111, 1115))
```

```{r}
#| label: tor-bayes
bayes_prop_test(x = c(3568, 766), n = c(4111, 1115))
```

About 15 to 21 points more often for the employed, by either test, and more than a 99% probability the employed rate is higher. But the two groups differ in other ways. `checks` counts how many of six police databases (prior arrests, convictions, parole, and so on) had the person's name:

```{r}
#| label: tor-checks
arrests |> summarize(avg_checks = mean(checks), people = n(), .by = employed)
```

People without a job had more of a record, and a record is a reason to hold someone. So compare like with like: keep only people with **no record at all**, and ask again.

```{r}
#| label: tor-clean
arrests |> filter(checks == 0) |> count(employed, released)
prop.test(x = c(1526, 166), n = c(1650, 201))
bayes_prop_test(x = c(1526, 166), n = c(1650, 201))
```

**Where is zero?** Start with *Arrows between groups*, then switch between *Everyone* and *No record only*: with only 201 people in the not-employed group, its curve gets wider, and so does the gap's pile, but it never reaches zero.

```{r}
#| label: tor-widget
#| echo: false
#| output: asis
ge_widget(
  ge_prop("Everyone", A = "Employed", B = "Not employed", out = "released",
          xa = 3568, na = 4111, xb = 766, nb = 1115,
          left = "employed released less often", right = "employed released more often",
          rerun = "another 5,226 arrests", unit_group = "arrests"),
  ge_prop("No record only", A = "Employed", B = "Not employed", out = "released",
          xa = 1526, na = 1650, xb = 166, nb = 201,
          left = "employed released less often", right = "employed released more often",
          rerun = "another 1,851 arrests with no record", unit_group = "arrests"),
  default = 2)
```

**Say it.** Among people with clean records, the employed were released about 10 points more often (92.5% vs 82.6%; interval about 4 to 16 points, credible interval 5.1 to 15.8 points, more than a 99% probability the employed rate is higher). The gap shrank from 18 points to 10 when we compared like with like, but it did not disappear. What this cannot tell you: whether some other difference (age, citizenship, where the arrest happened) explains the rest.

## 3 · Neighborhoods: violence and income

Chicago's 77 community areas. `high_violence` marks the 38 areas whose violent-crime rate is above the city's median. The outcome is `median_income`, a number in dollars, so the test is `t.test()`. One row is one **area**, not one person. First, name the groups and put High violence first:

```{r}
#| label: chi-groups
areas <- areas |>
  mutate(violence = factor(if_else(high_violence, "High violence", "Other areas"),
                           levels = c("High violence", "Other areas")))

areas |>
  summarize(avg_income = mean(median_income), areas = n(), .by = violence) |>
  arrange(violence)
```

**See it.** Each dot is one community area:

```{r}
#| label: chi-anim
#| code-fold: true
set.seed(705)
jit <- runif(nrow(areas), -0.18, 0.18)
stage_labels <- c("All 77 areas", "Split by violence", "The two averages")
chi_pts <- map_dfr(1:3, \(s) areas |>
  mutate(x = if (s == 1) 1.5 + jit else if_else(violence == "High violence", 1, 2) + jit,
         stage = stage_labels[s])) |>
  mutate(stage = factor(stage, levels = stage_labels),
         hover = paste0(community, "<br>$", format(round(median_income), big.mark = ",")))
chi_means <- areas |>
  summarize(avg = mean(median_income), .by = violence)
chi_means <- map_dfr(1:3, \(s) chi_means |>
  mutate(x = if (s < 3) -1 else if_else(violence == "High violence", 1, 2),
         stage = stage_labels[s])) |>
  mutate(stage = factor(stage, levels = stage_labels),
         txt = paste0("Average<br>$", format(round(avg), big.mark = ",")))
chi_gap <- tibble(stage = factor(stage_labels, levels = stage_labels),
                  x = c(-1, -1, 1.5), y = 175000, txt = "Gap: about $35,700")

plot_ly(height = 470) |>
  add_markers(data = chi_pts, x = ~x, y = ~median_income, frame = ~stage, ids = ~community,
              color = ~violence, colors = c("firebrick", "steelblue"),
              marker = list(size = 10, opacity = 0.8), text = ~hover, hoverinfo = "text") |>
  add_markers(data = chi_means, x = ~x, y = ~avg, frame = ~stage, ids = ~violence,
              marker = list(symbol = "line-ew-open", size = 60, color = "navy",
                            line = list(width = 4, color = "navy")),
              showlegend = FALSE, hoverinfo = "none") |>
  add_text(data = chi_means, x = ~x + 0.33, y = ~avg, text = ~txt, frame = ~stage,
           textfont = list(size = 14, color = "navy"), showlegend = FALSE, hoverinfo = "none") |>
  add_text(data = chi_gap, x = ~x, y = ~y, text = ~txt, frame = ~stage,
           textfont = list(size = 18, color = "navy"), showlegend = FALSE, hoverinfo = "none") |>
  layout(xaxis = list(title = "", range = c(0.4, 2.6), tickvals = c(1, 2),
                      ticktext = c("High violence", "Other areas")),
         yaxis = list(title = "Median household income", tickprefix = "$", range = c(0, 190000)),
         legend = list(orientation = "h", y = 1.08)) |>
  animation_opts(frame = 1800, transition = 900, redraw = FALSE) |>
  animation_slider(currentvalue = list(prefix = "")) |>
  config(displayModeBar = FALSE)
```

```{r}
#| label: chi-ttest
t.test(median_income ~ violence, data = areas)
```

**Where is zero?** Start with *Arrows between groups*: the two curves are where each group's true average income could be, and the arrows between them never come close to pointing the other way.

```{r}
#| label: chi-widget
#| echo: false
#| output: asis
ge_widget(
  ge_mean("Income", A = "High violence", B = "Other areas",
          a = areas$median_income[areas$violence == "High violence"],
          b = areas$median_income[areas$violence == "Other areas"],
          left = "high-violence areas poorer", right = "high-violence areas richer",
          rerun = "another 77 areas like these", unit_group = "areas"),
  default = 2)
```

**Say it.** High-violence areas minus other areas: about **$25,000 to $47,000 lower** median household income ($54,355 vs $90,088 on average), p = 0.00000001. The unit is dollars because the column is dollars. What this cannot tell you: which way it runs (does poverty breed violence, or does violence drive income out?), and anything about individual people. Each row is an area.

## 4 · Use of force: body cameras over time

The *Washington Post*'s database of every person shot and killed by on-duty police in the U.S., 2015 to 2024. `body_camera` is TRUE when news or police reports said a camera recorded the incident. One row is one person killed. Group: before and after 2020.

```{r}
#| label: wapo-period
shootings <- shootings |>
  mutate(period = if_else(year(date) >= 2020, "2020 to 2024", "2015 to 2019"))

shootings |> count(period, body_camera)
```

2020 to 2024: 1,205 of 1,205 + 4,302 = **5,507** (21.9%). 2015 to 2019: 594 of 594 + 4,329 = **4,923** (12.1%).

**Year by year** first, because two five-year averages can hide a lot:

```{r}
#| label: wapo-anim
#| code-fold: true
yearly <- shootings |>
  mutate(year = year(date)) |>
  summarize(pct = 100 * mean(body_camera), .by = c(year, period)) |>
  arrange(year)
yr_labels <- c(as.character(2015:2024), "The two periods")
bwc_anim <- map_dfr(seq_along(yr_labels), \(s) yearly |>
  mutate(shown = if_else(s <= 10 & year > 2014 + s, 0, pct), step = yr_labels[s])) |>
  mutate(step = factor(step, levels = yr_labels),
         txt = if_else(shown > 0, paste0(round(shown), "%"), ""))
per <- shootings |> summarize(pct = 100 * mean(body_camera), .by = period)
bwc_lines <- map_dfr(seq_along(yr_labels), \(s) per |>
  mutate(x = if (s < 11) 2010 else if_else(period == "2015 to 2019", 2017, 2022),
         step = yr_labels[s])) |>
  mutate(step = factor(step, levels = yr_labels),
         ytxt = if_else(period == "2015 to 2019", 19.5, 35),
         txt = paste0(period, ": ", round(pct, 1), "%"))

plot_ly(height = 470) |>
  add_bars(data = bwc_anim, x = ~year, y = ~shown, frame = ~step, color = ~period,
           colors = c("steelblue", "firebrick"), text = ~txt, textposition = "outside",
           hoverinfo = "none") |>
  add_markers(data = bwc_lines, x = ~x, y = ~pct, frame = ~step, ids = ~period,
              marker = list(symbol = "line-ew-open", size = 200, color = "navy",
                            line = list(width = 4, color = "navy")),
              showlegend = FALSE, hoverinfo = "none") |>
  add_text(data = bwc_lines, x = ~x, y = ~ytxt, text = ~txt, frame = ~step,
           textfont = list(size = 15, color = "navy"), showlegend = FALSE, hoverinfo = "none") |>
  layout(xaxis = list(title = "", range = c(2014.4, 2024.6), dtick = 1),
         yaxis = list(title = "Body camera reported (%)", range = c(0, 38)),
         legend = list(orientation = "h", y = 1.08)) |>
  animation_opts(frame = 900, transition = 400, redraw = FALSE) |>
  animation_slider(currentvalue = list(prefix = "")) |>
  config(displayModeBar = FALSE)
```

The rise is real but not smooth: 2023 hit 30%, and 2024 fell back to 20%. Comparing two five-year periods averages over those bumps; that is a choice, and an honest write-up says so. Now both tests on the two periods:

```{r}
#| label: wapo-tests
prop.test(x = c(1205, 594), n = c(5507, 4923))
bayes_prop_test(x = c(1205, 594), n = c(5507, 4923))
```

**Where is zero?** Start with the *More-data dial*: even with a quarter of the data, the interval stays well clear of zero. With 10,430 cases, a gap this size could not hide.

```{r}
#| label: wapo-widget
#| echo: false
#| output: asis
ge_widget(
  ge_prop("Body cameras", A = "2020 to 2024", B = "2015 to 2019", out = "with a camera",
          xa = 1205, na = 5507, xb = 594, nb = 4923,
          left = "cameras less common after 2020", right = "cameras more common after 2020",
          rerun = "another 10,430 shootings", unit_group = "shootings"),
  default = 4)
```

**Say it.** About 12 in 100 fatal shootings had known camera footage before 2020, and about 22 in 100 after (8 to 11 points higher, by either test). Close to doubling, and still fewer than 1 in 4. What this cannot tell you: whether cameras changed anything about the shootings, or how many cameras existed but went unreported. With 10,430 cases, the p-value was never in doubt. The size is the whole story.

## Four questions, one recipe

| | Test | Answer | The lesson |
|---|---|---|---|
| Reentry | `prop.test()` + Bayes | 22% vs 31%; p = 0.063; 97% probably lower | "Not significant" is not "no effect" |
| Pretrial | `prop.test()` + Bayes, then again after `filter()` | 18 points; 10 among clean records | Groups differ in other ways. Compare like with like |
| Neighborhoods | `t.test()` | $25k to $47k lower | Say the size in real units. Areas are not people |
| Use of force | `prop.test()` + Bayes | 12% to 22% | With a huge n, the p-value adds nothing; the size is the story |

## Try a variation

1. **Rossi:** were men released on **parole** (`paro`) re-arrested less often? Count, then both tests, then say it.
2. **Toronto:** were **citizens** (`citizen`) released more often than non-citizens? Then ask again among people with `checks == 0`. Does the gap shrink the way the employment gap did?
3. **Chicago areas:** run `t.test(pct_vacant ~ violence, data = areas)`. How much higher is the share of vacant housing in high-violence areas, in percentage points?
4. **Shootings:** was a body camera reported more often when the person was in a mental health crisis (`was_mental_illness_related`)? Count, then both tests, and say the size in plain words.
