---
title: "Demo Walkthrough · Session 1"
subtitle: "First R commands, levels of measurement, and why the scale controls the math"
---

::: {.callout-note appearance="simple"}
**What this page is.** The Session 1 lecture demos, written down step by step. Every chunk below is self-contained — copy any of them into your own R session and they run as-is. Use this page to catch up after a missed class, or to re-run what happened on screen at your own pace.
:::

## What you'll build

1. Your first R commands: arithmetic, vectors, and the functions you'll use all semester
2. Logical indexing — asking a vector a question and keeping only the rows that say yes
3. Factors, R's container for categorical data
4. The "scale determines the math" demos: why a *mean zone* is nonsense, and why °C and age play by different rules
5. The 2×2 of plots that match each level of measurement
6. A five-verb dplyr tour, and a first look at the real Chicago data we'll use all semester

## Setup

Everything below needs the tidyverse, plus **patchwork** for one plot-arranging trick near the end (`install.packages("patchwork")` once, if you don't have it):

```{r}
#| label: setup
library(tidyverse)
library(patchwork)
set.seed(705)

crimes <- read_csv("https://smourtgos.github.io/crju705/data/chicago-crimes-2025.csv",
                   show_col_types = FALSE)
```

## The case study: catching a bank robber

Session 1 opened with a serial offender hitting 40 banks across a city's five zones, and a police department with nothing to go on but *when* and *where* each robbery happened. A crime analyst turned that thin dataset into a logistic regression predicting which zone would be hit next; the model pointed to the West Zone, units were deployed there, and the robber was arrested days later approaching a bank. The animated timeline, map, and deployment visuals are built with tools from later in the course — watch them in the [Session 1 slides](../slides/week-01-stats-is-awesome.qmd); by Session 12 you'll be able to build the model yourself.

One more thing worth naming: you already reasoned like a **Bayesian** while watching that story. You started with a vague hunch spread across five zones (a *prior*), treated each new robbery as *evidence* that reshaped the hunch, read the model's risk-by-zone output as an *updated belief given the data*, and then made a deployment decision under quantified uncertainty. That whole loop has a formal name — **Bayesian updating**. The formula shows up in Session 3, and the full machinery in Session 6, side by side with the other great tradition, the frequentist one.

The rest of this page is the part of Session 1 you can run right now: your first R commands, and the levels-of-measurement demos.

## 1 · Arithmetic, assignment, and your first functions

R is a calculator first:

```{r}
#| label: arithmetic
3 + 5
```

The arrow `<-` **assigns** a value to a name. `1:10` builds the vector 1, 2, 3, …, 10, and functions like `sum()`, `mean()`, and `sd()` eat whole vectors at once:

```{r}
#| label: vectors
x <- 1:10
sum(x); mean(x); sd(x)
```

Every function has a help page — get in the habit of pulling them up constantly:

```{r}
#| label: help
#| eval: false
?mean
```

## 2 · Logical indexing — ask the data a question

Comparing a vector to a value asks the question element by element:

```{r}
#| label: logical-test
x > 5
```

Put that question inside square brackets and R keeps only the values that answered TRUE:

```{r}
#| label: logical-subset
x[x > 5]
```

This one move — *build a logical test, use it to subset* — is the ancestor of every `filter()` you'll ever write.

## 3 · Characters and factors

Text lives in **character** vectors; categories that will be *analyzed* belong in **factors**, which know their full set of possible levels:

```{r}
#| label: factors
student_names <- c("Alice", "Bob", "Carol")
is.character(student_names)

f <- factor(c("red", "blue", "red"))
f
```

Notice the printout: a factor reports its `Levels` — the categories that exist, whether or not each one appears in this particular vector.

## 4 · Scale determines the math

Session 1's central claim: a variable's **level of measurement** (nominal, ordinal, interval, ratio) determines what math is allowed. To demonstrate, simulate a small robbery-style dataset with one variable at each level:

```{r}
#| label: rob-sim
set.seed(58)   # same seed as the lecture, so these numbers match the slides
rob <- tibble(
  zone = factor(sample(c("North", "South", "East", "West", "Downtown"), 200, TRUE)),
  prior_robberies_in_zone = rpois(200, 3),    # discrete count (ratio)
  risk_likert = sample(1:7, 200, TRUE),       # ordinal
  temp_c = rnorm(200, 20, 5),                 # interval
  age = pmax(18, round(rnorm(200, 40, 12))),  # ratio
  caught = factor(rbinom(200, 1, 0.35), labels = c("No", "Yes"))
)
```

### Nominal: the "mean zone" is meaningless

R will happily convert factor levels to their underlying codes and average them. That doesn't make the answer mean anything:

```{r}
#| label: nominal-wrong
# WRONG: the "mean zone" is meaningless
rob$zone |> as.numeric() |> mean()
```

3.17 of… what? Zone 3.17 doesn't exist — the numbers are just labels in disguise. The *right* math for nominal data is counts and percentages:

```{r}
#| label: nominal-right
# RIGHT: counts and percentages
rob |>
  count(zone) |>
  mutate(pct = 100 * n / sum(n))
```

### Ordinal: order yes, arithmetic no

A 1–7 risk rating is *ordered*, but the gaps between points aren't guaranteed equal — so summarize with the median and IQR, and measure association with a rank-based (Spearman) correlation:

```{r}
#| label: ordinal-summaries
rob |>
  summarize(med = median(risk_likert), iqr = IQR(risk_likert))

cor(rob$risk_likert, rob$prior_robberies_in_zone, method = "spearman")
```

(That correlation is essentially zero — as it should be, since we simulated these two variables independently. A good sanity check.)

### Interval vs. ratio: the true-zero test

Temperature in °C is **interval**: differences are meaningful, but there's no true zero — 0 °C is just where water freezes. Watch what unit conversion does to the mean:

```{r}
#| label: interval-demo
mean(rob$temp_c)             # about 20 °C
mean(rob$temp_c * 1.8 + 32)  # the same temperatures, in °F
```

Same warmth, different number. "30 °C is 10 °C warmer than 20 °C" is fine; "20 °C is *twice as hot* as 10 °C" is meaningless — the ratio depends on which scale you happen to be using.

```{r}
#| label: ratio-demo
mean(rob$age)
```

Age is **ratio**: it has a true zero, so a 40-year-old really *is* twice as old as a 20-year-old. Differences *and* ratios both work.

## 5 · Plots that match the scale

The same logic governs graphics. One plot per level of measurement, arranged in a 2×2 with patchwork (`p1 + p2` puts plots side by side; `/` stacks rows):

```{r}
#| label: scale-plots
#| fig-width: 10
#| fig-height: 6
p1 <- ggplot(rob, aes(zone)) + geom_bar(fill = "firebrick") +
  labs(title = "Nominal: bar chart", x = NULL)
p2 <- ggplot(rob, aes(factor(risk_likert, levels = 1:7))) +
  geom_bar(fill = "navy") +
  labs(title = "Ordinal: ordered bars", x = "Risk (1–7)")
p3 <- ggplot(rob, aes(age)) + geom_histogram(binwidth = 5, fill = "steelblue") +
  labs(title = "Continuous (ratio): histogram")
p4 <- ggplot(rob, aes(prior_robberies_in_zone)) + geom_bar(fill = "goldenrod") +
  labs(title = "Discrete counts: bars", x = "Prior robberies")

(p1 + p2) / (p3 + p4)
```

Categories get bars; continuous measurements get histograms; discrete counts get one bar per whole number. If you ever catch yourself histogramming a nominal variable, stop and check the scale.

## 6 · Five dplyr verbs you'll use every week

The tidyverse works on **tibbles** (modern data frames). The `|>` symbol is the **pipe**: `x |> f()` means `f(x)` — "take x, *then* do f." (Older code, including the Stanton book, writes it `%>%`; they work the same for everything we do.)

```{r}
#| label: dplyr-tibble
cars_data <- as_tibble(mtcars, rownames = "model")
head(cars_data, 4)
```

`select()` picks **columns**; `filter()` picks **rows**:

```{r}
#| label: dplyr-select-filter
cars_data |> select(model, mpg, cyl, hp) |> head(4)
cars_data |> filter(cyl == 6) |> head(4)
```

`mutate()` builds new columns; `summarize()` collapses many rows to a few numbers — here, per group with `.by`:

```{r}
#| label: dplyr-mutate-summarize
cars_data |>
  mutate(mpg_per_hp = mpg / hp) |>
  select(model, mpg, hp, mpg_per_hp) |>
  head(4)

cars_data |>
  summarize(avg_mpg = mean(mpg), n = n(), .by = cyl)
```

## 7 · Coming attractions: real data, all semester

Starting in Session 2, the running example is **real**: a 30,000-incident sample of every crime reported in Chicago in 2025 (already loaded in Setup). Here's the count-and-bar-chart pattern from this session, applied to it:

```{r}
#| label: chicago-teaser
#| fig-height: 4.5
crimes |>
  count(primary_type, sort = TRUE) |>
  slice_head(n = 10) |>
  ggplot(aes(n, fct_reorder(primary_type, n))) +
  geom_col(fill = "firebrick") +
  labs(title = "Ten most common reported crimes, Chicago 2025",
       subtitle = "30,000-incident sample · Chicago Police Department data portal",
       x = "Incidents", y = NULL)
```

Theft tops the list at about 7,000 incidents. And note what kind of variable `primary_type` is — nominal — which is exactly why this is a bar chart and not a histogram.

## Try a variation

1. Build a vector of seven ages with `c()`. Compute the mean and SD, subset the values over 30, then use `?seq` to figure out how to generate `seq(0, 1, by = 0.2)`.
2. Using the built-in `mpg` data (type `mpg` in the console — ggplot2 ships it), find the top-3 models by `hwy` and the average `cty` by `class`. *(Hints: `arrange(desc(...))`, `slice_head()`, `summarize(..., .by = ...)`.)*
3. In the Chicago chunk, swap `primary_type` for `location_description`. Where does crime actually happen in Chicago?
4. The ordinal trap: run `educ <- rep(1:5, each = 50)` and `income <- rep(c(1, 2, 3, 4, 1000), each = 50)`, then `cor(educ, income)`. Pearson's r looks impressive — why is it misleading here? *(Hint: are the education codes equally spaced in any real sense?)*
